---
title: "COMAL project"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#load necessary libraries
library(sf)
library(reshape2)
library(ggcorrplot)
library(leaflet)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(tidyterra)
library(mapview)
library(spData)
library(gganimate)
library(htmlwidgets)
library(webshot)
library(chirps)
library(viridis)
library(patchwork)
library(GGally)
library(spdep)
library(INLA)
library(leafsync)
library(SpatialEpi)
library(geodata)
library(car)
library(gstat)
library(nngeo)
library(tidygeocoder)
library(malariaAtlas)
library(countrycode)
library(rmapshaper)
library(tidytext)
```

```{r}
# Function: fill NA values in one column using nearest non‑NA neighbor
fill_by_nearest <- function(sf_obj, colname) {

  # rows with non‑missing values
  non_na <- sf_obj %>% filter(!is.na(.data[[colname]]))

  # rows with missing values
  na_rows <- sf_obj %>% filter(is.na(.data[[colname]]))

  # if no NA: return unchanged
  if (nrow(na_rows) == 0) return(sf_obj)

  # find nearest non‑NA feature for each NA row
  nearest_id <- st_nearest_feature(na_rows, non_na)

  # extract the corresponding values
  replacement_values <- non_na[[colname]][nearest_id]

  # fill the NA values
  sf_obj[[colname]][is.na(sf_obj[[colname]])] <- replacement_values

  sf_obj
}
```



```{r}
#loading the shape files
shp_ben <- read_sf("data/ben_data/BEN_shape_file/gadm41_BEN_3.shp")
shp_con <- read_sf("data/cong_data/CON_shape_file/cog_admin2.shp")
shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")


#renaming the districts

shp_ben$NAME_3[shp_ben$NAME_3=="Gakpè"]="Gakpe"
shp_ben$NAME_3[shp_ben$NAME_3=="Tokpa Domé"]="Tokpa-dome"
shp_ben$NAME_3[shp_ben$NAME_3=="Ouakpé-Daho"]="Houakpe-daho"
shp_ben$NAME_3[shp_ben$NAME_3=="Aganmalomé"]="Aganmanlome"
shp_ben$NAME_3[shp_ben$NAME_3=="Agonkanmè"]="Agonkamey"

ben_vect <- vect(shp_ben)
cmr_vect <- vect(shp_cmr)
con_vect <- vect(shp_con)
```


```{r}
# Loading the data

BEN_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 1)

CON_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 2)

CMR_df <- read_excel("data/cmr_data/Cameroon2.xlsx")

GAB_df <- read_excel("data/gab_data/Gabon2.xlsx")

```


# Cycles Threshold (CT) analysis




```{r}

```



# Extract the longitude and latitude of households using gps column for Cameroon data

```{r}
## function to extract the GPS location for Cameroon data

convert_custom_dms <- function(x) {
  hemi <- substr(x, 1, 1)
  x <- substring(x, 2)

  apos <- regexpr("'", x)[1]

  if (hemi == "N") {
    deg <- as.numeric(substr(x, 1, apos - 4))
  } else {
    deg <- as.numeric(substr(x, 1, apos - 3))
  }

  min <- as.numeric(substr(x, apos - 2, apos - 1))
  sec <- as.numeric(gsub("\"", "", substr(x, apos + 1, nchar(x))))

  dec <- deg + min/60 + sec/3600

  if (hemi %in% c("S", "W")) dec <- -dec
  return(dec)
}

CMR_df$lat <- sapply(strsplit(CMR_df$gps, " "), function(x) convert_custom_dms(x[1]))
CMR_df$lon <- sapply(strsplit(CMR_df$gps, " "), function(x) convert_custom_dms(x[2]))


CMR_df_1 <- CMR_df %>% 
  dplyr::select(ID, lon, lat, gender, age, pcr_pm, pcr_pf) %>% 
  mutate(
    lng = lon,
    tot_participants = 1,
    tot_infected_pm  = if_else(pcr_pm == "pos", 1, 0),
    tot_infected_pf  = if_else(pcr_pf == "pos", 1, 0)
  )%>% 
  group_by(lng, lat) %>% 
  summarise(age_lower = min(age),
            age_upper = max(age),
            Number_of_participant_tot  = sum(tot_participants, na.rm = TRUE),
            Number_of_infected_tot_pm  = sum(tot_infected_pm, na.rm = TRUE),
            PR_Pm_tot = Number_of_infected_tot_pm  / Number_of_participant_tot ,
            Number_of_infected_tot_pf  = sum(tot_infected_pf, na.rm = TRUE),
            PR_Pf_tot = Number_of_infected_tot_pf  / Number_of_participant_tot ,
            .groups = "drop") %>% 
  mutate(country = "Cameroon")

# Group households into clusters

households <- st_as_sf(
    CMR_df_1,
    coords = c("lng","lat"),
    crs = 4326
)

households <- st_transform(households, 32632)

grid_cmr_cluster <- st_make_grid(
    households,
    cellsize = 1000,
    square = TRUE
)

grid_cmr_cluster <- st_sf(
  grid_id = seq_along(grid_cmr_cluster),
  geometry = grid_cmr_cluster
)

households_grid <- st_join(
  households,
  grid_cmr_cluster,
  join = st_intersects
)

cluster_data_cmr <-
  households_grid %>%
  st_drop_geometry() %>% 
  group_by(grid_id) %>%
  summarise(age_lower = min(age_lower),
         age_upper = max(age_upper),
         Number_of_participant_tot  = sum(Number_of_participant_tot, na.rm = TRUE),
         Number_of_infected_tot_pm  = sum(Number_of_infected_tot_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm  / Number_of_participant_tot ,
         Number_of_infected_tot_pf  = sum(Number_of_infected_tot_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf  / Number_of_participant_tot ,
         .groups = "drop") %>% 
  mutate(country = "Cameroon")


cluster_data_cmr_sf <-
  grid_cmr_cluster  %>% 
  inner_join(cluster_data_cmr, by = "grid_id")

cluster_data_cmr_sf <- st_transform(cluster_data_cmr_sf, 4326)
cluster_data_cmr_sf <- st_centroid(cluster_data_cmr_sf)

coords_cmr <- st_coordinates(cluster_data_cmr_sf)

cluster_data_cmr_sf$lng <- coords_cmr[,1]
cluster_data_cmr_sf$lat <- coords_cmr[,2]



CMR_df_1_sf <- st_as_sf(CMR_df_1, coords = c("lng", "lat"), crs = 4326)
cluster_data_cmr_sf <- st_as_sf(cluster_data_cmr_sf, coords = c("lng", "lat"), crs = 4326)



raw_pr_cmr <- ggplot() +
  geom_sf(data = shp_cmr, fill = "white", color = "black") +
  geom_point(data = cluster_data_cmr_sf, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/raw_obs_PR_cmr.pdf", plot = raw_pr_cmr, width = 6, height = 4, dpi = 300)

sampled_site_cmr <- shp_cmr %>% 
  filter(Admin2 == "Bankim")

sample_district <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_point(data = cluster_data_cmr_sf, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence (Bankim)",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/raw_obs_PR_sampled_location_cmr.pdf", plot = sample_district, width = 6, height = 4, dpi = 300)


mapview(shp_cmr,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(CMR_df_1_sf,
        zcol = "PR_Pm_tot",
        cex = "Number_tot_participants",
        col.regions = colorRampPalette(c("blue", "orange")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```


# Find the GPS coordinates for sampled districts in Benin and aggregation of our data at the level of village 

```{r}
village_coords_ben <- tribble(
  ~Municipality, ~District,     ~Village,            ~Match_confidence,   ~lat,     ~lng,
  "Ouidah",      "Gakpe",       "Lokohoue",           "fuzzy_match",      6.443353,  2.132558,#
  "Ouidah",      "Gakpe",       "Kindjitokpa",        "fuzzy_match",      6.461650,  2.148980,#
  "Kpomasse",    "Tokpa-dome",  "Lokogbo-Gnonwa",     "exact_match",      6.471386,  1.984651,#
  "Ouidah",      "Houakpe-daho","Gbezoume",           "exact_match",      6.344216,  1.962808,#
  "Ouidah",      "Gakpe",       "Sehlinhoue",         "fuzzy_match",      6.416572,  2.173980,#
  "Kpomasse",    "Aganmanlome", "Hessa",              "fuzzy_match",      6.495617,  2.053485,#
  "Kpomasse",    "Aganmanlome", "KOUZOUME",           "fuzzy_match",      6.462216,  2.017743,#
  "Ouidah",      "Gakpe",       "Agossouhouihoué",    "fuzzy_match",      6.421688,  2.144019,#
  "Ouidah",      "Gakpe",       "Ahossihoue",         "fuzzy_match",      6.452980,  2.167627,#
  "Ouidah",      "Houakpe-daho","Gbehonou",           "exact_match",      6.339606,  2.005831,#
  "Kpomasse",    "Aganmanlome", "KOUGBEDJI",          "fuzzy_match",      6.473206,  2.077346,#
  "Ouidah",      "Houakpe-daho","Toligbé",            "exact_match",      6.327061,  2.027671,#
  "Ouidah",      "Houakpe-daho","Houakpe-daho centre","exact_match",      6.329397,  2.047797,#
  "Kpomasse",    "Aganmanlome", "Aganmanlome Centre", "fuzzy_match",      6.477114,  2.049987,#
  "Ouidah",      "Gakpe",       "TOKOLI",             "fuzzy_match",      6.431493,  2.156038,#
  "Kpomasse",    "Aganmanlome", "Lokossa",            "fuzzy_match",      6.478105,  2.027037,#
  "Kpomasse",    "Agonkamey",   "Assogbenou Kpevi",   "exact_match",      6.366429,  2.027881,#
  "Ouidah",      "Houakpe-daho","Seyigbe",            "exact_match",      6.324844,  2.041879,#
  "Kpomasse",    "Agonkamey",   "Assogbenou daho",    "exact_match",      6.355088,  2.001053 #
)

BEN_1 <- BEN_df %>% 
  mutate(Village = case_when(Village == "HESSA" ~ "Hessa",
                             TRUE ~ Village)) %>% 
  dplyr::select(Village, Sex, `Age (year)`, `qRT-qPCR Pm positive`, `qRT-qPCR Pf positive`) %>%
  rename(age = `Age (year)`) %>% 
  mutate(Pm_positive = case_when( `qRT-qPCR Pm positive`=="x"~"yes",
                                  TRUE~"no"),
         Pf_positive = case_when( `qRT-qPCR Pf positive`=="x"~"yes",
                                  TRUE~"no"),
         participants = 1,
         infected_pm  = case_when(Pm_positive=="yes"~1,
                              TRUE~0),
         infected_pf  = case_when(Pf_positive=="yes"~1,
                              TRUE~0)) %>% 
  group_by(Village) %>% 
  summarise(age_lower = min(age, na.rm = TRUE),
         age_upper = max(age, na.rm = TRUE),
         Number_of_participant_tot = sum(participants, na.rm = TRUE),
         Number_of_infected_tot_pm = sum(infected_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm/Number_of_participant_tot,
         Number_of_infected_tot_pf = sum(infected_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf/Number_of_participant_tot,
         .groups = "drop") %>%  
  distinct()
  


BEN_1_GPS <- BEN_1 %>% 
  left_join(village_coords_ben %>% dplyr::select(Village, lat, lng, Match_confidence))

spatial_data_ben <- st_as_sf(BEN_1_GPS, coords = c("lng", "lat"), crs = 4326)

PR_obs_ben <- ggplot() +
  geom_sf(data = shp_ben, fill = "white", color = "black") +
  geom_point(data = BEN_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/PR_p_malariae_observed.pdf", plot = PR_obs_ben, width = 6, height = 4, dpi = 300)


sampled_site_ben <- shp_ben %>% 
  filter(NAME_3 %in% c("Tokpa-dome", "Aganmanlome", "Savi", "Kpomassè", "Gakpe", 
                       "Arrondissement III", "Arrondissement I", "Houakpe-daho",
                       "Agonkamey", "Arrondissement II") & NAME_2 %in% c("Kpomassè", "Ouidah"))

Raw_PR_sampled_location_ben <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_point(data = BEN_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Raw_PR_sampled_location.pdf", plot = Raw_PR_sampled_location_ben,
       width = 6, height = 4, dpi = 300)


mapview(shp_ben,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(spatial_data_ben,
        zcol = "PR_Pm_tot",
        cex = "Number_of_participant_tot",
        col.regions = colorRampPalette(c("blue", "orange")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```


# Find the GPS coordinates for sampled districts in Congo and aggregation of our data at the level of village 

```{r}
village_coords_cong <- tribble(
   ~District,     ~Match_confidence,     ~lat,                  ~lng,
    "Ntoula",      "fuzzy_match",      -4.36,                 15.15,#
    "Djoumouna",   "exact_match",      -4.37601066518807,     15.15963007611599,#
    "Mayanga",     "exact_match",      -4.284712200560734,    15.186363375868131,#
  )

CONG_1 <- CON_df %>% 
  dplyr::select(District, Sex, `Age (year)`, `qRT-qPCR Pm positive`, `qRT-qPCR Pf positive`) %>% 
  rename(age = `Age (year)`) %>% 
  mutate(Pm_positive = case_when( `qRT-qPCR Pm positive`=="x"~"yes",
                                  TRUE~"no"),
         Pf_positive = case_when( `qRT-qPCR Pf positive`=="x"~"yes",
                                  TRUE~"no"),
         participants = 1,
         infected_pm = case_when(Pm_positive=="yes"~1,
                              TRUE~0),
         infected_pf = case_when(Pf_positive=="yes"~1)) %>% 
  group_by(District) %>% 
  summarise(age_lower = min(age, na.rm = TRUE),
         age_upper = max(age, na.rm = TRUE),
         Number_of_participant_tot = sum(participants, na.rm = TRUE),
         Number_of_infected_tot_pm = sum(infected_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm/Number_of_participant_tot,
         Number_of_infected_tot_pf = sum(infected_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf/Number_of_participant_tot,
         .groups = "drop") %>% 
  distinct()
  


CONG_1_GPS <- CONG_1 %>% 
  left_join(village_coords_cong %>% dplyr::select(District, lat, lng, Match_confidence))

spatial_data_cong <- st_as_sf(CONG_1_GPS, coords = c("lng", "lat"), crs = 4326)
```


```{r}
raw_pr_cong <- ggplot() +
  geom_sf(data = shp_con, fill = "white", color = "black") +
  geom_point(data = CONG_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/raw_obs_PR.pdf", plot = raw_pr_cong, width = 6, height = 4, dpi = 300)

sampled_site_cong <- shp_con %>% 
  filter(adm2_name %in% c("Brazzaville", "Goma-Tsetse"))

sampled_location_pr_cong <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_point(data = CONG_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/raw_obs_PR_sampled_location.pdf", plot = sampled_location_pr_cong, width = 6, height = 4, dpi = 300)

mapview(shp_con,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(spatial_data_cong,
        zcol = "PR_Pm_tot",
        cex = "Number_of_participant_tot",
        col.regions = colorRampPalette(c("blue", "red")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```


#visualizing the raster files for the analysis
```{r}
land_cover_water <- rast("data/ben_data/landuse/WorldCover_water_30s.tif")
land_cover_trees <- rast("data/ben_data/landuse/WorldCover_trees_30s.tif")
land_cover_builts <- rast("data/ben_data/landuse/WorldCover_built_30s.tif")
land_cover_cropland <- rast("data/ben_data/landuse/WorldCover_cropland_30s.tif")
land_cover_wetland <- rast("data/ben_data/landuse/WorldCover_wetland_30s.tif")

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")

precip_ratser_2021 <- list.files("data/CHIPRS 2021/", pattern = "chirps", full.names = TRUE)
annual_temperature_climate <- rast("data/Africa_data/ERA5_LAND_DAILY_temperature_2m_2021-01-01_2021-12-31_-39p5213to41p4698N_-22p2966to54p0315E.temperature_2m.tif")

path1 <- "data/CHIRPS_PENTAD_precipitation_2021-01-01_2021-12-31_0p9311to13p9448N_8p2516to17p0602E.precipitation.tif"

rainfall_2021 <- rast(path1)


precip_ratser_2021_data <- rast(precip_ratser_2021)
names(precip_ratser_2021_data) <- seq(as.Date("2022-01-01"),as.Date("2022-12-31"),
                          by = "months") %>% 
  format("%b %Y")

precip_clean <- ifel(precip_ratser_2021_data < 0, NA, precip_ratser_2021_data)

precip_ratser_2021_data_annual <- sum(precip_clean, na.rm = TRUE)
names(precip_ratser_2021_data_annual) <- "annual_rainfall"

# global_spatial_limit_3857 <- project(global_spatial_limit, crs(shp_africa))

shp_africa <- read_sf("data/Africa_data/shp_africa/Africa_simplified.shp")

pf_limit_africa <- crop(global_spatial_limit, vect(shp_africa))
pf_limit_africa <- mask(pf_limit_africa, vect(shp_africa))

precip_ratser_2021_data_annual <- crop(precip_ratser_2021_data_annual, vect(shp_africa))
precip_ratser_2021_data_annual <- mask(precip_ratser_2021_data_annual, vect(shp_africa))

annual_temperature_climate  <- crop(annual_temperature_climate , vect(shp_africa))
annual_temperature_climate  <- mask(annual_temperature_climate , vect(shp_africa))
names(annual_temperature_climate) <- "temperature"

africa_water <- crop(land_cover_water, vect(shp_africa))
africa_water <- mask(africa_water, vect(shp_africa))

africa_trees <- crop(land_cover_trees, vect(shp_africa))
africa_trees <- mask(africa_trees, vect(shp_africa))

africa_builts <- crop(land_cover_builts, vect(shp_africa))
africa_builts <- mask(africa_builts, vect(shp_africa))

africa_cropland <- crop(land_cover_cropland, vect(shp_africa))
africa_cropland <- mask(africa_cropland, vect(shp_africa))

africa_wetland <- crop(land_cover_wetland, vect(shp_africa))
africa_wetland <- mask(africa_wetland, vect(shp_africa))

terra::plot(pf_limit_africa)
terra::plot(africa_wetland)
terra::plot(africa_cropland)
terra::plot(africa_builts)
terra::plot(africa_trees)
terra::plot(africa_water)



#asign a value for the class NA

pf_limit_africa2 <- classify(pf_limit_africa,
                             matrix(c(NA, NA, 0), ncol=3, byrow=TRUE))

lev <- data.frame(
  value = c(0, 1, 2),
  category = c("No transmission", "Unstable transmission", "Stable transmission")
)

levels(pf_limit_africa2) <- lev

pf_limit_africa2 <- crop(pf_limit_africa2, vect(shp_africa %>% st_transform(4326)))
pf_limit_africa2 <- mask(pf_limit_africa2, vect(shp_africa %>% st_transform(4326)))

terra::plot(pf_limit_africa2)

names(pf_limit_africa2) <- "Transmission level 2010 of Pf"

cats(pf_limit_africa2)

shp_africa_country <- shp_africa %>% 
  st_make_valid() %>% 
  group_by(country) %>% 
  summarise(geometry = st_union(geometry))

pf_limit_2010 <- ggplot(shp_africa_country)+
  geom_spatraster(data = pf_limit_africa2, mapping = aes(fill = `Transmission level 2010 of Pf`))+
  scale_fill_manual(values = c(
  "No transmission" = "grey80",
  "Unstable transmission" = "orange",
  "Stable transmission" = "red"), na.value =  "transparent", na.translate = FALSE)+
  geom_sf(fill = NA)+
  theme_void()+
  labs(title = "Limit of transmission of P. falciparum in Africa (2010)")

ggsave("Outputs/africa_output/limit_trans_pf.png", 
       plot = pf_limit_2010, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot(shp_africa_country)+
  geom_spatraster(data = precip_ratser_2021_data_annual, mapping = aes(fill = annual_rainfall))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
  colours = brewer.pal(9, "Blues"),
  breaks = c(0, 250, 500, 1000, 1500, 2000, 3000, 5000),
  na.value = "transparent",
  name = "Annual rainfall (mm)")+
  theme_void()+
  labs(title = "annual rainfall in 2021")
  
ggsave("Outputs/africa_output/annual_rainfall_2021_chirps.png", 
       plot = annual_rainfall, width = 6, height = 4, dpi = 300)


annual_temperature_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = annual_temperature_climate, mapping = aes(fill = temperature))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
  colours = rev(brewer.pal(9, "RdYlGn")),
  breaks = c(0, 10, 15, 20, 25, 30, 35),
  na.value = "transparent",
  name = "Annual mean temp (°C)")+
  theme_void()+
  labs(title = "annual temperature in 2021")

ggsave("Outputs/africa_output/annual_temperature_2021_climate.png", 
       plot = annual_temperature_2021, width = 6, height = 4, dpi = 300)


wetland_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = africa_wetland, mapping = aes(fill = wetland))+
  geom_sf(fill = NA)+
  scale_fill_gradientn(
  colours = viridis::magma(10),
  trans = "log10",
  limits = c(1e-6, 1),
  oob = scales::squish,   # 0 to 1e-6
  na.value = "transparent",
  name = "Wetland fraction (log10)")+
  theme_void()+
  labs(title = "Wetland in 2021")

ggsave("Outputs/africa_output/wetland_2021.png", 
       plot = wetland_2021, width = 6, height = 4, dpi = 300)


africa_water_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = africa_water , mapping = aes(fill = water))+
  geom_sf(fill = NA)+
  scale_fill_gradientn(
  colours = viridis::plasma(10),
  trans = "log10",
  limits = c(1e-6, 1),
  oob = scales::squish,
  na.value = "transparent",
  name = "Water fraction (log10)")+
  theme_void()+
  labs(title = "Water in 2021")

ggsave("Outputs/africa_output/africa_water_2021.png", 
       plot = africa_water_2021, width = 6, height = 4, dpi = 300)

africa_trees_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = africa_trees , mapping = aes(fill = trees))+
  geom_sf(fill = NA)+
  scale_fill_gradientn(
  colours = viridis::cividis(10),
  trans = "log10",
  limits = c(1e-6, 1),
  oob = scales::squish,
  na.value = "transparent",
  name = "Tree fraction (log10)")+
  theme_void()+
  labs(title = "Trees in 2021")

ggsave("Outputs/africa_output/africa_trees_2021.png", 
       plot = africa_trees_2021, width = 6, height = 4, dpi = 300)

africa_builts_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = africa_builts , mapping = aes(fill = built))+
  geom_sf(fill = NA)+
  scale_fill_gradientn(
  colours = viridis::rocket(10),
  trans = "log10",
  limits = c(1e-6, 1),
  oob = scales::squish,
  na.value = "transparent",
  name = "Built-up fraction (log10)")+
  theme_void()+
  labs(title = "Built-up in 2021")

ggsave("Outputs/africa_output/africa_builts_2021.png", 
       plot = africa_builts_2021, width = 6, height = 4, dpi = 300)

africa_cropland_2021 <- ggplot(shp_africa_country)+
  geom_spatraster(data = africa_cropland , mapping = aes(fill = cropland))+
  geom_sf(fill = NA)+
  scale_fill_gradientn(
  colours = viridis::inferno(10),
  trans = "log10",
  limits = c(1e-6, 1),
  oob = scales::squish,
  na.value = "transparent",
  name = "Cropland fraction (log10)")+
  theme_void()+
  labs(title = "Cropland in 2021")

ggsave("Outputs/africa_output/africa_cropland_2021.png", 
       plot = africa_cropland_2021, width = 6, height = 4, dpi = 300)

```



# loading covariates and extract them at the sample location Benin

```{r}

ndvi_path <- "data/ben_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)



ITN_intervention_2021 <- rast("data/ITN_2000/2025_GBD2024_Africa_ITN_2021.tif")

ben_pop <- rast("data/ben_data/ben_pd_2020_1km_UNadj.tif")

elevation_ben <- rast("data/ben_data/elevation/BEN_elv_msk.tif")
ben_ndvi_stack<- rast(ndvi_files)

ben_land_cover_water <- crop(land_cover_water, ben_vect)
ben_land_cover_water <- mask(ben_land_cover_water, ben_vect)

ben_land_cover_built <- crop(land_cover_builts, ben_vect)
ben_land_cover_built <- mask(ben_land_cover_built, ben_vect)

ben_land_cover_cropland <- crop(land_cover_cropland, ben_vect)
ben_land_cover_cropland <- mask(ben_land_cover_cropland, ben_vect)

ben_land_cover_trees <- crop(land_cover_trees, ben_vect)
ben_land_cover_trees <- mask(ben_land_cover_trees, ben_vect)

ben_land_cover_wetland <- crop(land_cover_wetland, ben_vect)
ben_land_cover_wetland <- mask(ben_land_cover_wetland, ben_vect)

ben_rainfall_2021 <- crop(rainfall_2021, ben_vect)
ben_rainfall_2021 <- mask(ben_rainfall_2021, ben_vect)

ben_temperature_2021 <- crop(annual_temperature_climate , ben_vect)
ben_temperature_2021 <- mask(ben_temperature_2021, ben_vect)


# Benin case

# create a circle around the GPS coordinates extracted manually to ensure of 
# reasonable extracted value for the population density

village_coords_ben_1 <- village_coords_ben %>% select(Village, lat, lng)

village_coords_ben_1 <- st_as_sf(village_coords_ben_1, coords = c("lng", "lat"),
                                 crs = 4326)
# project to a metric system
village_coords_ben_1_projected <- st_transform(village_coords_ben_1, crs = 32631)

# create 1000 buffer around each village point
# the distance can be adjusted based on the result obtained
village_coords_ben_1_projected_buffer <- st_buffer(village_coords_ben_1_projected,
                                                   dist = 1000)

# project back to match the raster's crs if they differ

village_coords_ben_1_projected_buffer_crs <- st_transform(village_coords_ben_1_projected_buffer,
                                                          crs = crs(ben_pop))

ben_pop_density_2020_df_buffer <- terra::extract(ben_pop, 
                                                 village_coords_ben_1_projected_buffer_crs,
                                                 fun = mean,
                                                 na.rm = TRUE)


ben_ITN_intervention_2021  <- crop(ITN_intervention_2021, ben_vect)
ben_ITN_intervention_2021  <- mask(ben_ITN_intervention_2021, ben_vect)

# ben_pop  <- project(ben_pop,  st_crs(spatial_data_ben)$wkt)
# ben_rainfall_2021 <- project(ben_rainfall_2021, st_crs(spatial_data_ben)$wkt)
# ben_temperature_2021 <- project(ben_temperature_2021, st_crs(spatial_data_ben)$wkt)
# ben_ndvi_stack <- project(ben_ndvi_stack, st_crs(spatial_data_ben)$wkt)
# elevation_ben   <- project(elevation_ben,   st_crs(spatial_data_ben)$wkt)
# 
# ben_land_cover_water   <- project(ben_land_cover_water,   st_crs(spatial_data_ben)$wkt)
# ben_land_cover_built   <- project(ben_land_cover_built,   st_crs(spatial_data_ben)$wkt)
# ben_land_cover_cropland   <- project(ben_land_cover_cropland,   st_crs(spatial_data_ben)$wkt)
# ben_land_cover_trees   <- project(ben_land_cover_trees,   st_crs(spatial_data_ben)$wkt)
# ben_land_cover_wetland   <- project(ben_land_cover_wetland,   st_crs(spatial_data_ben)$wkt)


ben_ITN_intervention_2021_df <- terra::extract(ben_ITN_intervention_2021,  spatial_data_ben)
ben_pop_density_2020_df <- terra::extract(ben_pop,  spatial_data_ben)
ben_rainfall_2021_df    <- terra::extract(ben_rainfall_2021, spatial_data_ben)
ben_temperature_2021_df <- terra::extract(ben_temperature_2021, spatial_data_ben)
ndvi_vals_ben_monthly <- terra::extract(ben_ndvi_stack, spatial_data_ben)
ndvi_mean_ben <- apply(ndvi_vals_ben_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_ben_df <- terra::extract(elevation_ben, spatial_data_ben)
ben_land_cover_water_df <- terra::extract(ben_land_cover_water, spatial_data_ben)
ben_land_cover_built_df <- terra::extract(ben_land_cover_built, spatial_data_ben)
ben_land_cover_cropland_df <- terra::extract(ben_land_cover_cropland, spatial_data_ben)
ben_land_cover_trees_df <- terra::extract(ben_land_cover_trees, spatial_data_ben)
ben_land_cover_wetland_df <- terra::extract(ben_land_cover_wetland, spatial_data_ben)



spatial_data_ben_sf <- spatial_data_ben %>% 
  mutate(pop_density_2020 = ben_pop_density_2020_df[,2],
         annual_rainfall_2021 = ben_rainfall_2021_df[,2],
         annual_temperature_2021 = ben_temperature_2021_df[,2],
         vegetation_index = ndvi_mean_ben,
         elevation = elevation_ben_df[,2],
         lc_water = ben_land_cover_water_df[,2],
         lc_built = ben_land_cover_built_df[,2],
         lc_trees = ben_land_cover_trees_df[,2],
         lc_wetland = ben_land_cover_wetland_df[,2],
         lc_cropland = ben_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = ben_ITN_intervention_2021_df[,2])


# distance to water

bbox <- st_bbox(shp_ben)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)

rivers_valid <- st_transform(rivers_valid, st_crs(shp_ben))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_ben))

# nearest river segment for each point
nn_riv <- st_nn(spatial_data_ben_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(spatial_data_ben_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

spatial_data_ben_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

spatial_data_ben_df <- spatial_data_ben_sf %>% 
  st_drop_geometry() 


missing_rows <- which(!complete.cases(spatial_data_ben_df))
spatial_data_ben_df[missing_rows, ]

# find nearest neighbor for each point
nn <- st_nn(spatial_data_ben_sf, spatial_data_ben_sf, k = 5)   # k=2 because first neighbor is itself

get_first_non_na <- function(values, neighbors) {
  for (i in neighbors) {
    if (!is.na(values[i])) return(values[i])
  }
  return(NA)   
}

var <- "annual_temperature_2021"
values <- spatial_data_ben_df[[var]]

missing <- which(is.na(values))

for (i in missing) {
  spatial_data_ben_df[[var]][i] <- get_first_non_na(values, nn[[i]])
}

spatial_data_ben_sf$annual_temperature_2021 <- spatial_data_ben_df$annual_temperature_2021

st_write(spatial_data_ben_sf, "data/ben_data/spatial_data_ben_sf.gpkg", append = FALSE)


# Congo case

Cong_pop <- rast("data/cong_data/cog_pd_2020_1km_UNadj.tif")

elevation_Cong <- rast("data/cong_data/elevation/COG_elv_msk.tif")
ndvi_path <- "data/cong_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

Cong_ndvi_stack <- rast(ndvi_files)

# create a circle around the GPS coordinates extracted manually to ensure of 
# reasonable extracted value for the population density

village_coords_cong_1 <- village_coords_cong %>% select(District, lat, lng)

village_coords_cong_1 <- st_as_sf(village_coords_cong_1, coords = c("lng", "lat"),
                                 crs = 4326)
# project to a metric system

village_coords_cong_1_projected <- st_transform(village_coords_cong_1, crs = 32631)

# create 1000 buffer around each village point
# the distance can be adjusted based on the result obtained

village_coords_cong_1_projected_buffer <- st_buffer(village_coords_cong_1_projected,
                                                   dist = 3000)

# project back to match the raster's crs if they differ

village_coords_cong_1_projected_buffer_crs <- st_transform(village_coords_cong_1_projected_buffer,
                                                          crs = crs(Cong_pop))
cong_pop_density_2020_df_buffer <- terra::extract(Cong_pop, 
                                                 village_coords_cong_1_projected_buffer_crs,
                                                 fun = mean,
                                                 na.rm = TRUE)

Cong_land_cover_water <- crop(land_cover_water, con_vect)
Cong_land_cover_water <- mask(cmr_land_cover_water, con_vect)

Cong_land_cover_built <- crop(land_cover_builts, con_vect)
Cong_land_cover_built <- mask(cmr_land_cover_built, con_vect)

Cong_land_cover_cropland <- crop(land_cover_cropland, con_vect)
Cong_land_cover_cropland <- mask(cmr_land_cover_cropland, con_vect)

Cong_land_cover_trees <- crop(land_cover_trees, con_vect)
Cong_land_cover_trees <- mask(cmr_land_cover_trees, con_vect)

Cong_land_cover_wetland <- crop(land_cover_wetland, con_vect)
Cong_land_cover_wetland <- mask(cmr_land_cover_wetland, con_vect)


Cong_rainfall_2021 <- crop(rainfall_2021, con_vect)
Cong_rainfall_2021 <- mask(Cong_rainfall_2021, con_vect)

Cong_temperature_2021 <- crop(annual_temperature_climate , con_vect)
Cong_temperature_2021 <- mask(Cong_temperature_2021, con_vect)


Cong_ITN_intervention_2021  <- crop(ITN_intervention_2021,  con_vect)
Cong_ITN_intervention_2021  <- mask(ITN_intervention_2021,  con_vect)

# Cong_pop  <- project(Cong_pop,  st_crs(spatial_data_cong)$wkt)
# Cong_rainfall_2021 <- project(ben_rainfall_2021, st_crs(spatial_data_cong)$wkt)
# Cong_temperature_2021 <- project(ben_temperature_2021, st_crs(spatial_data_cong)$wkt)
# Cong_ndvi_stack <- project(Cong_ndvi_stack, st_crs(spatial_data_cong)$wkt)
# elevation_Cong   <- project(elevation_Cong,   st_crs(spatial_data_cong)$wkt)
# Cong_land_cover_water   <- project(Cong_land_cover_water,   st_crs(spatial_data_cong)$wkt)
# Cong_land_cover_built   <- project(Cong_land_cover_built,   st_crs(spatial_data_cong)$wkt)
# Cong_land_cover_cropland   <- project(Cong_land_cover_cropland,   st_crs(spatial_data_cong)$wkt)
# Cong_land_cover_trees   <- project(Cong_land_cover_trees,   st_crs(spatial_data_cong)$wkt)
# Cong_land_cover_wetland   <- project(Cong_land_cover_wetland,   st_crs(spatial_data_cong)$wkt)

Cong_ITN_intervention_2021_df <- terra::extract(Cong_ITN_intervention_2021,  spatial_data_cong)
Cong_pop_density_2020_df <- terra::extract(Cong_pop,  spatial_data_cong)
Cong_rainfall_2021_df    <- terra::extract(Cong_rainfall_2021, spatial_data_cong)
Cong_temperature_2021_df <- terra::extract(Cong_temperature_2021, spatial_data_cong)
ndvi_vals_Cong_monthly <- terra::extract(Cong_ndvi_stack, spatial_data_cong)
ndvi_mean_Cong <- apply(ndvi_vals_Cong_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_Cong_df <- terra::extract(elevation_Cong, spatial_data_cong)
Cong_land_cover_water_df <- terra::extract(Cong_land_cover_water, spatial_data_cong)
Cong_land_cover_built_df <- terra::extract(Cong_land_cover_built, spatial_data_cong)
Cong_land_cover_cropland_df <- terra::extract(Cong_land_cover_cropland, spatial_data_cong)
Cong_land_cover_trees_df <- terra::extract(Cong_land_cover_trees, spatial_data_cong)
Cong_land_cover_wetland_df <- terra::extract(Cong_land_cover_wetland, spatial_data_cong)

shp_cong_country <- shp_con %>% 
  group_by(adm0_name) %>% 
  summarise(geometry = st_union(geometry))


bbox <- st_bbox(shp_con)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)

rivers_valid <- st_transform(rivers_valid, st_crs(shp_con))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_con))


# nearest river segment for each point
nn_riv <- st_nn(spatial_data_cong, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(spatial_data_cong, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

spatial_data_cong$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

spatial_data_cong <- spatial_data_cong %>% 
  mutate(pop_density_2020 = cong_pop_density_2020_df_buffer[,2],
         annual_rainfall_2021 = Cong_rainfall_2021_df[,2],
         annual_temperature_2021 = Cong_temperature_2021_df[,2],
         vegetation_index = ndvi_mean_Cong,
         elevation = elevation_Cong_df[,2],
         lc_water = Cong_land_cover_water_df[,2],
         lc_built = Cong_land_cover_built_df[,2],
         lc_trees = Cong_land_cover_trees_df[,2],
         lc_wetland = Cong_land_cover_wetland_df[,2],
         lc_cropland = Cong_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = Cong_ITN_intervention_2021_df[,2])

spatial_data_cong_df <- spatial_data_cong %>% 
  st_drop_geometry() 

st_write(spatial_data_cong, "data/cong_data/spatial_data_cong_sf.gpkg", append = FALSE)


```


```{r}
# Cameroon case

cmr_pop <- rast("data/cmr_data/cmr_pd_2011_1km_UNadj.tif")

elevation_cmr <- rast("data/cmr_data/elevation/CMR_elv_msk.tif")
ndvi_path <- "data/cmr_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

cmr_ndvi_stack <- rast(ndvi_files)

cmr_land_cover_water <- crop(land_cover_water, cmr_vect)
cmr_land_cover_water <- mask(cmr_land_cover_water, cmr_vect)

cmr_land_cover_built <- crop(land_cover_builts, cmr_vect)
cmr_land_cover_built <- mask(cmr_land_cover_built, cmr_vect)

cmr_land_cover_cropland <- crop(land_cover_cropland, cmr_vect)
cmr_land_cover_cropland <- mask(cmr_land_cover_cropland, cmr_vect)

cmr_land_cover_trees <- crop(land_cover_trees, cmr_vect)
cmr_land_cover_trees <- mask(cmr_land_cover_trees, cmr_vect)

cmr_land_cover_wetland <- crop(land_cover_wetland, cmr_vect)
cmr_land_cover_wetland <- mask(cmr_land_cover_wetland, cmr_vect)

cmr_rainfall_2021 <- crop(rainfall_2021, cmr_vect)
cmr_rainfall_2021 <- mask(cmr_rainfall_2021, cmr_vect)

cmr_temperature_2021 <- crop(annual_temperature_climate , cmr_vect)
cmr_temperature_2021 <- mask(cmr_temperature_2021, cmr_vect)



cmr_ITN_intervention_2021  <- crop(ITN_intervention_2021,  cmr_vect)
cmr_ITN_intervention_2021  <- mask(ITN_intervention_2021,  cmr_vect)

# cmr_pop  <- project(cmr_pop,  st_crs(CMR_df_1_sf)$wkt)
# cmr_rainfall_2021 <- project(ben_rainfall_2021, st_crs(CMR_df_1_sf)$wkt)
# cmr_temperature_2021 <- project(ben_temperature_2021, st_crs(CMR_df_1_sf)$wkt)
# cmr_ndvi_stack <- project(cmr_ndvi_stack, st_crs(CMR_df_1_sf)$wkt)
# elevation_cmr   <- project(elevation_cmr,   st_crs(CMR_df_1_sf)$wkt)
# cmr_land_cover_water   <- project(cmr_land_cover_water,   st_crs(CMR_df_1_sf)$wkt)
# cmr_land_cover_built   <- project(cmr_land_cover_built,   st_crs(CMR_df_1_sf)$wkt)
# cmr_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(CMR_df_1_sf)$wkt)
# cmr_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(CMR_df_1_sf)$wkt)
# cmr_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(CMR_df_1_sf)$wkt)

cmr_ITN_intervention_2021_df <- terra::extract(cmr_ITN_intervention_2021,  CMR_df_1_sf)
cmr_pop_density_2020_df <- terra::extract(cmr_pop,  CMR_df_1_sf)
cmr_rainfall_2021_df    <- terra::extract(cmr_rainfall_2021, CMR_df_1_sf)
cmr_temperature_2021_df <- terra::extract(cmr_temperature_2021, CMR_df_1_sf)
ndvi_vals_cmr_monthly <- terra::extract(cmr_ndvi_stack, CMR_df_1_sf)
ndvi_mean_cmr <- apply(ndvi_vals_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cmr_df <- terra::extract(elevation_cmr, CMR_df_1_sf)
cmr_land_cover_water_df <- terra::extract(cmr_land_cover_water, CMR_df_1_sf)
cmr_land_cover_built_df <- terra::extract(cmr_land_cover_built, CMR_df_1_sf)
cmr_land_cover_cropland_df <- terra::extract(cmr_land_cover_cropland, CMR_df_1_sf)
cmr_land_cover_trees_df <- terra::extract(cmr_land_cover_trees, CMR_df_1_sf)
cmr_land_cover_wetland_df <- terra::extract(cmr_land_cover_wetland, CMR_df_1_sf)

shp_cmr_country <- shp_cmr %>% 
  mutate(country = "Cameroon")%>% 
  group_by(country) %>% 
  summarise(geometry = st_union(geometry))


bbox <- st_bbox(shp_cmr)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)


rivers_valid <- st_transform(rivers_valid, st_crs(shp_cmr))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_cmr))


# nearest river segment for each point
nn_riv <- st_nn(CMR_df_1_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(CMR_df_1_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

CMR_df_1_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

CMR_df_1_sf <- CMR_df_1_sf %>% 
  mutate(pop_density_2020 = cmr_pop_density_2020_df[,2],
         annual_rainfall_2021 = cmr_rainfall_2021_df[,2],
         annual_temperature_2021 = cmr_temperature_2021_df[,2],
         vegetation_index = ndvi_mean_cmr,
         elevation = elevation_cmr_df[,2],
         lc_water = cmr_land_cover_water_df[,2],
         lc_built = cmr_land_cover_built_df[,2],
         lc_trees = cmr_land_cover_trees_df[,2],
         lc_wetland = cmr_land_cover_wetland_df[,2],
         lc_cropland = cmr_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = cmr_ITN_intervention_2021_df[,2])

CMR_df_1_sf_df <- CMR_df_1_sf %>% 
  st_drop_geometry() 


st_write(CMR_df_1_sf, "data/cmr_data/CMR_df_1_sf.gpkg", append = FALSE)


# Cameroon case with clusters of households

# cmr_cluster_pop  <- project(cmr_pop,  st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_rainfall_2021 <- project(ben_rainfall_2021, st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_temperature_2021 <- project(ben_temperature_2021, st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_ndvi_stack <- project(cmr_ndvi_stack, st_crs(cluster_data_cmr_sf)$wkt)
# elevation_cluster_cmr   <- project(elevation_cmr,   st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_land_cover_water   <- project(cmr_land_cover_water,   st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_land_cover_built   <- project(cmr_land_cover_built,   st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(cluster_data_cmr_sf)$wkt)
# cmr_cluster_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(cluster_data_cmr_sf)$wkt)

cmr_cluster_ITN_intervention_2021_df <- terra::extract(cmr_ITN_intervention_2021,
                                                       cluster_data_cmr_sf)
cmr_cluster_pop_density_2020_df <- terra::extract(cmr_pop,  cluster_data_cmr_sf)
cmr_cluster_rainfall_2021_df    <- terra::extract(cmr_rainfall_2021, cluster_data_cmr_sf)
cmr_cluster_temperature_2021_df <- terra::extract(cmr_temperature_2021, cluster_data_cmr_sf)
ndvi_vals_cluster_cmr_monthly <- terra::extract(cmr_ndvi_stack, cluster_data_cmr_sf)
ndvi_mean_cluster_cmr <- apply(ndvi_vals_cluster_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cluster_cmr_df <- terra::extract(elevation_cmr, cluster_data_cmr_sf)
cmr_cluster_land_cover_water_df <- terra::extract(cmr_land_cover_water, cluster_data_cmr_sf)
cmr_cluster_land_cover_built_df <- terra::extract(cmr_land_cover_built, cluster_data_cmr_sf)
cmr_cluster_land_cover_cropland_df <- terra::extract(cmr_land_cover_cropland, cluster_data_cmr_sf)
cmr_cluster_land_cover_trees_df <- terra::extract(cmr_land_cover_trees, cluster_data_cmr_sf)
cmr_cluster_land_cover_wetland_df <- terra::extract(cmr_land_cover_wetland, cluster_data_cmr_sf)


# nearest river segment for each point
nn_riv <- st_nn(cluster_data_cmr_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(cluster_data_cmr_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

cluster_data_cmr_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

cluster_data_cmr_sf <- cluster_data_cmr_sf %>% 
  mutate(pop_density_2020 = cmr_cluster_pop_density_2020_df[,2],
         annual_rainfall_2021 = cmr_cluster_rainfall_2021_df[,2],
         annual_temperature_2021 = cmr_cluster_temperature_2021_df[,2],
         vegetation_index = ndvi_mean_cluster_cmr,
         elevation = elevation_cluster_cmr_df[,2],
         lc_water = cmr_cluster_land_cover_water_df[,2],
         lc_built = cmr_cluster_land_cover_built_df[,2],
         lc_trees = cmr_cluster_land_cover_trees_df[,2],
         lc_wetland = cmr_cluster_land_cover_wetland_df[,2],
         lc_cropland = cmr_cluster_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = cmr_cluster_ITN_intervention_2021_df[,2]) %>% 
  dplyr::select(-c(grid_id, lat, lng))


st_write(cluster_data_cmr_sf, "data/cmr_data/cmr_cluster_df_1_sf.gpkg", append = FALSE)

```


# visualization of covariate that shows bigger variation within the sampled village
```{r}
# Benin

spatial_data_ben_sf <- st_read("data/ben_data/spatial_data_ben_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Benin)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/dist_water_viz_ben.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Benin)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/pop_density_viz_ben.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Benin)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/annual_rainfall_viz_ben.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Benin)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/annual_temperature_viz_ben.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Benin)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Vegetation_index_viz_ben.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)

elevation_ben <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Benin)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/elevation_viz_ben.pdf", plot = elevation_ben, width = 6, height = 4, dpi = 300)



```


```{r}
# Congo

spatial_data_cong <- st_read("data/cong_data/spatial_data_cong_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Congo)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/dist_water_viz_cong.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Congo)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/pop_density_viz_cong.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Congo)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/annual_rainfall_viz_cong.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Congo)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/annual_temperature_vizcong.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Congo)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Vegetation_index_viz_cong.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)


elevation_cong <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Congo)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/elevation_viz_cong.pdf", plot = elevation_cong, width = 6, height = 4, dpi = 300)


```


```{r}
# Cameroon

spatial_data_cmr <- st_read("data/cmr_data/cmr_cluster_df_1_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Cameroon)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/dist_water_viz_cmr.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Cameroon)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/pop_density_viz_cmr.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Cameroon)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/annual_rainfall_viz_cmr.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Cameroon)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/annual_temperature_viz_cmr.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Cameroon)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Vegetation_index_viz_cmr.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)


elevation_cmr <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Cameroon)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/elevation_viz_cmr.pdf", plot = elevation_cmr, width = 6, height = 4, dpi = 300)


```



```{r}
# Data for modelling

sampled_data_join_sf <- spatial_data_ben_sf %>% dplyr::select(-c(Village)) %>%
  mutate(country = "Benin") %>% 
  bind_rows(spatial_data_cong %>% dplyr::select(-c(District)) %>% 
              mutate(country = "Congo")) %>% 
    bind_rows(spatial_data_cmr) 



numer_var_ben <- spatial_data_ben_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "annual_ITN_usage_2021",
  "elevation", "vegetaion_index", "pop_density_2020", "dist_to_water", "lc_water",
  "lc_trees", "lc_built", "lc_cropland", "lc_wetland"))



Ben <- cor(numer_var_ben, use = "pairwise.complete.obs")
corrplot(
  Ben,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


numer_var_cong <- spatial_data_cong %>% 
  st_drop_geometry() %>% 
  select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

Cong <- cor(numer_var_cong, use = "pairwise.complete.obs")
corrplot(
  Cong,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

```



```{r}
num_vars <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees", "annual_ITN_usage_2021",
  "lc_built", "lc_cropland", "lc_wetland"))

num_vars_1 <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c("dist_to_water", "PR_Pm_tot"))

M1 <- cor(num_vars_1, use = "pairwise.complete.obs")

corrplot(
  M1,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

var1 <- ggcorrplot(
  M,
  type = "full",
  legend.title = "Correlation Matrix",
  lab = TRUE,
  lab_size = 3,
  colors = c("blue", "white", "red"),
  outline.color = "white",
  show.diag = TRUE       
) +
  scale_x_discrete(name = NULL) +
  scale_y_discrete(name = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),      
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Outputs/correlation_matrix.png", var1, dpi = 300, width = 6, height = 4)

num_vars_selected <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

M_selected <- cor(num_vars_selected, use = "pairwise.complete.obs")

corrplot(
  M_selected,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + 
    annual_temperature_2021 + elevation + vegetaion_index + pop_density_2020 + 
    dist_to_water + lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021,
  data = sampled_data_join_sf
)

vif(fit)

fit1 <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water
  + lc_trees + lc_built + lc_cropland + lc_wetland , 
  data = sampled_data_join_sf
)

vif(fit1)

```

# variogram, autocorrelation and prior ajustment

```{r}
# Project to meters (important!)
sf_data <- st_transform(sampled_data_join_sf, 3857)


sf_data$logit_prev <- qlogis(
  pmin(pmax(sf_data$PR_Pm_tot, 1e-6), 1 - 1e-6)
)
# Extract coordinates
coords <- st_coordinates(sf_data)

# Build gstat object
g <- gstat(id = "pm", formula = logit_prev ~ 1, data = sf_data)

# Empirical variogram

v <- variogram(g,  cutoff = 100000, width = 1000)  # adjust cutoff to country size
# fit an exponential model
v.fit <- fit.variogram(v, model = vgm("Sph"))

# plot empirical + fitted curve
plot(v, v.fit)

plot(v)

```

```{r}
fit_cov <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 +
    dist_to_water + lc_water + lc_trees + lc_built +
    lc_cropland + lc_wetland,
  data = sampled_data_join_sf
)

sampled_data_join_sf$resid <- residuals(fit_cov)


sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

# neighbors within, say, 50 km
nb <- dnearneigh(coords, d1 = 0, d2 = 50000)
lw <- nb2listw(nb, style = "W")

moran.test(sf_data$resid, lw)


```

The observed Moran’s I is almost identical to the expected value under randomness.

The p‑value is very high : no evidence of spatial autocorrelation.

The negative sign is tiny : not meaningful.


```{r}
sampled_data_join_sf$country_id <- as.integer(factor(sampled_data_join_sf$country))  # 1, 2, 3

sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

coords_A <- coords[sf_data$country_id == 1, ]
coords_B <- coords[sf_data$country_id == 2, ]
coords_C <- coords[sf_data$country_id == 3, ]

country_A <- st_transform(shp_ben, 3857)
country_B <- st_transform(shp_cmr, 3857)
country_C <- st_transform(shp_con, 3857)

country_A_union <- st_union(country_A)
country_B_union <- st_union(country_B)
country_C_union <- st_union(country_C)

country_A_simplified <- st_simplify(country_A_union, dTolerance = 10000)
country_B_simplified <- st_simplify(country_B_union, dTolerance = 10000)
country_C_simplified <- st_simplify(country_C_union, dTolerance = 10000)


# Extract geometry
geom_A <- st_geometry(country_A_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_A, "MULTIPOLYGON")) {
  geom_A <- geom_A[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_A <- geom_A[[1]]

# Convert to matrix with X,Y
boundary_A <- as.matrix(outer_ring_A)[, 1:2, drop = FALSE]



# country_B is an sf polygon or multipolygon
geom_B <- st_geometry(country_B_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_B, "MULTIPOLYGON")) {
  geom_B <- geom_B[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_B <- geom_B[[1]]

# Convert to matrix with X,Y
boundary_B <- as.matrix(outer_ring_B)[, 1:2, drop = FALSE]

# Extract geometry
geom_C <- st_geometry(country_C_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_C, "MULTIPOLYGON")) {
  geom_C <- geom_C[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_C <- geom_C[[1]]

# Convert to matrix with X,Y
boundary_C <- as.matrix(outer_ring_C)[, 1:2, drop = FALSE]


# mesh_A <- inla.mesh.2d(loc = coords_A, max.edge = c(5000, 20000))
# mesh_B <- inla.mesh.2d(loc = coords_B, max.edge = c(5000, 20000))
# mesh_C <- inla.mesh.2d(loc = coords_C, max.edge = c(5000, 20000))

mesh_A <- inla.mesh.2d(
  loc      = coords_A,
  boundary = list(boundary_A),
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_B <- inla.mesh.2d(
  loc = coords_B,
  boundary   = list(boundary_B) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_C <- inla.mesh.2d(
  loc = coords_C,
  boundary   = list(boundary_C) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

# # Extract triangle vertex indices
# tri <- mesh_C$graph$tv
# 
# 
# # Build edges from triangles
# edges_list <- list()
# 
# for (i in 1:nrow(tri)) {
#   idx <- tri[i, ]
#   # Each triangle has 3 edges
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(1,2)], 1:2])
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(2,3)], 1:2])
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(3,1)], 1:2])
# }
# 
# mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))
# 
# coords_df_C <- as.data.frame(coords_C)
# colnames(coords_df_C) <- c("X", "Y")
# 
# ggplot() +
#   geom_sf(data = country_C_simplified, fill = "white", color = "black", size = 0.8) +
#   geom_sf(data = mesh_edges_sf, color = "grey60", size = 0.3) +
#   geom_point(data = coords_df_C, aes(X, Y), color = "red", size = 1.5) +
#   theme_minimal() +
#   ggtitle("Mesh A with Country Boundary")

```


```{r}
# Build Matérn SPDE model of the spatial field on mesh

spde_A <- inla.spde2.pcmatern(
  mesh = mesh_A,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)
)

spde_B <- inla.spde2.pcmatern(
  mesh = mesh_B,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)   # 0.12 came from the squared root of the sill because it 
)                              # represent the standard deviance while the sill represent the variance.

spde_C <- inla.spde2.pcmatern(
  mesh = mesh_C,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)   
) 

# Computes the interpolation weights that map your SPDE spatial field (defined on mesh nodes) to your actual malaria data locations

# 1. Create indicator weights (1 if the row belongs to the country, 0 if not)
# 'coords' must be the full 36-row matrix of ALL coordinates
w_A <- as.numeric(sf_data$country_id == 1)
w_B <- as.numeric(sf_data$country_id == 2)
w_C <- as.numeric(sf_data$country_id == 3)

# 2. Build the full matrices automatically. 
# Passing the full 'coords' matrix ensures every matrix has exactly 36 rows.
# The 'weights' argument forces rows from other countries to be completely zero.
A_A_full <- inla.spde.make.A(mesh = mesh_A, loc = coords, weights = w_A)
A_B_full <- inla.spde.make.A(mesh = mesh_B, loc = coords, weights = w_B)
A_C_full <- inla.spde.make.A(mesh = mesh_C, loc = coords, weights = w_C)

A_fixed <- Diagonal(n = nrow(sf_data))
A_spatial <- list(A_A_full, A_B_full, A_C_full)


A_A_full <- as(A_A_full, "dgCMatrix")
A_B_full <- as(A_B_full, "dgCMatrix")
A_C_full <- as(A_C_full, "dgCMatrix")

```

# Build stack

```{r}

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "vegetation_index",
                   "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sf_data[[var]], na.rm = TRUE)
  sd_val   <- sd(sf_data[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sf_data[[var]] <- (sf_data[[var]] - mean_val) / sd_val
}

################################################################################
#        checking for overdispersion in our model
################################################################################

# 1. Calculate observed prevalence for each of the 36 villages
village_p <- sf_data$Number_of_infected_tot_pf / sf_data$Number_of_participant_tot

# 2. Calculate the global overall prevalence
global_p <- sum(sf_data$Number_of_infected_tot_pf) / sum(sf_data$Number_of_participant_tot)

# 3. Get the average number of participants per village
mean_N <- mean(sf_data$Number_of_participant_tot)

# 4. Calculate the Observed vs. Theoretical Variance
observed_variance    <- var(village_p)
theoretical_variance <- (global_p * (1 - global_p)) / mean_N

# 5. Compute the Raw Dispersion Ratio
raw_dispersion_ratio <- observed_variance / theoretical_variance
print(raw_dispersion_ratio)

# sf_data <- sf_data %>% 
#   mutate(pop_density_2020 = as.numeric(scale(pop_density_2020)),
#          annual_rainfall_2021 = as.numeric(scale(annual_rainfall_2021)),
#          annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
#          vegetaion_index = as.numeric(scale(vegetaion_index)),
#          dist_to_water  = as.numeric(scale(dist_to_water)),
#          elevation = as.numeric(scale(elevation)))

# Model building

effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sf_data$annual_rainfall_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept             = 1,
  annual_temperature_2021  = sf_data$annual_temperature_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetation_index      = sf_data$vegetation_index,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sf_data$elevation,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_country <- list(
  country_id = sf_data$country_id
)

effects_A <- list(field_A = 1:spde_A$n.spde)
effects_B <- list(field_B = 1:spde_B$n.spde)
effects_C <- list(field_C = 1:spde_C$n.spde)

A_fixed   <- Diagonal(nrow(sf_data))
A_country <- Diagonal(nrow(sf_data))


stack_test_rainfall <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_rainfall,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_temperature <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_temperature,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_vegetation_index <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_vegetation_index,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_elevation <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_elevation,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_baseline <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_baseline,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

```


# Fit our models

```{r}
# Penalize complexity which control how much countries differ from each other through
# the country level random effect
hyper_country <- list(
  prec = list(prior = "pc.prec", param = c(0.5, 0.5))
)

formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetation_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stack_test_rainfall),
  Ntrials = inla.stack.data(stack_test_rainfall)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_rainfall),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stack_test_temperature),
  Ntrials = inla.stack.data(stack_test_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stack_test_vegetation_index),
  Ntrials = inla.stack.data(stack_test_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stack_test_elevation),
  Ntrials = inla.stack.data(stack_test_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stack_test_baseline),
  Ntrials = inla.stack.data(stack_test_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

# ## diasgnostic on the distance to water
# 
# res_simple <- inla(
#   Number_of_infected_tot_pm ~ dist_to_water,
#   family = "binomial",
#   data = sf_data,
#   Ntrials = sf_data$Number_of_participant_tot
# )
# 
# res_simple$summary.fixed


```


The model’s hyperparameters show that country-level differences are almost negligible, meaning the covariates already explain most variation between Benin and Congo. The spatial field in Benin is clearly identifiable: it has a short correlation range (~7 km) and a moderate spatial variance, so spatial structure genuinely improves the fit there. In contrast, Congo’s field is extremely uncertain—with only three locations, the model cannot reliably estimate its spatial range or variance, so any spatial effect in Congo should be interpreted cautiously. The practical consequence is that predictions and spatial maps will be trustworthy in Benin, while Congo spatial surface will be weak, noisy, and driven mostly by priors rather than data. 


# Comparing the different models

```{r}
# Model 1:  Rainfall

# estimated values for the different covariates

res_rainfall$summary.fixed

# Hyper parameters examination

res_rainfall$summary.hyperpar

# DIC

res_rainfall$dic$dic

# WAIC

res_rainfall$waic$waic

# CPO

sum(log(res_rainfall$cpo$cpo), na.rm = TRUE)

summary(res_rainfall$cpo$cpo)

# PIT(probability integral transform) histogram

hist(res_rainfall$cpo$pit,
     breaks = 20,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

obs <- inla.stack.data(stack_test_rainfall)$y /
       inla.stack.data(stack_test_rainfall)$Ntrials

idx.est <- inla.stack.index(stack_test_rainfall, "est")$data

pred.est <- res_rainfall$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, rainfall model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual rainfall model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 2:  temperature

# estimated values for the different covariates

res_temperature$summary.fixed

# Hyper parameters examination

res_temperature$summary.hyperpar

# DIC

res_temperature$dic$dic

# WAIC

res_temperature$waic$waic

# CPO

sum(log(res_temperature$cpo$cpo), na.rm = TRUE)

summary(res_temperature$cpo$cpo)

# PIT histogram

hist(res_temperature$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_temperature, "est")$data

pred.est <- res_temperature$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title =  "Observed vs predicted, temperature model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual temperature model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```

```{r}
# Model 3:  vegetation index

# estimated values for the different covariates

res_vegetation_index$summary.fixed

# Hyper parameters examination

res_vegetation_index$summary.hyperpar

# DIC

res_vegetation_index$dic$dic

# WAIC

res_vegetation_index$waic$waic

# CPO

sum(log(res_vegetation_index$cpo$cpo), na.rm = TRUE)

summary(res_vegetation_index$cpo$cpo)

# PIT histogram

hist(res_vegetation_index$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_vegetation_index, "est")$data

pred.est <- res_vegetation_index$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, vegetation index model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual vegetation index model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 4:  elevation

# estimated values for the different covariates

res_elevation$summary.fixed

# Hyper parameters examination

res_elevation$summary.hyperpar

# DIC

res_elevation$dic$dic

# WAIC

res_elevation$waic$waic

# CPO

sum(log(res_elevation$cpo$cpo), na.rm = TRUE)

summary(res_elevation$cpo$cpo)

# PIT histogram

hist(res_elevation$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_elevation, "est")$data

pred.est <- res_elevation$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, elevation model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual elevation model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```


```{r}
# Model 5:  baseline model

# estimated values for the different covariates

res_baseline$summary.fixed

# Hyper parameters examination

res_baseline$summary.hyperpar

# DIC

res_baseline$dic$dic

# WAIC

res_baseline$waic$waic

# CPO

sum(log(res_baseline$cpo$cpo), na.rm = TRUE)

summary(res_baseline$cpo$cpo)

# PIT histogram

hist(res_baseline$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_baseline, "est")$data

pred.est <- res_baseline$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, baseline model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual baseline model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```
## Conclusion on models comparison (P. malaria model)

The rainfall model is the right choice because it consistently shows the strongest performance across all diagnostics. It has the lowest DIC and WAIC, the best CPO, the tightest observed‑vs‑predicted fit, and it reduces the spatial field variance more than any other covariate, meaning rainfall explains a large part of the spatial structure instead of leaving it to the latent field. In short, rainfall improves prediction, reduces unexplained spatial variation, and aligns with malaria biology making it the most reliable and scientifically justified model to continue with.

```{r}
idx_est <- inla.stack.index(stack_test_temperature, "est")$data

fitted_est <- res_temperature$summary.fitted.values[idx_est, "mean"]

fitted_est_25 <- res_temperature$summary.fitted.values[idx_est, "0.025quant"]
fitted_est_75 <- res_temperature$summary.fitted.values[idx_est, "0.975quant"]


fitted_data <- cbind(
  id        = idx_est, 
  observed  = sf_data$PR_Pm_tot,
  fitted    = fitted_est,
  fitted_25 = fitted_est_25,
  fitted_75 = fitted_est_75
)

fitted_data <- as.data.frame(fitted_data)
```

# Model examination
```{r}
# observed vs fitted

ggplot(fitted_data, aes(x = observed, y = fitted)) +
  geom_point(color = "blue", size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "red", lwd = 1.2) +
  labs(x = "Observed proportion",
       y = "Fitted prevalence",
       title = "Observed vs Fitted (INLA)") +
  theme_minimal()

# residual

fitted_data$residuals <- fitted_data$observed - fitted_data$fitted

ggplot(fitted_data, aes(x = 1:nrow(fitted_data), y = residuals)) +
  geom_point(color = "darkgreen", size = 2) +
  geom_hline(yintercept = 0, color = "red", lwd = 1.2) +
  labs(x = "Location index",
       y = "Residual (observed - fitted)",
       title = "Residuals of INLA fitted values") +
  theme_minimal()

# SPDE hyper parameters

res_temperature$summary.hyperpar


```

## Pf model examination
- The observed vs fitted values plot shows that the smooth, shrunk estimates of the 
true prevalence, which are generally consistent with the observed but less extreme.
the number of points above and below the red line are roughly the same which means
the model is not over estimating nor under estimating.

- The residual plot shows a rough horizontal cloud indicating that the model is 
stable and most residuals are between -0.1 and 0.1 meaning that the model fit well.
we also have some moderate residuals around 0.3 and -0.3 meaning the model smooth 
appropriately. Finally, the number of point above and below the horizontal line 
are almost the same, which means the model is unbiased. 

- The country_id precision is approximately ~15 which means there is some 
meaningful difference between countries after accounting for covariates + spatial 
fields.

- For all the three field, the range is ~7km to ~9km with a very wide credible interval,
meaning areas withing these distances are spatially correlated or influence each 
other but the model is unsure to which extend because the uncertainty is high. 
This means that the prediction will be smooth but not overly flat.

- All the three spatial field also have a very small standard deviation 0.021 to
0.027, which means spatial variation exist but covariates explain most of the 
variation and spatial field refine the prediction instead of dominating it.

# Creating the grid for prevalence estimation

```{r}
countries1_sf <- rbind(shp_ben %>% dplyr::select(COUNTRY, NAME_3, geometry) %>% 
                        rename(country = COUNTRY,
                               district = NAME_3), shp_con %>% dplyr::select(adm0_name,
                                                                      adm2_name, geometry) %>% 
                        rename(country = adm0_name,
                               district = adm2_name)) 


countries_sf <- rbind(countries1_sf, shp_cmr %>% dplyr::select(Admin2, geometry) %>% 
                        mutate(country = "Cameroon") %>% 
                        rename(district = Admin2))

countries_sf1 <- rbind(countries1_sf, shp_cmr %>% dplyr::select(Admin2, geometry) %>% 
                        mutate(country = "Cameroon") %>% 
                        rename(district = Admin2))

countries_sf1$country_id <- as.integer(factor(countries_sf1$country))


# 1. Reproject countries to a metric CRS
countries_m <- st_transform(countries_sf1, 3857)

# 2. Create bbox in the same CRS
bbox_m <- st_bbox(countries_m) %>% st_as_sfc()

# 3. Build 5 km grid (now 5000 means 5000 meters)
grid_5km <- st_make_grid(
  bbox_m,
  cellsize = 5000,
  what = "centers"
)

grid_sf <- st_sf(geometry = grid_5km)

# 4. Intersect
grid_sf <- st_intersection(grid_sf, countries_m)

grid_vect <- vect(countries_sf %>% st_transform(4326))
```



# Extraction of the covariates on the grid cells
```{r}
# Benin case

grid_pop <- raster::merge(ben_pop, Cong_pop, cmr_pop)
grid_elevation <- raster::merge(elevation_ben, elevation_cmr, elevation_Cong)
grid_NDVI <- raster::merge(ben_ndvi_stack, cmr_ndvi_stack, Cong_ndvi_stack)

# land_cover_water <- terra::project(land_cover_water, st_crs(grid_sf)$wkt)
# land_cover_builts <- terra::project(land_cover_builts, st_crs(grid_sf)$wkt)
# land_cover_cropland <- terra::project(land_cover_cropland, st_crs(grid_sf)$wkt)
# land_cover_trees <- terra::project(land_cover_trees, st_crs(grid_sf)$wkt)
# land_cover_wetland <- terra::project(land_cover_wetland, st_crs(grid_sf)$wkt)

grid_land_cover_water <- crop(land_cover_water, grid_vect)
grid_land_cover_water <- mask(grid_land_cover_water, grid_vect)

grid_land_cover_built <- crop(land_cover_builts, grid_vect)
grid_land_cover_built <- mask(grid_land_cover_built, grid_vect)

grid_land_cover_cropland <- crop(land_cover_cropland, grid_vect)
grid_land_cover_cropland <- mask(grid_land_cover_cropland, grid_vect)

grid_land_cover_trees <- crop(land_cover_trees, grid_vect)
grid_land_cover_trees <- mask(grid_land_cover_trees, grid_vect)

grid_land_cover_wetland <- crop(land_cover_wetland, grid_vect)
grid_land_cover_wetland <- mask(grid_land_cover_wetland, grid_vect)

grid_rainfall_2021 <- crop(rainfall_2021, grid_vect)
grid_rainfall_2021 <- mask(grid_rainfall_2021, grid_vect)

grid_temperature_2021 <- crop(annual_temperature_climate, grid_vect)
grid_temperature_2021 <- mask(grid_temperature_2021, grid_vect)


grid_ITN_usage_2021 <- crop(ITN_intervention_2021, grid_vect)
grid_ITN_usage_2021 <- mask(grid_ITN_usage_2021, grid_vect)

# grid_ITN_usage_2021  <- project(ITN_intervention_2021,  st_crs(grid_sf)$wkt)
# grid_pop  <- project(grid_pop,  st_crs(grid_sf)$wkt)
# grid_rainfall_2021 <- project(ben_rainfall_2021, st_crs(grid_sf)$wkt)
# grid_temperature_2021 <- project(ben_temperature_2021, st_crs(grid_sf)$wkt)
# grid_ndvi_stack <- project(grid_NDVI, st_crs(grid_sf)$wkt)
# elevation_grid   <- project(grid_elevation,   st_crs(grid_sf)$wkt)
# grid_land_cover_water   <- project(grid_land_cover_water,   st_crs(grid_sf)$wkt)
# grid_land_cover_built   <- project(grid_land_cover_built,   st_crs(grid_sf)$wkt)
# grid_land_cover_cropland   <- project(grid_land_cover_cropland,   st_crs(grid_sf)$wkt)
# grid_land_cover_trees   <- project(grid_land_cover_trees,   st_crs(grid_sf)$wkt)
# grid_land_cover_wetland   <- project(grid_land_cover_wetland,   st_crs(grid_sf)$wkt)



grid_ITN_usage_2021_df <- terra::extract(grid_ITN_usage_2021,  grid_sf%>% st_transform(4326))
grid_pop_density_2020_df <- terra::extract(grid_pop,  grid_sf%>% st_transform(4326))
grid_rainfall_2021_df    <- terra::extract(grid_rainfall_2021, grid_sf%>% st_transform(4326))
grid_temperature_2021_df <- terra::extract(grid_temperature_2021, grid_sf%>% st_transform(4326))
ndvi_vals_grid_monthly <- terra::extract(grid_NDVI, grid_sf%>% st_transform(4326))
ndvi_mean_grid <- apply(ndvi_vals_grid_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_grid_df <- terra::extract(grid_elevation, grid_sf%>% st_transform(4326))
grid_land_cover_water_df <- terra::extract(grid_land_cover_water, grid_sf%>% st_transform(4326))
grid_land_cover_built_df <- terra::extract(grid_land_cover_built, grid_sf%>% st_transform(4326))
grid_land_cover_cropland_df <- terra::extract(grid_land_cover_cropland, grid_sf%>% st_transform(4326))
grid_land_cover_trees_df <- terra::extract(grid_land_cover_trees, grid_sf%>% st_transform(4326))
grid_land_cover_wetland_df <- terra::extract(grid_land_cover_wetland, grid_sf%>% st_transform(4326))

# Distance to water

bbox <- st_bbox(countries_sf)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)



# nearest river segment for each point
nn_riv <- st_nn(grid_sf %>% st_transform(4326), rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(grid_sf%>% st_transform(4326), lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

grid_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

grid_sf <- grid_sf %>% 
  mutate(annual_rainfall_2021 = grid_rainfall_2021_df[,2],
         annual_temperature_2021 = grid_temperature_2021_df[,2],
         elevation               = elevation_grid_df[,2],
         vegetation_index = ndvi_mean_grid,
         pop_density_2020 = grid_pop_density_2020_df[,2],
         lc_cropland = grid_land_cover_cropland_df[,2],
         lc_water = grid_land_cover_water_df[,2],
         lc_wetland = grid_land_cover_wetland_df[,2],
         lc_built = grid_land_cover_built_df[,2],
         lc_trees = grid_land_cover_trees_df[,2],
         annual_ITN_usage_2021 = grid_ITN_usage_2021_df[,2]) 



numeric_cols <- names(grid_sf)[sapply(grid_sf, is.numeric)]

for (col in numeric_cols) {
  grid_sf <- fill_by_nearest(grid_sf, col)
}


st_write(grid_sf, "data/grid_data_for_prediction_sf.gpkg", append = FALSE)
```


```{r}
grid_sf <- read_sf("data/grid_data_for_prediction_sf.gpkg")

# Scale the prediction grid using the EXACT same parameters as in the fitting model
# (This ensures a grid cell is measured relative to our 36 observed points)
for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf[[var]] <- (grid_sf[[var]] - mean_val) / sd_val
}


# coords_pred <- st_coordinates(grid_sf)
# 
# coords_pred_A <- coords_pred[grid_sf$country_id == 1, ]
# coords_pred_B <- coords_pred[grid_sf$country_id == 2, ]
# coords_pred_C <- coords_pred[grid_sf$country_id == 3, ]
# 
# A_pred_A <- inla.spde.make.A(mesh = mesh_A, loc = coords_pred_A)
# A_pred_B <- inla.spde.make.A(mesh = mesh_B, loc = coords_pred_B)
# A_pred_C <- inla.spde.make.A(mesh = mesh_C, loc = coords_pred_C)
# 
# A_pred_A_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_A$n.spde)
# A_pred_B_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_B$n.spde)
# A_pred_C_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_C$n.spde)
# 
# A_pred_A_full[grid_sf$country_id == 1, ] <- as.matrix(A_pred_A)
# A_pred_B_full[grid_sf$country_id == 2, ] <- as.matrix(A_pred_B)
# A_pred_C_full[grid_sf$country_id == 3, ] <- as.matrix(A_pred_C)


# 1. Get the full coordinates matrix for all 37,515 grid cells
coords_pred <- st_coordinates(grid_sf)

# 2. Create 0/1 indicator weights across ALL 37,515 rows
w_pred_A <- as.numeric(grid_sf$country_id == 1)
w_pred_B <- as.numeric(grid_sf$country_id == 2)
w_pred_C <- as.numeric(grid_sf$country_id == 3)

# 3. Build native, optimized Sparse Matrices (dgCMatrix) automatically.
# Passing 'coords_pred' means every matrix instantly has exactly 37,515 rows.
# 'weights' forces cells outside the country to stay zero efficiently.
A_pred_A_full <- inla.spde.make.A(mesh = mesh_A, loc = coords_pred, weights = w_pred_A)
A_pred_B_full <- inla.spde.make.A(mesh = mesh_B, loc = coords_pred, weights = w_pred_B)
A_pred_C_full <- inla.spde.make.A(mesh = mesh_C, loc = coords_pred, weights = w_pred_C)

# Verify they are sparse structures
class(A_pred_A_full) # Should output "dgCMatrix"

effects_fixed_pred <- data.frame(
  intercept               = 1,
  # annual_rainfall_2021    = grid_sf$annual_rainfall_2021,
  annual_temperature_2021 = grid_sf$annual_temperature_2021,
  pop_density_2020        = grid_sf$pop_density_2020,
  # annual_ITN_usage_2021   = grid_sf$annual_ITN_usage_2021,
  dist_to_water           = grid_sf$dist_to_water,
  lc_water                = grid_sf$lc_water,
  lc_trees                = grid_sf$lc_trees,
  lc_built                = grid_sf$lc_built,
  lc_cropland             = grid_sf$lc_cropland,
  lc_wetland              = grid_sf$lc_wetland)

effects_country_pred <- list(country_id = grid_sf$country_id)

effects_A_pred <- list(field_A = 1:spde_A$n.spde)
effects_B_pred <- list(field_B = 1:spde_B$n.spde)
effects_C_pred <- list(field_C = 1:spde_C$n.spde)

stack_pred <- inla.stack(
  data = list(y = NA,
    Ntrials = NA),
  A = list(
    Diagonal(nrow(grid_sf)),      # fixed effects
    Diagonal(nrow(grid_sf)),      # country iid
    A_pred_A_full,                # spatial field A
    A_pred_B_full,                # spatial field B
    A_pred_C_full                 # A_pred_B_full C
  ),
  effects = list(
    effects_fixed_pred,
    effects_country_pred,
    effects_A_pred,
    effects_B_pred,
    effects_C_pred
  ),
  tag = "pred"
)

stack_full <- inla.stack(stack_test_temperature, stack_pred)

res_pred <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stack_full),
  Ntrials = inla.stack.data(stack_full)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_full),
    compute = TRUE,
    link = 1     
  ),
  control.fixed = list(mean = 0, prec = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)

#idx_pred <- inla.stack.index(stack_full, "pred")$data


set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stack_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]                   # linear predictor for grid cells
  plogis(lp)                                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)



# v1 <- res_pred$summary.fixed
# v2 <- res_pred$summary.fixed

```

## P. malaria model interpretation

### clustered model
Across the three‑country study area, Plasmodium malariae prevalence is strongly
influenced by environmental conditions. Rainfall and wetland show clear 
positive associations with infection risk, indicating that humid and water‑rich
ecosystems provide favorable breeding conditions for malaria vectors. Built‑up
areas display a strong negative effect, reflecting reduced transmission in
urbanized landscapes where mosquito habitats are disrupted. Distance to water
shows a small but statistically meaningful positive effect. In contrast, 
cropland, population density, forest cover, and land‑cover water all have 
credible intervals that include zero, meaning the data do not provide sufficient
evidence to classify their influence as either protective or risk‑enhancing. 
Overall, rainfall and wetlands emerge as key drivers of P. malariae transmission,
with urbanization acting as an important suppressor, while several other 
covariates show uncertain or negligible effects.

### unclustered model
The updated model produces results that are broadly consistent with the previous
analysis, identifying rainfall and wetland environments as strong positive 
drivers of Plasmodium malariae prevalence, and built‑up areas as a strong 
negative predictor. However, the new model shows slightly reduced effect sizes
for rainfall and increased uncertainty for several covariates. Notably, the effect
of distance to water, previously statistically positive, now has a credible interval
that includes zero, indicating insufficient evidence for a clear association. 
Cropland, population density, forest cover, and land‑cover water continue to show
wide credible intervals that cross zero, meaning their effects remain statistically
uncertain. Overall, the main difference between the two models lies in the reduced 
certainty around distance to water and slightly adjusted effect sizes, while the 
key environmental drivers rainfall, wetlands, and urbanization remain robust.


# P. falciparum prevalence from Malaria Atlas Project (MAP)
```{r}
# Cameroon case

# Load the modeled prevalence from MAP

africa_pf_pr_2_10 <- rast("data/prevalence_2000/2025_GBD2024_Global_PfPR_2021.tif")

grid_sf_vect <- vect(grid_sf %>% st_transform(4326))

africa_pf_pr_2021 <- crop(africa_pf_pr_2_10, grid_sf_vect)
africa_pf_pr_2021 <- mask(africa_pf_pr_2021, grid_sf_vect)

grid_sf_map_prev <- terra::extract(africa_pf_pr_2021, grid_sf %>% st_transform(4326))

grid_sf_map_prev <- grid_sf_map_prev %>% 
  select(ID, Data, LCI, UCI)

grid_sf <- grid_sf %>% 
  mutate(PR_2_10_map     = grid_sf_map_prev[,2],
         PR_2_10_map_L   = grid_sf_map_prev[,3],
         PR_2_10_map_U   = grid_sf_map_prev[,4])

numeric_cols <- names(grid_sf)[sapply(grid_sf, is.numeric)]

for (col in numeric_cols) {
  grid_sf <- fill_by_nearest(grid_sf, col)
}

grid_sf$PR_all_age <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")

grid_sf$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map_U,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")

grid_sf$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map_L,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")



grid_sf$GID_0 <- ifelse(grid_sf$country_id == 1, "BEN",
                 ifelse(grid_sf$country_id == 2, "CMR",
                 ifelse(grid_sf$country_id == 3, "COG", NA)))


grid_benin <- grid_sf %>% filter(GID_0 == "BEN")
shp_ben1 <-  st_transform(shp_ben, st_crs(grid_benin))

grid_benin1 <- st_join(grid_benin, shp_ben1, join = st_intersects)

ben_data_pev <- grid_benin1 %>%
  group_by(NAME_2) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


ben_data_pev_map <- shp_ben %>% group_by(NAME_2) %>% summarise(geometry = st_union(geometry)) %>%
  left_join(ben_data_pev %>% st_drop_geometry(), by = "NAME_2")

ben_prev_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_pm.pdf",
       plot = ben_prev_1, width = 6, height = 4, dpi = 300)


ben_prev_uncertainty_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_uncertainty_pm.pdf",
       plot = ben_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_cmr <- grid_sf %>% filter(GID_0 == "CMR")
shp_cmr1 <-  st_transform(shp_cmr, st_crs(grid_cmr))

grid_cmr1 <- st_join(grid_cmr, shp_cmr1, join = st_intersects)

cmr_data_pev <- grid_cmr1 %>%
  group_by(Admin2) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


cmr_data_pev_map <- shp_cmr %>%
  left_join(cmr_data_pev %>% st_drop_geometry(), by = "Admin2")

cmr_prev_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_pm.pdf", 
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

cmr_prev_uncertainty_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_uncertainty_pm.pdf", 
       plot = cmr_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_con <- grid_sf %>% filter(GID_0 == "COG")
shp_con1 <-  st_transform(shp_con, st_crs(grid_con))

grid_con1 <- st_join(grid_con, shp_con1, join = st_intersects)

con_data_pev <- grid_con1 %>%
  group_by(adm2_name) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


con_data_pev_map <- shp_con %>%
  left_join(con_data_pev %>% st_drop_geometry(), by = "adm2_name")

con_prev_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/cong_estimated_prev_pm.pdf", 
       plot = con_prev_1, width = 6, height = 4, dpi = 300)


con_prev_uncertainty_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/con_estimated_prev_uncertainty_pm.pdf", 
       plot = con_prev_uncertainty_1, width = 6, height = 4, dpi = 300)
```


```{r}
cmr_prev_pf_MAP_estimates <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_MAP_pf_all_age.pdf",
       plot = cmr_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

cmr_prev_pf_MAP_estimates_uncertainty <- ggplot(cmr_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_uncertainty_MAP_pf_all_age.pdf",
       plot = cmr_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)


# Benin case

ben_prev_pf_MAP_estimates <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_MAP_pf_all_age.pdf",
       plot = ben_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

ben_prev_pf_MAP_estimates_uncertainty <- ggplot(ben_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_uncertainty_MAP_pf_all_age.png",
       plot = ben_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)

# Congo case

con_prev_pf_MAP_estimates <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_MAP_pf_all_age.pdf",
       plot = con_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

con_prev_pf_MAP_estimates_uncertainty <- ggplot(con_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_uncertainty_MAP_pf_all_age.png",
       plot = con_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)

```

#Comparison between COMAL estimates and MAP estimates

```{r}
# Divide the burden into classes

ben_data_pev_map <- ben_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))


cmr_data_pev_map <- cmr_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))

con_data_pev_map <- con_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))


# Rank correlation analysis

spearman_res_ben <- cor.test(ben_data_pev_map$prev_mean, ben_data_pev_map$PR_all_age, method = "spearman")

spearman_res_cmr <- cor.test(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "spearman")
spearman_res_con <- cor.test(con_data_pev_map$prev_mean, con_data_pev_map$PR_all_age, method = "spearman")

comal_trans_intens_cmr <- ggplot(cmr_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pf_pr_estimated/comal_trans_intens_cmr.png",
       plot = comal_trans_intens_cmr, width = 6, height = 4, dpi = 300)


map_trans_intens_cmr <- ggplot(cmr_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/map_trans_intens_cmr.png",
       plot = map_trans_intens_cmr, width = 6, height = 4, dpi = 300)


comal_trans_intens_ben <- ggplot(ben_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pf_pr_estimated/comal_trans_intens_ben.png",
       plot = comal_trans_intens_ben, width = 6, height = 4, dpi = 300)


map_trans_intens_ben <- ggplot(ben_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/map_trans_intens_ben.png",
       plot = map_trans_intens_ben, width = 6, height = 4, dpi = 300)


comal_trans_intens_cong <- ggplot(con_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pf_pr_estimated/comal_trans_intens_cong.png",
       plot = comal_trans_intens_cong, width = 6, height = 4, dpi = 300)


map_trans_intens_cong <- ggplot(con_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/map_trans_intens_cong.png",
       plot = map_trans_intens_cong, width = 6, height = 4, dpi = 300)


```


# prepare data for the continental level estimation

```{r}
shp_global <- read_sf("data/Africa_data/shape_files_1/geoBoundariesCGAZ_ADM2.shp")

custom_match <- c(
  XKX = "Europe",
  ESH = "Africa",
  `111` = "Africa",        # Abyei
  `112` = "Asia",          # Aksai Chin
  `113` = "Asia",          # China–India border
  `114` = "Asia",          # Demchok
  `115` = "Europe",        # Dragonja
  `116` = "Africa",        # Dramana-Shakatoe
  `117` = "South America", # Falkland Islands
  `118` = "Asia",          # Gaza Strip
  `119` = "Asia",          # Kalapani
  `120` = "South America", # Isla Brasilera
  `121` = "Asia",          # Siachen-Saltoro
  `122` = "Africa",        # Koualou
  `123` = "Asia",          # Liancourt Rocks
  `124` = "Asia",          # No Man's Land
  `125` = "Asia",          # Paracel Islands
  `126` = "Asia",          # Sanafir & Tiran Islands
  `127` = "Asia",          # Senkaku Islands
  `128` = "Asia",          # Spratly Islands
  `129` = "Asia"           # West Bank
)

shp_global$continent <- countrycode(
  shp_global$shapeGroup,
  origin = "iso3c",
  destination = "continent",
  custom_match = custom_match
)

shp_africa <- shp_global %>% 
  filter(continent=="Africa")

custom_match <- c(
  `111` = "Abyei",                   # Abyei
  `116` = "Dramana-Shakatoe",        # Dramana-Shakatoe
  `122` = "Koualou"                  # Koualou
  )


shp_africa$country <- countrycode(
  shp_africa$shapeGroup,
  origin = "iso3c",
  destination = "country.name",
  custom_match = custom_match
)

shp_africa_simple <- ms_simplify(
  shp_africa,
  keep = 0.05,          # keep 5% of the vertices
  keep_shapes = TRUE
)

st_write(
  shp_africa_simple,
  "data/Africa_data/shp_africa/Africa_simplified.shp",
  delete_layer = TRUE,
  append = TRUE
)


ggplot()+
  geom_sf(data = shp_africa_simple, fill = "grey", color = "black")

```

To avoid extrapolating predictions into areas where malaria transmission is not considered ecologically feasible, model predictions will be masked using the Plasmodium falciparum transmission limits provided by the Malaria Atlas Project. Although this study focuses on Plasmodium malariae, no continent-wide transmission limits are currently available for this species. Given that P. malariae is predominantly transmitted in areas where P. falciparum is endemic and frequently occurs as a co-infection, the P. falciparum transmission limits will be used as a proxy for malaria transmission suitability.


```{r}

shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp") %>% st_transform(3857)

# Dissolve all polygons into one
shp_africa <- st_union(shp_africa)
shp_africa <- st_sf(geometry = shp_africa)

bbox <- st_bbox(shp_africa)

grid_5km_africa <- st_make_grid(
  st_as_sfc(bbox),
  cellsize = 5000,
  what = "centers"
)

grid_sf_africa <- st_sf(geometry = grid_5km_africa)
grid_sf_africa <- st_filter(grid_sf_africa, shp_africa)

ndvi_path <- "data/Africa_data/NDVI_africa/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

path1 <- "data/Africa_data/climateEngine_download.precipitation.tif"
path2 <- "data/Africa_data/ERA5_LAND_DAILY_temperature_2m_2021-01-01_2021-12-31_-39p5213to41p4698N_-22p2966to54p0315E.temperature_2m.tif"


ITN_intervention_2021 <- rast("data/ITN_2000/2025_GBD2024_Africa_ITN_2021.tif")
africa_rainfall_2021 <- rast(path1)
africa_temperature_2021 <- rast(path2)
africa_pop <- rast("data/Africa_data/pop_density/gpw_v4_population_density_rev11_2020_30_sec.tif")
africa_elevation <- rast("data/Africa_data/elevation/wc2.1_30s_elev.tif")
africa_ndvi_stack<- rast(ndvi_files)



africa_land_cover_water <- crop(land_cover_water, vect(shp_africa %>% st_transform(4326)))
africa_land_cover_water <- mask(africa_land_cover_water, vect(shp_africa %>% st_transform(4326)))

africa_land_cover_built <- crop(land_cover_builts, vect(shp_africa %>% st_transform(4326)))
africa_land_cover_built <- mask(africa_land_cover_built, vect(shp_africa %>% st_transform(4326)))

africa_land_cover_cropland <- crop(land_cover_cropland, vect(shp_africa %>% st_transform(4326)))
africa_land_cover_cropland <- mask(africa_land_cover_cropland, vect(shp_africa %>% st_transform(4326)))

africa_land_cover_trees <- crop(land_cover_trees, vect(shp_africa %>% st_transform(4326)))
africa_land_cover_trees <- mask(africa_land_cover_trees, vect(shp_africa %>% st_transform(4326)))

africa_land_cover_wetland <- crop(land_cover_wetland, vect(shp_africa %>% st_transform(4326)))
africa_land_cover_wetland <- mask(africa_land_cover_wetland, vect(shp_africa %>% st_transform(4326)))

africa_ITN_intervention_2021 <- crop(ITN_intervention_2021, vect(shp_africa %>% st_transform(4326)))
africa_ITN_intervention_2021 <- mask(africa_ITN_intervention_2021, vect(shp_africa %>% st_transform(4326)))

africa_pop_2020 <- crop(africa_pop, vect(shp_africa %>% st_transform(4326)))
africa_pop_2020 <- mask(africa_pop_2020, vect(shp_africa %>% st_transform(4326)))

africa_rainfall_2021 <- crop(africa_rainfall_2021, vect(shp_africa %>% st_transform(4326)))
africa_rainfall_2021 <- mask(africa_rainfall_2021, vect(shp_africa %>% st_transform(4326)))

africa_temperature_2021 <- crop(africa_temperature_2021, vect(shp_africa %>% st_transform(4326)))
africa_temperature_2021 <- mask(africa_temperature_2021, vect(shp_africa %>% st_transform(4326)))

africa_ndvi_stack <- crop(africa_ndvi_stack, vect(shp_africa %>% st_transform(4326)))
africa_ndvi_stack <- mask(africa_ndvi_stack, vect(shp_africa %>% st_transform(4326)))

africa_elevation <- crop(africa_elevation, vect(shp_africa %>% st_transform(4326)))
africa_elevation <- mask(africa_elevation, vect(shp_africa %>% st_transform(4326)))



# africa_ITN_intervention_2021  <- project(ITN_intervention_2021,  st_crs(grid_sf_africa)$wkt)
# africa_pop  <- project(africa_pop,  st_crs(grid_sf_africa)$wkt)
# africa_rainfall_2021 <- project(africa_rainfall_2021, st_crs(grid_sf_africa)$wkt)
# africa_temperature_2021 <- project(africa_temperature_2021, st_crs(grid_sf_africa)$wkt)
# africa_ndvi_stack <- project(africa_ndvi_stack, st_crs(grid_sf_africa)$wkt)
# africa_elevation   <- project(africa_elevation,   st_crs(grid_sf_africa)$wkt)
# africa_land_cover_water   <- project(africa_land_cover_water,   st_crs(grid_sf_africa)$wkt)
# africa_land_cover_built   <- project(africa_land_cover_built,   st_crs(grid_sf_africa)$wkt)
# africa_land_cover_cropland   <- project(africa_land_cover_cropland,   st_crs(grid_sf_africa)$wkt)
# africa_land_cover_trees   <- project(africa_land_cover_trees,   st_crs(grid_sf_africa)$wkt)
# africa_land_cover_wetland   <- project(africa_land_cover_wetland,   st_crs(grid_sf_africa)$wkt)

africa_ITN_intervention_2021_df <- terra::extract(africa_ITN_intervention_2021, 
                                                  grid_sf_africa %>% st_transform(4326))
africa_pop_density_2020_df <- terra::extract(africa_pop_2020,  
                                             grid_sf_africa %>% st_transform(4326))
africa_rainfall_2021_df    <- terra::extract(africa_rainfall_2021, 
                                             grid_sf_africa %>% st_transform(4326))
africa_temperature_2021_df <- terra::extract(africa_temperature_2021, 
                                             grid_sf_africa %>% st_transform(4326))

ndvi_vals_africa_monthly <- terra::extract(africa_ndvi_stack, 
                                           grid_sf_africa %>% st_transform(4326))
ndvi_mean_africa <- apply(ndvi_vals_africa_monthly[,-1], 1, mean, na.rm = TRUE)

africa_elevation_df <- terra::extract(africa_elevation,
                                      grid_sf_africa %>% st_transform(4326))
africa_land_cover_water_df <- terra::extract(africa_land_cover_water, 
                                             grid_sf_africa %>% st_transform(4326))
africa_land_cover_built_df <- terra::extract(africa_land_cover_built,
                                             grid_sf_africa %>% st_transform(4326))
africa_land_cover_cropland_df <- terra::extract(africa_land_cover_cropland,
                                                grid_sf_africa %>% st_transform(4326))
africa_land_cover_trees_df <- terra::extract(africa_land_cover_trees,
                                             grid_sf_africa %>% st_transform(4326))
africa_land_cover_wetland_df <- terra::extract(africa_land_cover_wetland,
                                               grid_sf_africa %>% st_transform(4326))

shp_africa_summarise <- shp_africa %>% 
  group_by(continent) %>% 
  summarise(geometry = st_union(geometry))

# distance to water

# bbox <- st_bbox(grid_sf_africa)

rivers <- vect("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp")
lakes <- vect("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp")


africa4326 <- st_transform(shp_africa, 4326)
africa4326 <- vect(africa4326)

lakes_africa <- crop(lakes, africa4326)

r_template <- rast(
  ext(vect(st_as_sf(shp_africa))),
  resolution = 5000,
  crs = "EPSG:3857")

rivers <- project(rivers, "EPSG:3857")
lakes_africa <- project(lakes_africa, "EPSG:3857")

rivers2 <- rivers[rivers$ORD_STRA >= 3, ]

riv_r <- rasterize(
  rivers2,
  r_template,
  field = 1,
  background = NA)


lake_r <- rasterize(
  lakes_africa,
  r_template,
  field = 1,
  background = NA)

water <- cover(riv_r, lake_r)

dist_water <- distance(
    water,
    filename = "data/dist_to_water_5km.tif",
    overwrite = TRUE)

dist_water <- rast("data/dist_to_water_5km.tif")

grid_sf_africa$dist_to_water <- terra::extract(
    dist_water,
    vect(grid_sf_africa))[,2]


grid_sf_africa <- grid_sf_africa %>% 
  mutate(pop_density_2020 = africa_pop_density_2020_df[,2],
         annual_rainfall_2021 = africa_rainfall_2021_df[,2],
         annual_temperature_2021 = africa_temperature_2021_df[,2],
         vegetation_index = ndvi_mean_africa,
         elevation = africa_elevation_df[,2],
         lc_water = africa_land_cover_water_df[,2],
         lc_built = africa_land_cover_built_df[,2],
         lc_trees = africa_land_cover_trees_df[,2],
         lc_wetland = africa_land_cover_wetland_df[,2],
         lc_cropland = africa_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = africa_ITN_intervention_2021_df[,2])


# CMR_df_1_sf_df <- CMR_df_1_sf %>% 
#   st_drop_geometry() 
# 
# st_write(CMR_df_1_sf, "cmr_data/CMR_df_1_sf.gpkg", append = FALSE)


numeric_cols <- names(grid_sf_africa)[sapply(grid_sf_africa, is.numeric)]

for (col in numeric_cols) {
  grid_sf_africa <- fill_by_nearest(grid_sf_africa, col)
}



st_write(grid_sf_africa, "data/Africa_data/grid_data_for_prediction_sf.gpkg", append = FALSE)

st_write(sampled_data_join_sf, "data/Africa_data/sampled_data_join_sf.gpkg", append = FALSE)

```


```{r}
sampled_data_join_sf <- st_read("data/Africa_data/sampled_data_join_sf.gpkg")
grid_sf_africa <- st_read("data/Africa_data/grid_data_for_prediction_sf.gpkg")

```


```{r}
#sampled_data_join_sf$country_id <- ifelse(sampled_data_join_sf$country == "Benin", 1, 2)

Africa_union <- st_union(shp_africa )

sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

sampled_data_join_sf_sp     <- as(sampled_data_join_sf, "Spatial")
Africa_union_sp <- as(Africa_union, "Spatial")

coords_Africa <- coordinates(sampled_data_join_sf_sp)


mesh_Africa <- inla.mesh.2d(
  loc      = coords_Africa,
  boundary = inla.sp2segment(Africa_union_sp),
  max.edge = c(1500e3, 2000e3),   
  cutoff   = 250e3,
  offset   = c(100e3, 2000e3))

# Extract triangle vertex indices
tri <- mesh_Africa$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_Africa <- as.data.frame(coords_Africa)
colnames(coords_df_Africa) <- c("X", "Y")

ggplot() +
  geom_sf(data = mesh_edges_sf, fill = NA, color = "grey60", size = 0.3) +
  geom_sf(data = Africa_union, fill = NA, color = "black", size = 0.8) +
  geom_point(data = coords_df_Africa, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh for Africa with Boundary")
```


```{r}
spde_Africa <- inla.spde2.pcmatern(
  mesh = mesh_Africa,
  prior.range = c(100000, 0.5),
  prior.sigma = c(0.2, 0.5)
)

A_est <- inla.spde.make.A(mesh_Africa, loc = coords_Africa)

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "vegetation_index",
                   "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sampled_data_join_sf[[var]], na.rm = TRUE)
  sd_val   <- sd(sampled_data_join_sf[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sampled_data_join_sf[[var]] <- (sampled_data_join_sf[[var]] - mean_val) / sd_val
}


# sampled_data_join_sf <- sampled_data_join_sf %>% 
#   mutate(pop_density_2020 = as.numeric(scale(pop_density_2020)),
#          annual_rainfall_2021 = as.numeric(scale(annual_rainfall_2021)),
#          annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
#          vegetaion_index = as.numeric(scale(vegetaion_index)),
#          dist_to_water  = as.numeric(scale(dist_to_water)),
#          elevation = as.numeric(scale(elevation)))

```



```{r}
effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sampled_data_join_sf$annual_rainfall_2021,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept                = 1,
  annual_temperature_2021  = sampled_data_join_sf$annual_temperature_2021,
  pop_density_2020         = sampled_data_join_sf$pop_density_2020,
  dist_to_water            = sampled_data_join_sf$dist_to_water,
  lc_water                 = sampled_data_join_sf$lc_water,
  lc_trees                 = sampled_data_join_sf$lc_trees,
  lc_built                 = sampled_data_join_sf$lc_built,
  lc_cropland              = sampled_data_join_sf$lc_cropland,
  lc_wetland               = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sampled_data_join_sf$elevation,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


stk_est_rainfall <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_rainfall
  ),
  tag = "est"
)

stk_est_temperature <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_temperature
  ),
  tag = "est"
)


stk_est_vegetation_index <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_vegetation_index
  ),
  tag = "est"
)

stk_est_elevation <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_elevation
  ),
  tag = "est"
)


stk_est_baseline <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_baseline
  ),
  tag = "est"
)

```


```{r}
formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetation_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stk_est_rainfall  ),
  Ntrials = inla.stack.data(stk_est_rainfall )$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_rainfall ),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stk_est_temperature),
  Ntrials = inla.stack.data(stk_est_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stk_est_vegetation_index),
  Ntrials = inla.stack.data(stk_est_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stk_est_elevation),
  Ntrials = inla.stack.data(stk_est_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stk_est_baseline),
  Ntrials = inla.stack.data(stk_est_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


hist(res_temperature$cpo$pit,
     breaks = 20)
```


```{r}

samples <- inla.posterior.sample(
  n = 100,
  result = res_rainfall 
)

# extraction of the fix effect

fixed_names <- res_rainfall$names.fixed

fixed_idx <- match(
  paste0(fixed_names, ":1"),
  rownames(samples[[1]]$latent)
)

# extract the spatial effect

spatial_idx <- grep(
  "^spatial:",
  rownames(samples[[1]]$latent)
)



idx_est <- inla.stack.index(stk_est_rainfall, "est")$data

fitted_est <- res_rainfall$summary.fitted.values[idx_est, "mean"]

cbind(
  observed = sampled_data_join_sf$PR_Pm_tot,
  fitted   = fitted_est
)[1:36, ]
```


# Given the large files, let's break the grid of prediction into tiles

```{r}

for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf_africa[[var]] <- (grid_sf_africa[[var]] - mean_val) / sd_val
}


## Choose number of tiles
n_tiles  <- 200
tile_id  <- cut(seq_len(nrow(grid_sf_africa)), breaks = n_tiles, labels = FALSE)
n_samp   <- 200  # posterior samples per tile

# Storage
pred_mean_full <- numeric(nrow(grid_sf_africa))
pred_sd_full   <- numeric(nrow(grid_sf_africa))
pred_q05_full  <- numeric(nrow(grid_sf_africa))
pred_q95_full  <- numeric(nrow(grid_sf_africa))

for (tile in seq_len(n_tiles)) {
  cat("Tile", tile, "of", n_tiles, "\n")
  
  # 1) Subset grid
  grid_tile <- grid_sf_africa[tile_id == tile, ]
  
  # 2) A-matrix for this tile
  A_pred_tile <- inla.spde.make.A(
    mesh = mesh_Africa,
    loc  = st_coordinates(grid_tile)
  )
  
  # 3) Fixed effects for this tile
  effects_pred_tile <- data.frame(
    intercept = 1,
    annual_temperature_2021 = grid_tile$annual_temperature_2021,
    pop_density_2020        = grid_tile$pop_density_2020,
    dist_to_water           = grid_tile$dist_to_water,
    lc_cropland             = grid_tile$lc_cropland,
    lc_water                = grid_tile$lc_water,
    lc_wetland              = grid_tile$lc_wetland,
    lc_built                = grid_tile$lc_built,
    lc_trees                = grid_tile$lc_trees
  )
  
  # 4) Prediction stack for this tile
  stk_pred_tile <- inla.stack(
    data    = list(y = NA, Ntrials = 1),
    A       = list(A_pred_tile, 1),
    effects = list(
      spatial = 1:spde_Africa$n.spde,
      effects_pred_tile
    ),
    tag = "pred"
  )
  
  # 5) Full stack for this tile
  stk_full_tile <- inla.stack(stk_est_temperature, stk_pred_tile)
  
  # 6) Fit INLA with config = TRUE for posterior sampling
  res_pred_tile <- inla(
    formula_temperature,
    family = "binomial",
    data   = inla.stack.data(stk_full_tile),
    Ntrials = inla.stack.data(stk_full_tile)$Ntrials,
    control.predictor = list(
      A       = inla.stack.A(stk_full_tile),
      compute = TRUE,
      link    = 1
    ),
    control.fixed = list(mean = 0, prec = 0.2),
    control.compute = list(config = TRUE)
  )
  
  # 7) Posterior samples for this tile
  samples_tile <- inla.posterior.sample(n_samp, res_pred_tile)
  
  # 8) Identify latent indices (once per tile)
  latent_names <- rownames(samples_tile[[1]]$latent)
  
  fixed_base_names   <- rownames(res_pred_tile$summary.fixed)
  fixed_latent_names <- paste0(fixed_base_names, ":1")
  fixed_idx          <- match(fixed_latent_names, latent_names)
  
  spatial_idx <- grep("^spatial:", latent_names)
  
  # 9) Build linear predictor samples for this tile
  n_grid_tile <- nrow(grid_tile)
  X_pred_tile <- as.matrix(effects_pred_tile)
  eta_mat_tile <- matrix(NA_real_, nrow = n_grid_tile, ncol = n_samp)
  
  for (s in seq_len(n_samp)) {
    latent_s   <- samples_tile[[s]]$latent
    beta_s     <- latent_s[fixed_idx]
    spatial_s  <- latent_s[spatial_idx]
    
    eta_fixed_s   <- as.vector(X_pred_tile %*% beta_s)
    eta_spatial_s <- as.vector(A_pred_tile %*% spatial_s)
    
    eta_mat_tile[, s] <- eta_fixed_s + eta_spatial_s
  }
  
  # 10) Transform to probabilities and summarize
  prob_mat_tile <- 1 / (1 + exp(-eta_mat_tile))
  
  pred_mean_tile <- rowMeans(prob_mat_tile)
  pred_sd_tile   <- apply(prob_mat_tile, 1, sd)
  pred_q05_tile  <- apply(prob_mat_tile, 1, quantile, 0.05)
  pred_q95_tile  <- apply(prob_mat_tile, 1, quantile, 0.95)
  
  # 11) Store back into full vectors
  idx <- (tile_id == tile)
  pred_mean_full[idx] <- pred_mean_tile
  pred_sd_full[idx]   <- pred_sd_tile
  pred_q05_full[idx]  <- pred_q05_tile
  pred_q95_full[idx]  <- pred_q95_tile
}

grid_sf_africa$pred_mean <- pred_mean_full
grid_sf_africa$pred_sd   <- pred_sd_full
grid_sf_africa$pred_q05  <- pred_q05_full
grid_sf_africa$pred_q95  <- pred_q95_full



```


```{r}


# Build the A‑matrix for prediction locations
# A_pred <- inla.spde.make.A(
#   mesh = mesh_Africa,
#   loc = st_coordinates(grid_sf_africa))
# 
# #Build the prediction stack
# 
# effects_pred <- data.frame(
#   intercept = 1,
#   annual_temperature_2021 = grid_sf_africa$annual_temperature_2021,
#   pop_density_2020        = grid_sf_africa$pop_density_2020,
#   dist_to_water           = grid_sf_africa$dist_to_water,
#   lc_cropland             = grid_sf_africa$lc_cropland,
#   lc_water                = grid_sf_africa$lc_water,
#   lc_wetland              = grid_sf_africa$lc_wetland,
#   lc_built                = grid_sf_africa$lc_built,
#   lc_trees                = grid_sf_africa$lc_trees
# )
# 
# 
# stk_pred <- inla.stack(
#   data = list(y = NA, Ntrials = 1),
#   A    = list(A_pred, 1),
#   effects = list(
#     spatial = 1:spde_Africa$n.spde,
#     effects_pred),
#   tag = "pred"
# )
# 
# 
# # Combine estimation + prediction stacks
# stk_full <- inla.stack(stk_est_temperature, stk_pred)
# 
# # Fit the model again with the full stack
# res_pred <- inla(
#   formula_temperature,
#   family = "binomial",
#   data = inla.stack.data(stk_full),
#   Ntrials = inla.stack.data(stk_full)$Ntrials,
#   control.predictor = list(
#     A = inla.stack.A(stk_full),
#     compute = TRUE,
#     link    = 1
#   ),
#   control.fixed = list(mean = 0, prec = 1)
#   #control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
# )
# 
# 
# # Fixing the prediction error
# 
# n_samp <- 200  
# 
# samples <- inla.posterior.sample(n_samp, res_temperature)
# 
# # Names of all latent variables in the first sample
# latent_names <- rownames(samples[[1]]$latent)
# 
# # Fixed effects: use names from res_temperature$summary.fixed
# fixed_base_names <- rownames(res_temperature$summary.fixed)
# fixed_latent_names <- paste0(fixed_base_names, ":1")
# fixed_idx <- match(fixed_latent_names, latent_names)
# 
# # Spatial field: usually starts with "spatial"
# spatial_idx <- grep("^spatial", latent_names)
# 
# ## Loop over samples and build the linear predictor on the grid
# 
# n_grid <- nrow(grid_sf_africa)
# 
# eta_mat <- matrix(NA_real_, nrow = n_grid, ncol = n_samp)
# 
# X_pred <- as.matrix(effects_pred)  # design matrix for fixed effects
# 
# for (i in seq_len(n_samp)) {
# 
#   latent_i <- samples[[i]]$latent
# 
#   beta_i    <- latent_i[fixed_idx]
#   spatial_i <- latent_i[spatial_idx]
# 
#   # sanity checks
#   if (any(is.na(beta_i)))    stop("beta_i contains NA")
#   if (any(is.na(spatial_i))) stop("spatial_i contains NA")
# 
#   eta_fixed_i   <- as.vector(X_pred %*% beta_i)
#   eta_spatial_i <- as.vector(A_pred %*% spatial_i)
# 
#   eta_mat[, i] <- eta_fixed_i + eta_spatial_i
# }
# 
# ## Transform to probabilities and compute uncertainty summaries
# 
# prob_mat <- 1 / (1 + exp(-eta_mat))
# 
# pred_mean <- rowMeans(prob_mat)
# pred_sd   <- apply(prob_mat, 1, sd)
# pred_q05  <- apply(prob_mat, 1, quantile, 0.05)
# pred_q95  <- apply(prob_mat, 1, quantile, 0.95)
# 
# grid_sf_africa$pred_mean <- pred_mean
# grid_sf_africa$pred_sd   <- pred_sd
# grid_sf_africa$pred_q05  <- pred_q05
# grid_sf_africa$pred_q95  <- pred_q95


# Extract predictions for the grid

# index_pred <- inla.stack.index(stk_full, "pred")$data
# index_est <- inla.stack.index(stk_full, "est")$data
# 
# grid_sf_africa$pred_mean <- res_pred$summary.fitted.values[index_pred, "mean"]
# grid_sf_africa$pred_sd   <- res_pred$summary.fitted.values[index_pred, "sd"]
# grid_sf_africa$pred_q025 <- res_pred$summary.fitted.values[index_pred, "0.025quant"]
# grid_sf_africa$pred_q975 <- res_pred$summary.fitted.values[index_pred, "0.975quant"]

# Using the transmission limit to mask the prediction grid

grid_sf_africa_wgs84 <- st_transform(grid_sf_africa, 4326)
grid_vect <- vect(grid_sf_africa_wgs84)

df_grid_africa <- grid_sf_africa_wgs84 %>%
    mutate(
        x = st_coordinates(.)[,1],
        y = st_coordinates(.)[,2]
    ) %>%
    st_drop_geometry()

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")

# global_spatial_limit_3857 <- project(global_spatial_limit, crs(shp_africa))


pf_limit_africa <- crop(global_spatial_limit, vect(shp_africa %>% st_transform(4326)))
pf_limit_africa <- mask(pf_limit_africa, vect(shp_africa %>% st_transform(4326)))


grid_with_limit <- terra::extract(pf_limit_africa, grid_vect)

grid_masked <- cbind(df_grid_africa, grid_with_limit[,2])

grid_masked_1 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when(is.na(limit) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q05   = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q05),
         pred_q95   = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q95)) 

grid_masked_2 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when((is.na(limit) | limit == 1) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q05  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q05),
         pred_q95  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q95)) 


grid_masked_sf <- st_as_sf(
    grid_masked_1,
    coords = c("x", "y"),
    crs = 4326)

grid_masked_sf_1 <- st_as_sf(
    grid_masked_2,
    coords = c("x", "y"),
    crs = 4326)

# shp_africa1 <-  st_transform(shp_africa, st_crs(grid_masked_sf))
# 
# 
# grid_africa1 <- st_join(grid_masked_sf, shp_africa1%>% 
#   select(-c(shapeName, shapeID, shapeGroup, shapeType, continent, country)), 
#   join = st_intersects) 
# 
# Africa_data_pev <- grid_africa1 %>%
#   group_by(shapeID) %>%
#   summarise(
#     prev_mean   = mean(pred_mean, na.rm = TRUE),
#     prev_q025   = mean(pred_q025, na.rm = TRUE),
#     prev_q975   = mean(pred_q975, na.rm = TRUE),
#     CI_width    = mean(pred_q975 - pred_q025, na.rm = TRUE))
# 
# Africa_country_prev_tx <- country_tx_sf %>%
#   left_join(Africa_country_prev, by = "shapeName")
# 
# Africa_data_pev <- Africa_data_pev %>%
#   mutate(prev_mean = if_else(is.nan(prev_mean), NA_real_, prev_mean))
# 
# Africa_data_pev_map <- shp_africa %>%
#   left_join(Africa_data_pev %>% st_drop_geometry(), by = "shapeID")
# 

shp_africa_country <- shp_africa %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry))


shp_africa_countryw4326 <- shp_africa_country %>% st_transform(4326)
shp_africaw4326 <- shp_africa %>% st_transform(4326)


Africa_prev_1 <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africaw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_admin2_level_unstable1.pdf",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country_1 <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africa_countryw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_country_level_unstable.pdf",
       plot = Africa_prev_country_1, width = 6, height = 4, dpi = 300)


# setting unstable areas to 0

Africa_prev_2 <- ggplot() +
  geom_sf(data = grid_masked_sf_1,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africaw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_admin2_level1.pdf",
       plot = Africa_prev_2, width = 6, height = 4, dpi = 300)

Africa_prev_country_2 <- ggplot() +
  geom_sf(data = grid_masked_sf_1,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africa_countryw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_country_level.pdf",
       plot = Africa_prev_country_2, width = 6, height = 4, dpi = 300)

```



```{r}
set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stk_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]   # linear predictor for grid cells
  plogis(lp)                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)

```







```{r}

```


```{r}

```


```{r}

```



```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```







