---
title: "COMAL project"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load necessary libraries
library(sf)
library(reshape2)
library(ggcorrplot)
library(leaflet)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(tidyterra)
library(mapview)
library(spData)
library(gganimate)
library(htmlwidgets)
library(webshot)
library(chirps)
library(viridis)
library(patchwork)
library(GGally)
library(spdep)
library(INLA)
library(leafsync)
library(SpatialEpi)
library(geodata)
library(car)
library(gstat)
library(nngeo)
library(tidygeocoder)
library(malariaAtlas)
library(rmapshaper)
library(tidytext)
```


```{r}
#loading the shape files
shp_ben <- read_sf("data/ben_data/BEN_shape_file/gadm41_BEN_3.shp")
shp_con <- read_sf("data/cong_data/CON_shape_file/cog_admin2.shp")
shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")
shp_gab <- read_sf("data/gab_data/gab_shape/gadm41_GAB_2.shp")


#renaming the districts

shp_ben$NAME_3[shp_ben$NAME_3=="Gakpè"]="Gakpe"
shp_ben$NAME_3[shp_ben$NAME_3=="Tokpa Domé"]="Tokpa-dome"
shp_ben$NAME_3[shp_ben$NAME_3=="Ouakpé-Daho"]="Houakpe-daho"
shp_ben$NAME_3[shp_ben$NAME_3=="Aganmalomé"]="Aganmanlome"
shp_ben$NAME_3[shp_ben$NAME_3=="Agonkanmè"]="Agonkamey"

ben_vect <- vect(shp_ben)
cmr_vect <- vect(shp_cmr)
con_vect <- vect(shp_con)
gab_vect <- vect(shp_gab)
```



```{r}
# Data for modelling

# spatial_data_ben_sf <- st_read("data/ben_data/spatial_data_ben_sf.gpkg")
# spatial_data_cong   <- st_read("data/cong_data/spatial_data_cong_sf.gpkg")
# spatial_data_cmr    <- st_read("data/cmr_data/cmr_cluster_df_1_sf.gpkg")
# spatial_data_gab    <- st_read("data/gab_data/GAB_df_1_sf.gpkg")
# 
# sampled_data_join_sf <- spatial_data_ben_sf %>% dplyr::select(-c(Village)) %>%
#   mutate(country = "Benin") %>% 
#   bind_rows(spatial_data_cong %>% dplyr::select(-c(District)) %>% 
#               mutate(country = "Congo")) %>%
#   bind_rows(spatial_data_cmr %>% 
#               mutate(country = "Cameroon")) %>% 
#   bind_rows(spatial_data_gab%>% 
#               mutate(country = "Gabon")) %>% 
#   select(-Match_confidence)
#   
# st_write(sampled_data_join_sf, "data/Africa_data/sampled_data_join_sf.gpkg", append = FALSE)

sampled_data_join_sf <- st_read("data/Africa_data/sampled_data_join_sf.gpkg")


numer_var_ben <- spatial_data_ben_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "annual_ITN_usage_2021",
  "elevation", "vegetaion_index", "pop_density_2020", "dist_to_water", "lc_water",
  "lc_trees", "lc_built", "lc_cropland", "lc_wetland"))



Ben <- cor(numer_var_ben, use = "pairwise.complete.obs")
corrplot(
  Ben,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


numer_var_cong <- spatial_data_cong %>% 
  st_drop_geometry() %>% 
  select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

Cong <- cor(numer_var_cong, use = "pairwise.complete.obs")
corrplot(
  Cong,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

```

```{r}
num_vars <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees", "annual_ITN_usage_2021",
  "lc_built", "lc_cropland", "lc_wetland"))

num_vars_1 <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c("dist_to_water", "PR_Pm_tot"))

M1 <- cor(num_vars_1, use = "pairwise.complete.obs")

corrplot(
  M1,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

var1 <- ggcorrplot(
  M,
  type = "full",
  legend.title = "Correlation Matrix",
  lab = TRUE,
  lab_size = 3,
  colors = c("blue", "white", "red"),
  outline.color = "white",
  show.diag = TRUE       
) +
  scale_x_discrete(name = NULL) +
  scale_y_discrete(name = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),      
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Outputs/correlation_matrix.png", var1, dpi = 300, width = 6, height = 4)

num_vars_selected <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

M_selected <- cor(num_vars_selected, use = "pairwise.complete.obs")

corrplot(
  M_selected,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + 
    annual_temperature_2021 + elevation + vegetaion_index + pop_density_2020 + 
    dist_to_water + lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021,
  data = sampled_data_join_sf
)

vif(fit)

fit1 <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water
  + lc_trees + lc_built + lc_cropland + lc_wetland , 
  data = sampled_data_join_sf
)

vif(fit1)

```

# variogram, autocorrelation and prior ajustment

```{r}
# Project to meters (important!)
sf_data <- st_transform(sampled_data_join_sf, 3857)


sf_data$logit_prev <- qlogis(
  pmin(pmax(sf_data$PR_Pm_tot, 1e-6), 1 - 1e-6)
)
# Extract coordinates
coords <- st_coordinates(sf_data)

# Build gstat object
g <- gstat(id = "pm", formula = logit_prev ~ 1, data = sf_data)

# Empirical variogram

v <- variogram(g,  cutoff = 100000, width = 1000)  # adjust cutoff to country size
# fit an exponential model
v.fit <- fit.variogram(v, model = vgm("Sph"))

# plot empirical + fitted curve
plot(v, v.fit)

plot(v)

```

```{r}
fit_cov <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 +
    dist_to_water + lc_water + lc_trees + lc_built +
    lc_cropland + lc_wetland,
  data = sampled_data_join_sf
)

sampled_data_join_sf$resid <- residuals(fit_cov)


sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

# neighbors within, say, 50 km
nb <- dnearneigh(coords, d1 = 0, d2 = 50000)
lw <- nb2listw(nb, style = "W")

moran.test(sf_data$resid, lw)


```

The observed Moran’s I is almost identical to the expected value under
randomness.

The p‑value is very high : no evidence of spatial autocorrelation.

The negative sign is tiny : not meaningful.

```{r}
sampled_data_join_sf$country_id <- as.integer(factor(sampled_data_join_sf$country))  # 1, 2, 3

sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

coords_A <- coords[sf_data$country_id == 1, ]
coords_B <- coords[sf_data$country_id == 2, ]
coords_C <- coords[sf_data$country_id == 3, ]

country_A <- st_transform(shp_ben, 3857)
country_B <- st_transform(shp_cmr, 3857)
country_C <- st_transform(shp_con, 3857)

country_A_union <- st_union(country_A)
country_B_union <- st_union(country_B)
country_C_union <- st_union(country_C)

country_A_simplified <- st_simplify(country_A_union, dTolerance = 10000)
country_B_simplified <- st_simplify(country_B_union, dTolerance = 10000)
country_C_simplified <- st_simplify(country_C_union, dTolerance = 10000)


# Extract geometry
geom_A <- st_geometry(country_A_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_A, "MULTIPOLYGON")) {
  geom_A <- geom_A[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_A <- geom_A[[1]]

# Convert to matrix with X,Y
boundary_A <- as.matrix(outer_ring_A)[, 1:2, drop = FALSE]



# country_B is an sf polygon or multipolygon
geom_B <- st_geometry(country_B_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_B, "MULTIPOLYGON")) {
  geom_B <- geom_B[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_B <- geom_B[[1]]

# Convert to matrix with X,Y
boundary_B <- as.matrix(outer_ring_B)[, 1:2, drop = FALSE]

# Extract geometry
geom_C <- st_geometry(country_C_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_C, "MULTIPOLYGON")) {
  geom_C <- geom_C[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_C <- geom_C[[1]]

# Convert to matrix with X,Y
boundary_C <- as.matrix(outer_ring_C)[, 1:2, drop = FALSE]


# mesh_A <- inla.mesh.2d(loc = coords_A, max.edge = c(5000, 20000))
# mesh_B <- inla.mesh.2d(loc = coords_B, max.edge = c(5000, 20000))
# mesh_C <- inla.mesh.2d(loc = coords_C, max.edge = c(5000, 20000))

mesh_A <- inla.mesh.2d(
  loc      = coords_A,
  boundary = list(boundary_A),
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_B <- inla.mesh.2d(
  loc = coords_B,
  boundary   = list(boundary_B) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_C <- inla.mesh.2d(
  loc = coords_C,
  boundary   = list(boundary_C) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

# # Extract triangle vertex indices
# tri <- mesh_C$graph$tv
# 
# 
# # Build edges from triangles
# edges_list <- list()
# 
# for (i in 1:nrow(tri)) {
#   idx <- tri[i, ]
#   # Each triangle has 3 edges
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(1,2)], 1:2])
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(2,3)], 1:2])
#   edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(3,1)], 1:2])
# }
# 
# mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))
# 
# coords_df_C <- as.data.frame(coords_C)
# colnames(coords_df_C) <- c("X", "Y")
# 
# ggplot() +
#   geom_sf(data = country_C_simplified, fill = "white", color = "black", size = 0.8) +
#   geom_sf(data = mesh_edges_sf, color = "grey60", size = 0.3) +
#   geom_point(data = coords_df_C, aes(X, Y), color = "red", size = 1.5) +
#   theme_minimal() +
#   ggtitle("Mesh A with Country Boundary")

```

```{r}
# Build Matérn SPDE model of the spatial field on mesh

spde_A <- inla.spde2.pcmatern(
  mesh = mesh_A,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)
)

spde_B <- inla.spde2.pcmatern(
  mesh = mesh_B,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)   # 0.12 came from the squared root of the sill because it 
)                              # represent the standard deviance while the sill represent the variance.

spde_C <- inla.spde2.pcmatern(
  mesh = mesh_C,
  prior.range = c(50000, 0.5),
  prior.sigma = c(0.1, 0.5)   
) 

# Computes the interpolation weights that map your SPDE spatial field (defined on mesh nodes) to your actual malaria data locations

# 1. Create indicator weights (1 if the row belongs to the country, 0 if not)
# 'coords' must be the full 36-row matrix of ALL coordinates
w_A <- as.numeric(sf_data$country_id == 1)
w_B <- as.numeric(sf_data$country_id == 2)
w_C <- as.numeric(sf_data$country_id == 3)

# 2. Build the full matrices automatically. 
# Passing the full 'coords' matrix ensures every matrix has exactly 36 rows.
# The 'weights' argument forces rows from other countries to be completely zero.
A_A_full <- inla.spde.make.A(mesh = mesh_A, loc = coords, weights = w_A)
A_B_full <- inla.spde.make.A(mesh = mesh_B, loc = coords, weights = w_B)
A_C_full <- inla.spde.make.A(mesh = mesh_C, loc = coords, weights = w_C)

A_fixed <- Diagonal(n = nrow(sf_data))
A_spatial <- list(A_A_full, A_B_full, A_C_full)


A_A_full <- as(A_A_full, "dgCMatrix")
A_B_full <- as(A_B_full, "dgCMatrix")
A_C_full <- as(A_C_full, "dgCMatrix")

```

# Build stack

```{r}

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "vegetation_index",
                   "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sf_data[[var]], na.rm = TRUE)
  sd_val   <- sd(sf_data[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sf_data[[var]] <- (sf_data[[var]] - mean_val) / sd_val
}

################################################################################
#        checking for overdispersion in our model
################################################################################

# 1. Calculate observed prevalence for each of the 36 villages
village_p <- sf_data$Number_of_infected_tot_pf / sf_data$Number_of_participant_tot

# 2. Calculate the global overall prevalence
global_p <- sum(sf_data$Number_of_infected_tot_pf) / sum(sf_data$Number_of_participant_tot)

# 3. Get the average number of participants per village
mean_N <- mean(sf_data$Number_of_participant_tot)

# 4. Calculate the Observed vs. Theoretical Variance
observed_variance    <- var(village_p)
theoretical_variance <- (global_p * (1 - global_p)) / mean_N

# 5. Compute the Raw Dispersion Ratio
raw_dispersion_ratio <- observed_variance / theoretical_variance
print(raw_dispersion_ratio)

# sf_data <- sf_data %>% 
#   mutate(pop_density_2020 = as.numeric(scale(pop_density_2020)),
#          annual_rainfall_2021 = as.numeric(scale(annual_rainfall_2021)),
#          annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
#          vegetaion_index = as.numeric(scale(vegetaion_index)),
#          dist_to_water  = as.numeric(scale(dist_to_water)),
#          elevation = as.numeric(scale(elevation)))

# Model building

effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sf_data$annual_rainfall_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept             = 1,
  annual_temperature_2021  = sf_data$annual_temperature_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetation_index      = sf_data$vegetation_index,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sf_data$elevation,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_country <- list(
  country_id = sf_data$country_id
)

effects_A <- list(field_A = 1:spde_A$n.spde)
effects_B <- list(field_B = 1:spde_B$n.spde)
effects_C <- list(field_C = 1:spde_C$n.spde)

A_fixed   <- Diagonal(nrow(sf_data))
A_country <- Diagonal(nrow(sf_data))


stack_test_rainfall <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_rainfall,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_temperature <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_temperature,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_vegetation_index <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_vegetation_index,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_elevation <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_elevation,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_baseline <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_baseline,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

```

# Fit our models

```{r}
# Penalize complexity which control how much countries differ from each other through
# the country level random effect
hyper_country <- list(
  prec = list(prior = "pc.prec", param = c(0.5, 0.5))
)

formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetation_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stack_test_rainfall),
  Ntrials = inla.stack.data(stack_test_rainfall)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_rainfall),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stack_test_temperature),
  Ntrials = inla.stack.data(stack_test_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stack_test_vegetation_index),
  Ntrials = inla.stack.data(stack_test_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stack_test_elevation),
  Ntrials = inla.stack.data(stack_test_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stack_test_baseline),
  Ntrials = inla.stack.data(stack_test_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

# ## diasgnostic on the distance to water
# 
# res_simple <- inla(
#   Number_of_infected_tot_pm ~ dist_to_water,
#   family = "binomial",
#   data = sf_data,
#   Ntrials = sf_data$Number_of_participant_tot
# )
# 
# res_simple$summary.fixed


```

The model’s hyperparameters show that country-level differences are
almost negligible, meaning the covariates already explain most variation
between Benin and Congo. The spatial field in Benin is clearly
identifiable: it has a short correlation range (\~7 km) and a moderate
spatial variance, so spatial structure genuinely improves the fit there.
In contrast, Congo’s field is extremely uncertain—with only three
locations, the model cannot reliably estimate its spatial range or
variance, so any spatial effect in Congo should be interpreted
cautiously. The practical consequence is that predictions and spatial
maps will be trustworthy in Benin, while Congo spatial surface will be
weak, noisy, and driven mostly by priors rather than data.

# Comparing the different models

```{r}
# Model 1:  Rainfall

# estimated values for the different covariates

res_rainfall$summary.fixed

# Hyper parameters examination

res_rainfall$summary.hyperpar

# DIC

res_rainfall$dic$dic

# WAIC

res_rainfall$waic$waic

# CPO

sum(log(res_rainfall$cpo$cpo), na.rm = TRUE)

summary(res_rainfall$cpo$cpo)

# PIT(probability integral transform) histogram

hist(res_rainfall$cpo$pit,
     breaks = 20,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

obs <- inla.stack.data(stack_test_rainfall)$y /
       inla.stack.data(stack_test_rainfall)$Ntrials

idx.est <- inla.stack.index(stack_test_rainfall, "est")$data

pred.est <- res_rainfall$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, rainfall model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual rainfall model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 2:  temperature

# estimated values for the different covariates

res_temperature$summary.fixed

# Hyper parameters examination

res_temperature$summary.hyperpar

# DIC

res_temperature$dic$dic

# WAIC

res_temperature$waic$waic

# CPO

sum(log(res_temperature$cpo$cpo), na.rm = TRUE)

summary(res_temperature$cpo$cpo)

# PIT histogram

hist(res_temperature$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_temperature, "est")$data

pred.est <- res_temperature$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title =  "Observed vs predicted, temperature model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual temperature model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```

```{r}
# Model 3:  vegetation index

# estimated values for the different covariates

res_vegetation_index$summary.fixed

# Hyper parameters examination

res_vegetation_index$summary.hyperpar

# DIC

res_vegetation_index$dic$dic

# WAIC

res_vegetation_index$waic$waic

# CPO

sum(log(res_vegetation_index$cpo$cpo), na.rm = TRUE)

summary(res_vegetation_index$cpo$cpo)

# PIT histogram

hist(res_vegetation_index$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_vegetation_index, "est")$data

pred.est <- res_vegetation_index$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, vegetation index model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual vegetation index model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 4:  elevation

# estimated values for the different covariates

res_elevation$summary.fixed

# Hyper parameters examination

res_elevation$summary.hyperpar

# DIC

res_elevation$dic$dic

# WAIC

res_elevation$waic$waic

# CPO

sum(log(res_elevation$cpo$cpo), na.rm = TRUE)

summary(res_elevation$cpo$cpo)

# PIT histogram

hist(res_elevation$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_elevation, "est")$data

pred.est <- res_elevation$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, elevation model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual elevation model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 5:  baseline model

# estimated values for the different covariates

res_baseline$summary.fixed

# Hyper parameters examination

res_baseline$summary.hyperpar

# DIC

res_baseline$dic$dic

# WAIC

res_baseline$waic$waic

# CPO

sum(log(res_baseline$cpo$cpo), na.rm = TRUE)

summary(res_baseline$cpo$cpo)

# PIT histogram

hist(res_baseline$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_baseline, "est")$data

pred.est <- res_baseline$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, baseline model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual baseline model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```

## Conclusion on models comparison (P. malaria model)

The rainfall model is the right choice because it consistently shows the
strongest performance across all diagnostics. It has the lowest DIC and
WAIC, the best CPO, the tightest observed‑vs‑predicted fit, and it
reduces the spatial field variance more than any other covariate,
meaning rainfall explains a large part of the spatial structure instead
of leaving it to the latent field. In short, rainfall improves
prediction, reduces unexplained spatial variation, and aligns with
malaria biology making it the most reliable and scientifically justified
model to continue with.

```{r}
idx_est <- inla.stack.index(stack_test_temperature, "est")$data

fitted_est <- res_temperature$summary.fitted.values[idx_est, "mean"]

fitted_est_25 <- res_temperature$summary.fitted.values[idx_est, "0.025quant"]
fitted_est_75 <- res_temperature$summary.fitted.values[idx_est, "0.975quant"]


fitted_data <- cbind(
  id        = idx_est, 
  observed  = sf_data$PR_Pm_tot,
  fitted    = fitted_est,
  fitted_25 = fitted_est_25,
  fitted_75 = fitted_est_75
)

fitted_data <- as.data.frame(fitted_data)
```

# Model examination

```{r}
# observed vs fitted

ggplot(fitted_data, aes(x = observed, y = fitted)) +
  geom_point(color = "blue", size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "red", lwd = 1.2) +
  labs(x = "Observed proportion",
       y = "Fitted prevalence",
       title = "Observed vs Fitted (INLA)") +
  theme_minimal()

# residual

fitted_data$residuals <- fitted_data$observed - fitted_data$fitted

ggplot(fitted_data, aes(x = 1:nrow(fitted_data), y = residuals)) +
  geom_point(color = "darkgreen", size = 2) +
  geom_hline(yintercept = 0, color = "red", lwd = 1.2) +
  labs(x = "Location index",
       y = "Residual (observed - fitted)",
       title = "Residuals of INLA fitted values") +
  theme_minimal()

# SPDE hyper parameters

res_temperature$summary.hyperpar


```

## *P. f* model examination

-   The observed vs fitted values plot shows that the smooth, shrunk
    estimates of the true prevalence, which are generally consistent
    with the observed but less extreme. the number of points above and
    below the red line are roughly the same which means the model is not
    over estimating nor under estimating.

-   The residual plot shows a rough horizontal cloud indicating that the
    model is stable and most residuals are between -0.1 and 0.1 meaning
    that the model fit well. we also have some moderate residuals around
    0.3 and -0.3 meaning the model smooth appropriately. Finally, the
    number of point above and below the horizontal line are almost the
    same, which means the model is unbiased.

-   The country_id precision is approximately \~15 which means there is
    some meaningful difference between countries after accounting for
    covariates + spatial fields.

-   For all the three field, the range is \~7km to \~9km with a very
    wide credible interval, meaning areas withing these distances are
    spatially correlated or influence each other but the model is unsure
    to which extend because the uncertainty is high. This means that the
    prediction will be smooth but not overly flat.

-   All the three spatial field also have a very small standard
    deviation 0.021 to 0.027, which means spatial variation exist but
    covariates explain most of the variation and spatial field refine
    the prediction instead of dominating it.

# Creating the grid for prevalence estimation

```{r}

```

```{r}
grid_sf <- read_sf("data/grid_data_for_prediction_sf.gpkg")

# Scale the prediction grid using the EXACT same parameters as in the fitting model
# (This ensures a grid cell is measured relative to our 36 observed points)
for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf[[var]] <- (grid_sf[[var]] - mean_val) / sd_val
}


# coords_pred <- st_coordinates(grid_sf)
# 
# coords_pred_A <- coords_pred[grid_sf$country_id == 1, ]
# coords_pred_B <- coords_pred[grid_sf$country_id == 2, ]
# coords_pred_C <- coords_pred[grid_sf$country_id == 3, ]
# 
# A_pred_A <- inla.spde.make.A(mesh = mesh_A, loc = coords_pred_A)
# A_pred_B <- inla.spde.make.A(mesh = mesh_B, loc = coords_pred_B)
# A_pred_C <- inla.spde.make.A(mesh = mesh_C, loc = coords_pred_C)
# 
# A_pred_A_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_A$n.spde)
# A_pred_B_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_B$n.spde)
# A_pred_C_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_C$n.spde)
# 
# A_pred_A_full[grid_sf$country_id == 1, ] <- as.matrix(A_pred_A)
# A_pred_B_full[grid_sf$country_id == 2, ] <- as.matrix(A_pred_B)
# A_pred_C_full[grid_sf$country_id == 3, ] <- as.matrix(A_pred_C)


# 1. Get the full coordinates matrix for all 37,515 grid cells
coords_pred <- st_coordinates(grid_sf)

# 2. Create 0/1 indicator weights across ALL 37,515 rows
w_pred_A <- as.numeric(grid_sf$country_id == 1)
w_pred_B <- as.numeric(grid_sf$country_id == 2)
w_pred_C <- as.numeric(grid_sf$country_id == 3)

# 3. Build native, optimized Sparse Matrices (dgCMatrix) automatically.
# Passing 'coords_pred' means every matrix instantly has exactly 37,515 rows.
# 'weights' forces cells outside the country to stay zero efficiently.
A_pred_A_full <- inla.spde.make.A(mesh = mesh_A, loc = coords_pred, weights = w_pred_A)
A_pred_B_full <- inla.spde.make.A(mesh = mesh_B, loc = coords_pred, weights = w_pred_B)
A_pred_C_full <- inla.spde.make.A(mesh = mesh_C, loc = coords_pred, weights = w_pred_C)

# Verify they are sparse structures
class(A_pred_A_full) # Should output "dgCMatrix"

effects_fixed_pred <- data.frame(
  intercept               = 1,
  # annual_rainfall_2021    = grid_sf$annual_rainfall_2021,
  annual_temperature_2021 = grid_sf$annual_temperature_2021,
  pop_density_2020        = grid_sf$pop_density_2020,
  # annual_ITN_usage_2021   = grid_sf$annual_ITN_usage_2021,
  dist_to_water           = grid_sf$dist_to_water,
  lc_water                = grid_sf$lc_water,
  lc_trees                = grid_sf$lc_trees,
  lc_built                = grid_sf$lc_built,
  lc_cropland             = grid_sf$lc_cropland,
  lc_wetland              = grid_sf$lc_wetland)

effects_country_pred <- list(country_id = grid_sf$country_id)

effects_A_pred <- list(field_A = 1:spde_A$n.spde)
effects_B_pred <- list(field_B = 1:spde_B$n.spde)
effects_C_pred <- list(field_C = 1:spde_C$n.spde)

stack_pred <- inla.stack(
  data = list(y = NA,
    Ntrials = NA),
  A = list(
    Diagonal(nrow(grid_sf)),      # fixed effects
    Diagonal(nrow(grid_sf)),      # country iid
    A_pred_A_full,                # spatial field A
    A_pred_B_full,                # spatial field B
    A_pred_C_full                 # A_pred_B_full C
  ),
  effects = list(
    effects_fixed_pred,
    effects_country_pred,
    effects_A_pred,
    effects_B_pred,
    effects_C_pred
  ),
  tag = "pred"
)

stack_full <- inla.stack(stack_test_temperature, stack_pred)

res_pred <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stack_full),
  Ntrials = inla.stack.data(stack_full)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_full),
    compute = TRUE,
    link = 1     
  ),
  control.fixed = list(mean = 0, prec = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)

#idx_pred <- inla.stack.index(stack_full, "pred")$data


set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stack_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]                   # linear predictor for grid cells
  plogis(lp)                                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)



# v1 <- res_pred$summary.fixed
# v2 <- res_pred$summary.fixed

```

## P. malaria model interpretation

### clustered model

Across the three‑country study area, *Plasmodium malariae* prevalence is
strongly influenced by environmental conditions. Rainfall and wetland
show clear positive associations with infection risk, indicating that
humid and water‑rich ecosystems provide favorable breeding conditions
for malaria vectors. Built‑up areas display a strong negative effect,
reflecting reduced transmission in urbanized landscapes where mosquito
habitats are disrupted. Distance to water shows a small but
statistically meaningful positive effect. In contrast, cropland,
population density, forest cover, and land‑cover water all have credible
intervals that include zero, meaning the data do not provide sufficient
evidence to classify their influence as either protective or
risk‑enhancing. Overall, rainfall and wetlands emerge as key drivers of
*P. malariae* transmission, with urbanization acting as an important
suppressor, while several other covariates show uncertain or negligible
effects.

### non clustered model

The updated model produces results that are broadly consistent with the
previous analysis, identifying rainfall and wetland environments as
strong positive drivers of *Plasmodium malariae* prevalence, and
built‑up areas as a strong negative predictor. However, the new model
shows slightly reduced effect sizes for rainfall and increased
uncertainty for several covariates. Notably, the effect of distance to
water, previously statistically positive, now has a credible interval
that includes zero, indicating insufficient evidence for a clear
association. Cropland, population density, forest cover, and land‑cover
water continue to show wide credible intervals that cross zero, meaning
their effects remain statistically uncertain. Overall, the main
difference between the two models lies in the reduced certainty around
distance to water and slightly adjusted effect sizes, while the key
environmental drivers rainfall, wetlands, and urbanization remain
robust.

# *P. falciparum* prevalence from Malaria Atlas Project (MAP)

```{r}
# Cameroon case

# Load the modeled prevalence from MAP

africa_pf_pr_2_10 <- rast("data/prevalence_2000/2025_GBD2024_Global_PfPR_2021.tif")

grid_sf_vect <- vect(grid_sf %>% st_transform(4326))

africa_pf_pr_2021 <- crop(africa_pf_pr_2_10, grid_sf_vect)
africa_pf_pr_2021 <- mask(africa_pf_pr_2021, grid_sf_vect)

grid_sf_map_prev <- terra::extract(africa_pf_pr_2021, grid_sf %>% st_transform(4326))

grid_sf_map_prev <- grid_sf_map_prev %>% 
  select(ID, Data, LCI, UCI)

grid_sf <- grid_sf %>% 
  mutate(PR_2_10_map     = grid_sf_map_prev[,2],
         PR_2_10_map_L   = grid_sf_map_prev[,3],
         PR_2_10_map_U   = grid_sf_map_prev[,4])

numeric_cols <- names(grid_sf)[sapply(grid_sf, is.numeric)]

for (col in numeric_cols) {
  grid_sf <- fill_by_nearest(grid_sf, col)
}

grid_sf$PR_all_age <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")

grid_sf$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map_U,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")

grid_sf$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = grid_sf$PR_2_10_map_L,
  age_min_in   = rep(2, nrow(grid_sf)),
  age_max_in   = rep(10, nrow(grid_sf)),
  age_min_out  = rep(0, nrow(grid_sf)),
  age_max_out  = rep(99, nrow(grid_sf)),
  parameters   = "Pf_Smith2007")



grid_sf$GID_0 <- ifelse(grid_sf$country_id == 1, "BEN",
                 ifelse(grid_sf$country_id == 2, "CMR",
                 ifelse(grid_sf$country_id == 3, "COG", NA)))


grid_benin <- grid_sf %>% filter(GID_0 == "BEN")
shp_ben1 <-  st_transform(shp_ben, st_crs(grid_benin))

grid_benin1 <- st_join(grid_benin, shp_ben1, join = st_intersects)

ben_data_pev <- grid_benin1 %>%
  group_by(NAME_2) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


ben_data_pev_map <- shp_ben %>% group_by(NAME_2) %>% summarise(geometry = st_union(geometry)) %>%
  left_join(ben_data_pev %>% st_drop_geometry(), by = "NAME_2")

ben_prev_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_pm.pdf",
       plot = ben_prev_1, width = 6, height = 4, dpi = 300)


ben_prev_uncertainty_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_uncertainty_pm.pdf",
       plot = ben_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_cmr <- grid_sf %>% filter(GID_0 == "CMR")
shp_cmr1 <-  st_transform(shp_cmr, st_crs(grid_cmr))

grid_cmr1 <- st_join(grid_cmr, shp_cmr1, join = st_intersects)

cmr_data_pev <- grid_cmr1 %>%
  group_by(Admin2) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


cmr_data_pev_map <- shp_cmr %>%
  left_join(cmr_data_pev %>% st_drop_geometry(), by = "Admin2")

cmr_prev_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_pm.pdf", 
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

cmr_prev_uncertainty_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_uncertainty_pm.pdf", 
       plot = cmr_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_con <- grid_sf %>% filter(GID_0 == "COG")
shp_con1 <-  st_transform(shp_con, st_crs(grid_con))

grid_con1 <- st_join(grid_con, shp_con1, join = st_intersects)

con_data_pev <- grid_con1 %>%
  group_by(adm2_name) %>%
  summarise(
    prev_mean        = mean(prev_mean, na.rm = TRUE),
    PR_all_age       = mean(PR_all_age, na.rm = TRUE),
    prev_q025        = mean(prev_q025, na.rm = TRUE),
    prev_q975        = mean(prev_q975, na.rm = TRUE),
    PR_all_age_U     = mean(PR_all_age_U, na.rm = TRUE),
    PR_all_age_L     = mean(PR_all_age_L, na.rm = TRUE),
    CI_width         = mean(prev_q975 - prev_q025, na.rm = TRUE),
    CI_width_all_age = mean(PR_all_age_U - PR_all_age_L, na.rm = TRUE)
  )


con_data_pev_map <- shp_con %>%
  left_join(con_data_pev %>% st_drop_geometry(), by = "adm2_name")

con_prev_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/cong_estimated_prev_pm.pdf", 
       plot = con_prev_1, width = 6, height = 4, dpi = 300)


con_prev_uncertainty_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/con_estimated_prev_uncertainty_pm.pdf", 
       plot = con_prev_uncertainty_1, width = 6, height = 4, dpi = 300)
```

```{r}
cmr_prev_pf_MAP_estimates <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_MAP_pf_all_age.pdf",
       plot = cmr_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

cmr_prev_pf_MAP_estimates_uncertainty <- ggplot(cmr_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_uncertainty_MAP_pf_all_age.pdf",
       plot = cmr_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)


# Benin case

ben_prev_pf_MAP_estimates <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_MAP_pf_all_age.pdf",
       plot = ben_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

ben_prev_pf_MAP_estimates_uncertainty <- ggplot(ben_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_uncertainty_MAP_pf_all_age.png",
       plot = ben_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)

# Congo case

con_prev_pf_MAP_estimates <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_MAP_pf_all_age.pdf",
       plot = con_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

con_prev_pf_MAP_estimates_uncertainty <- ggplot(con_data_pev) +
  geom_sf(aes(fill = CI_width_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_uncertainty_MAP_pf_all_age.png",
       plot = con_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)

```

# Comparison between COMAL estimates and MAP estimates

```{r}
# Divide the burden into classes

ben_data_pev_map <- ben_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))


cmr_data_pev_map <- cmr_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))

con_data_pev_map <- con_data_pev_map %>% 
  mutate(trans_intensity_comal = cut(
    prev_mean,
    breaks = quantile(prev_mean, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE),
    trans_intensity_map = cut(
    PR_all_age,
    breaks = quantile(PR_all_age, probs = seq(0, 1, length.out = 4), na.rm = TRUE),
    labels = c("Low", "Medium", "High"),
    include.lowest = TRUE))


# Rank correlation analysis

spearman_res_ben <- cor.test(ben_data_pev_map$prev_mean, ben_data_pev_map$PR_all_age, method = "spearman")

spearman_res_cmr <- cor.test(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "spearman")
spearman_res_con <- cor.test(con_data_pev_map$prev_mean, con_data_pev_map$PR_all_age, method = "spearman")

comal_trans_intens_cmr <- ggplot(cmr_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pf_pr_estimated/comal_trans_intens_cmr.png",
       plot = comal_trans_intens_cmr, width = 6, height = 4, dpi = 300)


map_trans_intens_cmr <- ggplot(cmr_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/map_trans_intens_cmr.png",
       plot = map_trans_intens_cmr, width = 6, height = 4, dpi = 300)


comal_trans_intens_ben <- ggplot(ben_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pf_pr_estimated/comal_trans_intens_ben.png",
       plot = comal_trans_intens_ben, width = 6, height = 4, dpi = 300)


map_trans_intens_ben <- ggplot(ben_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/map_trans_intens_ben.png",
       plot = map_trans_intens_ben, width = 6, height = 4, dpi = 300)


comal_trans_intens_cong <- ggplot(con_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_comal), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "COMAL transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pf_pr_estimated/comal_trans_intens_cong.png",
       plot = comal_trans_intens_cong, width = 6, height = 4, dpi = 300)


map_trans_intens_cong <- ggplot(con_data_pev_map%>% drop_na())+
  geom_sf(aes(fill = trans_intensity_map), color = "white", size = 0.2)+
  scale_fill_brewer(palette = "YlOrRd", name = "Classe", na.value = "transparent")+
  labs(title = "MAP transmission classes (Quantiles)")+
  theme_void()+
  coord_sf()+
  theme_minimal()+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/map_trans_intens_cong.png",
       plot = map_trans_intens_cong, width = 6, height = 4, dpi = 300)


```

# prepare data for the continental level estimation

```{r}
shp_global <- read_sf("data/Africa_data/shape_files_1/geoBoundariesCGAZ_ADM2.shp")

custom_match <- c(
  XKX = "Europe",
  ESH = "Africa",
  `111` = "Africa",        # Abyei
  `112` = "Asia",          # Aksai Chin
  `113` = "Asia",          # China–India border
  `114` = "Asia",          # Demchok
  `115` = "Europe",        # Dragonja
  `116` = "Africa",        # Dramana-Shakatoe
  `117` = "South America", # Falkland Islands
  `118` = "Asia",          # Gaza Strip
  `119` = "Asia",          # Kalapani
  `120` = "South America", # Isla Brasilera
  `121` = "Asia",          # Siachen-Saltoro
  `122` = "Africa",        # Koualou
  `123` = "Asia",          # Liancourt Rocks
  `124` = "Asia",          # No Man's Land
  `125` = "Asia",          # Paracel Islands
  `126` = "Asia",          # Sanafir & Tiran Islands
  `127` = "Asia",          # Senkaku Islands
  `128` = "Asia",          # Spratly Islands
  `129` = "Asia"           # West Bank
)

shp_global$continent <- countrycode(
  shp_global$shapeGroup,
  origin = "iso3c",
  destination = "continent",
  custom_match = custom_match
)

shp_africa <- shp_global %>% 
  filter(continent=="Africa")

custom_match <- c(
  `111` = "Abyei",                   # Abyei
  `116` = "Dramana-Shakatoe",        # Dramana-Shakatoe
  `122` = "Koualou"                  # Koualou
  )


shp_africa$country <- countrycode(
  shp_africa$shapeGroup,
  origin = "iso3c",
  destination = "country.name",
  custom_match = custom_match
)

shp_africa_simple <- ms_simplify(
  shp_africa,
  keep = 0.05,          # keep 5% of the vertices
  keep_shapes = TRUE
)

st_write(
  shp_africa_simple,
  "data/Africa_data/shp_africa/Africa_simplified.shp",
  delete_layer = TRUE,
  append = TRUE
)


ggplot()+
  geom_sf(data = shp_africa_simple, fill = "grey", color = "black")

```

To avoid extrapolating predictions into areas where malaria transmission
is not considered ecologically feasible, model predictions will be
masked using the *Plasmodium falciparum* transmission limits provided by
the Malaria Atlas Project. Although this study focuses on *Plasmodium
malariae*, no continent-wide transmission limits are currently available
for this species. Given that *P. malariae* is predominantly transmitted
in areas where *P. falciparum* is endemic and frequently occurs as a
co-infection, the *P. falciparum* transmission limits will be used as a
proxy for malaria transmission suitability.

```{r}
sampled_data_join_sf <- st_read("data/Africa_data/sampled_data_join_sf.gpkg")
grid_sf_africa <- st_read("data/Africa_data/grid_data_for_prediction_sf.gpkg")


shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp") %>% st_transform(3857)

Africa_union <- st_union(shp_africa )

sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

sampled_data_join_sf_sp     <- as(sampled_data_join_sf, "Spatial")
Africa_union_sp <- as(Africa_union, "Spatial")

coords_Africa <- coordinates(sampled_data_join_sf_sp)


mesh_Africa <- inla.mesh.2d(
  loc      = coords_Africa,
  boundary = inla.sp2segment(Africa_union_sp),
  max.edge = c(1500e3, 2000e3),   
  cutoff   = 250e3,
  offset   = c(100e3, 2000e3))

# Extract triangle vertex indices
tri <- mesh_Africa$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_Africa <- as.data.frame(coords_Africa)
colnames(coords_df_Africa) <- c("X", "Y")

ggplot() +
  geom_sf(data = mesh_edges_sf, fill = NA, color = "grey60", size = 0.3) +
  geom_sf(data = Africa_union, fill = NA, color = "black", size = 0.8) +
  geom_point(data = coords_df_Africa, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh for Africa with Boundary")
```

```{r}
spde_Africa <- inla.spde2.pcmatern(
  mesh = mesh_Africa,
  prior.range = c(100000, 0.5),
  prior.sigma = c(0.2, 0.5)
)

A_est <- inla.spde.make.A(mesh_Africa, loc = coords_Africa)

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "vegetation_index",
                   "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sampled_data_join_sf[[var]], na.rm = TRUE)
  sd_val   <- sd(sampled_data_join_sf[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sampled_data_join_sf[[var]] <- (sampled_data_join_sf[[var]] - mean_val) / sd_val
}


```

```{r}
effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sampled_data_join_sf$annual_rainfall_2021,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept                = 1,
  annual_temperature_2021  = sampled_data_join_sf$annual_temperature_2021,
  pop_density_2020         = sampled_data_join_sf$pop_density_2020,
  dist_to_water            = sampled_data_join_sf$dist_to_water,
  lc_water                 = sampled_data_join_sf$lc_water,
  lc_trees                 = sampled_data_join_sf$lc_trees,
  lc_built                 = sampled_data_join_sf$lc_built,
  lc_cropland              = sampled_data_join_sf$lc_cropland,
  lc_wetland               = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sampled_data_join_sf$elevation,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


stk_est_rainfall <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_rainfall
  ),
  tag = "est"
)

stk_est_temperature <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_temperature
  ),
  tag = "est"
)


stk_est_vegetation_index <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_vegetation_index
  ),
  tag = "est"
)

stk_est_elevation <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_elevation
  ),
  tag = "est"
)


stk_est_baseline <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_baseline
  ),
  tag = "est"
)

```

```{r}
formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetation_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stk_est_rainfall  ),
  Ntrials = inla.stack.data(stk_est_rainfall )$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_rainfall ),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stk_est_temperature),
  Ntrials = inla.stack.data(stk_est_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stk_est_vegetation_index),
  Ntrials = inla.stack.data(stk_est_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stk_est_elevation),
  Ntrials = inla.stack.data(stk_est_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stk_est_baseline),
  Ntrials = inla.stack.data(stk_est_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


hist(res_temperature$cpo$pit,
     breaks = 20)
```

```{r}

samples <- inla.posterior.sample(
  n = 100,
  result = res_rainfall 
)

# extraction of the fix effect

fixed_names <- res_rainfall$names.fixed

fixed_idx <- match(
  paste0(fixed_names, ":1"),
  rownames(samples[[1]]$latent)
)

# extract the spatial effect

spatial_idx <- grep(
  "^spatial:",
  rownames(samples[[1]]$latent)
)



idx_est <- inla.stack.index(stk_est_rainfall, "est")$data

fitted_est <- res_rainfall$summary.fitted.values[idx_est, "mean"]

cbind(
  observed = sampled_data_join_sf$PR_Pm_tot,
  fitted   = fitted_est
)[1:36, ]
```

# Given the large files, let's break the grid of prediction into tiles

```{r}

for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf_africa[[var]] <- (grid_sf_africa[[var]] - mean_val) / sd_val
}


## Choose number of tiles
n_tiles  <- 200
tile_id  <- cut(seq_len(nrow(grid_sf_africa)), breaks = n_tiles, labels = FALSE)
n_samp   <- 200  # posterior samples per tile

# Storage
pred_mean_full <- numeric(nrow(grid_sf_africa))
pred_sd_full   <- numeric(nrow(grid_sf_africa))
pred_q05_full  <- numeric(nrow(grid_sf_africa))
pred_q95_full  <- numeric(nrow(grid_sf_africa))

for (tile in seq_len(n_tiles)) {
  cat("Tile", tile, "of", n_tiles, "\n")
  
  # 1) Subset grid
  grid_tile <- grid_sf_africa[tile_id == tile, ]
  
  # 2) A-matrix for this tile
  A_pred_tile <- inla.spde.make.A(
    mesh = mesh_Africa,
    loc  = st_coordinates(grid_tile)
  )
  
  # 3) Fixed effects for this tile
  effects_pred_tile <- data.frame(
    intercept = 1,
    annual_temperature_2021 = grid_tile$annual_temperature_2021,
    pop_density_2020        = grid_tile$pop_density_2020,
    dist_to_water           = grid_tile$dist_to_water,
    lc_cropland             = grid_tile$lc_cropland,
    lc_water                = grid_tile$lc_water,
    lc_wetland              = grid_tile$lc_wetland,
    lc_built                = grid_tile$lc_built,
    lc_trees                = grid_tile$lc_trees
  )
  
  # 4) Prediction stack for this tile
  stk_pred_tile <- inla.stack(
    data    = list(y = NA, Ntrials = 1),
    A       = list(A_pred_tile, 1),
    effects = list(
      spatial = 1:spde_Africa$n.spde,
      effects_pred_tile
    ),
    tag = "pred"
  )
  
  # 5) Full stack for this tile
  stk_full_tile <- inla.stack(stk_est_temperature, stk_pred_tile)
  
  # 6) Fit INLA with config = TRUE for posterior sampling
  res_pred_tile <- inla(
    formula_temperature,
    family = "binomial",
    data   = inla.stack.data(stk_full_tile),
    Ntrials = inla.stack.data(stk_full_tile)$Ntrials,
    control.predictor = list(
      A       = inla.stack.A(stk_full_tile),
      compute = TRUE,
      link    = 1
    ),
    control.fixed = list(mean = 0, prec = 0.2),
    control.compute = list(config = TRUE)
  )
  
  # 7) Posterior samples for this tile
  samples_tile <- inla.posterior.sample(n_samp, res_pred_tile)
  
  # 8) Identify latent indices (once per tile)
  latent_names <- rownames(samples_tile[[1]]$latent)
  
  fixed_base_names   <- rownames(res_pred_tile$summary.fixed)
  fixed_latent_names <- paste0(fixed_base_names, ":1")
  fixed_idx          <- match(fixed_latent_names, latent_names)
  
  spatial_idx <- grep("^spatial:", latent_names)
  
  # 9) Build linear predictor samples for this tile
  n_grid_tile <- nrow(grid_tile)
  X_pred_tile <- as.matrix(effects_pred_tile)
  eta_mat_tile <- matrix(NA_real_, nrow = n_grid_tile, ncol = n_samp)
  
  for (s in seq_len(n_samp)) {
    latent_s   <- samples_tile[[s]]$latent
    beta_s     <- latent_s[fixed_idx]
    spatial_s  <- latent_s[spatial_idx]
    
    eta_fixed_s   <- as.vector(X_pred_tile %*% beta_s)
    eta_spatial_s <- as.vector(A_pred_tile %*% spatial_s)
    
    eta_mat_tile[, s] <- eta_fixed_s + eta_spatial_s
  }
  
  # 10) Transform to probabilities and summarize
  prob_mat_tile <- 1 / (1 + exp(-eta_mat_tile))
  
  pred_mean_tile <- rowMeans(prob_mat_tile)
  pred_sd_tile   <- apply(prob_mat_tile, 1, sd)
  pred_q05_tile  <- apply(prob_mat_tile, 1, quantile, 0.05)
  pred_q95_tile  <- apply(prob_mat_tile, 1, quantile, 0.95)
  
  # 11) Store back into full vectors
  idx <- (tile_id == tile)
  pred_mean_full[idx] <- pred_mean_tile
  pred_sd_full[idx]   <- pred_sd_tile
  pred_q05_full[idx]  <- pred_q05_tile
  pred_q95_full[idx]  <- pred_q95_tile
}

grid_sf_africa$pred_mean <- pred_mean_full
grid_sf_africa$pred_sd   <- pred_sd_full
grid_sf_africa$pred_q05  <- pred_q05_full
grid_sf_africa$pred_q95  <- pred_q95_full



```

```{r}


# Build the A‑matrix for prediction locations
# A_pred <- inla.spde.make.A(
#   mesh = mesh_Africa,
#   loc = st_coordinates(grid_sf_africa))
# 
# #Build the prediction stack
# 
# effects_pred <- data.frame(
#   intercept = 1,
#   annual_temperature_2021 = grid_sf_africa$annual_temperature_2021,
#   pop_density_2020        = grid_sf_africa$pop_density_2020,
#   dist_to_water           = grid_sf_africa$dist_to_water,
#   lc_cropland             = grid_sf_africa$lc_cropland,
#   lc_water                = grid_sf_africa$lc_water,
#   lc_wetland              = grid_sf_africa$lc_wetland,
#   lc_built                = grid_sf_africa$lc_built,
#   lc_trees                = grid_sf_africa$lc_trees
# )
# 
# 
# stk_pred <- inla.stack(
#   data = list(y = NA, Ntrials = 1),
#   A    = list(A_pred, 1),
#   effects = list(
#     spatial = 1:spde_Africa$n.spde,
#     effects_pred),
#   tag = "pred"
# )
# 
# 
# # Combine estimation + prediction stacks
# stk_full <- inla.stack(stk_est_temperature, stk_pred)
# 
# # Fit the model again with the full stack
# res_pred <- inla(
#   formula_temperature,
#   family = "binomial",
#   data = inla.stack.data(stk_full),
#   Ntrials = inla.stack.data(stk_full)$Ntrials,
#   control.predictor = list(
#     A = inla.stack.A(stk_full),
#     compute = TRUE,
#     link    = 1
#   ),
#   control.fixed = list(mean = 0, prec = 1)
#   #control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
# )
# 
# 
# # Fixing the prediction error
# 
# n_samp <- 200  
# 
# samples <- inla.posterior.sample(n_samp, res_temperature)
# 
# # Names of all latent variables in the first sample
# latent_names <- rownames(samples[[1]]$latent)
# 
# # Fixed effects: use names from res_temperature$summary.fixed
# fixed_base_names <- rownames(res_temperature$summary.fixed)
# fixed_latent_names <- paste0(fixed_base_names, ":1")
# fixed_idx <- match(fixed_latent_names, latent_names)
# 
# # Spatial field: usually starts with "spatial"
# spatial_idx <- grep("^spatial", latent_names)
# 
# ## Loop over samples and build the linear predictor on the grid
# 
# n_grid <- nrow(grid_sf_africa)
# 
# eta_mat <- matrix(NA_real_, nrow = n_grid, ncol = n_samp)
# 
# X_pred <- as.matrix(effects_pred)  # design matrix for fixed effects
# 
# for (i in seq_len(n_samp)) {
# 
#   latent_i <- samples[[i]]$latent
# 
#   beta_i    <- latent_i[fixed_idx]
#   spatial_i <- latent_i[spatial_idx]
# 
#   # sanity checks
#   if (any(is.na(beta_i)))    stop("beta_i contains NA")
#   if (any(is.na(spatial_i))) stop("spatial_i contains NA")
# 
#   eta_fixed_i   <- as.vector(X_pred %*% beta_i)
#   eta_spatial_i <- as.vector(A_pred %*% spatial_i)
# 
#   eta_mat[, i] <- eta_fixed_i + eta_spatial_i
# }
# 
# ## Transform to probabilities and compute uncertainty summaries
# 
# prob_mat <- 1 / (1 + exp(-eta_mat))
# 
# pred_mean <- rowMeans(prob_mat)
# pred_sd   <- apply(prob_mat, 1, sd)
# pred_q05  <- apply(prob_mat, 1, quantile, 0.05)
# pred_q95  <- apply(prob_mat, 1, quantile, 0.95)
# 
# grid_sf_africa$pred_mean <- pred_mean
# grid_sf_africa$pred_sd   <- pred_sd
# grid_sf_africa$pred_q05  <- pred_q05
# grid_sf_africa$pred_q95  <- pred_q95


# Extract predictions for the grid

# index_pred <- inla.stack.index(stk_full, "pred")$data
# index_est <- inla.stack.index(stk_full, "est")$data
# 
# grid_sf_africa$pred_mean <- res_pred$summary.fitted.values[index_pred, "mean"]
# grid_sf_africa$pred_sd   <- res_pred$summary.fitted.values[index_pred, "sd"]
# grid_sf_africa$pred_q025 <- res_pred$summary.fitted.values[index_pred, "0.025quant"]
# grid_sf_africa$pred_q975 <- res_pred$summary.fitted.values[index_pred, "0.975quant"]

# Using the transmission limit to mask the prediction grid

grid_sf_africa_wgs84 <- st_transform(grid_sf_africa, 4326)
grid_vect <- vect(grid_sf_africa_wgs84)

df_grid_africa <- grid_sf_africa_wgs84 %>%
    mutate(
        x = st_coordinates(.)[,1],
        y = st_coordinates(.)[,2]
    ) %>%
    st_drop_geometry()

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")

# global_spatial_limit_3857 <- project(global_spatial_limit, crs(shp_africa))


pf_limit_africa <- crop(global_spatial_limit, vect(shp_africa %>% st_transform(4326)))
pf_limit_africa <- mask(pf_limit_africa, vect(shp_africa %>% st_transform(4326)))


grid_with_limit <- terra::extract(pf_limit_africa, grid_vect)

grid_masked <- cbind(df_grid_africa, grid_with_limit[,2])

grid_masked_1 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when(is.na(limit) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q05   = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q05),
         pred_q95   = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q95)) 

grid_masked_2 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when((is.na(limit) | limit == 1) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q05  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q05),
         pred_q95  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q95)) 


grid_masked_sf <- st_as_sf(
    grid_masked_1,
    coords = c("x", "y"),
    crs = 4326)

grid_masked_sf_1 <- st_as_sf(
    grid_masked_2,
    coords = c("x", "y"),
    crs = 4326)

# shp_africa1 <-  st_transform(shp_africa, st_crs(grid_masked_sf))
# 
# 
# grid_africa1 <- st_join(grid_masked_sf, shp_africa1%>% 
#   select(-c(shapeName, shapeID, shapeGroup, shapeType, continent, country)), 
#   join = st_intersects) 
# 
# Africa_data_pev <- grid_africa1 %>%
#   group_by(shapeID) %>%
#   summarise(
#     prev_mean   = mean(pred_mean, na.rm = TRUE),
#     prev_q025   = mean(pred_q025, na.rm = TRUE),
#     prev_q975   = mean(pred_q975, na.rm = TRUE),
#     CI_width    = mean(pred_q975 - pred_q025, na.rm = TRUE))
# 
# Africa_country_prev_tx <- country_tx_sf %>%
#   left_join(Africa_country_prev, by = "shapeName")
# 
# Africa_data_pev <- Africa_data_pev %>%
#   mutate(prev_mean = if_else(is.nan(prev_mean), NA_real_, prev_mean))
# 
# Africa_data_pev_map <- shp_africa %>%
#   left_join(Africa_data_pev %>% st_drop_geometry(), by = "shapeID")
# 

shp_africa_country <- shp_africa %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry))


shp_africa_countryw4326 <- shp_africa_country %>% st_transform(4326)
shp_africaw4326 <- shp_africa %>% st_transform(4326)


Africa_prev_1 <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africaw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_admin2_level_unstable1.pdf",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country_1 <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africa_countryw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_country_level_unstable.pdf",
       plot = Africa_prev_country_1, width = 6, height = 4, dpi = 300)


# setting unstable areas to 0

Africa_prev_2 <- ggplot() +
  geom_sf(data = grid_masked_sf_1,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africaw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_admin2_level1.pdf",
       plot = Africa_prev_2, width = 6, height = 4, dpi = 300)

Africa_prev_country_2 <- ggplot() +
  geom_sf(data = grid_masked_sf_1,
          aes(color = pred_mean),
          size = 0.1,
          alpha = 0.4) +
  geom_sf(data = shp_africa_countryw4326,
          fill = NA,
          color = "black",
          size = 0.05) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_country_level.pdf",
       plot = Africa_prev_country_2, width = 6, height = 4, dpi = 300)

```

```{r}
set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stk_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]   # linear predictor for grid cells
  plogis(lp)                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
