---
title: "COMAL project"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load necessary libraries
library(sf)
library(reshape2)
library(ggcorrplot)
library(leaflet)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(tidyterra)
library(mapview)
library(spData)
library(gganimate)
library(htmlwidgets)
library(webshot)
library(chirps)
library(viridis)
library(patchwork)
library(GGally)
library(spdep)
library(INLA)
library(leafsync)
library(SpatialEpi)
library(geodata)
library(car)
library(gstat)
library(nngeo)
library(tidygeocoder)
library(malariaAtlas)
library(rmapshaper)
library(tidytext)
library(AER) #over dispersion package
```


```{r}
#loading the shape files
shp_ben <- read_sf("data/ben_data/BEN_shape_file/gadm41_BEN_3.shp")
shp_con <- read_sf("data/cong_data/CON_shape_file/cog_admin2.shp")
shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")
shp_gab <- read_sf("data/gab_data/gab_shape/gadm41_GAB_2.shp")


#renaming the districts

shp_ben$NAME_3[shp_ben$NAME_3=="Gakpè"]="Gakpe"
shp_ben$NAME_3[shp_ben$NAME_3=="Tokpa Domé"]="Tokpa-dome"
shp_ben$NAME_3[shp_ben$NAME_3=="Ouakpé-Daho"]="Houakpe-daho"
shp_ben$NAME_3[shp_ben$NAME_3=="Aganmalomé"]="Aganmanlome"
shp_ben$NAME_3[shp_ben$NAME_3=="Agonkanmè"]="Agonkamey"

ben_vect <- vect(shp_ben)
cmr_vect <- vect(shp_cmr)
con_vect <- vect(shp_con)
gab_vect <- vect(shp_gab)

sampled_data_join_sf <- st_read("data/Africa_data/sampled_data_join_sf.gpkg")

```

# Overdispersion check

```{r}
#| label: Overdispersion

 # scaling of the varibles

sampled_data_join_sf_scale <- sampled_data_join_sf %>% 
  mutate(annual_rainfall_2021    = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
         pop_density_2020        = as.numeric(scale(pop_density_2020)),
         dist_to_water           = as.numeric(scale(dist_to_water)),
         elevation               = as.numeric(scale(elevation)),
         vegetation_index        = as.numeric(scale(vegetation_index)))


glm_binom1 <- glm(
  cbind(Number_of_infected_tot_pm, Number_of_participant_tot - Number_of_infected_tot_pm) ~ 1,
  family = "binomial",
  data = sampled_data_join_sf_scale
)

pearson_res1 <- residuals(glm_binom1, type = "pearson")

phi <- sum(pearson_res1^2) / glm_binom1$df.residual

phi

glm_binom2 <- glm(
  cbind(Number_of_infected_tot_pm, Number_of_participant_tot - Number_of_infected_tot_pm) ~
    annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water + lc_trees + lc_cropland +
    lc_built + lc_wetland,
  family = "binomial",
  data = sampled_data_join_sf_scale
)

pearson_res2 <- residuals(glm_binom2, type = "pearson")

phi <- sum(pearson_res2^2) / glm_binom2$df.residual

phi
```
- With no covariate, the dispersion coefficient is greater than 6 which is a 
clear sign of over dispersion and even when we add covariate, the coefficient 
is still high (3.5). Therefore, the variation in the observed data is more than what is
expected under the binomial model.



```{r}
num_vars <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetation_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees", "annual_ITN_usage_2021",
  "lc_built", "lc_cropland", "lc_wetland"))

num_vars_1 <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c("dist_to_water", "PR_Pm_tot"))

M1 <- cor(num_vars_1, use = "pairwise.complete.obs")

corrplot(
  M1,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

var1 <- ggcorrplot(
  M,
  type = "full",
  legend.title = "Correlation Matrix",
  lab = TRUE,
  lab_size = 3,
  colors = c("blue", "white", "red"),
  outline.color = "white",
  show.diag = TRUE       
) +
  scale_x_discrete(name = NULL) +
  scale_y_discrete(name = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),      
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Outputs/correlation_matrix.png", var1, dpi = 300, width = 6, height = 4)

num_vars_selected <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

M_selected <- cor(num_vars_selected, use = "pairwise.complete.obs")

corrplot(
  M_selected,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + 
    annual_temperature_2021 + elevation + vegetation_index + pop_density_2020 + 
    dist_to_water + lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021,
  data = sampled_data_join_sf
)

vif(fit)

fit1 <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water
  + lc_trees + lc_built + lc_cropland + lc_wetland + vegetation_index, 
  data = sampled_data_join_sf
)

vif(fit1)

```

# variogram, autocorrelation and prior ajustment

```{r}
# # Project to meters (important)
# 
# sf_data <- st_transform(sampled_data_join_sf, 3857)
# 
# # Empirical variogram
# 
# v <- variogram(PR_Pm_tot ~ 1, data = sf_data)  # adjust cutoff to country size
# # fit an exponential model
# v.fit <- fit.variogram(v, model = vgm("Sph"))
# 
# # plot empirical + fitted curve
# plot(v, v.fit)
# 
# plot(v)

```

```{r}
# fit_cov <- lm(
#   PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 +
#     dist_to_water + lc_water + lc_trees + lc_built +
#     lc_cropland + lc_wetland + vegetation_index,
#   data = sf_data
# )
# 
# sf_data$resid <- residuals(fit_cov)
# 
# coords <- st_coordinates(sf_data)
# 
# # neighbors within, say, 50 km
# nb <- dnearneigh(coords, d1 = 0, d2 = 50000)
# lw <- nb2listw(nb, style = "W")
# 
# moran.test(sf_data$resid, lw)


```

The observed Moran’s I is almost identical to the expected value under
randomness.

The p‑value is very high : no evidence of spatial autocorrelation.

The negative sign is tiny : not meaningful.


# prepare data for the continental level estimation

To avoid extrapolating predictions into areas where malaria transmission
is not considered ecologically feasible, model predictions will be
masked using the *Plasmodium falciparum* transmission limits provided by
the Malaria Atlas Project. Although this study focuses on *Plasmodium
malariae*, no continent-wide transmission limits are currently available
for this species. Given that *P. malariae* is predominantly transmitted
in areas where *P. falciparum* is endemic and frequently occurs as a
co-infection, the *P. falciparum* transmission limits will be used as a
proxy for malaria transmission suitability.

```{r}
sampled_data_join_sf <- st_read("data/Africa_data/sampled_data_join_sf.gpkg")
grid_sf_africa <- st_read("data/Africa_data/grid_data_for_prediction_sf.gpkg")


shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp") %>% st_transform(3857)

Africa_union <- st_union(shp_africa )

sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

sampled_data_join_sf_sp     <- as(sampled_data_join_sf, "Spatial")
Africa_union_sp <- as(Africa_union, "Spatial")

coords_Africa <- coordinates(sampled_data_join_sf_sp)


mesh_Africa <- inla.mesh.2d(
  loc      = coords_Africa,
  boundary = inla.sp2segment(Africa_union_sp),
  max.edge = c(1000e3, 2000e3),   
  cutoff   = 250e3,
  offset   = c(100e3, 2000e3))

# Extract triangle vertex indices
tri <- mesh_Africa$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_Africa <- as.data.frame(coords_Africa)
colnames(coords_df_Africa) <- c("X", "Y")

ggplot() +
  geom_sf(data = mesh_edges_sf, fill = NA, color = "grey60", size = 0.3) +
  geom_sf(data = Africa_union, fill = NA, color = "black", size = 0.8) +
  geom_point(data = coords_df_Africa, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh for Africa with Boundary")
```

```{r}
spde_Africa <- inla.spde2.pcmatern(
  mesh = mesh_Africa,
  prior.range = c(100000, 0.5),
  prior.sigma = c(0.2, 0.5))

A_est <- inla.spde.make.A(mesh_Africa, loc = coords_Africa)

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sampled_data_join_sf[[var]], na.rm = TRUE)
  sd_val   <- sd(sampled_data_join_sf[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sampled_data_join_sf[[var]] <- (sampled_data_join_sf[[var]] - mean_val) / sd_val
}


```

```{r}
effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sampled_data_join_sf$annual_rainfall_2021,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_rainfall1 <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sampled_data_join_sf$annual_rainfall_2021,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept                = 1,
  annual_temperature_2021  = sampled_data_join_sf$annual_temperature_2021,
  vegetation_index         = sampled_data_join_sf$vegetation_index,
  pop_density_2020         = sampled_data_join_sf$pop_density_2020,
  dist_to_water            = sampled_data_join_sf$dist_to_water,
  lc_water                 = sampled_data_join_sf$lc_water,
  lc_trees                 = sampled_data_join_sf$lc_trees,
  lc_built                 = sampled_data_join_sf$lc_built,
  lc_cropland              = sampled_data_join_sf$lc_cropland,
  lc_wetland               = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sampled_data_join_sf$elevation,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  vegetation_index      = sampled_data_join_sf$vegetation_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


stk_est_rainfall <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pf,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_rainfall
  ),
  tag = "est"
)


stk_est_rainfall1 <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pf,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_rainfall1
  ),
  tag = "est"
)

stk_est_temperature <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pf,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_temperature
  ),
  tag = "est"
)


stk_est_elevation <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pf,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_elevation
  ),
  tag = "est"
)


stk_est_baseline <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pf,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_baseline
  ),
  tag = "est"
)

```


```{r}
formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 + vegetation_index + pop_density_2020 + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_rainfall1 <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 + vegetation_index + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 + vegetation_index + pop_density_2020 + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation + vegetation_index + pop_density_2020 + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_baseline <- y ~ 
  -1 +
  intercept +
  vegetation_index + pop_density_2020 + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)



res_rainfall <- inla(
  formula_rainfall,
  family = "betabinomial",
  data = inla.stack.data(stk_est_rainfall),
  Ntrials = inla.stack.data(stk_est_rainfall)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_rainfall),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

res_rainfall1 <- inla(
  formula_rainfall1,
  family = "betabinomial",
  data = inla.stack.data(stk_est_rainfall1),
  Ntrials = inla.stack.data(stk_est_rainfall1)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_rainfall1),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "betabinomial",
  data = inla.stack.data(stk_est_temperature),
  Ntrials = inla.stack.data(stk_est_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "betabinomial",
  data = inla.stack.data(stk_est_elevation),
  Ntrials = inla.stack.data(stk_est_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "betabinomial",
  data = inla.stack.data(stk_est_baseline),
  Ntrials = inla.stack.data(stk_est_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


hist(res_elevation$cpo$pit,
     breaks = 6)
```

```{r}
idx_est <- inla.stack.index(stk_est_elevation, "est")$data

fitted_est <- res_elevation$summary.fitted.values[idx_est, "mean"]

df_resid <- as.data.frame(cbind(
  observed = sampled_data_join_sf$PR_Pm_tot,
  fitted   = fitted_est
))

sampled_data_join_sf$resid <- (df_resid$observed - df_resid$fitted) / sqrt(df_resid$fitted * (1 - df_resid$fitted) / sampled_data_join_sf$Number_of_participant_tot)

coords <- st_coordinates(sampled_data_join_sf)

nb <- dnearneigh(coords, d1 = 0, d2 = 50000)   # 100 km neighbours
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

moran.test(sampled_data_join_sf$resid,
           lw,
           alternative = "two.sided",
           zero.policy = TRUE)

plot(df_resid$observed, df_resid$fitted,
       xlab = "Observed prevalence",
       ylab = "Fitted prevalence")
  abline(0, 1, col = "red")
```

Model diagnostics indicated no evidence of substantial overfitting or underfitting. The PIT histogram was approximately uniform, suggesting good predictive calibration, while the leave-one-out cross-validation diagnostic showed no CPO failures `(sum(CPO failure) = 0)`, indicating stable predictive performance. Pearson residuals exhibited no significant residual spatial autocorrelation `(Moran's I = -0.082, p = 0.181)`, suggesting that the spatial random field adequately captured the remaining spatial dependence. In addition, the observed versus fitted prevalence plot showed predictions closely aligned with the 1:1 line, with no systematic tendency to over- or under-predict. Collectively, these diagnostics indicate that the fitted model is well calibrated and provides an adequate representation of the observed prevalence data.

# Given the large files, let's break the grid of prediction into tiles

```{r}

for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf_africa[[var]] <- (grid_sf_africa[[var]] - mean_val) / sd_val
}


# ## Choose number of tiles
# n_tiles  <- 100
# tile_id  <- cut(seq_len(nrow(grid_sf_africa)), breaks = n_tiles, labels = FALSE)
# n_samp   <- 200  # posterior samples per tile
# 
# # Storage
# pred_mean_full <- numeric(nrow(grid_sf_africa))
# pred_sd_full   <- numeric(nrow(grid_sf_africa))
# pred_q05_full  <- numeric(nrow(grid_sf_africa))
# pred_q95_full  <- numeric(nrow(grid_sf_africa))
# 
# for (tile in seq_len(n_tiles)) {
#   cat("Tile", tile, "of", n_tiles, "\n")
#   
#   # 1) Subset grid
#   grid_tile <- grid_sf_africa[tile_id == tile, ]
#   
#   # 2) A-matrix for this tile
#   A_pred_tile <- inla.spde.make.A(
#     mesh = mesh_Africa,
#     loc  = st_coordinates(grid_tile)
#   )
#   
#   # 3) Fixed effects for this tile
#   effects_pred_tile <- data.frame(
#     intercept               = 1,
#     elevation               = grid_tile$elevation,
#     vegetation_index        = grid_tile$vegetation_index,
#     pop_density_2020        = grid_tile$pop_density_2020,
#     dist_to_water           = grid_tile$dist_to_water,
#     lc_cropland             = grid_tile$lc_cropland,
#     lc_water                = grid_tile$lc_water,
#     lc_wetland              = grid_tile$lc_wetland,
#     lc_built                = grid_tile$lc_built,
#     lc_trees                = grid_tile$lc_trees
#   )
#   
#   # 4) Prediction stack for this tile
#   stk_pred_tile <- inla.stack(
#     data    = list(y = NA, Ntrials = NA),
#     A       = list(A_pred_tile, 1),
#     effects = list(
#       spatial = 1:spde_Africa$n.spde,
#       effects_pred_tile
#     ),
#     tag = "pred"
#   )
#   
#   # 5) Full stack for this tile
#   stk_full_tile <- inla.stack(stk_est_elevation, stk_pred_tile)
#   
#   # 6) Fit INLA with config = TRUE for posterior sampling
#   res_pred_tile <- inla(
#     formula_elevation,
#     family = "betabinomial",
#     data   = inla.stack.data(stk_full_tile),
#     Ntrials = inla.stack.data(stk_full_tile)$Ntrials,
#     control.predictor = list(
#       A       = inla.stack.A(stk_full_tile),
#       compute = TRUE,
#       link    = 1
#     ),
#     control.fixed = list(mean = 0, prec = 0.2),
#     control.compute = list(config = TRUE)
#   )
#   
#   # 7) Posterior samples for this tile
#   samples_tile <- inla.posterior.sample(n_samp, res_pred_tile)
#   
#   # 8) Identify latent indices (once per tile)
#   latent_names <- rownames(samples_tile[[1]]$latent)
#   
#   fixed_base_names   <- rownames(res_pred_tile$summary.fixed)
#   fixed_latent_names <- paste0(fixed_base_names, ":1")
#   fixed_idx          <- match(fixed_latent_names, latent_names)
#   
#   spatial_idx <- grep("^spatial:", latent_names)
#   
#   # 9) Build linear predictor samples for this tile
#   n_grid_tile <- nrow(grid_tile)
#   X_pred_tile <- as.matrix(effects_pred_tile)
#   eta_mat_tile <- matrix(NA_real_, nrow = n_grid_tile, ncol = n_samp)
#   
#   for (s in seq_len(n_samp)) {
#     latent_s   <- samples_tile[[s]]$latent
#     beta_s     <- latent_s[fixed_idx]
#     spatial_s  <- latent_s[spatial_idx]
#     
#     eta_fixed_s   <- as.vector(X_pred_tile %*% beta_s)
#     eta_spatial_s <- as.vector(A_pred_tile %*% spatial_s)
#     
#     eta_mat_tile[, s] <- eta_fixed_s + eta_spatial_s
#   }
#   
#   # 10) Transform to probabilities and summarize
#   prob_mat_tile <- 1 / (1 + exp(-eta_mat_tile))
#   
#   pred_mean_tile <- rowMeans(prob_mat_tile)
#   pred_sd_tile   <- apply(prob_mat_tile, 1, sd)
#   pred_q05_tile  <- apply(prob_mat_tile, 1, quantile, 0.05)
#   pred_q95_tile  <- apply(prob_mat_tile, 1, quantile, 0.95)
#   
#   # 11) Store back into full vectors
#   idx <- (tile_id == tile)
#   pred_mean_full[idx] <- pred_mean_tile
#   pred_sd_full[idx]   <- pred_sd_tile
#   pred_q05_full[idx]  <- pred_q05_tile
#   pred_q95_full[idx]  <- pred_q95_tile
# }

################################################################################
#               Another way of fitting the model                               #
################################################################################

res_elevation <- inla(
  formula_elevation,
  family = "betabinomial",
  data = inla.stack.data(stk_est_elevation),
  Ntrials = inla.stack.data(stk_est_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)

n_tiles <- 100
tile_id <- cut(seq_len(nrow(grid_sf_africa)),
               breaks = n_tiles,
               labels = FALSE)

pred_mean <- numeric(nrow(grid_sf_africa))
pred_sd   <- numeric(nrow(grid_sf_africa))
pred_q05  <- numeric(nrow(grid_sf_africa))
pred_q95  <- numeric(nrow(grid_sf_africa))

for(tile in seq_len(n_tiles)){
  cat("Tile", tile, "of", n_tiles, "\n")

  idx <- which(tile_id == tile)
  grid_tile <- grid_sf_africa[idx, ]

  A_pred <- inla.spde.make.A(
    mesh = mesh_Africa,
    loc = st_coordinates(grid_tile)
  )

  effects_pred <- data.frame(
    intercept               = 1,
    elevation               = grid_tile$elevation,
    vegetation_index        = grid_tile$vegetation_index,
    pop_density_2020        = grid_tile$pop_density_2020,
    dist_to_water           = grid_tile$dist_to_water,
    lc_cropland             = grid_tile$lc_cropland,
    lc_water                = grid_tile$lc_water,
    lc_wetland              = grid_tile$lc_wetland,
    lc_built                = grid_tile$lc_built,
    lc_trees                = grid_tile$lc_trees)

  stk_pred <- inla.stack(
    data = list(y = NA, Ntrials = NA),
    A = list(A_pred, 1),
    effects = list(
      spatial = 1:spde_Africa$n.spde,
      effects_pred
    ),
    tag = "pred"
  )

  stk_full <- inla.stack(stk_est_elevation, stk_pred)

  res_tile <- inla(
    formula_elevation,
    family = "betabinomial",
    data = inla.stack.data(stk_full),
    Ntrials = inla.stack.data(stk_full)$Ntrials,

    control.predictor = list(
      A = inla.stack.A(stk_full),
      compute = TRUE,
      link = 1
    ),

    control.mode = list(
      theta = res_elevation$mode$theta,
      restart = FALSE
    ),

    control.compute = list(config = TRUE)
  )

  pred_idx <- inla.stack.index(stk_full, "pred")$data

  eta_mat <- inla.posterior.sample.eval(
    function() APredictor[pred_idx],
    inla.posterior.sample(200, res_tile)
  )

  prob_mat <- plogis(eta_mat)

  pred_mean[idx] <- rowMeans(prob_mat)
  pred_sd[idx]   <- apply(prob_mat, 1, sd)
  pred_q05[idx]  <- apply(prob_mat, 1, quantile, 0.05)
  pred_q95[idx]  <- apply(prob_mat, 1, quantile, 0.95)

  gc()
}

grid_sf_africa$pred_mean <- pred_mean
grid_sf_africa$pred_sd   <- pred_sd
grid_sf_africa$pred_q05  <- pred_q05
grid_sf_africa$pred_q95  <- pred_q95
################################################################################
#                       End                                                    #
################################################################################

```

```{r}
grid_vect <- vect(grid_sf_africa)

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")


################################################################################
#        Rasterize the prediction grid for an easy visualization
################################################################################

limit_3857 <- project(global_spatial_limit, "EPSG:3857", method = "near")
limit_3857 <- crop(limit_3857, vect(shp_africa))
limit_3857 <- mask(limit_3857, vect(shp_africa))

grid_limit <- terra::extract(limit_3857, grid_vect)

grid_sf_africa$limit <- grid_limit[,2]


grid_sf_africa$limit[is.na(grid_sf_africa$limit)] <- 0


grid_sf_africa_mask_unstable <- grid_sf_africa |>
  mutate(
    pred_mean = ifelse(limit == 0, NA, pred_mean),
    pred_sd   = ifelse(limit == 0, NA, pred_sd),
    pred_q05  = ifelse(limit == 0, NA, pred_q05),
    pred_q95  = ifelse(limit == 0, NA, pred_q95)
  )

grid_sf_africa_mask_stable <- grid_sf_africa |>
  mutate(
    pred_mean = ifelse(limit %in% c(0,1), NA, pred_mean),
    pred_sd   = ifelse(limit %in% c(0,1), NA, pred_sd),
    pred_q05  = ifelse(limit %in% c(0,1), NA, pred_q05),
    pred_q95  = ifelse(limit %in% c(0,1), NA, pred_q95)
  )

v1 <- vect(grid_sf_africa_mask_unstable)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)


grid_sf_africa_mask_unstable_raster <- rasterize(
  v1,
  template,
  field = c("pred_mean", "pred_q05", "pred_q95")
)

v1 <- vect(grid_sf_africa_mask_stable)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)

grid_sf_africa_mask_stable_raster <- rasterize(
  v1,
  template,
  field = c("pred_mean", "pred_q05", "pred_q95")
)


################################################################################
#                                   End                                        #
################################################################################

shp_africa_country <- shp_africa %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry))

# Africa_prev_1 <- ggplot() +
#   geom_sf(data = grid_masked_sf,
#           aes(color = pred_mean),
#           rasterize = TRUE,
#           size = 0.1,
#           alpha = 0.4) +
#   geom_sf(data = shp_africaw4326,
#           fill = NA,
#           color = "black",
#           size = 0.05) +
#   scale_color_viridis_c(option = "magma", na.value = "white") +
#   theme_bw(base_size = 15) +
#   theme(axis.text.x = element_blank(),
#         axis.text.y = element_blank())

Africa_prev_1 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using COMAL data")
  

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level_unstable.png",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country_1 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using COMAL data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_unstable.pdf",
       plot = Africa_prev_country_1, width = 6, height = 4, dpi = 300)


grid_sf_africa_mask_unstable_raster$CI_width <- grid_sf_africa_mask_unstable_raster$pred_q95 - grid_sf_africa_mask_unstable_raster$pred_q05

Africa_prev_country_uncertainty <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = CI_width))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "Credible intervale width\n of the estimated prevalence")+
  theme_void()+
  labs(title = "Associated uncertainty to the \n annual prevalence of P. f in 2021 using COMAL data")


ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_uncertainty.pdf",
       plot = Africa_prev_country_uncertainty, width = 6, height = 4, dpi = 300)


# setting unstable areas to 0

Africa_prev_2 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using COMAL data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level.png",
       plot = Africa_prev_2, width = 6, height = 4, dpi = 300)

Africa_prev_country_2 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using COMAL data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level.pdf",
       plot = Africa_prev_country_2, width = 6, height = 4, dpi = 300)



```


```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
