---
title: "COMAL project"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#load necessary libraries
library(sf)
library(reshape2)
library(ggcorrplot)
library(leaflet)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(mapview)
library(spData)
library(gganimate)
library(htmlwidgets)
library(webshot)
library(chirps)
library(viridis)
library(patchwork)
library(GGally)
library(spdep)
library(INLA)
library(leafsync)
library(SpatialEpi)
library(geodata)
library(car)
library(gstat)
library(nngeo)
library(tidygeocoder)
library(malariaAtlas)
library(countrycode)
library(rmapshaper)
library(tidytext)
```



```{r}
#loading the shape files
shp_ben <- read_sf("data/ben_data/BEN_shape_file/gadm41_BEN_3.shp")
shp_con <- read_sf("data/cong_data/CON_shape_file/cog_admin2.shp")
shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")


#renaming the districts

shp_ben$NAME_3[shp_ben$NAME_3=="Gakpè"]="Gakpe"
shp_ben$NAME_3[shp_ben$NAME_3=="Tokpa Domé"]="Tokpa-dome"
shp_ben$NAME_3[shp_ben$NAME_3=="Ouakpé-Daho"]="Houakpe-daho"
shp_ben$NAME_3[shp_ben$NAME_3=="Aganmalomé"]="Aganmanlome"
shp_ben$NAME_3[shp_ben$NAME_3=="Agonkanmè"]="Agonkamey"

ben_vect <- vect(shp_ben)
cmr_vect <- vect(shp_cmr)
con_vect <- vect(shp_con)
```


```{r}
# Loading the data

BEN_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 1)

CON_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 2)

CMR_df <- read_excel("data/cmr_data/Cameroon2.xlsx")

GAB_df <- read_excel("data/gab_data/Gabon2.xlsx")

```


# Cycles Threshold (CT) analysis

```{r}
BEN_Ct <- BEN_df %>% 
  dplyr::select(Village, `Sample ID`, `Household ID`, Sex, `Age (year)`, Season, `Parasitaemia Pf (P/µl)`,
                `Parasitaemia Pm (P/µl)`, `qRT-qPCR Pf positive`, `qRT-qPCR Pm positive`,
                `Pf(Ct) mean`, `Pm(Ct) mean`) %>% 
  rename(Sample_ID        = `Sample ID`,
         Household_ID     = `Household ID`,
         Age              = `Age (year)`,
         Parasitaemia_Pf  = `Parasitaemia Pf (P/µl)`,
         Parasitaemia_Pm  = `Parasitaemia Pm (P/µl)`,
         Positive_pf      = `qRT-qPCR Pf positive`,
         Positive_pm      = `qRT-qPCR Pm positive`,
         Ct_value_pf      = `Pf(Ct) mean`,
         Ct_value_pm      = `Pm(Ct) mean`) %>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 2, 10, 20, 30, Inf),
                                labels = c("<2", "2<10", "10<20", "20<30", "30+"),
                                right  = FALSE),
         Parasitaemia_Pf  = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Positive_pf      = case_when(is.na(Positive_pf) ~ "neg",
                                     TRUE ~ "pos"),
         Positive_pm      = case_when(is.na(Positive_pm) ~ "neg",
                                     TRUE ~ "pos"),
         Ct_value_pf      = case_when(is.na(Ct_value_pf) ~ 0,
                                     TRUE ~ Ct_value_pf),
         Ct_value_pm      = case_when(is.na(Ct_value_pm) ~ 0,
                                     TRUE ~ Ct_value_pm),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

CONG_Ct <- CON_df %>% 
  dplyr::select(`Sample ID`, Sex, `Age (year)`,`Parasitaemia Pf (P/µl)`, 
                `Parasitaemia Pm (P/µl)`, `qRT-qPCR Pf positive`,
                `qRT-qPCR Pm positive`, `Pf(Ct) mean`, `Pm(Ct) mean`)%>% 
  rename(Sample_ID        = `Sample ID`,
         Age              = `Age (year)`,
         Parasitaemia_Pf  = `Parasitaemia Pf (P/µl)`,
         Parasitaemia_Pm  = `Parasitaemia Pm (P/µl)`,
         Positive_pf      = `qRT-qPCR Pf positive`,
         Positive_pm      = `qRT-qPCR Pm positive`,
         Ct_value_pf      = `Pf(Ct) mean`,
         Ct_value_pm      = `Pm(Ct) mean`)%>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 2, 10, 20, 30, Inf),
                                labels = c("<2", "2<10", "10<20", "20<30", "30+"),
                                right  = FALSE),
         Parasitaemia_Pf  = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Positive_pf      = case_when(is.na(Positive_pf) ~ "neg",
                                     TRUE ~ "pos"),
         Positive_pm      = case_when(is.na(Positive_pm) ~ "neg",
                                     TRUE ~ "pos"),
         Ct_value_pf      = case_when(is.na(Ct_value_pf) ~ 0,
                                     TRUE ~ Ct_value_pf),
         Ct_value_pm      = case_when(is.na(Ct_value_pm) ~ 0,
                                     TRUE ~ Ct_value_pm),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

CMR_Ct <- CMR_df %>% 
  dplyr::select(district, ID, Season, age, parasitemia_pf, parasitemia_pm, pcr_pf,
                pcr_pm, pf_ct, pm_ct) %>% 
  rename(Village         = district,
         Sample_ID       = ID,
         Age             = age,
         Parasitaemia_Pf = parasitemia_pf,
         Parasitaemia_Pm = parasitemia_pm,
         Positive_pf     = pcr_pf,
         Positive_pm     = pcr_pm,
         Ct_value_pf     = pf_ct,
         Ct_value_pm     = pm_ct) %>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 2, 10, 20, 30, Inf),
                                labels = c("<2", "2<10", "10<20", "20<30", "30+"),
                                right  = FALSE),
         Parasitaemia_Pf = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

GAB_Ct <- GAB_df %>% 
  dplyr::select(district, ID, season, age, parasitemia_pf, parasitemia_pm, pcr_pf, 
                pcr_pm, pf_ct, pm_ct) %>% 
  rename(Village         = district,
         Sample_ID       = ID,
         Season          = season,
         Age             = age,
         Parasitaemia_Pf = parasitemia_pf,
         Parasitaemia_Pm = parasitemia_pm,
         Positive_pf     = pcr_pf,
         Positive_pm     = pcr_pm,
         Ct_value_pf     = pf_ct,
         Ct_value_pm     = pm_ct)%>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 2, 10, 20, 30, Inf),
                                labels = c("<2", "2<10", "10<20", "20<30", "30+"),
                                right  = FALSE),
         Parasitaemia_Pf = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))
  


# Demographic and risk factor analysis

boxplot_Ct_Age_Ben <- ggplot(data = BEN_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/ben_output/Ct_Age_Ben.png", plot = boxplot_Ct_Age_Ben, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_CONG <- ggplot(data = CONG_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/cong_output/Ct_Age_CONG.png", plot = boxplot_Ct_Age_CONG, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_CMR <- ggplot(data = CMR_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/cmr_output/Ct_Age_CMR.png", plot = boxplot_Ct_Age_CMR, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_GAB <- ggplot(data = GAB_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/gab_output/Ct_Age_GAB.png", plot = boxplot_Ct_Age_GAB, width = 6, height = 4, dpi = 300)

```
- In Benin the `10<20` age group is the one with the lowest Ct value meaning they 
are the one with high parasitemia in the country while `30+` is the one with highest
value, then low parasitemia.

- In Congo it is still the same age group with the lowest Ct value and the one
with the highest value is `2<10`.

- In Cameroon `10<20` and `10<20` are the ones with the lowest Ct value and the 
with the highest value is `20<30`. For this case with two outliers for `2<10` age
group with Ct value greater than `50`.

- For Gabon we still have `10<20` and `10<20` age group which are the ones with 
the lowest Ct value whereas `<2`, `20<30` and `30+` are the ones with the highest 
Ct value.

```{r}
country_summary <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  group_by(Country) %>%
  summarise(
    n_tested = n(),
    n_positive_pm = sum(Positive_pm == "pos", na.rm = TRUE),
    prevalence_pm = n_positive_pm / n_tested,
    .groups = "drop"
  )


age_prev <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  group_by(Country, Age_group) %>%
  summarise(
    n = n(),
    positive = sum(Positive_pm == "pos", na.rm = TRUE),
    prevalence = positive / n,
    .groups = "drop"
  )



ggplot(age_prev,
       aes(x = reorder(Age_group, prevalence), y = prevalence, fill = Country)) +
  geom_col(position = "dodge")+
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Age group",
    y = "P. malariae prevalence",
    fill = "Country"
  ) +
  theme_minimal()


ggplot(age_prev,
       aes(x = reorder(Country, prevalence),
           y = prevalence,
           fill = Country)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ Age_group, scales = "free_x") +
  labs(
    x = "Country",
    y = "P. malariae prevalence"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
pm_positive <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  filter(Positive_pm == "pos")

ggplot(pm_positive,
       aes(x = Country, y = Parasitaemia_Pm)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    x = "Country",
    y = "P. malariae parasitaemia"
  ) +
  theme_minimal()


ggplot(pm_positive,
       aes(x = Country, y = Ct_value_pm)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    x = "Country",
    y = "P. malariae Ct value"
  )

ggplot(pm_positive,
       aes(x = Ct_value_pm, fill = Country)) +
  geom_density(alpha = 0.4) +
  theme_minimal()
```

- So at the country level, Cameroon has the highest Ct level and also the highest
prevalence which means high prevalence might not necessary mean severity of the 
disease but instead a hidden reservoir of the parasite transmission with low
density of parasite load. But if we look closely to countries like Benin, 
Gabon, Congo, we will see that Benin has the prevalence value but all those three
countries have almost the same level of Ct values which mean with the data we
have, we cannot draw an effective conclusion with the data we have.


```{r}
country_pm <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
)

country_pm %>%
  group_by(Country) %>%
  summarise(
    prevalence = mean(Positive_pm == "pos", na.rm = TRUE),
    mean_Ct = mean(Ct_value_pm[Positive_pm == "pos"], na.rm = TRUE),
    median_Ct = median(Ct_value_pm[Positive_pm == "pos"], na.rm = TRUE)
  )
```

```{r}
ggplot(pm_positive,
       aes(x = Age, y = Ct_value_pm)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ Country) +
  theme_minimal()


co_inf <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  count(Country,Positive_pf, Positive_pm)
```



# Extract the longitude and latitude of households using gps column for Cameroon data

```{r}
## function to extract the GPS location for Cameroon data

convert_custom_dms <- function(x) {
  hemi <- substr(x, 1, 1)
  x <- substring(x, 2)

  apos <- regexpr("'", x)[1]

  if (hemi == "N") {
    deg <- as.numeric(substr(x, 1, apos - 4))
  } else {
    deg <- as.numeric(substr(x, 1, apos - 3))
  }

  min <- as.numeric(substr(x, apos - 2, apos - 1))
  sec <- as.numeric(gsub("\"", "", substr(x, apos + 1, nchar(x))))

  dec <- deg + min/60 + sec/3600

  if (hemi %in% c("S", "W")) dec <- -dec
  return(dec)
}

CMR_df$lat <- sapply(strsplit(CMR_df$gps, " "), function(x) convert_custom_dms(x[1]))
CMR_df$lon <- sapply(strsplit(CMR_df$gps, " "), function(x) convert_custom_dms(x[2]))


CMR_df_1 <- CMR_df %>% 
  dplyr::select(ID, lon, lat, gender, age, pcr_pm, pcr_pf) %>% 
  mutate(
    lng = lon,
    tot_participants = 1,
    tot_infected_pm  = if_else(pcr_pm == "pos", 1, 0),
    tot_infected_pf  = if_else(pcr_pf == "pos", 1, 0)
  )%>% 
  group_by(lng, lat) %>% 
  summarise(age_lower = min(age),
            age_upper = max(age),
            Number_of_participant_tot  = sum(tot_participants, na.rm = TRUE),
            Number_of_infected_tot_pm  = sum(tot_infected_pm, na.rm = TRUE),
            PR_Pm_tot = Number_of_infected_tot_pm  / Number_of_participant_tot ,
            Number_of_infected_tot_pf  = sum(tot_infected_pf, na.rm = TRUE),
            PR_Pf_tot = Number_of_infected_tot_pf  / Number_of_participant_tot ,
            .groups = "drop") %>% 
  mutate(country = "Cameroon")

# Group households into clusters

households <- st_as_sf(
    CMR_df_1,
    coords = c("lng","lat"),
    crs = 4326
)

households <- st_transform(households, 32632)

grid_cmr_cluster <- st_make_grid(
    households,
    cellsize = 1000,
    square = TRUE
)

grid_cmr_cluster <- st_sf(
  grid_id = seq_along(grid_cmr_cluster),
  geometry = grid_cmr_cluster
)

households_grid <- st_join(
  households,
  grid_cmr_cluster,
  join = st_intersects
)

cluster_data_cmr <-
  households_grid %>%
  st_drop_geometry() %>% 
  group_by(grid_id) %>%
  summarise(age_lower = min(age_lower),
         age_upper = max(age_upper),
         Number_of_participant_tot  = sum(Number_of_participant_tot, na.rm = TRUE),
         Number_of_infected_tot_pm  = sum(Number_of_infected_tot_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm  / Number_of_participant_tot ,
         Number_of_infected_tot_pf  = sum(Number_of_infected_tot_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf  / Number_of_participant_tot ,
         .groups = "drop") %>% 
  mutate(country = "Cameroon")


cluster_data_cmr_sf <-
  grid_cmr_cluster  %>% 
  inner_join(cluster_data_cmr, by = "grid_id")

cluster_data_cmr_sf <- st_transform(cluster_data_cmr_sf, 4326)
cluster_data_cmr_sf <- st_centroid(cluster_data_cmr_sf)

coords_cmr <- st_coordinates(cluster_data_cmr_sf)

cluster_data_cmr_sf$lng <- coords_cmr[,1]
cluster_data_cmr_sf$lat <- coords_cmr[,2]



CMR_df_1_sf <- st_as_sf(CMR_df_1, coords = c("lng", "lat"), crs = 4326)
cluster_data_cmr_sf <- st_as_sf(cluster_data_cmr_sf, coords = c("lng", "lat"), crs = 4326)



raw_pr_cmr <- ggplot() +
  geom_sf(data = shp_cmr, fill = "white", color = "black") +
  geom_point(data = cluster_data_cmr_sf, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/raw_obs_PR_cmr.pdf", plot = raw_pr_cmr, width = 6, height = 4, dpi = 300)

sampled_site_cmr <- shp_cmr %>% 
  filter(Admin2 == "Bankim")

sample_district <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_point(data = cluster_data_cmr_sf, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence (Bankim)",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/raw_obs_PR_sampled_location_cmr.pdf", plot = sample_district, width = 6, height = 4, dpi = 300)


mapview(shp_cmr,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(CMR_df_1_sf,
        zcol = "PR_Pm_tot",
        cex = "Number_tot_participants",
        col.regions = colorRampPalette(c("blue", "orange")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```


# Find the GPS coordinates for sampled districts in Benin and aggregation of our data at the level of village 

```{r}
village_coords_ben <- tribble(
  ~Municipality, ~District,     ~Village,            ~Match_confidence,   ~lat,     ~lng,
  "Ouidah",      "Gakpe",       "Lokohoue",           "fuzzy_match",      6.443353,  2.132558,#
  "Ouidah",      "Gakpe",       "Kindjitokpa",        "fuzzy_match",      6.461650,  2.148980,#
  "Kpomasse",    "Tokpa-dome",  "Lokogbo-Gnonwa",     "exact_match",      6.471386,  1.984651,#
  "Ouidah",      "Houakpe-daho","Gbezoume",           "exact_match",      6.344216,  1.962808,#
  "Ouidah",      "Gakpe",       "Sehlinhoue",         "fuzzy_match",      6.416572,  2.173980,#
  "Kpomasse",    "Aganmanlome", "Hessa",              "fuzzy_match",      6.495617,  2.053485,#
  "Kpomasse",    "Aganmanlome", "KOUZOUME",           "fuzzy_match",      6.462216,  2.017743,#
  "Ouidah",      "Gakpe",       "Agossouhouihoué",    "fuzzy_match",      6.421688,  2.144019,#
  "Ouidah",      "Gakpe",       "Ahossihoue",         "fuzzy_match",      6.452980,  2.167627,#
  "Ouidah",      "Houakpe-daho","Gbehonou",           "exact_match",      6.339606,  2.005831,#
  "Kpomasse",    "Aganmanlome", "KOUGBEDJI",          "fuzzy_match",      6.473206,  2.077346,#
  "Ouidah",      "Houakpe-daho","Toligbé",            "exact_match",      6.327061,  2.027671,#
  "Ouidah",      "Houakpe-daho","Houakpe-daho centre","exact_match",      6.329397,  2.047797,#
  "Kpomasse",    "Aganmanlome", "Aganmanlome Centre", "fuzzy_match",      6.477114,  2.049987,#
  "Ouidah",      "Gakpe",       "TOKOLI",             "fuzzy_match",      6.431493,  2.156038,#
  "Kpomasse",    "Aganmanlome", "Lokossa",            "fuzzy_match",      6.478105,  2.027037,#
  "Kpomasse",    "Agonkamey",   "Assogbenou Kpevi",   "exact_match",      6.366429,  2.027881,#
  "Ouidah",      "Houakpe-daho","Seyigbe",            "exact_match",      6.324844,  2.041879,#
  "Kpomasse",    "Agonkamey",   "Assogbenou daho",    "exact_match",      6.355088,  2.001053 #
)

BEN_1 <- BEN_df %>% 
  mutate(Village = case_when(Village == "HESSA" ~ "Hessa",
                             TRUE ~ Village)) %>% 
  dplyr::select(Village, Sex, `Age (year)`, `qRT-qPCR Pm positive`, `qRT-qPCR Pf positive`) %>%
  rename(age = `Age (year)`) %>% 
  mutate(Pm_positive = case_when( `qRT-qPCR Pm positive`=="x"~"yes",
                                  TRUE~"no"),
         Pf_positive = case_when( `qRT-qPCR Pf positive`=="x"~"yes",
                                  TRUE~"no"),
         participants = 1,
         infected_pm  = case_when(Pm_positive=="yes"~1,
                              TRUE~0),
         infected_pf  = case_when(Pf_positive=="yes"~1,
                              TRUE~0)) %>% 
  group_by(Village) %>% 
  summarise(age_lower = min(age, na.rm = TRUE),
         age_upper = max(age, na.rm = TRUE),
         Number_of_participant_tot = sum(participants, na.rm = TRUE),
         Number_of_infected_tot_pm = sum(infected_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm/Number_of_participant_tot,
         Number_of_infected_tot_pf = sum(infected_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf/Number_of_participant_tot,
         .groups = "drop") %>%  
  distinct()
  


BEN_1_GPS <- BEN_1 %>% 
  left_join(village_coords_ben %>% dplyr::select(Village, lat, lng, Match_confidence))

spatial_data_ben <- st_as_sf(BEN_1_GPS, coords = c("lng", "lat"), crs = 4326)

PR_obs_ben <- ggplot() +
  geom_sf(data = shp_ben, fill = "white", color = "black") +
  geom_point(data = BEN_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/PR_p_malariae_observed.pdf", plot = PR_obs_ben, width = 6, height = 4, dpi = 300)


sampled_site_ben <- shp_ben %>% 
  filter(NAME_3 %in% c("Tokpa-dome", "Aganmanlome", "Savi", "Kpomassè", "Gakpe", 
                       "Arrondissement III", "Arrondissement I", "Houakpe-daho",
                       "Agonkamey", "Arrondissement II") & NAME_2 %in% c("Kpomassè", "Ouidah"))

Raw_PR_sampled_location_ben <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_point(data = BEN_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Raw_PR_sampled_location.pdf", plot = Raw_PR_sampled_location_ben,
       width = 6, height = 4, dpi = 300)


mapview(shp_ben,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(spatial_data_ben,
        zcol = "PR_Pm_tot",
        cex = "Number_of_participant_tot",
        col.regions = colorRampPalette(c("blue", "orange")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```


# Find the GPS coordinates for sampled districts in Benin and aggregation of our data at the level of village 

```{r}
village_coords_cong <- tribble(
   ~District,     ~Match_confidence,     ~lat,                  ~lng,
    "Ntoula",      "fuzzy_match",      -4.36,                 15.15,#
    "Djoumouna",   "exact_match",      -4.37601066518807,     15.15963007611599,#
    "Mayanga",     "exact_match",      -4.284712200560734,    15.186363375868131,#
  )

CONG_1 <- CON_df %>% 
  dplyr::select(District, Sex, `Age (year)`, `qRT-qPCR Pm positive`, `qRT-qPCR Pf positive`) %>% 
  rename(age = `Age (year)`) %>% 
  mutate(Pm_positive = case_when( `qRT-qPCR Pm positive`=="x"~"yes",
                                  TRUE~"no"),
         Pf_positive = case_when( `qRT-qPCR Pf positive`=="x"~"yes",
                                  TRUE~"no"),
         participants = 1,
         infected_pm = case_when(Pm_positive=="yes"~1,
                              TRUE~0),
         infected_pf = case_when(Pf_positive=="yes"~1)) %>% 
  group_by(District) %>% 
  summarise(age_lower = min(age, na.rm = TRUE),
         age_upper = max(age, na.rm = TRUE),
         Number_of_participant_tot = sum(participants, na.rm = TRUE),
         Number_of_infected_tot_pm = sum(infected_pm, na.rm = TRUE),
         PR_Pm_tot = Number_of_infected_tot_pm/Number_of_participant_tot,
         Number_of_infected_tot_pf = sum(infected_pf, na.rm = TRUE),
         PR_Pf_tot = Number_of_infected_tot_pf/Number_of_participant_tot,
         .groups = "drop") %>% 
  distinct()
  


CONG_1_GPS <- CONG_1 %>% 
  left_join(village_coords_cong %>% dplyr::select(District, lat, lng, Match_confidence))

spatial_data_cong <- st_as_sf(CONG_1_GPS, coords = c("lng", "lat"), crs = 4326)
```


```{r}
raw_pr_cong <- ggplot() +
  geom_sf(data = shp_con, fill = "white", color = "black") +
  geom_point(data = CONG_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/raw_obs_PR.pdf", plot = raw_pr_cong, width = 6, height = 4, dpi = 300)

sampled_site_cong <- shp_con %>% 
  filter(adm2_name %in% c("Brazzaville", "Goma-Tsetse"))

sampled_location_pr_cong <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_point(data = CONG_1_GPS, aes(x = lng, y = lat, size = Number_of_participant_tot, color = PR_Pm_tot)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Raw P. malariae prevalence",
       color = "Prevalence", size = "N tested") +
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/raw_obs_PR_sampled_location.pdf", plot = sampled_location_pr_cong, width = 6, height = 4, dpi = 300)

mapview(shp_con,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
mapview(spatial_data_cong,
        zcol = "PR_Pm_tot",
        cex = "Number_of_participant_tot",
        col.regions = colorRampPalette(c("blue", "red")),
        at = NULL,
        layer.name = "P. malariae prevalence")
```



# loading covariates and extract them at the sample location Benin

```{r}
path1 <- "data/CHIRPS_PENTAD_precipitation_2021-01-01_2021-12-31_0p9311to13p9448N_8p2516to17p0602E.precipitation.tif"
path2 <- "data/ERA5_LAND_DAILY_temperature_2m_2021-01-01_2021-12-31_1p1068to13p6033N_8p1637to16p8844E.temperature_2m.tif"
ndvi_path <- "data/ben_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

land_cover_water <- rast("data/ben_data/landuse/WorldCover_water_30s.tif")
land_cover_trees <- rast("data/ben_data/landuse/WorldCover_trees_30s.tif")
land_cover_builts <- rast("data/ben_data/landuse/WorldCover_built_30s.tif")
land_cover_cropland <- rast("data/ben_data/landuse/WorldCover_cropland_30s.tif")
land_cover_wetland <- rast("data/ben_data/landuse/WorldCover_wetland_30s.tif")


ITN_intervention_2021 <- rast("data/ITN_2000/2025_GBD2024_Africa_ITN_2021.tif")

ben_pop <- rast("data/ben_data/ben_pd_2020_1km_UNadj.tif")
ben_rainfall_2021 <- rast(path1)
ben_temperature_2021 <- rast(path2)
elevation_ben <- rast("data/ben_data/elevation/BEN_elv_msk.tif")
ben_ndvi_stack<- rast(ndvi_files)

ben_land_cover_water <- crop(land_cover_water, ben_vect)
ben_land_cover_water <- mask(cmr_land_cover_water, ben_vect)

ben_land_cover_built <- crop(land_cover_builts, ben_vect)
ben_land_cover_built <- mask(cmr_land_cover_built, ben_vect)

ben_land_cover_cropland <- crop(land_cover_cropland, ben_vect)
ben_land_cover_cropland <- mask(cmr_land_cover_cropland, ben_vect)

ben_land_cover_trees <- crop(land_cover_trees, ben_vect)
ben_land_cover_trees <- mask(cmr_land_cover_trees, ben_vect)

ben_land_cover_wetland <- crop(land_cover_wetland, ben_vect)
ben_land_cover_wetland <- mask(cmr_land_cover_wetland, ben_vect)


# Benin case

# create a circle around the GPS coordinates extracted manually to ensure of 
# reasonable extracted value for the population density

village_coords_ben_1 <- village_coords_ben %>% select(Village, lat, lng)

village_coords_ben_1 <- st_as_sf(village_coords_ben_1, coords = c("lng", "lat"),
                                 crs = 4326)
# project to a metric system
village_coords_ben_1_projected <- st_transform(village_coords_ben_1, crs = 32631)

# create 1000 buffer around each village point
# the distance can be adjusted based on the result obtained
village_coords_ben_1_projected_buffer <- st_buffer(village_coords_ben_1_projected,
                                                   dist = 1000)

# project back to match the raster's crs if they differ

village_coords_ben_1_projected_buffer_crs <- st_transform(village_coords_ben_1_projected_buffer,
                                                          crs = crs(ben_pop))

ben_pop_density_2020_df_buffer <- terra::extract(ben_pop, 
                                                 village_coords_ben_1_projected_buffer_crs,
                                                 fun = mean,
                                                 na.rm = TRUE)


ben_ITN_intervention_2021  <- project(ITN_intervention_2021,  st_crs(spatial_data_ben)$wkt)
ben_pop  <- project(ben_pop,  st_crs(spatial_data_ben)$wkt)
ben_rainfall_2021 <- project(ben_rainfall_2021, st_crs(spatial_data_ben)$wkt)
ben_temperature_2021 <- project(ben_temperature_2021, st_crs(spatial_data_ben)$wkt)
ben_ndvi_stack <- project(ben_ndvi_stack, st_crs(spatial_data_ben)$wkt)
elevation_ben   <- project(elevation_ben,   st_crs(spatial_data_ben)$wkt)

ben_land_cover_water   <- project(ben_land_cover_water,   st_crs(spatial_data_ben)$wkt)
ben_land_cover_built   <- project(ben_land_cover_built,   st_crs(spatial_data_ben)$wkt)
ben_land_cover_cropland   <- project(ben_land_cover_cropland,   st_crs(spatial_data_ben)$wkt)
ben_land_cover_trees   <- project(ben_land_cover_trees,   st_crs(spatial_data_ben)$wkt)
ben_land_cover_wetland   <- project(ben_land_cover_wetland,   st_crs(spatial_data_ben)$wkt)


ben_ITN_intervention_2021_df <- terra::extract(ben_ITN_intervention_2021,  spatial_data_ben)
ben_pop_density_2020_df <- terra::extract(ben_pop,  spatial_data_ben)
ben_rainfall_2021_df    <- terra::extract(ben_rainfall_2021, spatial_data_ben)
ben_temperature_2021_df <- terra::extract(ben_temperature_2021, spatial_data_ben)
ndvi_vals_ben_monthly <- terra::extract(ben_ndvi_stack, spatial_data_ben)
ndvi_mean_ben <- apply(ndvi_vals_ben_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_ben_df <- terra::extract(elevation_ben, spatial_data_ben)
ben_land_cover_water_df <- terra::extract(ben_land_cover_water, spatial_data_ben)
ben_land_cover_built_df <- terra::extract(ben_land_cover_built, spatial_data_ben)
ben_land_cover_cropland_df <- terra::extract(ben_land_cover_cropland, spatial_data_ben)
ben_land_cover_trees_df <- terra::extract(ben_land_cover_trees, spatial_data_ben)
ben_land_cover_wetland_df <- terra::extract(ben_land_cover_wetland, spatial_data_ben)



spatial_data_ben_sf <- spatial_data_ben %>% 
  mutate(pop_density_2020 = ben_pop_density_2020_df[,2],
         annual_rainfall_2021 = ben_rainfall_2021_df[,2],
         annual_temperature_2021 = ben_temperature_2021_df[,2],
         vegetaion_index = ndvi_mean_ben,
         elevation = elevation_ben_df[,2],
         lc_water = ben_land_cover_water_df[,2],
         lc_built = ben_land_cover_built_df[,2],
         lc_trees = ben_land_cover_trees_df[,2],
         lc_wetland = ben_land_cover_wetland_df[,2],
         lc_cropland = ben_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = ben_ITN_intervention_2021_df[,2])


# distance to water

bbox <- st_bbox(shp_ben)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)

rivers_valid <- st_transform(rivers_valid, st_crs(shp_ben))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_ben))

# nearest river segment for each point
nn_riv <- st_nn(spatial_data_ben_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(spatial_data_ben_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

spatial_data_ben_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

spatial_data_ben_df <- spatial_data_ben_sf %>% 
  st_drop_geometry() 


missing_rows <- which(!complete.cases(spatial_data_ben_df))
spatial_data_ben_df[missing_rows, ]

# find nearest neighbor for each point
nn <- st_nn(spatial_data_ben_sf, spatial_data_ben_sf, k = 5)   # k=2 because first neighbor is itself

get_first_non_na <- function(values, neighbors) {
  for (i in neighbors) {
    if (!is.na(values[i])) return(values[i])
  }
  return(NA)   
}

var <- "annual_temperature_2021"
values <- spatial_data_ben_df[[var]]

missing <- which(is.na(values))

for (i in missing) {
  spatial_data_ben_df[[var]][i] <- get_first_non_na(values, nn[[i]])
}

spatial_data_ben_sf$annual_temperature_2021 <- spatial_data_ben_df$annual_temperature_2021

st_write(spatial_data_ben_sf, "data/ben_data/spatial_data_ben_sf.gpkg", append = FALSE)


# Congo case

Cong_pop <- rast("data/cong_data/cog_pd_2020_1km_UNadj.tif")

elevation_Cong <- rast("data/cong_data/elevation/COG_elv_msk.tif")
ndvi_path <- "data/cong_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

Cong_ndvi_stack <- rast(ndvi_files)

# create a circle around the GPS coordinates extracted manually to ensure of 
# reasonable extracted value for the population density

village_coords_cong_1 <- village_coords_cong %>% select(District, lat, lng)

village_coords_cong_1 <- st_as_sf(village_coords_cong_1, coords = c("lng", "lat"),
                                 crs = 4326)
# project to a metric system

village_coords_cong_1_projected <- st_transform(village_coords_cong_1, crs = 32631)

# create 1000 buffer around each village point
# the distance can be adjusted based on the result obtained

village_coords_cong_1_projected_buffer <- st_buffer(village_coords_cong_1_projected,
                                                   dist = 3000)

# project back to match the raster's crs if they differ

village_coords_cong_1_projected_buffer_crs <- st_transform(village_coords_cong_1_projected_buffer,
                                                          crs = crs(Cong_pop))
cong_pop_density_2020_df_buffer <- terra::extract(Cong_pop, 
                                                 village_coords_cong_1_projected_buffer_crs,
                                                 fun = mean,
                                                 na.rm = TRUE)

Cong_land_cover_water <- crop(land_cover_water, con_vect)
Cong_land_cover_water <- mask(cmr_land_cover_water, con_vect)

Cong_land_cover_built <- crop(land_cover_builts, con_vect)
Cong_land_cover_built <- mask(cmr_land_cover_built, con_vect)

Cong_land_cover_cropland <- crop(land_cover_cropland, con_vect)
Cong_land_cover_cropland <- mask(cmr_land_cover_cropland, con_vect)

Cong_land_cover_trees <- crop(land_cover_trees, con_vect)
Cong_land_cover_trees <- mask(cmr_land_cover_trees, con_vect)

Cong_land_cover_wetland <- crop(land_cover_wetland, con_vect)
Cong_land_cover_wetland <- mask(cmr_land_cover_wetland, con_vect)


Cong_ITN_intervention_2021  <- project(ITN_intervention_2021,  st_crs(spatial_data_cong)$wkt)
Cong_pop  <- project(Cong_pop,  st_crs(spatial_data_cong)$wkt)
Cong_rainfall_2021 <- project(ben_rainfall_2021, st_crs(spatial_data_cong)$wkt)
Cong_temperature_2021 <- project(ben_temperature_2021, st_crs(spatial_data_cong)$wkt)
Cong_ndvi_stack <- project(Cong_ndvi_stack, st_crs(spatial_data_cong)$wkt)
elevation_Cong   <- project(elevation_Cong,   st_crs(spatial_data_cong)$wkt)
Cong_land_cover_water   <- project(Cong_land_cover_water,   st_crs(spatial_data_cong)$wkt)
Cong_land_cover_built   <- project(Cong_land_cover_built,   st_crs(spatial_data_cong)$wkt)
Cong_land_cover_cropland   <- project(Cong_land_cover_cropland,   st_crs(spatial_data_cong)$wkt)
Cong_land_cover_trees   <- project(Cong_land_cover_trees,   st_crs(spatial_data_cong)$wkt)
Cong_land_cover_wetland   <- project(Cong_land_cover_wetland,   st_crs(spatial_data_cong)$wkt)

Cong_ITN_intervention_2021_df <- terra::extract(Cong_ITN_intervention_2021,  spatial_data_cong)
Cong_pop_density_2020_df <- terra::extract(Cong_pop,  spatial_data_cong)
Cong_rainfall_2021_df    <- terra::extract(Cong_rainfall_2021, spatial_data_cong)
Cong_temperature_2021_df <- terra::extract(Cong_temperature_2021, spatial_data_cong)
ndvi_vals_Cong_monthly <- terra::extract(Cong_ndvi_stack, spatial_data_cong)
ndvi_mean_Cong <- apply(ndvi_vals_Cong_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_Cong_df <- terra::extract(elevation_Cong, spatial_data_cong)
Cong_land_cover_water_df <- terra::extract(Cong_land_cover_water, spatial_data_cong)
Cong_land_cover_built_df <- terra::extract(Cong_land_cover_built, spatial_data_cong)
Cong_land_cover_cropland_df <- terra::extract(Cong_land_cover_cropland, spatial_data_cong)
Cong_land_cover_trees_df <- terra::extract(Cong_land_cover_trees, spatial_data_cong)
Cong_land_cover_wetland_df <- terra::extract(Cong_land_cover_wetland, spatial_data_cong)

shp_cong_country <- shp_con %>% 
  group_by(adm0_name) %>% 
  summarise(geometry = st_union(geometry))


bbox <- st_bbox(shp_con)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)

rivers_valid <- st_transform(rivers_valid, st_crs(shp_con))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_con))


# nearest river segment for each point
nn_riv <- st_nn(spatial_data_cong, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(spatial_data_cong, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

spatial_data_cong$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

spatial_data_cong <- spatial_data_cong %>% 
  mutate(pop_density_2020 = cong_pop_density_2020_df_buffer[,2],
         annual_rainfall_2021 = Cong_rainfall_2021_df[,2],
         annual_temperature_2021 = Cong_temperature_2021_df[,2],
         vegetaion_index = ndvi_mean_Cong,
         elevation = elevation_Cong_df[,2],
         lc_water = Cong_land_cover_water_df[,2],
         lc_built = Cong_land_cover_built_df[,2],
         lc_trees = Cong_land_cover_trees_df[,2],
         lc_wetland = Cong_land_cover_wetland_df[,2],
         lc_cropland = Cong_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = Cong_ITN_intervention_2021_df[,2])

spatial_data_cong_df <- spatial_data_cong %>% 
  st_drop_geometry() 

st_write(spatial_data_cong, "data/cong_data/spatial_data_cong_sf.gpkg", append = FALSE)


```


```{r}
# Cameroon case

cmr_pop <- rast("data/cmr_data/cmr_pd_2011_1km_UNadj.tif")

elevation_cmr <- rast("data/cmr_data/elevation/CMR_elv_msk.tif")
ndvi_path <- "data/cmr_data/NDVI2021/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

cmr_ndvi_stack <- rast(ndvi_files)

cmr_land_cover_water <- crop(land_cover_water, cmr_vect)
cmr_land_cover_water <- mask(cmr_land_cover_water, cmr_vect)

cmr_land_cover_built <- crop(land_cover_builts, cmr_vect)
cmr_land_cover_built <- mask(cmr_land_cover_built, cmr_vect)

cmr_land_cover_cropland <- crop(land_cover_cropland, cmr_vect)
cmr_land_cover_cropland <- mask(cmr_land_cover_cropland, cmr_vect)

cmr_land_cover_trees <- crop(land_cover_trees, cmr_vect)
cmr_land_cover_trees <- mask(cmr_land_cover_trees, cmr_vect)

cmr_land_cover_wetland <- crop(land_cover_wetland, cmr_vect)
cmr_land_cover_wetland <- mask(cmr_land_cover_wetland, cmr_vect)


cmr_ITN_intervention_2021  <- project(ITN_intervention_2021,  st_crs(CMR_df_1_sf)$wkt)
cmr_pop  <- project(cmr_pop,  st_crs(CMR_df_1_sf)$wkt)
cmr_rainfall_2021 <- project(ben_rainfall_2021, st_crs(CMR_df_1_sf)$wkt)
cmr_temperature_2021 <- project(ben_temperature_2021, st_crs(CMR_df_1_sf)$wkt)
cmr_ndvi_stack <- project(cmr_ndvi_stack, st_crs(CMR_df_1_sf)$wkt)
elevation_cmr   <- project(elevation_cmr,   st_crs(CMR_df_1_sf)$wkt)
cmr_land_cover_water   <- project(cmr_land_cover_water,   st_crs(CMR_df_1_sf)$wkt)
cmr_land_cover_built   <- project(cmr_land_cover_built,   st_crs(CMR_df_1_sf)$wkt)
cmr_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(CMR_df_1_sf)$wkt)
cmr_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(CMR_df_1_sf)$wkt)
cmr_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(CMR_df_1_sf)$wkt)

cmr_ITN_intervention_2021_df <- terra::extract(cmr_ITN_intervention_2021,  CMR_df_1_sf)
cmr_pop_density_2020_df <- terra::extract(cmr_pop,  CMR_df_1_sf)
cmr_rainfall_2021_df    <- terra::extract(cmr_rainfall_2021, CMR_df_1_sf)
cmr_temperature_2021_df <- terra::extract(cmr_temperature_2021, CMR_df_1_sf)
ndvi_vals_cmr_monthly <- terra::extract(cmr_ndvi_stack, CMR_df_1_sf)
ndvi_mean_cmr <- apply(ndvi_vals_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cmr_df <- terra::extract(elevation_cmr, CMR_df_1_sf)
cmr_land_cover_water_df <- terra::extract(cmr_land_cover_water, CMR_df_1_sf)
cmr_land_cover_built_df <- terra::extract(cmr_land_cover_built, CMR_df_1_sf)
cmr_land_cover_cropland_df <- terra::extract(cmr_land_cover_cropland, CMR_df_1_sf)
cmr_land_cover_trees_df <- terra::extract(cmr_land_cover_trees, CMR_df_1_sf)
cmr_land_cover_wetland_df <- terra::extract(cmr_land_cover_wetland, CMR_df_1_sf)

shp_cmr_country <- shp_cmr %>% 
  mutate(country = "Cameroon")%>% 
  group_by(country) %>% 
  summarise(geometry = st_union(geometry))


bbox <- st_bbox(shp_cmr)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)


rivers_valid <- st_transform(rivers_valid, st_crs(shp_cmr))
lakes_valid  <- st_transform(lakes_valid,  st_crs(shp_cmr))


# nearest river segment for each point
nn_riv <- st_nn(CMR_df_1_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(CMR_df_1_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

CMR_df_1_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

CMR_df_1_sf <- CMR_df_1_sf %>% 
  mutate(pop_density_2020 = cmr_pop_density_2020_df[,2],
         annual_rainfall_2021 = cmr_rainfall_2021_df[,2],
         annual_temperature_2021 = cmr_temperature_2021_df[,2],
         vegetaion_index = ndvi_mean_cmr,
         elevation = elevation_cmr_df[,2],
         lc_water = cmr_land_cover_water_df[,2],
         lc_built = cmr_land_cover_built_df[,2],
         lc_trees = cmr_land_cover_trees_df[,2],
         lc_wetland = cmr_land_cover_wetland_df[,2],
         lc_cropland = cmr_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = cmr_ITN_intervention_2021_df[,2])

CMR_df_1_sf_df <- CMR_df_1_sf %>% 
  st_drop_geometry() 


st_write(CMR_df_1_sf, "data/cmr_data/CMR_df_1_sf.gpkg", append = FALSE)

# Cameroon case with clusters of households

cmr_land_cover_water <- crop(land_cover_water, cmr_vect)

cmr_land_cover_built <- crop(land_cover_builts, cmr_vect)

cmr_land_cover_cropland <- crop(land_cover_cropland, cmr_vect)

cmr_land_cover_trees <- crop(land_cover_trees, cmr_vect)

cmr_land_cover_wetland <- crop(land_cover_wetland, cmr_vect)

cmr_cluster_ITN_intervention_2021  <- project(ITN_intervention_2021, 
                                              st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_pop  <- project(cmr_pop,  st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_rainfall_2021 <- project(ben_rainfall_2021, st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_temperature_2021 <- project(ben_temperature_2021, st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_ndvi_stack <- project(cmr_ndvi_stack, st_crs(cluster_data_cmr_sf)$wkt)
elevation_cluster_cmr   <- project(elevation_cmr,   st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_land_cover_water   <- project(cmr_land_cover_water,   st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_land_cover_built   <- project(cmr_land_cover_built,   st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(cluster_data_cmr_sf)$wkt)
cmr_cluster_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(cluster_data_cmr_sf)$wkt)

cmr_cluster_ITN_intervention_2021_df <- terra::extract(cmr_cluster_ITN_intervention_2021,
                                                       cluster_data_cmr_sf)
cmr_cluster_pop_density_2020_df <- terra::extract(cmr_cluster_pop,  cluster_data_cmr_sf)
cmr_cluster_rainfall_2021_df    <- terra::extract(cmr_cluster_rainfall_2021, cluster_data_cmr_sf)
cmr_cluster_temperature_2021_df <- terra::extract(cmr_cluster_temperature_2021, cluster_data_cmr_sf)
ndvi_vals_cluster_cmr_monthly <- terra::extract(cmr_ndvi_stack, cluster_data_cmr_sf)
ndvi_mean_cluster_cmr <- apply(ndvi_vals_cluster_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cluster_cmr_df <- terra::extract(elevation_cluster_cmr, cluster_data_cmr_sf)
cmr_cluster_land_cover_water_df <- terra::extract(cmr_cluster_land_cover_water, cluster_data_cmr_sf)
cmr_cluster_land_cover_built_df <- terra::extract(cmr_cluster_land_cover_built, cluster_data_cmr_sf)
cmr_cluster_land_cover_cropland_df <- terra::extract(cmr_cluster_land_cover_cropland, cluster_data_cmr_sf)
cmr_cluster_land_cover_trees_df <- terra::extract(cmr_cluster_land_cover_trees, cluster_data_cmr_sf)
cmr_cluster_land_cover_wetland_df <- terra::extract(cmr_cluster_land_cover_wetland, cluster_data_cmr_sf)


# nearest river segment for each point
nn_riv <- st_nn(cluster_data_cmr_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(cluster_data_cmr_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

cluster_data_cmr_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

cluster_data_cmr_sf <- cluster_data_cmr_sf %>% 
  mutate(pop_density_2020 = cmr_cluster_pop_density_2020_df[,2],
         annual_rainfall_2021 = cmr_cluster_rainfall_2021_df[,2],
         annual_temperature_2021 = cmr_cluster_temperature_2021_df[,2],
         vegetaion_index = ndvi_mean_cluster_cmr,
         elevation = elevation_cluster_cmr_df[,2],
         lc_water = cmr_cluster_land_cover_water_df[,2],
         lc_built = cmr_cluster_land_cover_built_df[,2],
         lc_trees = cmr_cluster_land_cover_trees_df[,2],
         lc_wetland = cmr_cluster_land_cover_wetland_df[,2],
         lc_cropland = cmr_cluster_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = cmr_cluster_ITN_intervention_2021_df[,2]) %>% 
  dplyr::select(-c(grid_id, lat, lng))


st_write(cluster_data_cmr_sf, "data/cmr_data/cmr_cluster_df_1_sf.gpkg", append = FALSE)
```


# visualization of covariate that shows bigger variation within the sampled village
```{r}
# Benin

spatial_data_ben_sf <- st_read("data/ben_data/spatial_data_ben_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Benin)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/dist_water_viz_ben.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Benin)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/pop_density_viz_ben.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Benin)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/annual_rainfall_viz_ben.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Benin)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/annual_temperature_viz_ben.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Benin)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Vegetation_index_viz_ben.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)

elevation_ben <- ggplot() +
  geom_sf(data = sampled_site_ben, fill = "white", color = "black") +
  geom_sf(data = spatial_data_ben_sf, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Benin)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/elevation_viz_ben.pdf", plot = elevation_ben, width = 6, height = 4, dpi = 300)



```


```{r}
# congo

spatial_data_cong <- st_read("data/cong_data/spatial_data_cong_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Congo)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/dist_water_viz_cong.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Congo)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/pop_density_viz_cong.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Congo)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/annual_rainfall_viz_cong.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Congo)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/annual_temperature_vizcong.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Congo)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Vegetation_index_viz_cong.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)


elevation_cong <- ggplot() +
  geom_sf(data = sampled_site_cong, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cong, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Congo)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/elevation_viz_cong.pdf", plot = elevation_cong, width = 6, height = 4, dpi = 300)


```


```{r}
# Cameroon

spatial_data_cmr <- st_read("data/cmr_data/cmr_cluster_df_1_sf.gpkg")

dist_water <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = dist_to_water/1000)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Distance to water (km) in sampled site (Cameroon)",
       color = "Distance to water (km)") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/dist_water_viz_cmr.pdf", plot = dist_water, width = 6, height = 4, dpi = 300)

pop_density <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = pop_density_2020)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "population density in sampled site (Cameroon)",
       color = "population density") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/pop_density_viz_cmr.pdf", plot = pop_density, width = 6, height = 4, dpi = 300)

annual_rainfall <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = annual_rainfall_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual rainfall in sampled site (Cameroon)",
       color = "Annual rainfall") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/annual_rainfall_viz_cmr.pdf", plot = annual_rainfall, width = 6, height = 4, dpi = 300)

annual_temperature <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = annual_temperature_2021)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Annual temperature in sampled site (Cameroon)",
       color = "Annual temperature") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/annual_temperature_viz_cmr.pdf", plot = annual_temperature, width = 6, height = 4, dpi = 300)


Vegetation_index <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = vegetaion_index)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "Vegetation index in sampled site (Cameroon)",
       color = "Vegetation index") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Vegetation_index_viz_cmr.pdf", plot = Vegetation_index, width = 6, height = 4, dpi = 300)


elevation_cmr <- ggplot() +
  geom_sf(data = sampled_site_cmr, fill = "white", color = "black") +
  geom_sf(data = spatial_data_cmr, aes(color = elevation)) +
  scale_color_viridis_c(option = "C") +
  labs(title = "elevation in sampled site (Cameroon)",
       color = "elevation") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/elevation_viz_cmr.pdf", plot = elevation_cmr, width = 6, height = 4, dpi = 300)


```



```{r}
# Data for modelling

sampled_data_join_sf <- spatial_data_ben_sf %>% dplyr::select(-c(Village)) %>%
  mutate(country = "Benin") %>% 
  bind_rows(spatial_data_cong %>% dplyr::select(-c(District)) %>% 
              mutate(country = "Congo")) %>% 
    bind_rows(spatial_data_cmr) 



numer_var_ben <- spatial_data_ben_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "annual_ITN_usage_2021",
  "elevation", "vegetaion_index", "pop_density_2020", "dist_to_water", "lc_water",
  "lc_trees", "lc_built", "lc_cropland", "lc_wetland"))

Ben <- cor(numer_var_ben, use = "pairwise.complete.obs")
corrplot(
  Ben,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)


numer_var_cong <- spatial_data_cong %>% 
  st_drop_geometry() %>% 
  select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

Cong <- cor(numer_var_cong, use = "pairwise.complete.obs")
corrplot(
  Cong,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

```



```{r}
num_vars <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetaion_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees", "annual_ITN_usage_2021",
  "lc_built", "lc_cropland", "lc_wetland"))

M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

var1 <- ggcorrplot(
  M,
  type = "full",
  legend.title = "Correlation Matrix",
  lab = TRUE,
  lab_size = 3,
  colors = c("blue", "white", "red"),
  outline.color = "white",
  show.diag = TRUE       
) +
  scale_x_discrete(name = NULL) +
  scale_y_discrete(name = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),      
    panel.background = element_rect(fill = "white", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("Outputs/correlation_matrix.png", var1, dpi = 300, width = 6, height = 4)

num_vars_selected <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "pop_density_2020", "dist_to_water", "lc_water", "lc_trees",
  "lc_built", "lc_cropland", "lc_wetland"))

M_selected <- cor(num_vars_selected, use = "pairwise.complete.obs")

corrplot(
  M_selected,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + 
    annual_temperature_2021 + elevation + vegetaion_index + pop_density_2020 + 
    dist_to_water + lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021,
  data = sampled_data_join_sf
)

vif(fit)

fit1 <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water
  + lc_trees + lc_built + lc_cropland + lc_wetland , 
  data = sampled_data_join_sf
)

vif(fit1)

```

# variogram, autocorrelation and prior ajustment

```{r}
# Project to meters (important!)
sf_data <- st_transform(sampled_data_join_sf, 3857)


sf_data$logit_prev <- qlogis(
  pmin(pmax(sf_data$PR_Pm_tot, 1e-6), 1 - 1e-6)
)
# Extract coordinates
coords <- st_coordinates(sf_data)

# Build gstat object
g <- gstat(id = "pm", formula = logit_prev ~ 1, data = sf_data)

# Empirical variogram

v <- variogram(g,  cutoff = 100000, width = 1000)  # adjust cutoff to country size
# fit an exponential model
v.fit <- fit.variogram(v, model = vgm("Sph"))

# plot empirical + fitted curve
plot(v, v.fit)

plot(v)

```

```{r}
fit_cov <- lm(
  PR_Pm_tot ~ annual_rainfall_2021 + pop_density_2020 +
    dist_to_water + lc_water + lc_trees + lc_built +
    lc_cropland + lc_wetland,
  data = sampled_data_join_sf
)

sampled_data_join_sf$resid <- residuals(fit_cov)


sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

# neighbors within, say, 50 km
nb <- dnearneigh(coords, d1 = 0, d2 = 50000)
lw <- nb2listw(nb, style = "W")

moran.test(sf_data$resid, lw)


```

The observed Moran’s I is almost identical to the expected value under randomness.

The p‑value is very high : no evidence of spatial autocorrelation.

The negative sign is tiny : not meaningful.


```{r}
#sampled_data_join_sf$country_id <- ifelse(sampled_data_join_sf$country == "Benin", 1, 2)

sampled_data_join_sf$country_id <- as.integer(factor(sampled_data_join_sf$country))  # 1, 2

sf_data <- st_as_sf(sampled_data_join_sf, coords = c("lon", "lat"), crs = 4326)
sf_data <- st_transform(sf_data, 3857)

coords <- st_coordinates(sf_data)

coords_A <- coords[sf_data$country_id == 1, ]
coords_B <- coords[sf_data$country_id == 2, ]
coords_C <- coords[sf_data$country_id == 3, ]

country_A <- st_transform(shp_ben, 3857)
country_B <- st_transform(shp_cmr, 3857)
country_C <- st_transform(shp_con, 3857)

country_A_union <- st_union(country_A)
country_B_union <- st_union(country_B)
country_C_union <- st_union(country_C)

country_A_simplified <- st_simplify(country_A_union, dTolerance = 10000)
country_B_simplified <- st_simplify(country_B_union, dTolerance = 10000)
country_C_simplified <- st_simplify(country_C_union, dTolerance = 10000)


# Extract geometry
geom_A <- st_geometry(country_A_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_A, "MULTIPOLYGON")) {
  geom_A <- geom_A[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_A <- geom_A[[1]]

# Convert to matrix with X,Y
boundary_A <- as.matrix(outer_ring_A)[, 1:2, drop = FALSE]



# country_B is an sf polygon or multipolygon
geom_B <- st_geometry(country_B_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_B, "MULTIPOLYGON")) {
  geom_B <- geom_B[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_B <- geom_B[[1]]

# Convert to matrix with X,Y
boundary_B <- as.matrix(outer_ring_B)[, 1:2, drop = FALSE]

# Extract geometry
geom_C <- st_geometry(country_C_simplified)[[1]]

# If MULTIPOLYGON : take the first polygon
if (inherits(geom_C, "MULTIPOLYGON")) {
  geom_C <- geom_C[[1]]
}

# Extract the OUTER ring only (first ring)
outer_ring_C <- geom_C[[1]]

# Convert to matrix with X,Y
boundary_C <- as.matrix(outer_ring_C)[, 1:2, drop = FALSE]


# mesh_A <- inla.mesh.2d(loc = coords_A, max.edge = c(5000, 20000))
# mesh_B <- inla.mesh.2d(loc = coords_B, max.edge = c(5000, 20000))
# mesh_C <- inla.mesh.2d(loc = coords_C, max.edge = c(5000, 20000))

mesh_A <- inla.mesh.2d(
  loc      = coords_A,
  boundary = list(boundary_A),
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_B <- inla.mesh.2d(
  loc = coords_B,
  boundary   = list(boundary_B) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

mesh_C <- inla.mesh.2d(
  loc = coords_C,
  boundary   = list(boundary_C) ,
  max.edge = c(3000, 80000),   # inner ~3 km, outer ~80 km
  cutoff   = 25000,
  offset   = c(0, 100000)
)

# Extract triangle vertex indices
tri <- mesh_C$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_C$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_C <- as.data.frame(coords_C)
colnames(coords_df_C) <- c("X", "Y")

ggplot() +
  geom_sf(data = country_C_simplified, fill = "white", color = "black", size = 0.8) +
  geom_sf(data = mesh_edges_sf, color = "grey60", size = 0.3) +
  geom_point(data = coords_df_C, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh A with Country Boundary")

```


```{r}
# Build Matérn SPDE model of the spatial field on mesh

spde_A <- inla.spde2.pcmatern(
  mesh = mesh_A,
  prior.range = c(10000, 0.5),
  prior.sigma = c(0.02, 0.5)
)

spde_B <- inla.spde2.pcmatern(
  mesh = mesh_B,
  prior.range = c(10000, 0.5),
  prior.sigma = c(0.02, 0.5)   # 0.12 came from the squared root of the sill because it 
)                              # represent the standard deviance while the sill represent the variance.

spde_C <- inla.spde2.pcmatern(
  mesh = mesh_C,
  prior.range = c(10000, 0.5),
  prior.sigma = c(0.02, 0.5)   
) 

# Computes the interpolation weights that map your SPDE spatial field (defined on mesh nodes) to your actual malaria data locations

A_A <- inla.spde.make.A(
  mesh = mesh_A,
  loc = coords_A
)

A_B <- inla.spde.make.A(
  mesh = mesh_B,
  loc = coords_B
)

A_C <- inla.spde.make.A(
  mesh = mesh_C,
  loc = coords_C
)

A_A_full <- matrix(0, nrow = nrow(sf_data), ncol = spde_A$n.spde)
A_B_full <- matrix(0, nrow = nrow(sf_data), ncol = spde_B$n.spde)
A_C_full <- matrix(0, nrow = nrow(sf_data), ncol = spde_C$n.spde)

A_A_full[sf_data$country_id == 1, ] <- as.matrix(A_A)
A_B_full[sf_data$country_id == 2, ] <- as.matrix(A_B)
A_C_full[sf_data$country_id == 3, ] <- as.matrix(A_C)

A_fixed <- Diagonal(n = nrow(sf_data))
A_spatial <- list(A_A_full, A_B_full, A_C_full)


```

# Build stack

```{r}

# Standardization of our continuous variables

sf_data <- sf_data %>% 
  mutate(pop_density_2020 = as.numeric(scale(pop_density_2020)),
         annual_rainfall_2021 = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
         vegetaion_index = as.numeric(scale(vegetaion_index)),
         dist_to_water  = as.numeric(scale(dist_to_water)),
         elevation = as.numeric(scale(elevation)))

# Model building

effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sf_data$annual_rainfall_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept             = 1,
  annual_temperature_2021  = sf_data$annual_temperature_2021,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetaion_index       = sf_data$vegetaion_index,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sf_data$elevation,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sf_data$pop_density_2020,
  dist_to_water         = sf_data$dist_to_water,
  lc_water              = sf_data$lc_water,
  lc_trees              = sf_data$lc_trees,
  lc_built              = sf_data$lc_built,
  lc_cropland           = sf_data$lc_cropland,
  lc_wetland            = sf_data$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_country <- list(
  country_id = sf_data$country_id
)

effects_A <- list(field_A = 1:spde_A$n.spde)
effects_B <- list(field_B = 1:spde_B$n.spde)
effects_C <- list(field_C = 1:spde_C$n.spde)

A_fixed   <- Diagonal(nrow(sf_data))
A_country <- Diagonal(nrow(sf_data))


stack_test_rainfall <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_rainfall,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_temperature <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_temperature,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_vegetation_index <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_vegetation_index,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

stack_test_elevation <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_elevation,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)


stack_test_baseline <- inla.stack(
  data = list(
    y       = sf_data$Number_of_infected_tot_pm,
    Ntrials = sf_data$Number_of_participant_tot
  ),
  A = list(
    A_fixed,
    A_country,
    A_A_full,
    A_B_full,
    A_C_full
  ),
  effects = list(
    effects_fixed_baseline,
    effects_country,
    effects_A,
    effects_B,
    effects_C
  ),
  tag = "est"
)

```


# Fit our models

```{r}
# Penalize complexity which control how much countries differ from each other through
# the country level random effect
hyper_country <- list(
  prec = list(prior = "pc.prec", param = c(1, 0.01))
)

formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetaion_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(country_id, model="iid", hyper=hyper_country) +
  f(field_A, model=spde_A) +
  f(field_B, model=spde_B) +
  f(field_C, model=spde_C)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stack_test_rainfall),
  Ntrials = inla.stack.data(stack_test_rainfall)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_rainfall),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stack_test_temperature),
  Ntrials = inla.stack.data(stack_test_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stack_test_vegetation_index),
  Ntrials = inla.stack.data(stack_test_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stack_test_elevation),
  Ntrials = inla.stack.data(stack_test_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stack_test_baseline),
  Ntrials = inla.stack.data(stack_test_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_test_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


```


The model’s hyperparameters show that country-level differences are almost negligible, meaning the covariates already explain most variation between Benin and Congo. The spatial field in Benin is clearly identifiable: it has a short correlation range (~7 km) and a moderate spatial variance, so spatial structure genuinely improves the fit there. In contrast, Congo’s field is extremely uncertain—with only three locations, the model cannot reliably estimate its spatial range or variance, so any spatial effect in Congo should be interpreted cautiously. The practical consequence is that predictions and spatial maps will be trustworthy in Benin, while Congos spatial surface will be weak, noisy, and driven mostly by priors rather than data. 


# Comparing the different models

```{r}
# Model 1:  Rainfall

# estimated values for the different covariates

res_rainfall$summary.fixed

# Hyper parameters examination

res_rainfall$summary.hyperpar

# DIC

res_rainfall$dic$dic

# WAIC

res_rainfall$waic$waic

# CPO

sum(log(res_rainfall$cpo$cpo), na.rm = TRUE)

summary(res_rainfall$cpo$cpo)

# PIT histogram

hist(res_rainfall$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

obs <- inla.stack.data(stack_test_rainfall)$y /
       inla.stack.data(stack_test_rainfall)$Ntrials

idx.est <- inla.stack.index(stack_test_rainfall, "est")$data

pred.est <- res_rainfall$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, rainfall model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual rainfall model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 2:  temperature

# estimated values for the different covariates

res_temperature$summary.fixed

# Hyper parameters examination

res_temperature$summary.hyperpar

# DIC

res_temperature$dic$dic

# WAIC

res_temperature$waic$waic

# CPO

sum(log(res_temperature$cpo$cpo), na.rm = TRUE)

summary(res_temperature$cpo$cpo)

# PIT histogram

hist(res_temperature$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_temperature, "est")$data

pred.est <- res_temperature$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title =  "Observed vs predicted, temperature model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual temperature model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```

```{r}
# Model 3:  vegetation index

# estimated values for the different covariates

res_vegetation_index$summary.fixed

# Hyper parameters examination

res_vegetation_index$summary.hyperpar

# DIC

res_vegetation_index$dic$dic

# WAIC

res_vegetation_index$waic$waic

# CPO

sum(log(res_vegetation_index$cpo$cpo), na.rm = TRUE)

summary(res_vegetation_index$cpo$cpo)

# PIT histogram

hist(res_vegetation_index$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_vegetation_index, "est")$data

pred.est <- res_vegetation_index$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, vegetation index model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual vegetation index model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```

```{r}
# Model 4:  elevation

# estimated values for the different covariates

res_elevation$summary.fixed

# Hyper parameters examination

res_elevation$summary.hyperpar

# DIC

res_elevation$dic$dic

# WAIC

res_elevation$waic$waic

# CPO

sum(log(res_elevation$cpo$cpo), na.rm = TRUE)

summary(res_elevation$cpo$cpo)

# PIT histogram

hist(res_elevation$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_elevation, "est")$data

pred.est <- res_elevation$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, elevation model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual elevation model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()


```


```{r}
# Model 5:  baseline model

# estimated values for the different covariates

res_baseline$summary.fixed

# Hyper parameters examination

res_baseline$summary.hyperpar

# DIC

res_baseline$dic$dic

# WAIC

res_baseline$waic$waic

# CPO

sum(log(res_baseline$cpo$cpo), na.rm = TRUE)

summary(res_baseline$cpo$cpo)

# PIT histogram

hist(res_baseline$cpo$pit,
     breaks = 10,
     main = "PIT Histogram",
     xlab = "PIT")

# observed vs predicted prevalence

idx.est <- inla.stack.index(stack_test_baseline, "est")$data

pred.est <- res_baseline$summary.fitted.values$mean[idx.est]

predicted_observed <- data.frame(obs = obs, pred = pred.est)

ggplot(data.frame(obs = obs, pred = pred.est),
       aes(x = obs, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1) +
  labs(title = "Observed vs predicted, baseline model",
    x = "Observed prevalence",
    y = "Predicted prevalence"
  ) +
  theme_minimal()

# residual plot

residuals <- obs - pred.est

ggplot(data.frame(pred = pred.est, resid = residuals),
       aes(x = pred, y = resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", size = 1) +
  labs(title = "residual baseline model",
    x = "Predicted",
    y = "Residuals"
  ) +
  theme_minimal()

```
# Conclusion on models comparison

The rainfall model is the right choice because it consistently shows the strongest performance across all diagnostics. It has the lowest DIC and WAIC, the best CPO, the tightest observed‑vs‑predicted fit, and it reduces the spatial field variance more than any other covariate, meaning rainfall explains a large part of the spatial structure instead of leaving it to the latent field. In short, rainfall improves prediction, reduces unexplained spatial variation, and aligns with malaria biology making it the most reliable and scientifically justified model to continue with.

```{r}
idx_est <- inla.stack.index(stack_test_rainfall, "est")$data

fitted_est <- res_rainfall$summary.fitted.values[idx_est, "mean"]

cbind(
  observed = sf_data$PR_Pm_tot,
  fitted   = fitted_est
)[1:25, ]
```


# Creating the grid for prevalence estimation

```{r}
countries1_sf <- rbind(shp_ben %>% dplyr::select(COUNTRY, NAME_3, geometry) %>% 
                        rename(country = COUNTRY,
                               district = NAME_3), shp_con %>% dplyr::select(adm0_name,
                                                                      adm2_name, geometry) %>% 
                        rename(country = adm0_name,
                               district = adm2_name)) 


countries_sf <- rbind(countries1_sf, shp_cmr %>% dplyr::select(Admin2, geometry) %>% 
                        mutate(country = "Cameroon") %>% 
                        rename(district = Admin2))

countries_sf1 <- rbind(countries1_sf, shp_cmr %>% dplyr::select(Admin2, geometry) %>% 
                        mutate(country = "Cameroon") %>% 
                        rename(district = Admin2))%>% 
  st_transform(3857) 

countries_sf1$country_id <- as.integer(factor(countries_sf1$country))




bbox <- st_bbox(countries_sf1)

grid_5km <- st_make_grid(
  st_as_sfc(bbox),
  cellsize = 5000,
  what = "centers"
)

grid_sf <- st_sf(geometry = grid_5km)
grid_sf <- st_intersection(grid_sf, countries_sf1)

grid_vect <- vect(countries_sf)
```



# Extraction of the covariates on the grid cells
```{r}
# Benin case

grid_pop <- raster::merge(ben_pop, Cong_pop, cmr_pop)
grid_elevation <- raster::merge(elevation_ben, elevation_cmr, elevation_Cong)
grid_NDVI <- raster::merge(ben_ndvi_stack, cmr_ndvi_stack, Cong_ndvi_stack)

# land_cover_water <- terra::project(land_cover_water, st_crs(grid_sf)$wkt)
# land_cover_builts <- terra::project(land_cover_builts, st_crs(grid_sf)$wkt)
# land_cover_cropland <- terra::project(land_cover_cropland, st_crs(grid_sf)$wkt)
# land_cover_trees <- terra::project(land_cover_trees, st_crs(grid_sf)$wkt)
# land_cover_wetland <- terra::project(land_cover_wetland, st_crs(grid_sf)$wkt)

grid_land_cover_water <- crop(land_cover_water, grid_sf)

grid_land_cover_built <- crop(land_cover_builts, grid_vect)

grid_land_cover_cropland <- crop(land_cover_cropland, grid_vect)

grid_land_cover_trees <- crop(land_cover_trees, grid_vect)

grid_land_cover_wetland <- crop(land_cover_wetland, grid_vect)

grid_ITN_usage_2021  <- project(ITN_intervention_2021,  st_crs(grid_sf)$wkt)
grid_pop  <- project(grid_pop,  st_crs(grid_sf)$wkt)
grid_rainfall_2021 <- project(ben_rainfall_2021, st_crs(grid_sf)$wkt)
grid_temperature_2021 <- project(ben_temperature_2021, st_crs(grid_sf)$wkt)
grid_ndvi_stack <- project(grid_NDVI, st_crs(grid_sf)$wkt)
elevation_grid   <- project(grid_elevation,   st_crs(grid_sf)$wkt)
grid_land_cover_water   <- project(grid_land_cover_water,   st_crs(grid_sf)$wkt)
grid_land_cover_built   <- project(grid_land_cover_built,   st_crs(grid_sf)$wkt)
grid_land_cover_cropland   <- project(grid_land_cover_cropland,   st_crs(grid_sf)$wkt)
grid_land_cover_trees   <- project(grid_land_cover_trees,   st_crs(grid_sf)$wkt)
grid_land_cover_wetland   <- project(grid_land_cover_wetland,   st_crs(grid_sf)$wkt)



grid_ITN_usage_2021_df <- terra::extract(grid_ITN_usage_2021,  grid_sf)
grid_pop_density_2020_df <- terra::extract(grid_pop,  grid_sf)
grid_rainfall_2021_df    <- terra::extract(grid_rainfall_2021, grid_sf)
grid_temperature_2021_df <- terra::extract(grid_temperature_2021, grid_sf)
ndvi_vals_grid_monthly <- terra::extract(grid_ndvi_stack, grid_sf)
ndvi_mean_grid <- apply(ndvi_vals_grid_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_grid_df <- terra::extract(elevation_grid, grid_sf)
grid_land_cover_water_df <- terra::extract(grid_land_cover_water, grid_sf)
grid_land_cover_built_df <- terra::extract(grid_land_cover_built, grid_sf)
grid_land_cover_cropland_df <- terra::extract(grid_land_cover_cropland, grid_sf)
grid_land_cover_trees_df <- terra::extract(grid_land_cover_trees, grid_sf)
grid_land_cover_wetland_df <- terra::extract(grid_land_cover_wetland, grid_sf)

# Distance to water

bbox <- st_bbox(countries_sf)


rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)


rivers_valid <- st_transform(rivers_valid, st_crs(grid_sf))
lakes_valid  <- st_transform(lakes_valid,  st_crs(grid_sf))

# nearest river segment for each point
nn_riv <- st_nn(grid_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(grid_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)

grid_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

grid_sf <- grid_sf %>% 
  mutate(annual_rainfall_2021 = grid_rainfall_2021_df[,2],
         pop_density_2020 = grid_pop_density_2020_df[,2],
         lc_cropland = grid_land_cover_cropland_df[,2],
         lc_water = grid_land_cover_water_df[,2],
         lc_wetland = grid_land_cover_wetland_df[,2],
         lc_built = grid_land_cover_built_df[,2],
         lc_trees = grid_land_cover_trees_df[,2],
         annual_ITN_usage_2021 = grid_ITN_usage_2021_df[,2]) 

# Identify missing rows
missing_idx <- which(is.na(grid_sf$pop_density_2020))

# Compute nearest neighbors (k = 5 gives robustness)
nn <- st_nn(grid_sf, grid_sf, k = 5)

# Function: return first non‑missing neighbor
get_first_non_na <- function(values, neighbors) {
  for (i in neighbors) {
    if (!is.na(values[i])) return(values[i])
  }
  return(NA)
}

# Apply imputation
for (i in missing_idx) {
  grid_sf$pop_density_2020[i] <- get_first_non_na(grid_sf$pop_density_2020, nn[[i]])
}

st_write(grid_sf, "data/grid_data_for_prediction_sf.gpkg", append = FALSE)
```


```{r}
grid_sf <- read_sf("data/grid_data_for_prediction_sf.gpkg")


# grid_sf$dist_to_water <- scale(grid_sf$dist_to_water)
# grid_sf$pop_density_2020 <- scale(grid_sf$pop_density_2020)
# grid_sf$annual_rainfall_2021 <- scale(grid_sf$annual_rainfall_2021)
# grid_sf$annual_temperature_2021 <- scale(grid_sf$annual_temperature_2021)
# grid_sf$elevation <- scale(grid_sf$elevation)
# grid_sf$Vegetation_index <- scale(grid_sf$Vegetation_index)

grid_sf <- grid_sf %>% 
  mutate(dist_to_water            = as.numeric(scale(dist_to_water)),
         pop_density_2020         = as.numeric(scale(pop_density_2020)),
         annual_rainfall_2021     = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021  = as.numeric(scale(annual_temperature_2021)),
         elevation                = as.numeric(scale(elevation)),
         Vegetation_index         = as.numeric(scale(Vegetation_index)))

coords_pred <- st_coordinates(grid_sf)

coords_pred_A <- coords_pred[grid_sf$country_id == 1, ]
coords_pred_B <- coords_pred[grid_sf$country_id == 2, ]
coords_pred_C <- coords_pred[grid_sf$country_id == 3, ]

A_pred_A <- inla.spde.make.A(mesh = mesh_A, loc = coords_pred_A)
A_pred_B <- inla.spde.make.A(mesh = mesh_B, loc = coords_pred_B)
A_pred_C <- inla.spde.make.A(mesh = mesh_C, loc = coords_pred_C)

A_pred_A_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_A$n.spde)
A_pred_B_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_B$n.spde)
A_pred_C_full <- matrix(0, nrow = nrow(grid_sf), ncol = spde_C$n.spde)

A_pred_A_full[grid_sf$country_id == 1, ] <- as.matrix(A_pred_A)
A_pred_B_full[grid_sf$country_id == 2, ] <- as.matrix(A_pred_B)
A_pred_C_full[grid_sf$country_id == 3, ] <- as.matrix(A_pred_C)

effects_fixed_pred <- data.frame(
  intercept               = 1,
  annual_rainfall_2021    = grid_sf$annual_rainfall_2021,
  pop_density_2020        = grid_sf$pop_density_2020,
  # annual_ITN_usage_2021   = grid_sf$annual_ITN_usage_2021,
  dist_to_water           = grid_sf$dist_to_water,
  lc_water                = grid_sf$lc_water,
  lc_trees                = grid_sf$lc_trees,
  lc_built                = grid_sf$lc_built,
  lc_cropland             = grid_sf$lc_cropland,
  lc_wetland              = grid_sf$lc_wetland
)

effects_country_pred <- list(country_id = grid_sf$country_id)

effects_A_pred <- list(field_A = 1:spde_A$n.spde)
effects_B_pred <- list(field_B = 1:spde_B$n.spde)
effects_C_pred <- list(field_C = 1:spde_C$n.spde)

stack_pred <- inla.stack(
  data = list(y = NA,
    Ntrials = NA),
  A = list(
    Diagonal(nrow(grid_sf)),      # fixed effects
    Diagonal(nrow(grid_sf)),      # country iid
    A_pred_A_full,                # spatial field A
    A_pred_B_full,                # spatial field B
    A_pred_C_full                 # A_pred_B_full C
  ),
  effects = list(
    effects_fixed_pred,
    effects_country_pred,
    effects_A_pred,
    effects_B_pred,
    effects_C_pred
  ),
  tag = "pred"
)

stack_full <- inla.stack(stack_test_rainfall, stack_pred)

res_pred <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stack_full),
  Ntrials = inla.stack.data(stack_full)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stack_full),
    compute = TRUE,
    link = 1     
  ),
  control.fixed = list(mean = 0, prec = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE,
    config = TRUE)
)

#idx_pred <- inla.stack.index(stack_full, "pred")$data


set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stack_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]   # linear predictor for grid cells
  plogis(lp)                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)



grid_sf$GID_0 <- ifelse(grid_sf$country_id == 1, "BEN",
                 ifelse(grid_sf$country_id == 2, "CMR",
                 ifelse(grid_sf$country_id == 3, "COG", NA)))


grid_benin <- grid_sf %>% filter(GID_0 == "BEN")
shp_ben1 <-  st_transform(shp_ben, st_crs(grid_benin))

grid_benin1 <- st_join(grid_benin, shp_ben1, join = st_intersects)

ben_data_pev <- grid_benin1 %>%
  group_by(NAME_2) %>%
  summarise(
    prev_mean   = mean(prev_mean, na.rm = TRUE),
    prev_q025   = mean(prev_q025, na.rm = TRUE),
    prev_q975   = mean(prev_q975, na.rm = TRUE),
    CI_width    = mean(prev_q975 - prev_q025, na.rm = TRUE)
  )


ben_data_pev_map <- shp_ben %>% group_by(NAME_2) %>% summarise(geometry = st_union(geometry)) %>%
  left_join(ben_data_pev %>% st_drop_geometry(), by = "NAME_2")

ben_prev_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_pm.pdf",
       plot = ben_prev_1, width = 6, height = 4, dpi = 300)


ben_prev_uncertainty_1 <- ggplot(ben_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/Pm_pr_estimated/benin_estimated_prev_uncertainty_pm.pdf",
       plot = ben_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_cmr <- grid_sf %>% filter(GID_0 == "CMR")
shp_cmr1 <-  st_transform(shp_cmr, st_crs(grid_cmr))

grid_cmr1 <- st_join(grid_cmr, shp_cmr1, join = st_intersects)

cmr_data_pev <- grid_cmr1 %>%
  group_by(Admin2) %>%
  summarise(
    prev_mean   = mean(prev_mean, na.rm = TRUE),
    prev_q025   = mean(prev_q025, na.rm = TRUE),
    prev_q975   = mean(prev_q975, na.rm = TRUE),
    CI_width    = mean(prev_q975 - prev_q025, na.rm = TRUE)
  )


cmr_data_pev_map <- shp_cmr %>%
  left_join(cmr_data_pev %>% st_drop_geometry(), by = "Admin2")

cmr_prev_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_pm.pdf", 
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

cmr_prev_uncertainty_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/Pm_pr_estimated/cmr_estimated_prev_pm_uncertainty.pdf", 
       plot = cmr_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


grid_con <- grid_sf %>% filter(GID_0 == "COG")
shp_con1 <-  st_transform(shp_con, st_crs(grid_con))

grid_con1 <- st_join(grid_con, shp_con1, join = st_intersects)

con_data_pev <- grid_con1 %>%
  group_by(adm2_name) %>%
  summarise(
    prev_mean   = mean(prev_mean, na.rm = TRUE),
    prev_q025   = mean(prev_q025, na.rm = TRUE),
    prev_q975   = mean(prev_q975, na.rm = TRUE),
    CI_width    = mean(prev_q975 - prev_q025, na.rm = TRUE)
  )


con_data_pev_map <- shp_con %>%
  left_join(con_data_pev %>% st_drop_geometry(), by = "adm2_name")

con_prev_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/cong_estimated_prev_pm.pdf", 
       plot = con_prev_1, width = 6, height = 4, dpi = 300)


con_prev_uncertainty_1 <- ggplot(con_data_pev_map) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Credible interval width")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/Pm_pr_estimated/con_estimated_prev_pm_uncertainty.pdf", 
       plot = con_prev_uncertainty_1, width = 6, height = 4, dpi = 300)


# v1 <- res_pred$summary.fixed
# v2 <- res_pred$summary.fixed

```
# clustered model
Across the three‑country study area, Plasmodium malariae prevalence is strongly influenced by environmental conditions. Rainfall and wetland areas show clear positive associations with infection risk, indicating that humid and water‑rich ecosystems provide favourable breeding conditions for malaria vectors. Built‑up areas display a strong negative effect, reflecting reduced transmission in urbanised landscapes where mosquito habitats are disrupted. Distance to water shows a small but statistically meaningful positive effect. In contrast, cropland, population density, forest cover, and land‑cover water all have credible intervals that include zero, meaning the data do not provide sufficient evidence to classify their influence as either protective or risk‑enhancing. Overall, rainfall and wetlands emerge as key drivers of P. malariae transmission, with urbanisation acting as an important suppressor, while several other covariates show uncertain or negligible effects.

# unclustered model
The updated model produces results that are broadly consistent with the previous analysis, identifying rainfall and wetland environments as strong positive drivers of Plasmodium malariae prevalence, and built‑up areas as a strong negative predictor. However, the new model shows slightly reduced effect sizes for rainfall and increased uncertainty for several covariates. Notably, the effect of distance to water, previously statistically positive, now has a credible interval that includes zero, indicating insufficient evidence for a clear association. Cropland, population density, forest cover, and land‑cover water continue to show wide credible intervals that cross zero, meaning their effects remain statistically uncertain. Overall, the main difference between the two models lies in the reduced certainty around distance to water and slightly adjusted effect sizes, while the key environmental drivers rainfall, wetlands, and urbanisation remain robust.



```{r}
con_sf <- shp_con %>% st_union() %>% st_sf()
grid_con <- grid_sf %>% filter(country_id == 3)

con_sf <- st_transform(con_sf, st_crs(grid_con))


grid_con_poly <- st_make_grid(
  con_sf,
  cellsize = 5000,
  square = TRUE
)

grid_con_poly <- st_sf(geometry = grid_con_poly)

grid_con_poly <- st_intersection(grid_con_poly, con_sf)

#st_crs(grid_benin_poly) <- st_crs(grid_benin)

grid_con_poly$pred <- grid_con$pred[
  st_nearest_feature(grid_con_poly, grid_con)
]

#grid_poly_ben$pred <- grid_sf$pred[st_nearest_feature(grid_poly_ben, grid_sf)]

ggplot(grid_con_poly) +
  geom_sf(aes(fill = pred), color = NA) +
  scale_fill_viridis_c() +
  theme_minimal()
```

```{r}
# Load the modeled prevalence from MAP

africa_pf_pr_2_10 <- rast("data/prevalence_2000/2025_GBD2024_Global_PfPR_2021.tif")

# Cameroon case

cmr_bbox <- ext(shp_cmr)

africa_pf_pr_2_10_cmr <- crop(africa_pf_pr_2_10, cmr_bbox)

cmr_pr_raster_map_2021 <- project(
  africa_pf_pr_2_10_cmr,
  st_crs(shp_cmr)$wkt
)

cmr_admin2_pr <- terra::extract(
  cmr_pr_raster_map_2021,
  shp_cmr,
  fun = mean,
  exact = FALSE
)

cmr_map_pr_pf_2_10 <- shp_cmr %>% 
  mutate(pr_pf_2_10 = cmr_admin2_pr[,2],
         pr_pf_2_10_L = cmr_admin2_pr[,4],
         pr_pf_2_10_U = cmr_admin2_pr[,5],
         CI_width = pr_pf_2_10_U - pr_pf_2_10_L)

cmr_prev_pf_MAP_estimates <- ggplot(cmr_map_pr_pf_2_10) +
  geom_sf(aes(fill = pr_pf_2_10), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_MAP_pf_2_10.png",
       plot = cmr_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

cmr_prev_pf_MAP_estimates_uncertainty <- ggplot(cmr_map_pr_pf_2_10) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cmr_output/MAP_pr_modeled/cmr_pr_uncertainty_MAP_pf_2_10.png",
       plot = cmr_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)


# Benin case

ben_bbox <- ext(shp_ben)

africa_pf_pr_2_10_ben <- crop(africa_pf_pr_2_10, ben_bbox)

ben_pr_raster_map_2021 <- project(
  africa_pf_pr_2_10_ben,
  st_crs(shp_ben)$wkt
)

ben_admin2_pr <- terra::extract(
  ben_pr_raster_map_2021,
  shp_ben,
  fun = mean,
  exact = FALSE
)

ben_map_pr_pf_2_10 <- shp_ben %>% 
  mutate(pr_pf_2_10 = ben_admin2_pr[,2],
         pr_pf_2_10_L = ben_admin2_pr[,4],
         pr_pf_2_10_U = ben_admin2_pr[,5],
         CI_width = pr_pf_2_10_U - pr_pf_2_10_L)

ben_prev_pf_MAP_estimates <- ggplot(ben_map_pr_pf_2_10) +
  geom_sf(aes(fill = pr_pf_2_10), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_MAP_pf_2_10.png",
       plot = ben_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

ben_prev_pf_MAP_estimates_uncertainty <- ggplot(ben_map_pr_pf_2_10) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/ben_output/MAP_pr_modeled/ben_pr_uncertainty_MAP_pf_2_10.png",
       plot = ben_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)

# Congo case

Cong_bbox <- ext(shp_con)

africa_pf_pr_2_10_con <- crop(africa_pf_pr_2_10, Cong_bbox)

con_pr_raster_map_2021 <- project(
  africa_pf_pr_2_10_con,
  st_crs(shp_con)$wkt
)

cong_admin2_pr <- terra::extract(
  con_pr_raster_map_2021,
  shp_con,
  fun = mean,
  exact = FALSE
)

con_map_pr_pf_2_10 <- shp_con %>% 
  mutate(pr_pf_2_10 = cong_admin2_pr[,2],
         pr_pf_2_10_L = cong_admin2_pr[,4],
         pr_pf_2_10_U = cong_admin2_pr[,5],
         CI_width = pr_pf_2_10_U - pr_pf_2_10_L)

con_prev_pf_MAP_estimates <- ggplot(con_map_pr_pf_2_10) +
  geom_sf(aes(fill = pr_pf_2_10), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_MAP_pf_2_10.png",
       plot = con_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

con_prev_pf_MAP_estimates_uncertainty <- ggplot(con_map_pr_pf_2_10) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/cong_output/MAP_pr_modeled/con_pr_uncertainty_MAP_pf_2_10.png",
       plot = con_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)


```



```{r}
# benin
benin_sf <- shp_ben %>% st_union() %>% st_sf()
grid_benin <- grid_sf %>% filter(country_id == 1)

benin_sf <- st_transform(benin_sf, st_crs(grid_benin))


grid_benin_poly <- st_make_grid(
  benin_sf,
  cellsize = 5000,
  square = TRUE
)

grid_benin_poly <- st_sf(geometry = grid_benin_poly)

grid_benin_poly <- st_intersection(grid_benin_poly, benin_sf)

#st_crs(grid_benin_poly) <- st_crs(grid_benin)

grid_benin_poly$pred <- grid_benin$pred[
  st_nearest_feature(grid_benin_poly, grid_benin)
]


# congo

congo_sf <- shp_con %>% st_union() %>% st_sf()
grid_congo <- grid_sf %>% filter(country_id == 2)

congo_sf <- st_transform(congo_sf, st_crs(grid_congo))


grid_congo_poly <- st_make_grid(
  congo_sf,
  cellsize = 5000,
  square = TRUE
)

grid_congo_poly <- st_sf(geometry = grid_congo_poly)

grid_congo_poly <- st_intersection(grid_congo_poly, congo_sf)

#st_crs(grid_congo_poly) <- st_crs(grid_congo)

grid_congo_poly$pred <- grid_congo$pred[
  st_nearest_feature(grid_congo_poly, grid_congo)
]


```



# prepare data for the continental level estimation

```{r}
shp_global <- read_sf("data/Africa_data/shape_files_1/geoBoundariesCGAZ_ADM2.shp")

custom_match <- c(
  XKX = "Europe",
  ESH = "Africa",
  `111` = "Africa",        # Abyei
  `112` = "Asia",          # Aksai Chin
  `113` = "Asia",          # China–India border
  `114` = "Asia",          # Demchok
  `115` = "Europe",        # Dragonja
  `116` = "Africa",        # Dramana-Shakatoe
  `117` = "South America", # Falkland Islands
  `118` = "Asia",          # Gaza Strip
  `119` = "Asia",          # Kalapani
  `120` = "South America", # Isla Brasilera
  `121` = "Asia",          # Siachen-Saltoro
  `122` = "Africa",        # Koualou
  `123` = "Asia",          # Liancourt Rocks
  `124` = "Asia",          # No Man's Land
  `125` = "Asia",          # Paracel Islands
  `126` = "Asia",          # Sanafir & Tiran Islands
  `127` = "Asia",          # Senkaku Islands
  `128` = "Asia",          # Spratly Islands
  `129` = "Asia"           # West Bank
)

shp_global$continent <- countrycode(
  shp_global$shapeGroup,
  origin = "iso3c",
  destination = "continent",
  custom_match = custom_match
)

shp_africa <- shp_global %>% 
  filter(continent=="Africa")

custom_match <- c(
  `111` = "Abyei",                   # Abyei
  `116` = "Dramana-Shakatoe",        # Dramana-Shakatoe
  `122` = "Koualou"                  # Koualou
  )


shp_africa$country <- countrycode(
  shp_africa$shapeGroup,
  origin = "iso3c",
  destination = "country.name",
  custom_match = custom_match
)

shp_africa_simple <- ms_simplify(
  shp_africa,
  keep = 0.05,          # keep 5% of the vertices
  keep_shapes = TRUE
)

st_write(
  shp_africa_simple,
  "data/Africa_data/shp_africa/Africa_simplified.shp",
  delete_layer = TRUE,
  append = TRUE
)


ggplot()+
  geom_sf(data = shp_africa_simple, fill = "grey", color = "black")

```

To avoid extrapolating predictions into areas where malaria transmission is not considered ecologically feasible, model predictions will be masked using the Plasmodium falciparum transmission limits provided by the Malaria Atlas Project. Although this study focuses on Plasmodium malariae, no continent-wide transmission limits are currently available for this species. Given that P. malariae is predominantly transmitted in areas where P. falciparum is endemic and frequently occurs as a co-infection, the P. falciparum transmission limits will be used as a proxy for malaria transmission suitability.


```{r}

shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp") %>% st_transform(3857)

# Dissolve all polygons into one
shp_africa <- st_union(shp_africa)
shp_africa <- st_sf(geometry = shp_africa)

bbox <- st_bbox(shp_africa)

grid_5km_africa <- st_make_grid(
  st_as_sfc(bbox),
  cellsize = 5000,
  what = "centers"
)

grid_sf_africa <- st_sf(geometry = grid_5km_africa)
grid_sf_africa <- st_filter(grid_sf_africa, shp_africa)

ndvi_path <- "Africa_data/NDVI_africa/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

path1 <- "data/Africa_data/climateEngine_download.precipitation.tif"
path2 <- "data/Africa_data/ERA5_LAND_DAILY_temperature_2m_2021-01-01_2021-12-31_-39p5213to41p4698N_-22p2966to54p0315E.temperature_2m.tif"


ITN_intervention_2021 <- rast("data/ITN_2000/2025_GBD2024_Africa_ITN_2021.tif")
africa_rainfall_2021 <- rast(path1)
africa_temperature_2021 <- rast(path2)
africa_pop <- rast("data/Africa_data/pop_density/gpw_v4_population_density_rev11_2020_30_sec.tif")
africa_elevation <- rast("data/Africa_data/elevation/wc2.1_30s_elev.tif")
africa_ndvi_stack<- rast(ndvi_files)



africa_land_cover_water <- crop(land_cover_water, ben_vect)
africa_land_cover_water <- mask(cmr_land_cover_water, ben_vect)

africa_land_cover_built <- crop(land_cover_builts, ben_vect)
africa_land_cover_built <- mask(cmr_land_cover_built, ben_vect)

africa_land_cover_cropland <- crop(land_cover_cropland, ben_vect)
africa_land_cover_cropland <- mask(cmr_land_cover_cropland, ben_vect)

africa_land_cover_trees <- crop(land_cover_trees, ben_vect)
africa_land_cover_trees <- mask(cmr_land_cover_trees, ben_vect)

africa_land_cover_wetland <- crop(land_cover_wetland, ben_vect)
africa_land_cover_wetland <- mask(cmr_land_cover_wetland, ben_vect)


africa_ITN_intervention_2021  <- project(ITN_intervention_2021,  st_crs(grid_sf_africa)$wkt)
africa_pop  <- project(africa_pop,  st_crs(grid_sf_africa)$wkt)
africa_rainfall_2021 <- project(africa_rainfall_2021, st_crs(grid_sf_africa)$wkt)
africa_temperature_2021 <- project(africa_temperature_2021, st_crs(grid_sf_africa)$wkt)
africa_ndvi_stack <- project(africa_ndvi_stack, st_crs(grid_sf_africa)$wkt)
africa_elevation   <- project(africa_elevation,   st_crs(grid_sf_africa)$wkt)
africa_land_cover_water   <- project(africa_land_cover_water,   st_crs(grid_sf_africa)$wkt)
africa_land_cover_built   <- project(africa_land_cover_built,   st_crs(grid_sf_africa)$wkt)
africa_land_cover_cropland   <- project(africa_land_cover_cropland,   st_crs(grid_sf_africa)$wkt)
africa_land_cover_trees   <- project(africa_land_cover_trees,   st_crs(grid_sf_africa)$wkt)
africa_land_cover_wetland   <- project(africa_land_cover_wetland,   st_crs(grid_sf_africa)$wkt)

africa_ITN_intervention_2021_df <- terra::extract(africa_ITN_intervention_2021,  grid_sf_africa)
africa_pop_density_2020_df <- terra::extract(africa_pop,  grid_sf_africa)
africa_rainfall_2021_df    <- terra::extract(africa_rainfall_2021, grid_sf_africa)
africa_temperature_2021_df <- terra::extract(africa_temperature_2021, grid_sf_africa)
ndvi_vals_africa_monthly <- terra::extract(africa_ndvi_stack, grid_sf_africa)
ndvi_mean_africa <- apply(ndvi_vals_africa_monthly[,-1], 1, mean, na.rm = TRUE)
africa_elevation_df <- terra::extract(africa_elevation, grid_sf_africa)
africa_land_cover_water_df <- terra::extract(africa_land_cover_water, grid_sf_africa)
africa_land_cover_built_df <- terra::extract(africa_land_cover_built, grid_sf_africa)
africa_land_cover_cropland_df <- terra::extract(africa_land_cover_cropland, grid_sf_africa)
africa_land_cover_trees_df <- terra::extract(africa_land_cover_trees, grid_sf_africa)
africa_land_cover_wetland_df <- terra::extract(africa_land_cover_wetland, grid_sf_africa)

shp_africa_summarise <- shp_africa %>% 
  group_by(continent) %>% 
  summarise(geometry = st_union(geometry))

# distance to water

# bbox <- st_bbox(grid_sf_africa)

rivers <- vect("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp")
lakes <- vect("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp")


africa4326 <- st_transform(shp_africa, 4326)
africa4326 <- vect(africa4326)

lakes_africa <- crop(lakes, africa4326)

r_template <- rast(
  ext(vect(st_as_sf(shp_africa))),
  resolution = 5000,
  crs = "EPSG:3857")

rivers <- project(rivers, "EPSG:3857")
lakes_africa <- project(lakes_africa, "EPSG:3857")

rivers2 <- rivers[rivers$ORD_STRA >= 3, ]

riv_r <- rasterize(
  rivers2,
  r_template,
  field = 1,
  background = NA)


lake_r <- rasterize(
  lakes_africa,
  r_template,
  field = 1,
  background = NA)

water <- cover(riv_r, lake_r)

dist_water <- distance(
    water,
    filename = "data/dist_to_water_5km.tif",
    overwrite = TRUE)

grid_sf_africa$dist_to_water <- terra::extract(
    dist_water,
    vect(grid_sf_africa))[,2]


grid_sf_africa <- grid_sf_africa %>% 
  mutate(pop_density_2020 = africa_pop_density_2020_df[,2],
         annual_rainfall_2021 = africa_rainfall_2021_df[,2],
         annual_temperature_2021 = africa_temperature_2021_df[,2],
         vegetaion_index = ndvi_mean_africa,
         elevation = africa_elevation_df[,2],
         lc_water = africa_land_cover_water_df[,2],
         lc_built = africa_land_cover_built_df[,2],
         lc_trees = africa_land_cover_trees_df[,2],
         lc_wetland = africa_land_cover_wetland_df[,2],
         lc_cropland = africa_land_cover_cropland_df[,2],
         annual_ITN_usage_2021 = africa_ITN_intervention_2021_df[,2])


# CMR_df_1_sf_df <- CMR_df_1_sf %>% 
#   st_drop_geometry() 
# 
# st_write(CMR_df_1_sf, "cmr_data/CMR_df_1_sf.gpkg", append = FALSE)


nn <- st_nn(grid_sf_africa, grid_sf_africa, k = 200)

vars_to_fill <- names(grid_sf_africa)[
  sapply(grid_sf_africa, function(x) any(is.na(x)))
]

for (var in vars_to_fill) {
  grid_sf_africa[[var]] <- as.numeric(grid_sf_africa[[var]])
}

nn2 <- sapply(nn, function(x) x[2])


for (var in vars_to_fill) {
  missing <- is.na(grid_sf_africa[[var]])
  grid_sf_africa[[var]][missing] <- grid_sf_africa[[var]][ nn2[missing] ]
}


for (var in vars_to_fill) {
  missing <- is.na(grid_sf_africa[[var]])
  grid_sf_africa[[var]][missing] <- mean(grid_sf_africa[[var]], na.rm = TRUE)
}


st_write(grid_sf_africa, "data/Africa_data/grid_data_for_prediction_sf.gpkg", append = FALSE)

```


```{r}
grid_sf_africa <- st_read("data/Africa_data/grid_data_for_prediction_sf.gpkg")

```


```{r}
#sampled_data_join_sf$country_id <- ifelse(sampled_data_join_sf$country == "Benin", 1, 2)

Africa_union <- st_union(shp_africa )

sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

sampled_data_join_sf_sp     <- as(sampled_data_join_sf, "Spatial")
Africa_union_sp <- as(Africa_union, "Spatial")

coords_Africa <- coordinates(sampled_data_join_sf_sp)


mesh_Africa <- inla.mesh.2d(
  loc      = coords_Africa,
  boundary = inla.sp2segment(Africa_union_sp),
  max.edge = c(500e3, 800e3),   
  cutoff   = 250e3,
  offset   = c(100e3, 2000e3))

# Extract triangle vertex indices
tri <- mesh_Africa$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_Africa <- as.data.frame(coords_Africa)
colnames(coords_df_Africa) <- c("X", "Y")

ggplot() +
  geom_sf(data = mesh_edges_sf, fill = NA, color = "grey60", size = 0.3) +
  geom_sf(data = Africa_union, fill = NA, color = "black", size = 0.8) +
  geom_point(data = coords_df_Africa, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh for Africa with Boundary")
```


```{r}
spde_Africa <- inla.spde2.pcmatern(
  mesh = mesh_Africa,
  prior.range = c(10000, 0.5),
  prior.sigma = c(0.02, 0.5)
)

A_est <- inla.spde.make.A(mesh_Africa, loc = coords_Africa)

sampled_data_join_sf <- sampled_data_join_sf %>% 
  mutate(pop_density_2020 = as.numeric(scale(pop_density_2020)),
         annual_rainfall_2021 = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
         vegetaion_index = as.numeric(scale(vegetaion_index)),
         dist_to_water  = as.numeric(scale(dist_to_water)),
         elevation = as.numeric(scale(elevation)))

```



```{r}
effects_fixed_rainfall <- data.frame(
  intercept             = 1,
  annual_rainfall_2021  = sampled_data_join_sf$annual_rainfall_2021,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_temperature <- data.frame(
  intercept                = 1,
  annual_temperature_2021  = sampled_data_join_sf$annual_temperature_2021,
  pop_density_2020         = sampled_data_join_sf$pop_density_2020,
  dist_to_water            = sampled_data_join_sf$dist_to_water,
  lc_water                 = sampled_data_join_sf$lc_water,
  lc_trees                 = sampled_data_join_sf$lc_trees,
  lc_built                 = sampled_data_join_sf$lc_built,
  lc_cropland              = sampled_data_join_sf$lc_cropland,
  lc_wetland               = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_vegetation_index <- data.frame(
  intercept             = 1,
  vegetaion_index       = sampled_data_join_sf$vegetaion_index,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


effects_elevation      <- data.frame(
  intercept             = 1,
  elevation             = sampled_data_join_sf$elevation,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)

effects_fixed_baseline <- data.frame(
  intercept             = 1,
  pop_density_2020      = sampled_data_join_sf$pop_density_2020,
  dist_to_water         = sampled_data_join_sf$dist_to_water,
  lc_water              = sampled_data_join_sf$lc_water,
  lc_trees              = sampled_data_join_sf$lc_trees,
  lc_built              = sampled_data_join_sf$lc_built,
  lc_cropland           = sampled_data_join_sf$lc_cropland,
  lc_wetland            = sampled_data_join_sf$lc_wetland
  # annual_ITN_usage_2021 = sf_data$annual_ITN_usage_2021
)


stk_est_rainfall <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_rainfall
  ),
  tag = "est"
)

stk_est_temperature <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_temperature
  ),
  tag = "est"
)


stk_est_vegetation_index <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_vegetation_index
  ),
  tag = "est"
)

stk_est_elevation <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_elevation
  ),
  tag = "est"
)


stk_est_baseline <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$Number_of_infected_tot_pm,
    Ntrials = sampled_data_join_sf$Number_of_participant_tot
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_baseline
  ),
  tag = "est"
)

```


```{r}
formula_rainfall <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_temperature <- y ~ 
  -1 +
  intercept +
  annual_temperature_2021 +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_vegetation_index  <- y ~ 
  -1 +
  intercept +
  vegetaion_index +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_elevation <- y ~ 
  -1 +
  intercept +
  elevation +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)

formula_baseline <- y ~ 
  -1 +
  intercept +
  pop_density_2020 +
  dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + 
  f(spatial, model = spde_Africa)



res_rainfall <- inla(
  formula_rainfall,
  family = "binomial",
  data = inla.stack.data(stk_est_rainfall  ),
  Ntrials = inla.stack.data(stk_est_rainfall )$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_rainfall ),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_temperature <- inla(
  formula_temperature,
  family = "binomial",
  data = inla.stack.data(stk_est_temperature),
  Ntrials = inla.stack.data(stk_est_temperature)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_temperature),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_vegetation_index <- inla(
  formula_vegetation_index,
  family = "binomial",
  data = inla.stack.data(stk_est_vegetation_index),
  Ntrials = inla.stack.data(stk_est_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


res_elevation <- inla(
  formula_elevation,
  family = "binomial",
  data = inla.stack.data(stk_est_elevation),
  Ntrials = inla.stack.data(stk_est_elevation)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_elevation),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

res_baseline <- inla(
  formula_baseline,
  family = "binomial",
  data = inla.stack.data(stk_est_baseline),
  Ntrials = inla.stack.data(stk_est_baseline)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_baseline),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

```


```{r}

samples <- inla.posterior.sample(
  n = 100,
  result = res_rainfall 
)

# extraction of the fix effect

fixed_names <- res_rainfall$names.fixed

fixed_idx <- match(
  paste0(fixed_names, ":1"),
  rownames(samples[[1]]$latent)
)

# extract the spatial effect

spatial_idx <- grep(
  "^spatial:",
  rownames(samples[[1]]$latent)
)



idx_est <- inla.stack.index(stk_est_rainfall, "est")$data

fitted_est <- res_rainfall$summary.fitted.values[idx_est, "mean"]

cbind(
  observed = sampled_data_join_sf$PR_Pm_tot,
  fitted   = fitted_est
)[1:30, ]
```


```{r}

grid_sf_africa <- grid_sf_africa %>% 
  mutate(annual_rainfall_2021       = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021    = as.numeric(scale(annual_temperature_2021)),
         annual_ITN_usage_2021      = as.numeric(scale(annual_ITN_usage_2021)),
         vegetaion_index            = as.numeric(scale(vegetaion_index)),
         elevation                  = as.numeric(scale(elevation)),
         dist_to_water              = as.numeric(scale(dist_to_water)),
         pop_density_2020           = as.numeric(scale(pop_density_2020)))

# Build the A‑matrix for prediction locations
A_pred <- inla.spde.make.A(
  mesh = mesh_Africa,
  loc = st_coordinates(grid_sf_africa))

#Build the prediction stack

stk_pred <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid_sf_africa$annual_rainfall_2021,
      pop_density      = grid_sf_africa$pop_density_2020,
      dist_to_water    = grid_sf_africa$dist_to_water,
      lc_cropland      = grid_sf_africa$lc_cropland,
      lc_water         = grid_sf_africa$lc_water,
      lc_wetland       = grid_sf_africa$lc_wetland,
      lc_built         = grid_sf_africa$lc_built,
      lc_trees         = grid_sf_africa$lc_trees
    )
  ),
  tag = "pred"
)


# Combine estimation + prediction stacks
stk_full <- inla.stack(stk_est_rainfall, stk_pred)


# Fit the model again with the full stack
res_pred <- inla(
  formula_rainfall,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full),
  control.predictor = list(
    A = inla.stack.A(stk_full),
    compute = TRUE,
    link    = 1
  ),
  control.fixed = list(mean = 0, prec = 1),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


# Extract predictions for the grid
index_pred <- inla.stack.index(stk_full, "pred")$data

grid_sf_africa$pred_mean <- res_pred$summary.fitted.values[index_pred, "mean"]
grid_sf_africa$pred_sd   <- res_pred$summary.fitted.values[index_pred, "sd"]
grid_sf_africa$pred_q025 <- res_pred$summary.fitted.values[index_pred, "0.025quant"]
grid_sf_africa$pred_q975 <- res_pred$summary.fitted.values[index_pred, "0.975quant"]

# Using the transmission limit to mask the prediction grid

grid_sf_africa_wgs84 <- st_transform(grid_sf_africa, 4326)
grid_vect <- vect(grid_sf_africa_wgs84)

df_grid_africa <- grid_sf_africa_wgs84 %>%
    mutate(
        x = st_coordinates(.)[,1],
        y = st_coordinates(.)[,2]
    ) %>%
    st_drop_geometry()

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")

# global_spatial_limit_3857 <- project(global_spatial_limit, crs(shp_africa))


pf_limit_africa <- crop(global_spatial_limit, vect(shp_africa %>% st_transform(4326)))
pf_limit_africa <- mask(pf_limit_africa, vect(shp_africa %>% st_transform(4326)))


grid_with_limit <- terra::extract(pf_limit_africa, grid_vect)

grid_masked <- cbind(df_grid_africa, grid_with_limit[,2])

grid_masked_1 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when(is.na(limit) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q025  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q025),
         pred_q975  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q975)) 

grid_masked_2 <- grid_masked %>% 
  rename(limit = `grid_with_limit[, 2]`) %>% 
  mutate(limit = case_when((is.na(limit) | limit == 1) ~ 0,
                           TRUE ~ limit),
         pred_mean  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_mean),
         pred_sd    = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_sd),
         pred_q025  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q025),
         pred_q975  = case_when(limit == 0 ~ NA,
                                TRUE ~ pred_q975)) 


grid_masked_sf <- st_as_sf(
    grid_masked_1,
    coords = c("x", "y"),
    crs = 4326)

grid_masked_sf_1 <- st_as_sf(
    grid_masked_2,
    coords = c("x", "y"),
    crs = 4326)

# shp_africa1 <-  st_transform(shp_africa, st_crs(grid_masked_sf))
# 
# 
# grid_africa1 <- st_join(grid_masked_sf, shp_africa1%>% 
#   select(-c(shapeName, shapeID, shapeGroup, shapeType, continent, country)), 
#   join = st_intersects) 
# 
# Africa_data_pev <- grid_africa1 %>%
#   group_by(shapeID) %>%
#   summarise(
#     prev_mean   = mean(pred_mean, na.rm = TRUE),
#     prev_q025   = mean(pred_q025, na.rm = TRUE),
#     prev_q975   = mean(pred_q975, na.rm = TRUE),
#     CI_width    = mean(pred_q975 - pred_q025, na.rm = TRUE))
# 
# Africa_country_prev_tx <- country_tx_sf %>%
#   left_join(Africa_country_prev, by = "shapeName")
# 
# Africa_data_pev <- Africa_data_pev %>%
#   mutate(prev_mean = if_else(is.nan(prev_mean), NA_real_, prev_mean))
# 
# Africa_data_pev_map <- shp_africa %>%
#   left_join(Africa_data_pev %>% st_drop_geometry(), by = "shapeID")
# 

shp_africa_country <- shp_africa %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry))


shp_africa_countryw4326 <- shp_africa_country %>% st_transform(4326)
shp_africaw4326 <- shp_africa %>% st_transform(4326)


Africa_prev_1 <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.8,
          alpha = 0.8) +
  geom_sf(data = shp_africaw4326,
          fill = NA,
          color = "black",
          size = 0.1) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_admin2_level_instable.png",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country <- ggplot() +
  geom_sf(data = grid_masked_sf,
          aes(color = pred_mean),
          size = 0.8,
          alpha = 0.8) +
  geom_sf(data = shp_africa_countryw4326,
          fill = NA,
          color = "black",
          size = 0.1) +
  scale_color_viridis_c(option = "magma", na.value = "white") +
  theme_bw(base_size = 15) +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("Outputs/africa_output/2021_africa_pr_malariae_map_country_level.png",
       plot = Africa_prev_country, width = 6, height = 4, dpi = 300)

```

```{r}

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")

# global_spatial_limit_3857 <- project(global_spatial_limit, crs(shp_africa))


pf_limit_africa <- crop(global_spatial_limit, vect(shp_africa %>% st_transform(4326)))
pf_limit_africa <- mask(pf_limit_africa, vect(shp_africa %>% st_transform(4326)))

df_limit_africa <- as.data.frame(pf_limit_africa, xy = TRUE, na.rm = TRUE)


writeRaster(
    pf_limit_africa,
    "data/Africa_data/spatial_limit_of_transmission/Pf_limit_Africa.tif",
    overwrite = TRUE
)

plot(pf_limit_africa)


countries <- c("Algeria", "Morocco", "Tunisia", "Egypt")

shp_africa_simple[shp_africa_simple$country %in% countries, ]

# Convert raster to data frame for ggplot
r_df <- as.data.frame(pf_limit_africa, xy = TRUE, na.rm = FALSE)

ggplot() +
  geom_raster(
    data = r_df,
    aes(x = x, y = y, fill = `2010_Pf_Limits_Decompressed`)
  ) +
  geom_sf(
    data = shp_africa_simple,
    fill = NA,
    color = "red",
    size = 0.6
  ) +
  scale_fill_viridis_c(option = "magma", na.value = "white") +
  coord_sf() +
  theme_minimal(base_size = 18) +
  labs(
    title = "Global Spatial Limit with Africa Boundary",
    fill  = "Value"
  )
```


```{r}
set.seed(123)
samples <- inla.posterior.sample(100, res_pred)

idx_pred <- inla.stack.index(stk_full, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]   # linear predictor for grid cells
  plogis(lp)                 # convert logit  prevalence
})


grid_sf$prev_mean   <- rowMeans(pred_mat)
grid_sf$prev_sd     <- apply(pred_mat, 1, sd)
grid_sf$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid_sf$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)

```







```{r}

```


```{r}

```


```{r}

```



```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```







