---
title: "cmrchmarking exercise"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#load necessary libraries
library(malariaAtlas)
library(sf)
library(leaflet)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(mapview)
library(spData)
library(gganimate)
library(htmlwidgets)
library(webshot)
library(chirps)
library(viridis)
library(patchwork)
library(GGally)
library(spdep)
library(INLA)
library(leafsync)
library(SpatialEpi)
library(geodata)
library(tidygeocoder)
library(nngeo)
library(car)
library(gstat)
library(raster)
library(FNN)
```


```{r}
#set the working directory

setwd("C:/Users/USER/Documents/Research project tubingen/R_project_tuebingen/")
```

```{r}
# Downloading the malaria dataset
# library(malariaAtlas)
# d <- getPR(country = "Cameroon", species = "BOTH")
# ggplot2::autoplot(d)
# 
# write_csv(d, "cmr_data/cmr_map_PR.csv")

```


```{r}
GPS_cmr <- st_read("cmr_data/GPS location 2011/CMGE61FL.shp")

head(GPS_cmr)
names(GPS_cmr)

```


```{r}
GPS_cmr_coords <- GPS_cmr %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_cmr)[, 1],
    latitude  = st_coordinates(GPS_cmr)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  filter(latitude != 0 & longitude != 0)
```

# loading the extracted MAP data

```{r}
raw_prevalence_cmr <- read.csv("cmr_data/cmr_map_PR.csv")

raw_prevalence_cmr_with_coord <- raw_prevalence_cmr %>% 
  filter(!is.na(longitude) & !is.na(latitude))

raw_prevalence_cmr_without_coord <- raw_prevalence_cmr %>% 
  filter(is.na(longitude) & is.na(latitude)) %>% 
  dplyr::select(-c(longitude, latitude))


map_without_coords_cmr <- raw_prevalence_cmr_without_coord %>%
  mutate(cluster_num = as.integer(substr(dhs_id, 9, 16)))

map_recovered_cmr <- map_without_coords_cmr %>%
  left_join(GPS_cmr_coords, by = c("cluster_num" = "DHSCLUST"))

cat("Recovered:", sum(!is.na(map_recovered_cmr$longitude)), "\n")
cat("Still missing:", sum(is.na(map_recovered_cmr$longitude)), "\n")
```
```{r}
map_full_cmr <- bind_rows(
  raw_prevalence_cmr_with_coord %>% filter(!is.na(longitude) & !is.na(latitude)),
  map_recovered_cmr %>% filter(!is.na(longitude) & !is.na(latitude))
)

write_csv(map_full_cmr, file = "cmr_data/cmr_MAP_full_raw_prevalence.csv")
```

# Filter for the year 2011 for our analysis

```{r}
map_2011_cmr <- map_full_cmr %>%
  filter(year_start == 2011) %>%
  filter(!is.na(longitude) & !is.na(latitude))

map_2011_cmr_sf <- st_as_sf(map_2011_cmr, coords = c("longitude", "latitude"), crs = 4326)

plot(st_geometry(map_2011_cmr_sf), pch = 16, cex = 0.5, main = "2011 Survey Points — Cameroon")
```

```{r}
coords <- cbind(map_2011_cmr$longitude, map_2011_cmr$latitude)

mesh <- inla.mesh.2d(
  loc      = coords,
  max.edge = c(0.5, 2),
  cutoff   = 0.05,
  offset   = c(0.5, 1)
)

# Extract triangle vertices
triangles <- mesh$graph$tv

# Build polygons
mesh_list <- lapply(1:nrow(triangles), function(i) {
  coords_i <- mesh$loc[triangles[i, ], 1:2]
  coords_i <- rbind(coords_i, coords_i[1, ])  # close polygon
  st_polygon(list(coords_i))
})

mesh_sf <- st_as_sf(st_sfc(mesh_list))
coords_sf <- st_as_sf(
  data.frame(x = coords[,1], y = coords[,2]),
  coords = c("x", "y"),
  crs = NA
)


# Plot
Mesh <- ggplot() +
  geom_sf(data = mesh_sf, fill = NA, color = "grey40") +
  geom_sf(data = coords_sf, color = "red", size = 0.8) +
  theme_minimal()

# ggsave("cmr_data/mesh_map_data.png", plot = Mesh, width = 8, height = 10, dpi = 150)
```

# Extraction of the covariates at the sampled location 
```{r}
path1 <- "cmr_data/chirps_2011/CHIRPS_PENTAD_precipitation_2011-01-01_2011-12-31_-5p8264to13p9448N_0p6930to19p1696E.precipitation.tif"
path2 <- "cmr_data/ERA5_temperature/ERA5_LAND_DAILY_temperature_2m_2011-01-01_2011-12-31_-5p4765to13p9448N_0p6930to19p3454E.temperature_2m.tif"
ndvi_path <- "cmr_data/NDVI2011/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

cmr_pop <- rast("cmr_data/cmr_pd_2011_1km_UNadj.tif")
cmr_rainfall_2011 <- rast(path1)
cmr_temperature_2011 <- rast(path2)
elevation_cmr <- rast("cmr_data/elevation/cmr_elv_msk.tif")
cmr_land_cover_water <- rast("ben_data/landuse/WorldCover_water_30s.tif")
cmr_land_cover_built <- rast("ben_data/landuse/WorldCover_built_30s.tif")
cmr_land_cover_cropland <- rast("ben_data/landuse/WorldCover_cropland_30s.tif")
cmr_land_cover_trees <- rast("ben_data/landuse/WorldCover_trees_30s.tif")
cmr_land_cover_wetland <- rast("ben_data/landuse/WorldCover_wetland_30s.tif")
cmr_ndvi_stack<- rast(ndvi_files)


cmr_pop  <- project(cmr_pop,  st_crs(map_2011_cmr_sf)$wkt)
cmr_rainfall_2011 <- project(cmr_rainfall_2011, st_crs(map_2011_cmr_sf)$wkt)
cmr_temperature_2011 <- project(cmr_temperature_2011, st_crs(map_2011_cmr_sf)$wkt)
cmr_ndvi_stack <- project(cmr_ndvi_stack, st_crs(map_2011_cmr_sf)$wkt)
elevation_cmr   <- project(elevation_cmr,   st_crs(map_2011_cmr_sf)$wkt)
cmr_land_cover_water   <- project(cmr_land_cover_water,   st_crs(map_2011_cmr_sf)$wkt)
cmr_land_cover_built   <- project(cmr_land_cover_built,   st_crs(map_2011_cmr_sf)$wkt)
cmr_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(map_2011_cmr_sf)$wkt)
cmr_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(map_2011_cmr_sf)$wkt)
cmr_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(map_2011_cmr_sf)$wkt)

cmr_pop_density_2011_df <- terra::extract(cmr_pop,  map_2011_cmr_sf)
cmr_rainfall_2011_df    <- terra::extract(cmr_rainfall_2011, map_2011_cmr_sf)
cmr_temperature_2011_df <- terra::extract(cmr_temperature_2011, map_2011_cmr_sf)
ndvi_vals_cmr_monthly <- terra::extract(cmr_ndvi_stack, map_2011_cmr_sf)
ndvi_mean_cmr <- apply(ndvi_vals_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cmr_df <- terra::extract(elevation_cmr, map_2011_cmr_sf)
cmr_land_cover_water_df <- terra::extract(cmr_land_cover_water, map_2011_cmr_sf)
cmr_land_cover_built_df <- terra::extract(cmr_land_cover_built, map_2011_cmr_sf)
cmr_land_cover_cropland_df <- terra::extract(cmr_land_cover_cropland, map_2011_cmr_sf)
cmr_land_cover_trees_df <- terra::extract(cmr_land_cover_trees, map_2011_cmr_sf)
cmr_land_cover_wetland_df <- terra::extract(cmr_land_cover_wetland, map_2011_cmr_sf)


# distance to water

bbox <- st_bbox(shp_cmr)

rivers <- st_read("HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

rivers <- ms_simplify(rivers, keep = 0.3, keep_shapes = TRUE)
lakes  <- ms_simplify(lakes,  keep = 0.3, keep_shapes = TRUE)

rivers <- st_transform(rivers, st_crs(map_2011_cmr_sf))
lakes  <- st_transform(lakes,  st_crs(map_2011_cmr_sf))

rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

#water_cmr <- st_union(rivers_valid, lakes_valid)
#dist_matrix <- st_distance(map_2011_cmr_sf, water)

shp_cmr_country <- shp_cmr %>%
  mutate(country = "CMR") %>% 
  group_by(country) %>% 
    summarise(geometry=st_union(geometry))

rivers_cmr <- st_intersection(rivers_valid, shp_cmr_country)
lakes_cmr  <- st_intersection(lakes_valid,  shp_cmr_country)


# nearest river segment for each point
nn_riv <- st_nn(map_2011_cmr_sf, rivers_cmr, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(map_2011_cmr_sf, lakes_cmr, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)


map_2011_cmr_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)


cmr_pf_data_2011 <- map_2011_cmr_sf %>% 
  st_drop_geometry()%>% 
  dplyr::select(site_id, lower_age, upper_age, examined, positive, pr, species, dist_to_water) %>% 
  mutate(pop_density = cmr_pop_density_2011_df[,2],
         rainfall = cmr_rainfall_2011_df[,2],
         temperature = cmr_temperature_2011_df[,2],
         elevation = elevation_cmr_df[,2],
         vegetation_index = ndvi_mean_cmr,
         landcover_water = cmr_land_cover_water_df[,2],
         landcover_trees = cmr_land_cover_trees_df[,2],
         landcover_builts = cmr_land_cover_built_df[,2],
         landcover_cropland = cmr_land_cover_cropland_df[,2],
         landcover_wetland = cmr_land_cover_wetland_df[,2]) %>% 
  filter(species == "P. falciparum")


missing_rows <- which(!complete.cases(cmr_pf_data_2011))
cmr_pf_data_2011[missing_rows, ]

# find nearest neighbor for each point
nn <- st_nn(map_2011_cmr_sf, map_2011_cmr_sf, k = 2)   # k=2 because first neighbor is itself

for (var in c("pop_density", "temperature", "elevation")) {
  missing <- is.na(cmr_pf_data_2011[[var]])
  cmr_pf_data_2011[[var]][missing] <- cmr_pf_data_2011[[var]][ sapply(nn[missing], `[`, 2) ]
}

map_2011_cmr_sf <- map_2011_cmr_sf %>% 
  mutate(pop_density = cmr_pf_data_2011$pop_density,
         rainfall = cmr_pf_data_2011$rainfall,
         temperature = cmr_pf_data_2011$temperature,
         elevation = cmr_pf_data_2011$elevation,
         vegetation_index = cmr_pf_data_2011$vegetation_index,
         landcover_water = cmr_pf_data_2011$landcover_water,
         landcover_trees = cmr_pf_data_2011$landcover_trees,
         landcover_builts = cmr_pf_data_2011$landcover_builts,
         landcover_cropland = cmr_pf_data_2011$landcover_cropland,
         landcover_wetland = cmr_pf_data_2011$landcover_wetland) %>% 
  filter(species == "P. falciparum")

```

# Check for multicollinearity

```{r}
num_vars <- cmr_pf_data_2011 %>% 
  dplyr::select(c(
  "rainfall", "temperature", "elevation", "vegetation_index",
  "pop_density", "dist_to_water", "landcover_water", "landcover_trees",
  "landcover_builts", "landcover_cropland", "landcover_wetland"))

M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  positive / examined ~ rainfall + temperature + elevation +
    vegetation_index + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland,
  data = cmr_pf_data_2011
)

vif(fit)

fit1 <- lm(
  positive / examined ~ rainfall + temperature + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland,
  data = cmr_pf_data_2011
)

vif(fit1)
```
We can see that elevation and temperature are highly correlated r = 0.8. So in the first place we will continue with temperature is more directly interpretable for malaria biology. We also notice a strong correlation between land cover (trees) and vegetation index. Here we will also go with vegetation index because it is also one of the covariates than is often use for malaria spatial modelling for it biological interpretability.


# Autocorrelation check
```{r}
# create neighbors based on k nearest neighbors (e.g. k = 4)
coords <- st_coordinates(map_2011_cmr_sf)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

# Moran's I on prevalence
moran.test(map_2011_cmr_sf$pr, lw)
```

# Variogram
```{r}
map_2011_cmr_sf_sp <- as(map_2011_cmr_sf, "Spatial")

vgm_emp <- variogram(pr ~ 1, data = map_2011_cmr_sf_sp)   # prevalence vs distance
plot(vgm_emp)

vgm_mod <- fit.variogram(
  vgm_emp,
  model = vgm("Mat", nugget = 0.025, psill = 0.55, range = 100, kappa = 0.55)
)
vgm_mod

plot(vgm_emp, vgm_mod)
```

```{r}

crs_proj <- 32633  

pts      <- st_transform(map_2011_cmr_sf, crs_proj)
country  <- st_transform(shp_cmr_country, crs_proj)

pts_sp     <- as(pts, "Spatial")
country_sp <- as(country, "Spatial")

mesh <- inla.mesh.2d(
  loc      = coordinates(pts_sp),
  boundary = inla.sp2segment(country_sp),
  max.edge = c(50e3, 200e3),   # 50 km near data, 200 km outer
  cutoff   = 10e3,             # min distance between points (10 km)
  offset   = c(50e3, 150e3)    # extend mesh 50–150 km beyond country
)

# Extract mesh triangles
mesh_tri <- mesh$graph$tv   # triangle vertices (indices)

# Convert each triangle to a polygon
triangles <- lapply(1:nrow(mesh_tri), function(i) {
  coords <- mesh$loc[mesh_tri[i, ], 1:2]
  coords <- rbind(coords, coords[1, ])  # close polygon
  st_polygon(list(coords))
})

mesh_sf <- st_sf(
  geometry = st_sfc(triangles),
  crs = st_crs(pts)   # use same CRS as your points
)


ggplot() +
  geom_sf(data = mesh_sf, fill = NA, color = "grey70", size = 0.2) +
  geom_sf(data = country, fill = NA, color = "black", size = 0.4) +
  geom_sf(data = pts, color = "red", size = 2) +
  theme_minimal() +
  labs(title = "INLA Mesh over Country Boundary")

```

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  alpha = 2,
  prior.range = c(120e3, 0.5),  # P(range < 120 km) = 0.5
  prior.sigma = c(0.22, 0.5)      # weakly informative
)

A_est <- inla.spde.make.A(mesh, loc = coordinates(pts_sp))

pts <- pts %>% mutate(
  rainfall = as.numeric(scale(rainfall)),
  temperature = as.numeric(scale(temperature)),
  dist_to_water = as.numeric(scale(dist_to_water)),
  pop_density = as.numeric(scale(pop_density)),
  vegetation_index = as.numeric(scale(vegetation_index)),
  elevation        = as.numeric(scale(elevation))
)

# pts$rainfall <- scale(pts$rainfall)
# pts$pop_density <- scale(pts$pop_density)
# pts$dist_to_water <- scale(pts$dist_to_water)
# pts$temperature <- scale(pts$temperature)
# pts$vegetation_index <- scale(pts$vegetation_index)
# pts$elevation <- scale(pts$elevation)


stk_est <- inla.stack(
  data = list(
    y       = pts$positive,
    Ntrials = pts$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
    intercept        = 1,
    rainfall         = pts$rainfall,
    temperature      = pts$temperature,
    pop_density      = pts$pop_density,
    dist_to_water    = pts$dist_to_water,
    lc_water         = pts$landcover_water,
    lc_trees         = pts$landcover_trees,
    lc_built         = pts$landcover_builts,
    lc_wetland       = pts$landcover_wetland,
    lc_cropland      = pts$landcover_cropland
    )
  ),
  tag = "est"
)

```

# Build the model

```{r}

formula <- y ~ -1 + intercept +
  rainfall +
  temperature +
  pop_density +
  dist_to_water +
  lc_water +
  lc_cropland+
  lc_wetland +
  lc_built +
  lc_trees  +
  f(spatial, model = spde)


res <- inla(
  formula,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_est),
  control.predictor = list(
    A = inla.stack.A(stk_est),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(
    dic = TRUE,
    waic = TRUE,
    cpo = TRUE
  )
)

res$summary.fixed

res$summary.hyperpar


res$dic$dic
res$waic$waic

spatial_field <- res$summary.random$spatial
head(spatial_field)

```

```{r}
idx_est <- inla.stack.index(stk_est, "est")$data

fitted_est <- res$summary.fitted.values[idx_est, "mean"]

cbind(
  observed = pts$positive / pts$examined,
  fitted   = fitted_est
)[1:30, ]
```


# Build grid for prediction and the prediction model

```{r}
# resolution in meters (5000 m = 5 km)
res <- 5000  

# Create grid covering bounding box of Cameroon
grid <- st_make_grid(
  country,
  cellsize = res,
  what = "centers"
)

grid <- st_sf(geometry = grid)

# Keep only grid points inside Cameroon
grid <- st_intersection(grid, country)


# grid_ITN_usage_2021  <- project(ITN_intervention_2021,  st_crs(grid)$wkt)
grid_pop  <- project(cmr_pop,  st_crs(grid)$wkt)
grid_rainfall_2021 <- project(cmr_rainfall_2011, st_crs(grid)$wkt)
grid_temperature_2021 <- project(cmr_temperature_2011, st_crs(grid)$wkt)
grid_ndvi_stack <- project(cmr_ndvi_stack, st_crs(grid)$wkt)
elevation_grid   <- project(elevation_cmr,   st_crs(grid)$wkt)
grid_land_cover_water   <- project(cmr_land_cover_water,   st_crs(grid)$wkt)
grid_land_cover_built   <- project(cmr_land_cover_built,   st_crs(grid)$wkt)
grid_land_cover_cropland   <- project(cmr_land_cover_cropland,   st_crs(grid)$wkt)
grid_land_cover_trees   <- project(cmr_land_cover_trees,   st_crs(grid)$wkt)
grid_land_cover_wetland   <- project(cmr_land_cover_wetland,   st_crs(grid)$wkt)



# grid_ITN_usage_2021_df <- terra::extract(grid_ITN_usage_2021,  grid)
grid_pop_density_2020_df <- terra::extract(grid_pop,  grid)
grid_rainfall_2021_df    <- terra::extract(grid_rainfall_2021, grid)
grid_temperature_2021_df <- terra::extract(grid_temperature_2021, grid)
ndvi_vals_grid_monthly <- terra::extract(grid_ndvi_stack, grid)
ndvi_mean_grid <- apply(ndvi_vals_grid_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_grid_df <- terra::extract(elevation_grid, grid)
grid_land_cover_water_df <- terra::extract(grid_land_cover_water, grid)
grid_land_cover_built_df <- terra::extract(grid_land_cover_built, grid)
grid_land_cover_cropland_df <- terra::extract(grid_land_cover_cropland, grid)
grid_land_cover_trees_df <- terra::extract(grid_land_cover_trees, grid)
grid_land_cover_wetland_df <- terra::extract(grid_land_cover_wetland, grid)


# nearest river segment for each point
rivers_cmr <- rivers_cmr %>%  st_transform(crs = st_crs(grid))
lakes_cmr <- lakes_cmr %>% st_transform(crs = st_crs(grid))

nn_riv <- st_nn(grid, rivers_cmr, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(grid, lakes_cmr, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)


grid$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

grid <- grid %>% 
  mutate(annual_rainfall_2021    = grid_rainfall_2021_df[,2],
         annual_temperature_2021 = grid_temperature_2021_df[,2],
         pop_density_2020        = grid_pop_density_2020_df[,2],
         elevation               = elevation_grid_df[,2],
         vegetation_index        = ndvi_mean_grid,
         lc_cropland             = grid_land_cover_cropland_df[,2],
         lc_water                = grid_land_cover_water_df[,2],
         lc_wetland              = grid_land_cover_wetland_df[,2],
         lc_built                = grid_land_cover_built_df[,2],
         lc_trees                = grid_land_cover_trees_df[,2]) 

grid_df <- grid %>% 
  st_drop_geometry()

missing_rows <- which(!complete.cases(grid_df))
grid_df[missing_rows, ]

# find nearest neighbor for each point
nn <- st_nn(grid, grid, k = 50)   # k=2 because first neighbor is itself

nn2 <- vapply(nn, function(x) x[2], integer(1))


for (var in c("pop_density_2020", "annual_temperature_2021", "elevation", "vegetation_index")) {
  missing <- is.na(grid_df[[var]])
  grid_df[[var]][missing] <- grid_df[[var]][ nn2[missing] ]
}

grid <- grid %>% 
  mutate(annual_rainfall_2021    = grid_df$annual_rainfall_2021,
         annual_temperature_2021 = grid_df$annual_temperature_2021,
         pop_density_2020        = grid_df$pop_density_2020,
         elevation               = grid_df$elevation,
         vegetation_index        = grid_df$vegetation_index,
         lc_cropland             = grid_df$lc_cropland,
         lc_water                = grid_df$lc_water,
         lc_wetland              = grid_df$lc_wetland,
         lc_built                = grid_df$lc_built,
         lc_trees                = grid_df$lc_trees) 

grid <- grid %>% 
  mutate(annual_rainfall_2021    = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
         pop_density_2020        = as.numeric(scale(pop_density_2020)),
         dist_to_water           = as.numeric(scale(dist_to_water)),
         elevation               = as.numeric(scale(elevation)),
         vegetation_index        = as.numeric(scale(vegetation_index)))

st_write(grid, "cmr_data/cmr_map_grid.gpkg", append = FALSE)


# Build the A‑matrix for prediction locations
A_pred <- inla.spde.make.A(
  mesh = mesh,
  loc = st_coordinates(grid)
)

#Build the prediction stack

stk_pred <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid$annual_rainfall_2021,
      temperature      = grid$annual_temperature_2021,
      pop_density      = grid$pop_density_2020,
      dist_to_water    = grid$dist_to_water,
      lc_cropland      = grid$lc_cropland,
      lc_water         = grid$lc_water,
      lc_wetland       = grid$lc_wetland,
      lc_built         = grid$lc_built,
      lc_trees         = grid$lc_trees
    )
  ),
  tag = "pred"
)


# Combine estimation + prediction stacks
stk_full <- inla.stack(stk_est, stk_pred)


# Fit the model again with the full stack
res_pred <- inla(
  formula,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full),
  control.predictor = list(
    A = inla.stack.A(stk_full),
    compute = TRUE,
    link = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)


# Extract predictions for the grid
index_pred <- inla.stack.index(stk_full, "pred")$data

grid$pred_mean <- res_pred$summary.fitted.values[index_pred, "mean"]
grid$pred_sd   <- res_pred$summary.fitted.values[index_pred, "sd"]
grid$pred_q025 <- res_pred$summary.fitted.values[index_pred, "0.025quant"]
grid$pred_q975 <- res_pred$summary.fitted.values[index_pred, "0.975quant"]

shp_cmr1 <-  st_transform(shp_cmr, st_crs(grid))

grid_cmr1 <- st_join(grid, shp_cmr1, join = st_intersects)

cmr_data_pev <- grid_cmr1 %>%
  group_by(Admin2) %>%
  summarise(
    prev_mean   = mean(pred_mean, na.rm = TRUE),
    prev_q025   = mean(pred_q025, na.rm = TRUE),
    prev_q975   = mean(pred_q975, na.rm = TRUE),
    CI_width    = mean(pred_q975 - pred_q025, na.rm = TRUE))

cmr_data_pev_map <- shp_cmr %>%
  left_join(cmr_data_pev %>% st_drop_geometry(), by = "Admin2")

cmr_prev_1 <- ggplot(cmr_data_pev_map) +
  geom_sf(aes(fill = prev_mean), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Estimated Prevalence")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("cmr_data/MAP_pr_modeled/2011_cmr_pr_test_pf_all_age.png",
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

```


```{r}
# Load the modeled prevalence from MAP

africa_pf_pr_2_10 <- rast("prevalence_2000/2025_GBD2024_Global_PfPR_2011.tif")

# Cameroon case

cmr_bbox <- ext(shp_cmr)

africa_pf_pr_2_10_cmr <- crop(africa_pf_pr_2_10, cmr_bbox)

cmr_pr_raster_map_2011 <- project(
  africa_pf_pr_2_10_cmr,
  st_crs(shp_cmr)$wkt
)

cmr_admin2_pr <- terra::extract(
  cmr_pr_raster_map_2011,
  shp_cmr,
  fun = mean,
  exact = FALSE
)

cmr_map_pr_pf_2_10 <- shp_cmr %>% 
  mutate(pr_pf_2_10 = cmr_admin2_pr[,2],
         pr_pf_2_10_L = cmr_admin2_pr[,4],
         pr_pf_2_10_U = cmr_admin2_pr[,5],
         CI_width = pr_pf_2_10_U - pr_pf_2_10_L)

cmr_map_pr_pf_2_10$PR_all_age <- mapply(
  convertPrevalence,
  prev         = cmr_map_pr_pf_2_10$pr_pf_2_10,
  age_min_in   = rep(2, nrow(cmr_map_pr_pf_2_10)),
  age_max_in   = rep(10, nrow(cmr_map_pr_pf_2_10)),
  age_min_out  = rep(0, nrow(cmr_map_pr_pf_2_10)),
  age_max_out  = rep(99, nrow(cmr_map_pr_pf_2_10)),
  parameters   = "Pf_Smith2007")

cmr_map_pr_pf_2_10$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = cmr_map_pr_pf_2_10$pr_pf_2_10_U,
  age_min_in   = rep(2, nrow(cmr_map_pr_pf_2_10)),
  age_max_in   = rep(10, nrow(cmr_map_pr_pf_2_10)),
  age_min_out  = rep(0, nrow(cmr_map_pr_pf_2_10)),
  age_max_out  = rep(99, nrow(cmr_map_pr_pf_2_10)),
  parameters   = "Pf_Smith2007")

cmr_map_pr_pf_2_10$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = cmr_map_pr_pf_2_10$pr_pf_2_10_L,
  age_min_in   = rep(2, nrow(cmr_map_pr_pf_2_10)),
  age_max_in   = rep(10, nrow(cmr_map_pr_pf_2_10)),
  age_min_out  = rep(0, nrow(cmr_map_pr_pf_2_10)),
  age_max_out  = rep(99, nrow(cmr_map_pr_pf_2_10)),
  parameters   = "Pf_Smith2007")




cmr_prev_pf_MAP_estimates <- ggplot(cmr_map_pr_pf_2_10) +
  geom_sf(aes(fill = PR_all_age), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence of Pf (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("cmr_data/MAP_pr_modeled/2011_cmr_pr_MAP_pf_all_age.pdf",
       plot = cmr_prev_pf_MAP_estimates, width = 6, height = 4, dpi = 300)

cmr_prev_pf_MAP_estimates_uncertainty <- ggplot(cmr_map_pr_pf_2_10) +
  geom_sf(aes(fill = CI_width), color = NA) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Prevalence estimates uncertainty of P.f (2-10)")+
  theme_bw(base_size = 15)+
  theme(axis.text.x = element_text)+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

ggsave("cmr_data/MAP_pr_modeled/cmr_pr_uncertainty_MAP_pf_2_10.png",
       plot = cmr_prev_pf_MAP_estimates_uncertainty, width = 6, height = 4, dpi = 300)


```



# Extract MAP data

```{r}
#cor(cmr_data_pev_map$prev_mean, cmr_map_pr_pf_2_10$PR_all_age, use = "complete.obs")

cmr_data_pev_map <- cmr_data_pev_map %>% 
  mutate(prev_mean = case_when(is.na(prev_mean) ~ mean(prev_mean, na.rm = TRUE),
                               TRUE ~ prev_mean))

cmr_data_pev_map$PR_all_age <- cmr_map_pr_pf_2_10$PR_all_age
#-----------------------------------------------------------
# 1. BASIC AGREEMENT METRICS
#-----------------------------------------------------------

# Bias
bias <- mean(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age, na.rm = TRUE)

# MAE
mae <- mean(abs(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age), na.rm = TRUE)

# RMSE
rmse <- sqrt(mean((cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age)^2, na.rm = TRUE))

# Pearson correlation
pearson_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "pearson")

# Spearman rank correlation
spearman_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "spearman")

# #-----------------------------------------------------------
# # 2. SPATIAL AGREEMENT METRICS
# #-----------------------------------------------------------
# 
# # Create spatial weights
# nb <- poly2nb(cmr_data_pev_map)
# lw <- nb2listw(nb, style = "W")
# 
# # Spatial cross-correlation (correlation of lagged values)
# lag_model <- lag.listw(lw, cmr_data_pev_map$prev_mean)
# lag_map   <- lag.listw(lw, cmr_data_pev_map$PR_all_age)
# spatial_cross_cor <- cor(lag_model, lag_map, use = "complete.obs")
# 
# # Moran's I of residuals
# residuals <- cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age
# moran_residuals <- moran.test(residuals, lw)
# 
# #-----------------------------------------------------------
# # 3. SEMIVARIOGRAM COMPARISON
# #-----------------------------------------------------------
# 
# library(gstat)
# 
# # Convert sf to sp for gstat
# admin2_sp <- as(cmr_data_pev_map, "Spatial")
# 
# vg_model <- variogram(prev_mean ~ 1, admin2_sp)
# vg_map   <- variogram(PR_all_age ~ 1, admin2_sp)
# 
# #-----------------------------------------------------------
# # 4. DISTRIBUTIONAL COMPARISON
# #-----------------------------------------------------------
# 
# # Q–Q plot data
# qq_data <- data.frame(
#   model = quantile(cmr_data_pev_map$prev_mean, probs = seq(0,1,0.01), na.rm = TRUE),
#   map   = quantile(cmr_data_pev_map$PR_all_age, probs = seq(0,1,0.01), na.rm = TRUE)
# )
# 
# # Distribution overlap (simple measure)
# dens_model <- density(cmr_data_pev_map$prev_mean, na.rm = TRUE)
# dens_map   <- density(cmr_data_pev_map$PR_all_age, na.rm = TRUE)
# 
# # Overlap integral
# x_common <- seq(min(dens_model$x, dens_map$x), max(dens_model$x, dens_map$x), length.out = 1000)
# y_model  <- approx(dens_model$x, dens_model$y, x_common)$y
# y_map    <- approx(dens_map$x, dens_map$y, x_common)$y
# distribution_overlap <- sum(pmin(y_model, y_map)) * (x_common[2] - x_common[1])
# 
# #-----------------------------------------------------------
# # 5. CLASSIFICATION AGREEMENT (TRANSMISSION STRATA)
# #-----------------------------------------------------------
# 
# # Define strata
# cut_points <- c(0, 0.05, 0.30, 1)
# labels <- c("low", "medium", "high")
# 
# cmr_data_pev_map <- cmr_data_pev_map %>%
#   mutate(
#     class_model = cut(prev_mean, breaks = cut_points, labels = labels, include.lowest = TRUE),
#     class_map   = cut(PR_all_age, breaks = cut_points, labels = labels, include.lowest = TRUE)
#   )
# 
# # Confusion matrix
# confusion_matrix <- table(cmr_data_pev_map$class_model, cmr_data_pev_map$class_map)
# 
# # Kappa statistic
# library(irr)
# kappa_stat <- kappam.fleiss(as.data.frame(confusion_matrix))
# 
# #-----------------------------------------------------------
# # 6. ADMIN2 AGGREGATED COMPARISONS
# #-----------------------------------------------------------
# 
# # Admin2 mean prevalence (already aggregated)
# admin2_comparison <- cmr_data_pev_map %>%
#   st_drop_geometry() %>%
#   select(admin2, pfpr_model, pfpr_map_all_age)
# 
# # Rank correlation of districts
# rank_cor <- cor(
#   rank(cmr_data_pev_map$pfpr_model),
#   rank(cmr_data_pev_map$pfpr_map_all_age),
#   method = "spearman"
# )
# 
# #-----------------------------------------------------------
# # 7. POPULATION-WEIGHTED PREVALENCE (if pop column exists)
# #-----------------------------------------------------------
# 
# if("population" %in% colnames(cmr_data_pev_map)) {
#   pwpr_model <- sum(cmr_data_pev_map$pfpr_model * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
#   
#   pwpr_map <- sum(cmr_data_pev_map$pfpr_map_all_age * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
# }
# 
# #-----------------------------------------------------------
# # 8. UNCERTAINTY COMPARISON (if your model has CI)
# #-----------------------------------------------------------
# 
# if("model_lower" %in% colnames(cmr_data_pev_map) &&
#    "model_upper" %in% colnames(cmr_data_pev_map)) {
#   
#   coverage_probability <- mean(
#     cmr_data_pev_map$pfpr_map_all_age >= cmr_data_pev_map$model_lower &
#       cmr_data_pev_map$pfpr_map_all_age <= cmr_data_pev_map$model_upper,
#     na.rm = TRUE
#   )
# }
# 
# #-----------------------------------------------------------
# # PRINT RESULTS
# #-----------------------------------------------------------
# 
# list(
#   bias = bias,
#   mae = mae,
#   rmse = rmse,
#   pearson_cor = pearson_cor,
#   spearman_cor = spearman_cor,
#   spatial_cross_cor = spatial_cross_cor,
#   moran_residuals = moran_residuals,
#   semivariogram_model = vg_model,
#   semivariogram_map = vg_map,
#   distribution_overlap = distribution_overlap,
#   confusion_matrix = confusion_matrix,
#   kappa_stat = kappa_stat,
#   admin2_comparison = admin2_comparison,
#   rank_cor = rank_cor
# )
# 

```
Overall, our model tends to overestimate malaria prevalence compared to MAP (bias ≈ 0.064), with typical differences of about 9–12 percentage points (MAE and RMSE). The moderate Pearson correlation (~0.55) shows that our model captures only part of MAP’s spatial variation, while the slightly higher Spearman correlation (~0.60) indicates that it does a somewhat better job ranking districts from low to high transmission than matching exact values. Taken together, these results suggest that oour model broadly identifies the same high‑burden and low‑burden areas as MAP but differs noticeably in magnitude, with some districts showing larger discrepancies. This is a solid foundation for a first geospatial model, but it also highlights clear opportunities to refine covariates, spatial structure, or calibration to better align with MAP’s established patterns.

```{r}

```


```{r}
# cmr_map_data <- read.csv("cmr_data/cmr_map_PR.csv")
# 
# cmr_map_data_clean <- cmr_map_data %>%
#   filter(!is.na(longitude) & !is.na(latitude)) %>% 
#   filter(species=="P. falciparum") 
# 
# # Convert to sf spatial object
# cmr_map_sf <- st_as_sf(
#   cmr_map_data_clean,
#   coords = c("longitude", "latitude"),
#   crs    = 4326
# )

# Check
# plot(st_geometry(cmr_map_sf), main = "MAP Survey Points")
# 
# shp_cmr <- st_transform(shp_cmr, crs=4326)
# 
# plot(st_geometry(shp_cmr))
# plot(st_geometry(cmr_map_sf), col = "red", pch = 16, cex = 0.5, add = TRUE)
# 
# ggplot()+
#   geom_sf(data = shp_cmr, fill=NA, color="black")+
#   geom_sf(data=cmr_map_sf, color="red", size=0.2)


# cmr_map_with_district <- st_join(
#   cmr_map_sf,
#   shp_cmr[, c("Admin1", "Admin2")],
#   join = st_within
# )
# 
# cmr_points_per_district <- cmr_map_with_district %>%
#   st_drop_geometry() %>%
#   group_by(Admin2) %>%
#   summarise(n_surveys = n()) %>%
#   arrange(desc(n_surveys))
# 
# #print(cmr_points_per_district)
# 
# 
# cmr_district_prevalence <- cmr_map_with_district %>%
#   st_drop_geometry() %>%
#   filter(!is.na(Admin2)) %>%       
#   group_by(Admin1, Admin2) %>%
#   summarise(n_examined     = sum(examined,  na.rm = TRUE),
#             n_positive     = sum(positive,  na.rm = TRUE),
#             raw_prevalence = n_positive / n_examined,
#             raw_pr         = sum(pr),
#             n_surveys      = n(),
#             year_min       = min(year_start, na.rm = TRUE),
#             year_max       = max(year_end,   na.rm = TRUE),
#             .groups = "drop")
# 
# 
# #2011 data
# 
# cmr_map_2011 <- cmr_map_data_clean %>%
#   filter(year_start == 2011 | year_end==2011) %>% 
#   filter(species=="P. falciparum")
# 
# 
# cmr_map_2011_sf <- st_as_sf(
#   cmr_map_2011,
#   coords = c("longitude", "latitude"),
#   crs    = 4326
# )
# 
# 
# cmr_map_with_district_2011 <- st_join(
#   cmr_map_2011_sf,
#   shp_cmr[, c("Admin1", "Admin2")],
#   join = st_within
# )
# 
# 
# cmr_observed_districts_2011 <- cmr_map_with_district_2011 %>%
#   st_drop_geometry() %>%
#   group_by(Admin1, Admin2) %>%
#   summarise(
#     n_points   = n(),
#     n_examined = sum(examined, na.rm = TRUE),
#     n_positive = sum(positive, na.rm = TRUE),
#     .groups    = "drop"
#   ) %>%
#   mutate(raw_prevalence = n_positive / n_examined)
# 
# ggplot()+
#   geom_sf(data = shp_cmr, fill=NA, color="black")+
#   geom_sf(data = shp_cmr[shp_cmr$Admin2 %in% cmr_observed_districts_2011$Admin2, ], color="red")+
#   geom_sf(data = cmr_map_2011_sf, color = "blue", size=2)
# 
# 
# ### extract the observed districts as well as their neighbors
# observed_names <- cmr_observed_districts_2011$Admin2
# observed_districts_sf <- shp_cmr %>%
#   filter(str_squish(Admin2) %in% observed_names)
# 
# nb_queen <- poly2nb(shp_cmr, queen = TRUE)
# # Get row indices of observed districts in the full shape file
# observed_indices <- which(
#   str_squish(shp_cmr$Admin2) %in% observed_names
# )
# 
# 
# # Get neighbor indices for each observed district
# neighbor_indices <- list()
# 
# for(i in observed_indices){
#   neighbors_i <- nb_queen[[i]]
#   cat("District:", shp_cmr$Admin2[i], 
#       "Number of neighbors:", length(neighbors_i), "\n")
#   neighbor_indices[[i]] <- neighbors_i
# }
# 
# # Combine all neighbor indices into one vector
# all_neighbor_indices <- unique(unlist(neighbor_indices))
# # Remove observed districts from neighbor list
# # (in case an observed district is neighbor of another observed district)
# pure_neighbor_indices <- setdiff(all_neighbor_indices, observed_indices)
# # Combine observed + neighbor indices
# study_area_indices <- unique(c(observed_indices, pure_neighbor_indices))
# 
# # Extract study area from full shapefile
# study_area_sf <- shp_cmr[study_area_indices, ] %>%
#   mutate(district_type = ifelse(str_squish(Admin2) %in% observed_names,
#       "Observed", "Neighbor"))
# 
# # Visualization of our study area
# 
# ggplot() +
#   # Full country in background
#   geom_sf(data = shp_cmr, fill = "white", color = "grey80") +
#   # Study area
#   geom_sf(
#     data = study_area_sf,
#     aes(fill = district_type),
#     color = "grey40"
#   ) +
#   scale_fill_manual(
#     values = c("Observed" = "#d73027", "Neighbor" = "#fee090"),
#     name   = "District type"
#   ) +
#   labs(
#     title    = "Study Area: Observed and Neighbor Districts",
#     subtitle = "Cameroon 2011: Queen Contiguity"
#   ) +
#   theme_minimal()+
#   theme(
#     axis.text  = element_blank(),
#     axis.ticks = element_blank()
#   )
# 
# # Merge aggregated prevalence into study area
# study_area_sf <- study_area_sf %>%
#   left_join(
#     cmr_observed_districts_2011,
#     by = c("Admin2")
#   ) %>%
#   mutate(
#     # Neighbor districts have NA for observed data — correct
#     n_examined = ifelse(is.na(n_examined), NA, n_examined),
#     n_positive = ifelse(is.na(n_positive), NA, n_positive)
#   )


```



```{r}
## Building the adjacency matrix for our model

study_area_sf <- study_area_sf %>%
  mutate(district_id = 1:nrow(.)) 

nb_study <- poly2nb(study_area_sf, queen = TRUE)

# # Summary
# summary(nb_study)
# 
# nb_check <- nb2listw(nb_study, style = "B", zero.policy = TRUE)
# 
# 
components <- n.comp.nb(nb_study)
cat("Number of connected components:", components$nc, "\n")

# Which districts belong to which component
study_area_sf$component <- components$comp.id
print(study_area_sf[, c("Admin2", "district_type", "component")])

nb2INLA("study_area_adj.graph", nb_study)

file.exists("study_area_adj.graph")


### Visualize the study area

coords <- st_coordinates(st_centroid(study_area_sf))
nb_lines <- map2_dfr(
  .x = seq_along(nb_study),
  .y = nb_study,
  .f = function(i, neighbors) {
    if (length(neighbors) == 0) return(NULL)

    map_dfr(neighbors, function(j) {
      st_linestring(rbind(coords[i, ], coords[j, ])) |> 
        st_sfc(crs = st_crs(study_area_sf)) |> 
        st_sf(from = i, to = j, geometry = _)
    })
  }
)


ggplot() +
  # District polygons
  geom_sf(
    data = study_area_sf,
    aes(fill = district_type),
    color = "black",
    size = 0.3
  ) +
  
  # Neighbor links
  geom_sf(
    data = nb_lines,
    color = "blue",
    size = 0.7
  ) +
  
  # District labels
  geom_text(
    data = cbind(study_area_sf, coords),
    aes(X, Y, label = Admin2),
    size = 3
  ) +
  
  scale_fill_manual(
    values = c("Observed" = "#d73027", "Neighbor" = "#fee090")
  ) +
  
  labs(
    title = "Queen Contiguity Neighborhood Structure",
    fill = "District Type"
  ) +
  
  theme_minimal()+
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

study_area_sf <- study_area_sf %>%
  mutate(
    component_factor = as.factor(component)
  )

```


```{r}
#building our model

model_data <- study_area_sf %>%
  st_drop_geometry() %>%
  dplyr::select(district_id, Admin2, district_type, 
                component, n_examined, n_positive) %>%
  mutate(
    # R-INLA needs numeric outcome
    n_examined = as.integer(n_examined),
    n_positive = as.integer(n_positive),
    
    # District ID for spatial random effect
    district_id_unstr = district_id  # separate copy for unstructured effect
  )

# Define formula
# No covariates — only intercept + spatial + unstructured random effects
formula_null <- n_positive ~ 1 +
  f(district_id,
    model         = "besag",        # spatially structured ICAR
    graph         = "study_area_adj.graph",
    scale.model   = TRUE,
    constr        = TRUE,
    hyper         = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  f(district_id_unstr,
    model = "iid",                  # unstructured random effect
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Fit the model
model_null <- inla(
  formula_null,
  family          = "binomial",
  Ntrials         = model_data$n_examined,
  data            = model_data,
  control.predictor = list(
    compute = TRUE,
    link    = 1          # use the link function for predictions
  ),
  control.compute = list(
    dic  = TRUE,
    waic = TRUE,
    cpo  = TRUE
  )
)

# Summary
summary(model_null)


# Extract posterior mean and 95% credible intervals
model_data$prev_mean  <- model_null$summary.fitted.values$mean
model_data$prev_lower <- model_null$summary.fitted.values$`0.025quant`
model_data$prev_upper <- model_null$summary.fitted.values$`0.975quant`

# Check results
model_data %>%
  dplyr::select(Admin2, district_type, 
                n_examined, n_positive,
                prev_mean, prev_lower, prev_upper) %>%
  mutate(across(c(prev_mean, prev_lower, prev_upper), 
                ~ round(., 3)))

study_area_results <- study_area_sf %>%
  left_join(
    model_data %>% dplyr::select(district_id, prev_mean, 
                                  prev_lower, prev_upper),
    by = "district_id"
  )

ggplot() +
  geom_sf(data = shp_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data = study_area_results,
    aes(fill = prev_mean),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "Prevalence",
    option = "magma",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Estimated Malaria Prevalence — Null Spatial Model",
    subtitle = "BYM2 without covariates — Cameroon 2011"
  ) +
  theme_minimal()

# Map uncertainty (width of credible interval)
study_area_results <- study_area_results %>%
  mutate(ci_width = prev_upper - prev_lower)

ggplot() +
  geom_sf(data = shp_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data = study_area_results,
    aes(fill = ci_width),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "CI Width",
    option = "rocket",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Model Uncertainty — Null Spatial Model",
    subtitle = "Width of 95% Credible Interval"
  ) +
  theme_minimal()

```


```{r}
# merging the extracted covariates with the study areas data

# Extract district names from your study area
study_districts <- study_area_sf %>%
  st_drop_geometry() %>%
  dplyr::select(district_id, Admin2, 
                district_type, component,
                n_examined, n_positive)

# loading the covariates data

rainfall_cmr_2011 <- read.csv("cmr_data/cmr_rainfall_2011.csv")
temperature_cmr_2011 <- read.csv("cmr_data/cmr_temperature_2011.csv")
vegetation_index_cmr_2011 <- read.csv("cmr_data/cmr_vegetation_index_2011.csv")
population_density_cmr_2011 <- read.csv("cmr_data/cmr_pop_density_2011.csv")
elevation_data_cmr <- read.csv("cmr_data/cmr_elevation_data.csv")
land_cover_data_cmr <- read.csv("cmr_data/cmr_landcover_per_district.csv")
distance_to_water_cmr <- read.csv("cmr_data/cmr_distance_to_water.csv")


# Filter each CSV to keep only study area districts
# then merge all together

model_data <- study_districts %>%
  
  # Rainfall
  left_join(
    rainfall_cmr_2011 %>% dplyr::select(District, mean_rainfall),
    by = c("Admin2" = "District")
  ) %>%
  
  # Temperature
  left_join(
    temperature_cmr_2011 %>% dplyr::select(District, mean_temperature),
    by = c("Admin2" = "District")
  ) %>%
  
  # NDVI
  left_join(
    vegetation_index_cmr_2011 %>% dplyr::select(District, NDVI),
    by = c("Admin2" = "District")
  ) %>%
  
  # Elevation
  left_join(
    elevation_data_cmr %>% dplyr::select(District, mean_elevation),
    by = c("Admin2" = "District")
  ) %>%
  
  # Population density
  left_join(
    population_density_cmr_2011 %>% dplyr::select(District, pop_density_2011),
    by = c("Admin2" = "District")
  ) %>%
  
  # Distance to water
  left_join(
    distance_to_water_cmr %>% dplyr::select(District, dist_to_water_km),
    by = c("Admin2" = "District")
  ) %>%
  
  # Land cover
  left_join(
    land_cover_data_cmr %>% dplyr::select(District, prop_trees, prop_water,
                                 prop_wetland, prop_cropland, prop_built),
    by = c("Admin2" = "District")
  )

#save the data
write_csv(model_data, "cmr_data/cmr_study_area_model_data_2011.csv")

covariate_cols <- c("mean_rainfall", "mean_temperature",
                    "NDVI", "mean_elevation", "pop_density_2011",
                    "dist_to_water_km", "prop_trees", "prop_water",
                    "prop_wetland", "prop_cropland", "prop_built")

# Standardize to mean 0 and SD 1
model_data <- model_data %>%
  mutate(across(
    all_of(covariate_cols),
    ~ scale(.) %>% as.vector(),
    .names = "{.col}_std"
  ))

# Add unstructured random effect ID
model_data <- model_data %>%
  mutate(district_id_unstr = district_id)

# Check standardized values
model_data %>%
  dplyr::select(Admin2, ends_with("_std")) %>%
  head()


cov_matrix <- model_data %>%
  dplyr::select(all_of(covariate_cols)) %>%
  cor(use = "complete.obs")

print(round(cov_matrix, 2))

corrplot(cov_matrix,
         method      = "color",
         type        = "upper",
         tl.cex      = 0.7,
         addCoef.col = "black",
         number.cex  = 0.6)


final_covariates <- c("mean_rainfall", "mean_temperature",
                      "NDVI", "dist_to_water_km",
                      "prop_wetland", "prop_cropland", "prop_built")

# Check correlations again
final_cor <- model_data %>%
  dplyr::select(all_of(final_covariates)) %>%
  cor(use = "complete.obs")

corrplot(final_cor,
         method      = "color",
         type        = "upper",
         tl.cex      = 0.7,
         addCoef.col = "black",
         number.cex  = 0.6)

```


```{r}
#run the model with covariates

# Add unstructured ID
model_data <- model_data %>%
  mutate(district_id_unstr = district_id)

# Define formula with all 7 covariates
formula_full <- n_positive ~ 1 +
  # Fixed effects — standardized covariates
  mean_rainfall_std +
  mean_temperature_std +
  NDVI_std +
  dist_to_water_km_std +
  prop_wetland_std +
  prop_cropland_std +
  prop_built_std +
  # Spatially structured random effect
  f(district_id,
    model       = "besag",
    graph       = "study_area_adj.graph",
    scale.model = TRUE,
    constr      = TRUE,
    hyper       = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  # Unstructured random effect
  f(district_id_unstr,
    model = "iid",
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Fit the model
model_full <- inla(
  formula_full,
  family            = "binomial",
  Ntrials           = model_data$n_examined,
  data              = model_data,
  control.predictor = list(
    compute = TRUE,
    link    = 1
  ),
  control.compute   = list(
    dic  = TRUE,
    waic = TRUE,
    cpo  = TRUE
  )
)

# Summary
summary(model_full)


# Use informative priors based on malaria literature
# This prevents the model from wandering into implausible territory

formula_full_informed <- n_positive ~ 1 +
  mean_rainfall_std +
  mean_temperature_std +
  NDVI_std +
  dist_to_water_km_std +
  prop_wetland_std +
  prop_cropland_std +
  prop_built_std +
  f(district_id,
    model       = "besag",
    graph       = "study_area_adj.graph",
    scale.model = TRUE,
    constr      = TRUE,
    hyper       = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  f(district_id_unstr,
    model = "iid",
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Tighter priors on fixed effects
model_full_informed <- inla(
  formula_full_informed,
  family            = "binomial",
  Ntrials           = model_data$n_examined,
  data              = model_data,
  control.fixed     = list(
    mean.intercept = 0,
    prec.intercept = 1,     # tighter prior on intercept
    mean           = 0,
    prec           = 1      # tighter prior on all coefficients
                            # equivalent to SD = 1 on logit scale
  ),
  control.predictor = list(compute = TRUE, link = 1),
  control.compute   = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_full_informed)

```


```{r}
#extract the result of the full informed model

# Extract fitted values
model_data$prev_mean  <- model_full_informed$summary.fitted.values$mean
model_data$prev_lower <- model_full_informed$summary.fitted.values$`0.025quant`
model_data$prev_upper <- model_full_informed$summary.fitted.values$`0.975quant`
model_data$ci_width   <- model_data$prev_upper - model_data$prev_lower

# Check observed vs fitted for observed districts
model_data %>%
  filter(district_type == "Observed") %>%
  mutate(
    raw_prev    = round(n_positive / n_examined, 3),
    fitted_prev = round(prev_mean, 3),
    lower       = round(prev_lower, 3),
    upper       = round(prev_upper, 3)
  ) %>%
  dplyr::select(Admin2, raw_prev, fitted_prev, lower, upper)

model_data %>%
  filter(district_type == "Neighbor") %>%
  mutate(
    fitted_prev = round(prev_mean,  3),
    lower       = round(prev_lower, 3),
    upper       = round(prev_upper, 3),
    ci_width    = round(ci_width,   3)
  ) %>%
  dplyr::select(Admin2, component, fitted_prev,
                lower, upper, ci_width) %>%
  arrange(component, desc(fitted_prev))

# map the result

# Merge results to spatial object
study_area_results <- study_area_sf %>%
  left_join(
    model_data %>% dplyr::select(district_id, prev_mean,
                                  prev_lower, prev_upper, ci_width),
    by = "district_id"
  )

# Map 1 — Posterior mean prevalence
ggplot() +
  geom_sf(data = districts_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data  = study_area_results,
    aes(fill = prev_mean),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "Prevalence",
    option = "magma",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Estimated Malaria Prevalence",
    subtitle = "BYM2 with informative priors — Cameroon 2011"
  ) +
  theme_minimal()

# Map 2 — Uncertainty
ggplot() +
  geom_sf(data = districts_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data  = study_area_results,
    aes(fill = ci_width),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "CI Width",
    option = "rocket",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Model Uncertainty",
    subtitle = "Width of 95% Credible Interval"
  ) +
  theme_minimal()

```

# Extract the modelled prevalence from MAP

```{r}
available_rasters <- listRaster()
print(available_rasters)

available_rasters[grep("Pf|falciparum|PR", 
                        available_rasters$dataset_id,
                        ignore.case = TRUE), ]

# Get Cameroon shapefile
cmr_shp <- getShp(ISO = "CMR", admin_level = "admin0")

# Download MAP PfPR surface using correct dataset_id
map_raster <- getRaster(
  dataset_id = "Malaria__202206_Global_Pf_Parasite_Rate",
  shp        = cmr_shp,
  year       = 2011
)


plot(map_raster, main = "MAP PfPR2-10 — Cameroon 2011")

# Extract MAP prevalence per study area district
map_extracted <- extract(
  map_raster,
  vect(study_area_sf),
  fun   = "mean",
  na.rm = TRUE
)

model_data$map_prev <- map_extracted[, 2]

# Compare
comparison <- model_data %>%
  dplyr::select(Admin2, district_type,
                prev_mean, prev_lower, prev_upper,
                map_prev) %>%
  mutate(
    prev_mean  = round(prev_mean, 3),
    map_prev   = round(map_prev,  3),
    difference = round(prev_mean - map_prev, 3),
    map_within_ci = map_prev >= prev_lower &
                    map_prev <= prev_upper
  )

```
