---
title: "cmrchmarking exercise"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#load necessary libraries
library(malariaAtlas)
library(sf)
library(leaflet)
library(mapview)
library(htmlwidgets)
library(webshot2)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(spData)
library(gganimate)
library(viridis)
library(GGally)
library(spdep)
library(INLA)
library(SpatialEpi)
library(geodata)
library(tidygeocoder)
library(nngeo)
library(car)
library(gstat)
library(raster)
library(FNN)
library(rmapshaper)
library(tidyterra)
library(fmesher)
```


```{r}
## Function: fill NA values in one column using nearest non‑NA neighbor
fill_by_nearest <- function(sf_obj, colname) {
  
  # rows with non‑missing values
  non_na <- sf_obj %>% filter(!is.na(.data[[colname]]))
  
  # rows with missing values
  na_rows <- sf_obj %>% filter(is.na(.data[[colname]]))
  
  # if no NA: return unchanged
  if (nrow(na_rows) == 0) return(sf_obj)
  
  # find nearest non‑NA feature for each NA row
  nearest_id <- st_nearest_feature(na_rows, non_na)
  
  # extract the corresponding values
  replacement_values <- non_na[[colname]][nearest_id]
  
  # fill the NA values
  sf_obj[[colname]][is.na(sf_obj[[colname]])] <- replacement_values
  
  sf_obj
}
```


```{r}
# Downloading the malaria dataset
# library(malariaAtlas)
# d <- getPR(country = "Cameroon", species = "BOTH")
# ggplot2::autoplot(d)
# 
# write_csv(d, "cmr_data/cmr_map_PR.csv")

```


```{r}
GPS_cmr_2011 <- st_read("data/cmr_data/GPS location 2011/CMGE61FL.shp")
GPS_cmr_2022 <- st_read("data/cmr_data/GPS location 2022/CMGE81FL.shp")


shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")

cmr_vect <- vect(shp_cmr)

```


```{r}
GPS_cmr_coords_2011 <- GPS_cmr_2011 %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_cmr_2011)[, 1],
    latitude  = st_coordinates(GPS_cmr_2011)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  filter(latitude != 0 & longitude != 0)

GPS_cmr_coords_2022 <- GPS_cmr_2022 %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_cmr_2022)[, 1],
    latitude  = st_coordinates(GPS_cmr_2022)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  filter(latitude != 0 & longitude != 0)
```

# loading the extracted MAP data

```{r}
raw_prevalence_cmr <- read.csv("data/cmr_data/cmr_map_PR.csv")

raw_prevalence_cmr_with_coord <- raw_prevalence_cmr %>% 
  filter(!is.na(longitude) & !is.na(latitude))

raw_prevalence_cmr_without_coord <- raw_prevalence_cmr %>% 
  filter(is.na(longitude) & is.na(latitude)) %>% 
  dplyr::select(-c(longitude, latitude))


map_without_coords_cmr <- raw_prevalence_cmr_without_coord %>%
  mutate(cluster_num = as.integer(substr(dhs_id, 9, 16)))

map_recovered_cmr_2011 <- map_without_coords_cmr %>%
  left_join(GPS_cmr_coords_2011, by = c("cluster_num" = "DHSCLUST"))


map_recovered_cmr_2022 <- map_without_coords_cmr %>%
  left_join(GPS_cmr_coords_2022, by = c("cluster_num" = "DHSCLUST"))

cat("Recovered:", sum(!is.na(map_recovered_cmr_2011$longitude)), "\n")
cat("Still missing:", sum(is.na(map_recovered_cmr_2011$longitude)), "\n")
```



```{r}
map_full_cmr_2011 <- bind_rows(
  raw_prevalence_cmr_with_coord %>% filter(!is.na(longitude) & !is.na(latitude)),
  map_recovered_cmr_2011 %>% filter(!is.na(longitude) & !is.na(latitude))
)

map_full_cmr_2011 <- map_full_cmr_2011 %>%
  filter(year_start == 2011) %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  filter(species == "P. falciparum")


write_csv(map_full_cmr_2011, file = "data/cmr_data/cmr_MAP_full_raw_prevalence_2011.csv")

map_full_cmr_2022 <- bind_rows(
  raw_prevalence_cmr_with_coord %>% filter(!is.na(longitude) & !is.na(latitude)),
  map_recovered_cmr_2022 %>% filter(!is.na(longitude) & !is.na(latitude))
)

map_full_cmr_2022 <- map_full_cmr_2022 %>%
  filter(year_start == 2022) %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  filter(species == "P. falciparum")


write_csv(map_full_cmr_2022, file = "data/cmr_data/cmr_MAP_full_raw_prevalence_2022.csv")
```

# Filter for the year 2011 for our analysis

```{r}
map_2011_cmr <- st_read("data/cmr_data/cmr_MAP_full_raw_prevalence_2011.csv")

map_2011_cmr_sf <- st_as_sf(map_2011_cmr, coords = c("longitude", "latitude"), crs = 4326)


breaks_pr <- c(0, 0.05, 0.10, 0.20, 0.40, 0.60, 1)
labels_pr <- c(
  "0–5%", "5–10%", "10–20%", "20–40%", "40–60%", ">60%"
)

colors_pr <- c(
  "#ffffcc",  # pale yellow (very low)
  "#ffeda0",  # light yellow
  "#feb24c",  # orange
  "#fd8d3c",  # deep orange
  "#f03b20",  # red
  "#bd0026"   # dark red (very high)
)

map_2011_cmr_sf$pr <- as.numeric(map_2011_cmr_sf$pr)

map_2011_cmr_sf$pr_class <- cut(
  map_2011_cmr_sf$pr,
  breaks = breaks_pr,
  labels = labels_pr,
  include.lowest = TRUE
)

mapviewOptions(viewer.suppress = T)
map <- mapview(shp_cmr,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
  mapview(map_2011_cmr_sf,
          zcol = "pr_class",             # categorical variable
          cex = "examined",
          col.regions = colors_pr,       # one color per class
          layer.name = "P. malariae prevalence",
          legend = TRUE)


# Save widget manually
htmlwidgets::saveWidget(map@map, "Outputs/cmr_output/raw_prevalence_map.html", selfcontained = FALSE)

# Capture screenshot with webshot2
webshot2::webshot("Outputs/cmr_output/raw_prevalence_map.html",
                  "Outputs/cmr_output/raw_prevalence_map.png", vwidth = 1200, vheight = 900)
# mapshot2(map, file = "Outputs/cmr_output/raw_prevalence_map.png")
# mapshot2(map, "Outputs/cmr_output/raw_prevalence_map.html", selfcontained = TRUE)


plot(st_geometry(map_2011_cmr_sf), pch = 16, cex = 0.5, main = "2011 Survey Points — Cameroon")
```

# Extraction of the covariates  
```{r}
path1 <- "data/cmr_data/chirps_2011/CHIRPS_PENTAD_precipitation_2011-01-01_2011-12-31_-5p8264to13p9448N_0p6930to19p1696E.precipitation.tif"
path2 <- "data/cmr_data/ERA5_temperature/ERA5_LAND_DAILY_temperature_2m_2011-01-01_2011-12-31_-5p4765to13p9448N_0p6930to19p3454E.temperature_2m.tif"
ndvi_path <- "data/cmr_data/NDVI2011/"
# Load all 12 monthly NDVI tif files
ndvi_files <- list.files(
  ndvi_path,
  pattern    = "_1_km_monthly_NDVI.*\\.tif$",
  full.names = TRUE
)

cmr_ndvi_stack<- rast(ndvi_files)


cmr_pop <- rast("data/cmr_data/cmr_pd_2011_1km_UNadj.tif")
cmr_rainfall_2011 <- rast(path1)
cmr_temperature_2011 <- rast(path2)
elevation_cmr <- rast("data/cmr_data/elevation/cmr_elv_msk.tif")
land_cover_water <- rast("data/ben_data/landuse/WorldCover_water_30s.tif")
land_cover_trees <- rast("data/ben_data/landuse/WorldCover_trees_30s.tif")
land_cover_builts <- rast("data/ben_data/landuse/WorldCover_built_30s.tif")
land_cover_cropland <- rast("data/ben_data/landuse/WorldCover_cropland_30s.tif")
land_cover_wetland <- rast("data/ben_data/landuse/WorldCover_wetland_30s.tif")
cmr_ITN_2011  <- rast("data/ITN_2000/2025_GBD2024_Africa_ITN_2011.tif")
cmr_ITN_2011  <- rast("data/ITN_2000_2/2026_GBD2025_Africa_ITN_Use_Rate_2011.tif")
cmr_IRS_2011  <- rast("data/IRS_2000/2025_GBD2024_Africa_IRS_2011.tif")
cmr_effective_treat_2011 <- rast("data/Effective_treat_2000_2025/2026_GBD2025_Global_Antimalarial_EFT_2011.tif")

################################################################################
#            Extraction of the covariates at the sampled location              #
################################################################################

cmr_pop  <- crop(cmr_pop,  cmr_vect)
cmr_pop  <- mask(cmr_pop,  cmr_vect)

cmr_rainfall_2011 <- crop(cmr_rainfall_2011, cmr_vect)
cmr_rainfall_2011 <- mask(cmr_rainfall_2011, cmr_vect)

cmr_temperature_2011 <- crop(cmr_temperature_2011, cmr_vect)
cmr_temperature_2011 <- mask(cmr_temperature_2011, cmr_vect)

cmr_ndvi_stack <- crop(cmr_ndvi_stack, cmr_vect)
cmr_ndvi_stack <- mask(cmr_ndvi_stack, cmr_vect)

elevation_cmr   <- crop(elevation_cmr, cmr_vect)
elevation_cmr   <- mask(elevation_cmr, cmr_vect)

cmr_land_cover_water     <- crop(land_cover_water, cmr_vect)
cmr_land_cover_water     <- mask(cmr_land_cover_water, cmr_vect)

cmr_land_cover_built     <- crop(land_cover_builts, cmr_vect)
cmr_land_cover_built     <- mask(cmr_land_cover_built, cmr_vect)

cmr_land_cover_cropland  <- crop(land_cover_cropland, cmr_vect)
cmr_land_cover_cropland  <- mask(cmr_land_cover_cropland, cmr_vect)

cmr_land_cover_trees     <- crop(land_cover_trees, cmr_vect)
cmr_land_cover_trees     <- mask(cmr_land_cover_trees, cmr_vect)

cmr_land_cover_wetland   <- crop(land_cover_wetland, cmr_vect)
cmr_land_cover_wetland   <- mask(cmr_land_cover_wetland, cmr_vect)

cmr_ITN_2011   <- crop(cmr_ITN_2011, cmr_vect)
cmr_ITN_2011   <- mask(cmr_ITN_2011, cmr_vect)

cmr_IRS_2011   <- crop(cmr_IRS_2011, cmr_vect)
cmr_IRS_2011   <- mask(cmr_IRS_2011, cmr_vect)

cmr_effective_treat_2011   <- crop(cmr_effective_treat_2011, cmr_vect)
cmr_effective_treat_2011   <- mask(cmr_effective_treat_2011, cmr_vect)

cmr_pop_density_2011_df <- terra::extract(cmr_pop,  map_2011_cmr_sf)
cmr_rainfall_2011_df    <- terra::extract(cmr_rainfall_2011, map_2011_cmr_sf)
cmr_temperature_2011_df <- terra::extract(cmr_temperature_2011, map_2011_cmr_sf)
ndvi_vals_cmr_monthly <- terra::extract(cmr_ndvi_stack, map_2011_cmr_sf)
ndvi_mean_cmr <- apply(ndvi_vals_cmr_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_cmr_df <- terra::extract(elevation_cmr, map_2011_cmr_sf)
cmr_land_cover_water_df <- terra::extract(cmr_land_cover_water, map_2011_cmr_sf)
cmr_land_cover_built_df <- terra::extract(cmr_land_cover_built, map_2011_cmr_sf)
cmr_land_cover_cropland_df <- terra::extract(cmr_land_cover_cropland, map_2011_cmr_sf)
cmr_land_cover_trees_df <- terra::extract(cmr_land_cover_trees, map_2011_cmr_sf)
cmr_land_cover_wetland_df <- terra::extract(cmr_land_cover_wetland, map_2011_cmr_sf)
cmr_ITN_2011_df   <- terra::extract(cmr_ITN_2011, map_2011_cmr_sf)
cmr_IRS_2011_df   <- terra::extract(cmr_IRS_2011, map_2011_cmr_sf)
cmr_effective_treat_2011_df   <- terra::extract(cmr_effective_treat_2011, map_2011_cmr_sf)


# distance to water

bbox <- st_bbox(shp_cmr)

rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))


rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)


# nearest river segment for each point
nn_riv <- st_nn(map_2011_cmr_sf, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(map_2011_cmr_sf, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)


map_2011_cmr_sf$dist_to_water <- pmin(dist_riv_min, dist_lake_min)


cmr_pf_data_2011 <- map_2011_cmr_sf %>% 
  dplyr::select(site_id, lower_age, upper_age, examined, positive, pr, species, dist_to_water) %>% 
  mutate(pop_density = cmr_pop_density_2011_df[,2],
         rainfall = cmr_rainfall_2011_df[,2],
         temperature = cmr_temperature_2011_df[,2],
         elevation = elevation_cmr_df[,2],
         vegetation_index = ndvi_mean_cmr,
         landcover_water = cmr_land_cover_water_df[,2],
         landcover_trees = cmr_land_cover_trees_df[,2],
         landcover_builts = cmr_land_cover_built_df[,2],
         landcover_cropland = cmr_land_cover_cropland_df[,2],
         landcover_wetland = cmr_land_cover_wetland_df[,2],
         ITN_2011         = cmr_ITN_2011_df[,2],
         effective_treat  = cmr_effective_treat_2011_df[,2]) %>% 
  filter(species == "P. falciparum")


numeric_cols <- names(cmr_pf_data_2011)[sapply(cmr_pf_data_2011, is.numeric)]

for (col in numeric_cols) {
  cmr_pf_data_2011 <- fill_by_nearest(cmr_pf_data_2011, col)
}

st_write(cmr_pf_data_2011, 
         "data/cmr_data/benchmarking_spatial_data_cmr_sf.gpkg", append = FALSE)

################################################################################
#              build prediction grid  and covariates extract                   #
################################################################################

# resolution in meters (5000 m = 5 km)
res <- 5000  

box_cmr <- st_bbox(shp_cmr %>% st_transform(3857)) %>% st_as_sfc()
# Create grid covering bounding box of Cameroon
grid <- st_make_grid(
  box_cmr,
  cellsize = res,
  what = "centers"
)

grid <- st_sf(geometry = grid)

# Keep only grid points inside Cameroon
grid <- st_intersection(grid, shp_cmr%>% st_transform(3857))

cmr_vect <- vect(grid %>% st_transform(4326))

# grid_ITN_usage_2021  <- project(ITN_intervention_2021,  st_crs(grid)$wkt)
grid_pop  <- crop(cmr_pop,  cmr_vect)
grid_pop  <- mask(cmr_pop,  cmr_vect)

grid_rainfall_2021 <- crop(cmr_rainfall_2011, cmr_vect)
grid_rainfall_2021 <- mask(cmr_rainfall_2011, cmr_vect)

grid_temperature_2021 <- crop(cmr_temperature_2011, cmr_vect)
grid_temperature_2021 <- mask(cmr_temperature_2011, cmr_vect)

grid_ndvi_stack <- crop(cmr_ndvi_stack, cmr_vect)
grid_ndvi_stack <- mask(cmr_ndvi_stack, cmr_vect)

elevation_grid   <- crop(elevation_cmr, cmr_vect)
elevation_grid   <- mask(elevation_cmr, cmr_vect)

grid_land_cover_water   <- crop(land_cover_water, cmr_vect)
grid_land_cover_water   <- mask(grid_land_cover_water, cmr_vect)

grid_land_cover_built   <- crop(land_cover_builts, cmr_vect)
grid_land_cover_built   <- mask(grid_land_cover_built, cmr_vect)

grid_land_cover_cropland   <- crop(land_cover_cropland, cmr_vect)
grid_land_cover_cropland   <- mask(grid_land_cover_cropland, cmr_vect)

grid_land_cover_trees   <- crop(land_cover_trees, cmr_vect)
grid_land_cover_trees   <- mask(grid_land_cover_trees, cmr_vect)

grid_land_cover_wetland   <- crop(land_cover_wetland, cmr_vect)
grid_land_cover_wetland   <- mask(grid_land_cover_wetland, cmr_vect)

grid_ITN   <- crop(cmr_ITN_2011, cmr_vect)
grid_ITN   <- mask(grid_ITN, cmr_vect)

grid_IRS   <- crop(cmr_IRS_2011, cmr_vect)
grid_IRS   <- mask(grid_IRS, cmr_vect)

grid_effective_treat   <- crop(cmr_effective_treat_2011, cmr_vect)
grid_effective_treat   <- mask(grid_effective_treat, cmr_vect)


# grid_ITN_usage_2021_df <- terra::extract(grid_ITN_usage_2021,  grid)
grid_pop_density_2020_df <- terra::extract(grid_pop,  cmr_vect)
grid_rainfall_2021_df    <- terra::extract(grid_rainfall_2021, cmr_vect)
grid_temperature_2021_df <- terra::extract(grid_temperature_2021, cmr_vect)
ndvi_vals_grid_monthly <- terra::extract(grid_ndvi_stack, cmr_vect)
ndvi_mean_grid <- apply(ndvi_vals_grid_monthly[,-1], 1, mean, na.rm = TRUE)
elevation_grid_df <- terra::extract(elevation_grid, cmr_vect)
grid_land_cover_water_df <- terra::extract(grid_land_cover_water, cmr_vect)
grid_land_cover_built_df <- terra::extract(grid_land_cover_built, cmr_vect)
grid_land_cover_cropland_df <- terra::extract(grid_land_cover_cropland, cmr_vect)
grid_land_cover_trees_df <- terra::extract(grid_land_cover_trees, cmr_vect)
grid_land_cover_wetland_df <- terra::extract(grid_land_cover_wetland, cmr_vect)
grid_ITN_df                <- terra::extract(grid_ITN, cmr_vect)
grid_effective_treat_df    <- terra::extract(grid_effective_treat, cmr_vect)


# nearest river segment for each point

bbox <- st_bbox(grid %>% st_transform(4326))

rivers <- st_read("data/HydroRIVERS_v10_af_shp/HydroRIVERS_v10_af.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))

lakes  <- st_read("data/HydroLAKES_polys_v10_shp/HydroLAKES_polys_v10.shp",
                  wkt_filter = st_as_text(st_as_sfc(bbox)))


rivers_valid <- st_make_valid(rivers)
lakes_valid  <- st_make_valid(lakes)

rivers_valid <- ms_simplify(rivers_valid, keep = 0.3, keep_shapes = TRUE)
lakes_valid  <- ms_simplify(lakes_valid,  keep = 0.3, keep_shapes = TRUE)

rivers_valid <- rivers_valid %>%  st_transform(3857)
lakes_valid <- lakes_valid %>% st_transform(3857)

nn_riv <- st_nn(grid, rivers_valid, k = 1, returnDist = TRUE)
dist_riv_min <- sapply(nn_riv$dist, `[`, 1)

# nearest lake polygon for each point
nn_lake <- st_nn(grid, lakes_valid, k = 1, returnDist = TRUE)
dist_lake_min <- sapply(nn_lake$dist, `[`, 1)


grid$dist_to_water <- pmin(dist_riv_min, dist_lake_min)

grid <- grid %>% 
  mutate(rainfall            = grid_rainfall_2021_df[,2],
         temperature         = grid_temperature_2021_df[,2],
         pop_density         = grid_pop_density_2020_df[,2],
         elevation           = elevation_grid_df[,2],
         vegetation_index    = ndvi_mean_grid,
         landcover_cropland  = grid_land_cover_cropland_df[,2],
         landcover_water     = grid_land_cover_water_df[,2],
         landcover_wetland   = grid_land_cover_wetland_df[,2],
         landcover_builts    = grid_land_cover_built_df[,2],
         landcover_trees     = grid_land_cover_trees_df[,2],
         ITN_2011            = grid_ITN_df[,2],
         effective_treat     = grid_effective_treat_df[,2]) 

numeric_cols <- names(grid)[sapply(grid, is.numeric)]

for (col in numeric_cols) {
  grid <- fill_by_nearest(grid, col)
}

st_write(grid, "data/cmr_data/cmr_map_grid.gpkg", append = FALSE)





```

# Check for multicollinearity

```{r}
cmr_pf_data_2011 <- st_read("data/cmr_data/benchmarking_spatial_data_cmr_sf.gpkg")
grid <- st_read("data/cmr_data/cmr_map_grid.gpkg")
```


```{r}
num_vars <- cmr_pf_data_2011 %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "rainfall", "temperature", "elevation", "vegetation_index", "pop_density", 
  "dist_to_water", "landcover_water", "landcover_trees", "landcover_builts", 
  "landcover_cropland", "landcover_wetland", "ITN_2011", "effective_treat"))

M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  positive / examined ~ rainfall + temperature + elevation +
    vegetation_index + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland + ITN_2011 + effective_treat,
  data = cmr_pf_data_2011
)

vif(fit)

fit1 <- lm(
  positive / examined ~ rainfall + temperature + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland + ITN_2011 + effective_treat + vegetation_index,
  data = cmr_pf_data_2011
)

vif(fit1)
```
We can see that elevation and temperature are highly correlated r = 0.8. So in the first place we will continue with temperature is more directly interpretable for malaria biology. We also notice a strong correlation between land cover (trees) and vegetation index. Here we will also go with vegetation index because it is also one of the covariates than is often use for malaria spatial modelling for it biological interpretability.


# Autocorrelation check
```{r}
# create neighbors based on k nearest neighbors (e.g. k = 4)
coords <- st_coordinates(cmr_pf_data_2011)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

# Moran's I on prevalence
moran.test(cmr_pf_data_2011$pr, lw)
```
- Our malaria prevalence data is strongly spatially clustered a spatial model is absolutely required.

# Variogram
```{r}
map_2011_cmr_sf_sp <- cmr_pf_data_2011 %>% st_transform(3857)

vgm_emp <- variogram(pr ~ 1, data = map_2011_cmr_sf_sp)   # prevalence vs distance
plot(vgm_emp)

vgm_mod <- fit.variogram(
  vgm_emp,
  model = vgm("Mat", nugget = 0.025, psill = 0.05, range = 100000, kappa = 0.55)
)
vgm_mod

plot(vgm_emp, vgm_mod)
```

```{r}

pts      <- st_transform(cmr_pf_data_2011, 3857)
shp_cmr_m  <- st_transform(shp_cmr, 3857)


mesh <- fmesher::fm_mesh_2d_inla(
  loc      = st_coordinates(pts),
  boundary = fm_as_segm(shp_cmr_m),
  max.edge = c(50e3, 200e3),   # 50 km near data, 200 km outer
  cutoff   = 15e3,             # min distance between points (15 km)
  offset   = c(50e3, 150e3)    # extend mesh 50–150 km beyond country
)

# Extract mesh triangles
mesh_tri <- mesh$graph$tv   # triangle vertices (indices)

# Convert each triangle to a polygon
triangles <- lapply(1:nrow(mesh_tri), function(i) {
  coords <- mesh$loc[mesh_tri[i, ], 1:2]
  coords <- rbind(coords, coords[1, ])  # close polygon
  st_polygon(list(coords))
})

mesh_sf <- st_sf(
  geometry = st_sfc(triangles),
  crs = st_crs(pts)   # use same CRS as your points
)


ggplot() +
  geom_sf(data = mesh_sf, fill = NA, color = "grey70", size = 0.2) +
  geom_sf(data = shp_cmr, fill = NA, color = "black", size = 0.4) +
  geom_sf(data = pts, color = "red", size = 2) +
  theme_minimal() +
  labs(title = "INLA Mesh over Country Boundary")

```

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  alpha = 2,
  prior.range = c(90e3, 0.5),    # P(range < 120 km) = 0.5
  prior.sigma = c(0.22, 0.5)      # weakly informative
)

A_est <- inla.spde.make.A(mesh, loc = st_coordinates(pts))


#1. Variables to be scale
vars_to_scale <- c("rainfall", "temperature", "pop_density", 
                   "dist_to_water", "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(pts[[var]], na.rm = TRUE)
  sd_val   <- sd(pts[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  pts[[var]] <- (pts[[var]] - mean_val) / sd_val
}



stk_est <- inla.stack(
  data = list(
    y       = pts$positive,
    Ntrials = pts$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
    intercept        = 1,
    rainfall         = pts$rainfall,
    temperature      = pts$temperature,
    pop_density      = pts$pop_density,
    dist_to_water    = pts$dist_to_water,
    lc_water         = pts$landcover_water,
    lc_trees         = pts$landcover_trees,
    lc_built         = pts$landcover_builts,
    lc_wetland       = pts$landcover_wetland,
    lc_cropland      = pts$landcover_cropland
    )
  ),
  tag = "est"
)


stk_est_intervention <- inla.stack(
  data = list(
    y       = pts$positive,
    Ntrials = pts$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
    intercept        = 1,
    rainfall         = pts$rainfall,
    temperature      = pts$temperature,
    pop_density      = pts$pop_density,
    dist_to_water    = pts$dist_to_water,
    lc_water         = pts$landcover_water,
    lc_trees         = pts$landcover_trees,
    lc_built         = pts$landcover_builts,
    lc_wetland       = pts$landcover_wetland,
    lc_cropland      = pts$landcover_cropland,
    ITN              = pts$ITN_2011,
    effective_treat  = pts$effective_treat
    )
  ),
  tag = "est"
)

```

# Build the model

```{r}

formula <- y ~ -1 + intercept +
  rainfall +
  temperature +
  pop_density +
  dist_to_water +
  lc_water +
  lc_cropland+
  lc_wetland +
  lc_built +
  lc_trees  +
  f(spatial, model = spde)


formula_intervention <- y ~ -1 + intercept +
  rainfall +
  temperature +
  pop_density +
  dist_to_water +
  lc_water +
  lc_cropland+
  lc_wetland +
  lc_built +
  lc_trees  +
  ITN +
  effective_treat+
  f(spatial, model = spde)


res <- inla(
  formula,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_est),
  control.predictor = list(
    A = inla.stack.A(stk_est),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(
    dic = TRUE,
    waic = TRUE,
    cpo = TRUE
  )
)


res_intervention <- inla(
  formula_intervention,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_est_intervention),
  control.predictor = list(
    A = inla.stack.A(stk_est_intervention),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(
    dic = TRUE,
    waic = TRUE,
    cpo = TRUE
  )
)

res$summary.fixed

res$summary.hyperpar


res$dic$dic
res$waic$waic

res_intervention$summary.fixed
res_intervention$summary.hyperpar
```

#checking for overdispersion
```{r}
idx_est <- inla.stack.index(stk_est, "est")$data

fitted_est <- res$summary.fitted.values[idx_est, "mean"]

y <- pts$positive
n <- pts$examined

pearson_res <- (y - n * fitted_est) / sqrt(n * fitted_est * (1 - fitted_est))

n_obs <- length(y)
p_eff <- res$dic$p.eff   # effective number of parameters

phi <- sum(pearson_res^2) / (n_obs - p_eff)
phi

```

# Model diagnostic
```{r}
hist(res_intervention$cpo$pit,
     breaks = 20,
     main = "PIT Histogram",
     xlab = "PIT")

idx_est <- inla.stack.index(stk_est_intervention, "est")$data

fitted_est <- res_intervention$summary.fitted.values[idx_est, "mean"]

df_resid <- as.data.frame(cbind(
  observed = pts$positive / pts$examined,
  fitted   = fitted_est
))

pts$resid <- (df_resid$observed - df_resid$fitted) / sqrt(df_resid$fitted * (1 - df_resid$fitted) / pts$examined)

coords <- st_coordinates(pts)

nb <- dnearneigh(coords, d1 = 0, d2 = 28000)   # 100 km neighbours
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

moran.test(pts$resid,
           lw,
           alternative = "two.sided",
           zero.policy = TRUE)


plot(df_resid$observed, df_resid$fitted,
       xlab = "Observed prevalence",
       ylab = "Fitted prevalence")
abline(0, 1, col = "red")

```

The beta‑binomial SPDE model shows strong but short‑range spatial dependence (posterior range ~ 28 km, σ ~ 1.23). The PIT histogram indicates mild right‑skew, reflecting slight underestimation of very high observed prevalences, but overall calibration is acceptable. Residual Moran’s I is significantly negative (I ≈ –0.07, p ≈ 0.02), consistent with localized over‑ and under‑prediction patterns produced by the short spatial range. Compared to the binomial model, the beta‑binomial likelihood reduces residual spatial autocorrelation and improves PIT calibration. Overall, the model fits well, with only minor miscalibration in the upper tail.

# Build grid for prediction and the prediction model

```{r}
# Build the A‑matrix for prediction locations
A_pred <- inla.spde.make.A(
  mesh = mesh,
  loc = st_coordinates(grid)
)


# Scale the prediction grid using the EXACT same parameters as in the fitting model
# (This ensures a grid cell is measured relative to our 36 observed points)
for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid[[var]] <- (grid[[var]] - mean_val) / sd_val
}



#Build the prediction stack

stk_pred <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid$rainfall,
      temperature      = grid$temperature,
      pop_density      = grid$pop_density,
      dist_to_water    = grid$dist_to_water,
      lc_cropland      = grid$landcover_cropland,
      lc_water         = grid$landcover_water,
      lc_wetland       = grid$landcover_wetland,
      lc_built         = grid$landcover_builts,
      lc_trees         = grid$landcover_trees
    )
  ),
  tag = "pred"
)


stk_pred_intervention <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid$rainfall,
      temperature      = grid$temperature,
      pop_density      = grid$pop_density,
      dist_to_water    = grid$dist_to_water,
      lc_cropland      = grid$landcover_cropland,
      lc_water         = grid$landcover_water,
      lc_wetland       = grid$landcover_wetland,
      lc_built         = grid$landcover_builts,
      lc_trees         = grid$landcover_trees,
      ITN              = grid$ITN_2011,
      effective_treat  = grid$effective_treat
    )
  ),
  tag = "pred"
)


# Combine estimation + prediction stacks
stk_full <- inla.stack(stk_est, stk_pred)

stk_full_intervention <- inla.stack(stk_est_intervention, stk_pred)


# Fit the model again with the full stack
res_pred <- inla(
  formula,
  family = "betabinomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full),
  control.predictor = list(
    A = inla.stack.A(stk_full),
    compute = TRUE,
    link = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_pred_intervention <- inla(
  formula,
  family = "betabinomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full_intervention),
  control.predictor = list(
    A = inla.stack.A(stk_full_intervention),
    compute = TRUE,
    link = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


set.seed(123)
samples <- inla.posterior.sample(100, res_pred_intervention)

idx_pred <- inla.stack.index(stk_full_intervention, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]                   # linear predictor for grid cells
  plogis(lp)                                 # convert logit  prevalence
})


grid$prev_mean   <- rowMeans(pred_mat)
grid$prev_sd     <- apply(pred_mat, 1, sd)
grid$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)


v <- vect(grid)

template <- rast(
  ext = ext(v),
  resolution = 5000,
  crs = crs(v)
)


grid_rast <- rasterize(
  v,
  template,
  field = c("prev_mean", "prev_q025", "prev_q975")
)


cmr_prev_1 <- ggplot(shp_cmr) +
  geom_spatraster(data = grid_rast,
                  mapping = aes(fill = prev_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2011 with intervention")

ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_test_pf_all_age_with_intervention.pdf",
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

pfpr_2011_cmr <- rast("data/cmr_data/MAP_modelled_prevalence/202608_Global_Pf_Parasite_Rate_CMR_2011.tiff") 

names(pfpr_2011_cmr) <- c("Prev_mean", " ", "lci", "uci")

cmr_prev_2 <- ggplot(shp_cmr) +
  geom_spatraster(data = pfpr_2011_cmr,
                  mapping = aes(fill = Prev_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f (2-10years)")+
  theme_void()+
  labs(title = "annual prevalence of P. f (2-10years) in 2011")

ggsave("Outputs/cmr_output/MAP_pr_modeled/2011_cmr_pr_map_pf_2_10.png",
       plot = cmr_prev_2, width = 6, height = 4, dpi = 300)


cmr_vect <- vect(shp_cmr)

prev_dat_cmr_2011 <- terra::extract(pfpr_2011_cmr, cmr_vect)

prev_dat_cmr_2011_test <- terra::extract(grid_rast, vect(shp_cmr))

prev_summary <- prev_dat_cmr_2011 %>%
  group_by(ID) %>%
  summarise(
    PfPR_2_10   = mean(Prev_mean, na.rm = TRUE),
    PfPR_2_10_l = mean(lci, na.rm = TRUE),
    PfPR_2_10_u = mean(uci, na.rm = TRUE)
  )

prev_summary_test <- prev_dat_cmr_2011_test %>%
  group_by(ID) %>%
  summarise(
    PfPR_mean   = mean(prev_mean, na.rm = TRUE),
    PfPR_l      = mean(prev_q025, na.rm = TRUE),
    PfPR_u      = mean(prev_q975, na.rm = TRUE)
  )

prev_dat_cmr_2011_sf <- shp_cmr %>% 
  mutate(PfPR_2_10   =  prev_summary$PfPR_2_10,
         PfPR_2_10_l = prev_summary$PfPR_2_10_l,
         PfPR_2_10_u = prev_summary$PfPR_2_10_u)

prev_dat_cmr_2011_sf_test <- shp_cmr %>% 
  mutate(PfPR_mean   =  prev_summary_test$PfPR_mean,
         PfPR_l      = prev_summary_test$PfPR_l,
         PfPR_u      = prev_summary_test$PfPR_u)


# Convertion of the prevalence to all age group

n <- nrow(prev_dat_cmr_2011_sf)

prev_dat_cmr_2011_sf$PR_all_age <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(100, n),
  parameters   = "Pf_Smith2007"
)

prev_dat_cmr_2011_sf$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10_u,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")

prev_dat_cmr_2011_sf$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10_l,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")


breaks_prev <- c(0, 0.1, 0.2, 0.4, 0.5, 0.6, 0.8, 1)
labels_prev <- c(
  "0–10%", "10–20%", "20–40%", "40–50%",
  "50–60%", "60–80%", "80–100%"
)

colors_prev <- rev(brewer.pal(7, "RdYlGn"))



cmr_prev_3 <- ggplot(prev_dat_cmr_2011_sf) +
  geom_sf(aes(fill = cut(
    PR_all_age,
    breaks = breaks_prev,
    labels = labels_prev,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev,
    name = "Estimated prevalence"
  ) +
  labs(title = "Estimated P. falciparum in prevalence\n in 2011 from MAP") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15)
  )

ggsave("Outputs/cmr_output/MAP_pr_modeled/2011_cmr_pr_map_pf_all_age_admin2.png",
       plot = cmr_prev_3, width = 6, height = 4, dpi = 300)



cmr_prev_4 <- ggplot(prev_dat_cmr_2011_sf_test) +
  geom_sf(aes(fill = cut(
    PfPR_mean,
    breaks = breaks_prev,
    labels = labels_prev,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev,
    name = "Estimated prevalence"
  ) +
  labs(title = "Estimated P. falciparum in prevalence\n in 2011 usind our model (interventions included)") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15)
  )


ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_pf_all_age_admin2_with_intervention.png",
       plot = cmr_prev_4, width = 6, height = 4, dpi = 300)

prev_dat_cmr_2011_sf_test$PfPR_map <- prev_dat_cmr_2011_sf$PR_all_age
prev_dat_cmr_2011_sf_test$PfPR_map_u <- prev_dat_cmr_2011_sf$PR_all_age_U
prev_dat_cmr_2011_sf_test$PfPR_map_l <- prev_dat_cmr_2011_sf$PR_all_age_L

prev_dat_cmr_2011_sf_test <- prev_dat_cmr_2011_sf_test %>% 
  mutate(diff_PfPR = PfPR_mean - PfPR_map)

breaks_prev_diff <- c(-0.15, -0.05, 0.05, 0.1, 0.15, 0.2, 0.25, 30)
labels_prev_diff <- c(
  "-0.15%<-0.05%", "-0.05%<0.05%", "0.05%<0.10%", "0.10%<0.15%",
  "0.15%<0.20%", "0.20%<0.25%", "0.25<0.30%"
)

colors_prev_diff <- rev(brewer.pal(7, "RdYlGn"))

cmr_prev_5 <- ggplot(prev_dat_cmr_2011_sf_test) +
  geom_sf(aes(fill = cut(
    diff_PfPR,
    breaks = breaks_prev_diff,
    labels = labels_prev_diff,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev_diff,
    name = "Difference in prevalence"
  ) +
  labs(title = "Difference in prevalence with intervention (2011)") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15))


prev_dat_cmr_2011_sf_test$diff_class <- cut(
  prev_dat_cmr_2011_sf_test$diff_PfPR,
  breaks = breaks_prev_diff,
  labels = labels_prev_diff,
  include.lowest = TRUE
)

map1 <- mapview(
  prev_dat_cmr_2011_sf_test,
  zcol = "diff_class",
  col.regions = colors_prev_diff,
  layer.name = "Difference in prevalence (2011)",
  legend = TRUE
  )
  

mapshot(map1, 
        file = "Outputs/cmr_output/Pf_pr_estimated/difference_prev_MAP_vsour_model.html",
        selfcontained = T)

ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_diff_pf_all_age_admin2_with_intervention.png",
       plot = cmr_prev_5, width = 6, height = 4, dpi = 300)

```


# Extract MAP data

```{r}
#cor(cmr_data_pev_map$prev_mean, cmr_map_pr_pf_2_10$PR_all_age, use = "complete.obs")

cmr_data_pev_map <- cmr_data_pev_map %>% 
  mutate(prev_mean = case_when(is.na(prev_mean) ~ mean(prev_mean, na.rm = TRUE),
                               TRUE ~ prev_mean))

cmr_data_pev_map$PR_all_age <- cmr_map_pr_pf_2_10$PR_all_age
#-----------------------------------------------------------
# 1. BASIC AGREEMENT METRICS
#-----------------------------------------------------------

# Bias
bias <- mean(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age, na.rm = TRUE)

# MAE
mae <- mean(abs(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age), na.rm = TRUE)

# RMSE
rmse <- sqrt(mean((cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age)^2, na.rm = TRUE))

# Pearson correlation
pearson_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "pearson")

# Spearman rank correlation
spearman_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "spearman")

# #-----------------------------------------------------------
# # 2. SPATIAL AGREEMENT METRICS
# #-----------------------------------------------------------
# 
# # Create spatial weights
# nb <- poly2nb(cmr_data_pev_map)
# lw <- nb2listw(nb, style = "W")
# 
# # Spatial cross-correlation (correlation of lagged values)
# lag_model <- lag.listw(lw, cmr_data_pev_map$prev_mean)
# lag_map   <- lag.listw(lw, cmr_data_pev_map$PR_all_age)
# spatial_cross_cor <- cor(lag_model, lag_map, use = "complete.obs")
# 
# # Moran's I of residuals
# residuals <- cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age
# moran_residuals <- moran.test(residuals, lw)
# 
# #-----------------------------------------------------------
# # 3. SEMIVARIOGRAM COMPARISON
# #-----------------------------------------------------------
# 
# library(gstat)
# 
# # Convert sf to sp for gstat
# admin2_sp <- as(cmr_data_pev_map, "Spatial")
# 
# vg_model <- variogram(prev_mean ~ 1, admin2_sp)
# vg_map   <- variogram(PR_all_age ~ 1, admin2_sp)
# 
# #-----------------------------------------------------------
# # 4. DISTRIBUTIONAL COMPARISON
# #-----------------------------------------------------------
# 
# # Q–Q plot data
# qq_data <- data.frame(
#   model = quantile(cmr_data_pev_map$prev_mean, probs = seq(0,1,0.01), na.rm = TRUE),
#   map   = quantile(cmr_data_pev_map$PR_all_age, probs = seq(0,1,0.01), na.rm = TRUE)
# )
# 
# # Distribution overlap (simple measure)
# dens_model <- density(cmr_data_pev_map$prev_mean, na.rm = TRUE)
# dens_map   <- density(cmr_data_pev_map$PR_all_age, na.rm = TRUE)
# 
# # Overlap integral
# x_common <- seq(min(dens_model$x, dens_map$x), max(dens_model$x, dens_map$x), length.out = 1000)
# y_model  <- approx(dens_model$x, dens_model$y, x_common)$y
# y_map    <- approx(dens_map$x, dens_map$y, x_common)$y
# distribution_overlap <- sum(pmin(y_model, y_map)) * (x_common[2] - x_common[1])
# 
# #-----------------------------------------------------------
# # 5. CLASSIFICATION AGREEMENT (TRANSMISSION STRATA)
# #-----------------------------------------------------------
# 
# # Define strata
# cut_points <- c(0, 0.05, 0.30, 1)
# labels <- c("low", "medium", "high")
# 
# cmr_data_pev_map <- cmr_data_pev_map %>%
#   mutate(
#     class_model = cut(prev_mean, breaks = cut_points, labels = labels, include.lowest = TRUE),
#     class_map   = cut(PR_all_age, breaks = cut_points, labels = labels, include.lowest = TRUE)
#   )
# 
# # Confusion matrix
# confusion_matrix <- table(cmr_data_pev_map$class_model, cmr_data_pev_map$class_map)
# 
# # Kappa statistic
# library(irr)
# kappa_stat <- kappam.fleiss(as.data.frame(confusion_matrix))
# 
# #-----------------------------------------------------------
# # 6. ADMIN2 AGGREGATED COMPARISONS
# #-----------------------------------------------------------
# 
# # Admin2 mean prevalence (already aggregated)
# admin2_comparison <- cmr_data_pev_map %>%
#   st_drop_geometry() %>%
#   select(admin2, pfpr_model, pfpr_map_all_age)
# 
# # Rank correlation of districts
# rank_cor <- cor(
#   rank(cmr_data_pev_map$pfpr_model),
#   rank(cmr_data_pev_map$pfpr_map_all_age),
#   method = "spearman"
# )
# 
# #-----------------------------------------------------------
# # 7. POPULATION-WEIGHTED PREVALENCE (if pop column exists)
# #-----------------------------------------------------------
# 
# if("population" %in% colnames(cmr_data_pev_map)) {
#   pwpr_model <- sum(cmr_data_pev_map$pfpr_model * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
#   
#   pwpr_map <- sum(cmr_data_pev_map$pfpr_map_all_age * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
# }
# 
# #-----------------------------------------------------------
# # 8. UNCERTAINTY COMPARISON (if your model has CI)
# #-----------------------------------------------------------
# 
# if("model_lower" %in% colnames(cmr_data_pev_map) &&
#    "model_upper" %in% colnames(cmr_data_pev_map)) {
#   
#   coverage_probability <- mean(
#     cmr_data_pev_map$pfpr_map_all_age >= cmr_data_pev_map$model_lower &
#       cmr_data_pev_map$pfpr_map_all_age <= cmr_data_pev_map$model_upper,
#     na.rm = TRUE
#   )
# }
# 
# #-----------------------------------------------------------
# # PRINT RESULTS
# #-----------------------------------------------------------
# 
# list(
#   bias = bias,
#   mae = mae,
#   rmse = rmse,
#   pearson_cor = pearson_cor,
#   spearman_cor = spearman_cor,
#   spatial_cross_cor = spatial_cross_cor,
#   moran_residuals = moran_residuals,
#   semivariogram_model = vg_model,
#   semivariogram_map = vg_map,
#   distribution_overlap = distribution_overlap,
#   confusion_matrix = confusion_matrix,
#   kappa_stat = kappa_stat,
#   admin2_comparison = admin2_comparison,
#   rank_cor = rank_cor
# )
# 

```
Overall, our model tends to overestimate malaria prevalence compared to MAP (bias ≈ 0.064), with typical differences of about 9–12 percentage points (MAE and RMSE). The moderate Pearson correlation (~0.55) shows that our model captures only part of MAP’s spatial variation, while the slightly higher Spearman correlation (~0.60) indicates that it does a somewhat better job ranking districts from low to high transmission than matching exact values. Taken together, these results suggest that oour model broadly identifies the same high‑burden and low‑burden areas as MAP but differs noticeably in magnitude, with some districts showing larger discrepancies. This is a solid foundation for a first geospatial model, but it also highlights clear opportunities to refine covariates, spatial structure, or calibration to better align with MAP’s established patterns.

```{r}

```


```{r}
# cmr_map_data <- read.csv("cmr_data/cmr_map_PR.csv")
# 
# cmr_map_data_clean <- cmr_map_data %>%
#   filter(!is.na(longitude) & !is.na(latitude)) %>% 
#   filter(species=="P. falciparum") 
# 
# # Convert to sf spatial object
# cmr_map_sf <- st_as_sf(
#   cmr_map_data_clean,
#   coords = c("longitude", "latitude"),
#   crs    = 4326
# )

# Check
# plot(st_geometry(cmr_map_sf), main = "MAP Survey Points")
# 
# shp_cmr <- st_transform(shp_cmr, crs=4326)
# 
# plot(st_geometry(shp_cmr))
# plot(st_geometry(cmr_map_sf), col = "red", pch = 16, cex = 0.5, add = TRUE)
# 
# ggplot()+
#   geom_sf(data = shp_cmr, fill=NA, color="black")+
#   geom_sf(data=cmr_map_sf, color="red", size=0.2)


# cmr_map_with_district <- st_join(
#   cmr_map_sf,
#   shp_cmr[, c("Admin1", "Admin2")],
#   join = st_within
# )
# 
# cmr_points_per_district <- cmr_map_with_district %>%
#   st_drop_geometry() %>%
#   group_by(Admin2) %>%
#   summarise(n_surveys = n()) %>%
#   arrange(desc(n_surveys))
# 
# #print(cmr_points_per_district)
# 
# 
# cmr_district_prevalence <- cmr_map_with_district %>%
#   st_drop_geometry() %>%
#   filter(!is.na(Admin2)) %>%       
#   group_by(Admin1, Admin2) %>%
#   summarise(n_examined     = sum(examined,  na.rm = TRUE),
#             n_positive     = sum(positive,  na.rm = TRUE),
#             raw_prevalence = n_positive / n_examined,
#             raw_pr         = sum(pr),
#             n_surveys      = n(),
#             year_min       = min(year_start, na.rm = TRUE),
#             year_max       = max(year_end,   na.rm = TRUE),
#             .groups = "drop")
# 
# 
# #2011 data
# 
# cmr_map_2011 <- cmr_map_data_clean %>%
#   filter(year_start == 2011 | year_end==2011) %>% 
#   filter(species=="P. falciparum")
# 
# 
# cmr_map_2011_sf <- st_as_sf(
#   cmr_map_2011,
#   coords = c("longitude", "latitude"),
#   crs    = 4326
# )
# 
# 
# cmr_map_with_district_2011 <- st_join(
#   cmr_map_2011_sf,
#   shp_cmr[, c("Admin1", "Admin2")],
#   join = st_within
# )
# 
# 
# cmr_observed_districts_2011 <- cmr_map_with_district_2011 %>%
#   st_drop_geometry() %>%
#   group_by(Admin1, Admin2) %>%
#   summarise(
#     n_points   = n(),
#     n_examined = sum(examined, na.rm = TRUE),
#     n_positive = sum(positive, na.rm = TRUE),
#     .groups    = "drop"
#   ) %>%
#   mutate(raw_prevalence = n_positive / n_examined)
# 
# ggplot()+
#   geom_sf(data = shp_cmr, fill=NA, color="black")+
#   geom_sf(data = shp_cmr[shp_cmr$Admin2 %in% cmr_observed_districts_2011$Admin2, ], color="red")+
#   geom_sf(data = cmr_map_2011_sf, color = "blue", size=2)
# 
# 
# ### extract the observed districts as well as their neighbors
# observed_names <- cmr_observed_districts_2011$Admin2
# observed_districts_sf <- shp_cmr %>%
#   filter(str_squish(Admin2) %in% observed_names)
# 
# nb_queen <- poly2nb(shp_cmr, queen = TRUE)
# # Get row indices of observed districts in the full shape file
# observed_indices <- which(
#   str_squish(shp_cmr$Admin2) %in% observed_names
# )
# 
# 
# # Get neighbor indices for each observed district
# neighbor_indices <- list()
# 
# for(i in observed_indices){
#   neighbors_i <- nb_queen[[i]]
#   cat("District:", shp_cmr$Admin2[i], 
#       "Number of neighbors:", length(neighbors_i), "\n")
#   neighbor_indices[[i]] <- neighbors_i
# }
# 
# # Combine all neighbor indices into one vector
# all_neighbor_indices <- unique(unlist(neighbor_indices))
# # Remove observed districts from neighbor list
# # (in case an observed district is neighbor of another observed district)
# pure_neighbor_indices <- setdiff(all_neighbor_indices, observed_indices)
# # Combine observed + neighbor indices
# study_area_indices <- unique(c(observed_indices, pure_neighbor_indices))
# 
# # Extract study area from full shapefile
# study_area_sf <- shp_cmr[study_area_indices, ] %>%
#   mutate(district_type = ifelse(str_squish(Admin2) %in% observed_names,
#       "Observed", "Neighbor"))
# 
# # Visualization of our study area
# 
# ggplot() +
#   # Full country in background
#   geom_sf(data = shp_cmr, fill = "white", color = "grey80") +
#   # Study area
#   geom_sf(
#     data = study_area_sf,
#     aes(fill = district_type),
#     color = "grey40"
#   ) +
#   scale_fill_manual(
#     values = c("Observed" = "#d73027", "Neighbor" = "#fee090"),
#     name   = "District type"
#   ) +
#   labs(
#     title    = "Study Area: Observed and Neighbor Districts",
#     subtitle = "Cameroon 2011: Queen Contiguity"
#   ) +
#   theme_minimal()+
#   theme(
#     axis.text  = element_blank(),
#     axis.ticks = element_blank()
#   )
# 
# # Merge aggregated prevalence into study area
# study_area_sf <- study_area_sf %>%
#   left_join(
#     cmr_observed_districts_2011,
#     by = c("Admin2")
#   ) %>%
#   mutate(
#     # Neighbor districts have NA for observed data — correct
#     n_examined = ifelse(is.na(n_examined), NA, n_examined),
#     n_positive = ifelse(is.na(n_positive), NA, n_positive)
#   )


```



```{r}
## Building the adjacency matrix for our model

study_area_sf <- study_area_sf %>%
  mutate(district_id = 1:nrow(.)) 

nb_study <- poly2nb(study_area_sf, queen = TRUE)

# # Summary
# summary(nb_study)
# 
# nb_check <- nb2listw(nb_study, style = "B", zero.policy = TRUE)
# 
# 
components <- n.comp.nb(nb_study)
cat("Number of connected components:", components$nc, "\n")

# Which districts belong to which component
study_area_sf$component <- components$comp.id
print(study_area_sf[, c("Admin2", "district_type", "component")])

nb2INLA("study_area_adj.graph", nb_study)

file.exists("study_area_adj.graph")


### Visualize the study area

coords <- st_coordinates(st_centroid(study_area_sf))
nb_lines <- map2_dfr(
  .x = seq_along(nb_study),
  .y = nb_study,
  .f = function(i, neighbors) {
    if (length(neighbors) == 0) return(NULL)

    map_dfr(neighbors, function(j) {
      st_linestring(rbind(coords[i, ], coords[j, ])) |> 
        st_sfc(crs = st_crs(study_area_sf)) |> 
        st_sf(from = i, to = j, geometry = _)
    })
  }
)


ggplot() +
  # District polygons
  geom_sf(
    data = study_area_sf,
    aes(fill = district_type),
    color = "black",
    size = 0.3
  ) +
  
  # Neighbor links
  geom_sf(
    data = nb_lines,
    color = "blue",
    size = 0.7
  ) +
  
  # District labels
  geom_text(
    data = cbind(study_area_sf, coords),
    aes(X, Y, label = Admin2),
    size = 3
  ) +
  
  scale_fill_manual(
    values = c("Observed" = "#d73027", "Neighbor" = "#fee090")
  ) +
  
  labs(
    title = "Queen Contiguity Neighborhood Structure",
    fill = "District Type"
  ) +
  
  theme_minimal()+
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank()
  )

study_area_sf <- study_area_sf %>%
  mutate(
    component_factor = as.factor(component)
  )

```


```{r}
#building our model

model_data <- study_area_sf %>%
  st_drop_geometry() %>%
  dplyr::select(district_id, Admin2, district_type, 
                component, n_examined, n_positive) %>%
  mutate(
    # R-INLA needs numeric outcome
    n_examined = as.integer(n_examined),
    n_positive = as.integer(n_positive),
    
    # District ID for spatial random effect
    district_id_unstr = district_id  # separate copy for unstructured effect
  )

# Define formula
# No covariates — only intercept + spatial + unstructured random effects
formula_null <- n_positive ~ 1 +
  f(district_id,
    model         = "besag",        # spatially structured ICAR
    graph         = "study_area_adj.graph",
    scale.model   = TRUE,
    constr        = TRUE,
    hyper         = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  f(district_id_unstr,
    model = "iid",                  # unstructured random effect
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Fit the model
model_null <- inla(
  formula_null,
  family          = "binomial",
  Ntrials         = model_data$n_examined,
  data            = model_data,
  control.predictor = list(
    compute = TRUE,
    link    = 1          # use the link function for predictions
  ),
  control.compute = list(
    dic  = TRUE,
    waic = TRUE,
    cpo  = TRUE
  )
)

# Summary
summary(model_null)


# Extract posterior mean and 95% credible intervals
model_data$prev_mean  <- model_null$summary.fitted.values$mean
model_data$prev_lower <- model_null$summary.fitted.values$`0.025quant`
model_data$prev_upper <- model_null$summary.fitted.values$`0.975quant`

# Check results
model_data %>%
  dplyr::select(Admin2, district_type, 
                n_examined, n_positive,
                prev_mean, prev_lower, prev_upper) %>%
  mutate(across(c(prev_mean, prev_lower, prev_upper), 
                ~ round(., 3)))

study_area_results <- study_area_sf %>%
  left_join(
    model_data %>% dplyr::select(district_id, prev_mean, 
                                  prev_lower, prev_upper),
    by = "district_id"
  )

ggplot() +
  geom_sf(data = shp_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data = study_area_results,
    aes(fill = prev_mean),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "Prevalence",
    option = "magma",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Estimated Malaria Prevalence — Null Spatial Model",
    subtitle = "BYM2 without covariates — Cameroon 2011"
  ) +
  theme_minimal()

# Map uncertainty (width of credible interval)
study_area_results <- study_area_results %>%
  mutate(ci_width = prev_upper - prev_lower)

ggplot() +
  geom_sf(data = shp_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data = study_area_results,
    aes(fill = ci_width),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "CI Width",
    option = "rocket",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Model Uncertainty — Null Spatial Model",
    subtitle = "Width of 95% Credible Interval"
  ) +
  theme_minimal()

```


```{r}
# merging the extracted covariates with the study areas data

# Extract district names from your study area
study_districts <- study_area_sf %>%
  st_drop_geometry() %>%
  dplyr::select(district_id, Admin2, 
                district_type, component,
                n_examined, n_positive)

# loading the covariates data

rainfall_cmr_2011 <- read.csv("cmr_data/cmr_rainfall_2011.csv")
temperature_cmr_2011 <- read.csv("cmr_data/cmr_temperature_2011.csv")
vegetation_index_cmr_2011 <- read.csv("cmr_data/cmr_vegetation_index_2011.csv")
population_density_cmr_2011 <- read.csv("cmr_data/cmr_pop_density_2011.csv")
elevation_data_cmr <- read.csv("cmr_data/cmr_elevation_data.csv")
land_cover_data_cmr <- read.csv("cmr_data/cmr_landcover_per_district.csv")
distance_to_water_cmr <- read.csv("cmr_data/cmr_distance_to_water.csv")


# Filter each CSV to keep only study area districts
# then merge all together

model_data <- study_districts %>%
  
  # Rainfall
  left_join(
    rainfall_cmr_2011 %>% dplyr::select(District, mean_rainfall),
    by = c("Admin2" = "District")
  ) %>%
  
  # Temperature
  left_join(
    temperature_cmr_2011 %>% dplyr::select(District, mean_temperature),
    by = c("Admin2" = "District")
  ) %>%
  
  # NDVI
  left_join(
    vegetation_index_cmr_2011 %>% dplyr::select(District, NDVI),
    by = c("Admin2" = "District")
  ) %>%
  
  # Elevation
  left_join(
    elevation_data_cmr %>% dplyr::select(District, mean_elevation),
    by = c("Admin2" = "District")
  ) %>%
  
  # Population density
  left_join(
    population_density_cmr_2011 %>% dplyr::select(District, pop_density_2011),
    by = c("Admin2" = "District")
  ) %>%
  
  # Distance to water
  left_join(
    distance_to_water_cmr %>% dplyr::select(District, dist_to_water_km),
    by = c("Admin2" = "District")
  ) %>%
  
  # Land cover
  left_join(
    land_cover_data_cmr %>% dplyr::select(District, prop_trees, prop_water,
                                 prop_wetland, prop_cropland, prop_built),
    by = c("Admin2" = "District")
  )

#save the data
write_csv(model_data, "cmr_data/cmr_study_area_model_data_2011.csv")

covariate_cols <- c("mean_rainfall", "mean_temperature",
                    "NDVI", "mean_elevation", "pop_density_2011",
                    "dist_to_water_km", "prop_trees", "prop_water",
                    "prop_wetland", "prop_cropland", "prop_built")

# Standardize to mean 0 and SD 1
model_data <- model_data %>%
  mutate(across(
    all_of(covariate_cols),
    ~ scale(.) %>% as.vector(),
    .names = "{.col}_std"
  ))

# Add unstructured random effect ID
model_data <- model_data %>%
  mutate(district_id_unstr = district_id)

# Check standardized values
model_data %>%
  dplyr::select(Admin2, ends_with("_std")) %>%
  head()


cov_matrix <- model_data %>%
  dplyr::select(all_of(covariate_cols)) %>%
  cor(use = "complete.obs")

print(round(cov_matrix, 2))

corrplot(cov_matrix,
         method      = "color",
         type        = "upper",
         tl.cex      = 0.7,
         addCoef.col = "black",
         number.cex  = 0.6)


final_covariates <- c("mean_rainfall", "mean_temperature",
                      "NDVI", "dist_to_water_km",
                      "prop_wetland", "prop_cropland", "prop_built")

# Check correlations again
final_cor <- model_data %>%
  dplyr::select(all_of(final_covariates)) %>%
  cor(use = "complete.obs")

corrplot(final_cor,
         method      = "color",
         type        = "upper",
         tl.cex      = 0.7,
         addCoef.col = "black",
         number.cex  = 0.6)

```


```{r}
#run the model with covariates

# Add unstructured ID
model_data <- model_data %>%
  mutate(district_id_unstr = district_id)

# Define formula with all 7 covariates
formula_full <- n_positive ~ 1 +
  # Fixed effects — standardized covariates
  mean_rainfall_std +
  mean_temperature_std +
  NDVI_std +
  dist_to_water_km_std +
  prop_wetland_std +
  prop_cropland_std +
  prop_built_std +
  # Spatially structured random effect
  f(district_id,
    model       = "besag",
    graph       = "study_area_adj.graph",
    scale.model = TRUE,
    constr      = TRUE,
    hyper       = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  # Unstructured random effect
  f(district_id_unstr,
    model = "iid",
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Fit the model
model_full <- inla(
  formula_full,
  family            = "binomial",
  Ntrials           = model_data$n_examined,
  data              = model_data,
  control.predictor = list(
    compute = TRUE,
    link    = 1
  ),
  control.compute   = list(
    dic  = TRUE,
    waic = TRUE,
    cpo  = TRUE
  )
)

# Summary
summary(model_full)


# Use informative priors based on malaria literature
# This prevents the model from wandering into implausible territory

formula_full_informed <- n_positive ~ 1 +
  mean_rainfall_std +
  mean_temperature_std +
  NDVI_std +
  dist_to_water_km_std +
  prop_wetland_std +
  prop_cropland_std +
  prop_built_std +
  f(district_id,
    model       = "besag",
    graph       = "study_area_adj.graph",
    scale.model = TRUE,
    constr      = TRUE,
    hyper       = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  ) +
  f(district_id_unstr,
    model = "iid",
    hyper = list(
      prec = list(prior = "pc.prec", param = c(1, 0.01))
    )
  )

# Tighter priors on fixed effects
model_full_informed <- inla(
  formula_full_informed,
  family            = "binomial",
  Ntrials           = model_data$n_examined,
  data              = model_data,
  control.fixed     = list(
    mean.intercept = 0,
    prec.intercept = 1,     # tighter prior on intercept
    mean           = 0,
    prec           = 1      # tighter prior on all coefficients
                            # equivalent to SD = 1 on logit scale
  ),
  control.predictor = list(compute = TRUE, link = 1),
  control.compute   = list(dic = TRUE, waic = TRUE, cpo = TRUE)
)

summary(model_full_informed)

```


```{r}
#extract the result of the full informed model

# Extract fitted values
model_data$prev_mean  <- model_full_informed$summary.fitted.values$mean
model_data$prev_lower <- model_full_informed$summary.fitted.values$`0.025quant`
model_data$prev_upper <- model_full_informed$summary.fitted.values$`0.975quant`
model_data$ci_width   <- model_data$prev_upper - model_data$prev_lower

# Check observed vs fitted for observed districts
model_data %>%
  filter(district_type == "Observed") %>%
  mutate(
    raw_prev    = round(n_positive / n_examined, 3),
    fitted_prev = round(prev_mean, 3),
    lower       = round(prev_lower, 3),
    upper       = round(prev_upper, 3)
  ) %>%
  dplyr::select(Admin2, raw_prev, fitted_prev, lower, upper)

model_data %>%
  filter(district_type == "Neighbor") %>%
  mutate(
    fitted_prev = round(prev_mean,  3),
    lower       = round(prev_lower, 3),
    upper       = round(prev_upper, 3),
    ci_width    = round(ci_width,   3)
  ) %>%
  dplyr::select(Admin2, component, fitted_prev,
                lower, upper, ci_width) %>%
  arrange(component, desc(fitted_prev))

# map the result

# Merge results to spatial object
study_area_results <- study_area_sf %>%
  left_join(
    model_data %>% dplyr::select(district_id, prev_mean,
                                  prev_lower, prev_upper, ci_width),
    by = "district_id"
  )

# Map 1 — Posterior mean prevalence
ggplot() +
  geom_sf(data = districts_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data  = study_area_results,
    aes(fill = prev_mean),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "Prevalence",
    option = "magma",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Estimated Malaria Prevalence",
    subtitle = "BYM2 with informative priors — Cameroon 2011"
  ) +
  theme_minimal()

# Map 2 — Uncertainty
ggplot() +
  geom_sf(data = districts_cmr, fill = "grey95", color = "grey80") +
  geom_sf(
    data  = study_area_results,
    aes(fill = ci_width),
    color = "grey40"
  ) +
  scale_fill_viridis_c(
    name   = "CI Width",
    option = "rocket",
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Model Uncertainty",
    subtitle = "Width of 95% Credible Interval"
  ) +
  theme_minimal()

```

# Extract the modelled prevalence from MAP

```{r}
available_rasters <- listRaster()
print(available_rasters)

available_rasters[grep("Pf|falciparum|PR", 
                        available_rasters$dataset_id,
                        ignore.case = TRUE), ]

# Get Cameroon shapefile
cmr_shp <- getShp(ISO = "CMR", admin_level = "admin0")

# Download MAP PfPR surface using correct dataset_id
map_raster <- getRaster(
  dataset_id = "Malaria__202206_Global_Pf_Parasite_Rate",
  shp        = cmr_shp,
  year       = 2011
)


plot(map_raster, main = "MAP PfPR2-10 — Cameroon 2011")

# Extract MAP prevalence per study area district
map_extracted <- extract(
  map_raster,
  vect(study_area_sf),
  fun   = "mean",
  na.rm = TRUE
)

model_data$map_prev <- map_extracted[, 2]

# Compare
comparison <- model_data %>%
  dplyr::select(Admin2, district_type,
                prev_mean, prev_lower, prev_upper,
                map_prev) %>%
  mutate(
    prev_mean  = round(prev_mean, 3),
    map_prev   = round(map_prev,  3),
    difference = round(prev_mean - map_prev, 3),
    map_within_ci = map_prev >= prev_lower &
                    map_prev <= prev_upper
  )

```
