---
title: "cmrchmarking exercise"
author: "Ndim Handson NSANI"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#load necessary libraries
library(malariaAtlas)
library(sf)
library(leaflet)
library(mapview)
library(htmlwidgets)
library(webshot2)
library(ggplot2)
library(RColorBrewer)
library(leafpop)
library(readr)
library(tidyverse)
library(readxl)
library(corrplot)
library(MODISTools)
library(terra)
library(spData)
library(gganimate)
library(viridis)
library(GGally)
library(spdep)
library(INLA)
library(SpatialEpi)
library(geodata)
library(tidygeocoder)
library(nngeo)
library(car)
library(gstat)
library(raster)
library(FNN)
library(rmapshaper)
library(tidyterra)
library(fmesher)
```


```{r}
## Function: fill NA values in one column using nearest non‑NA neighbor
fill_by_nearest <- function(sf_obj, colname) {
  
  # rows with non‑missing values
  non_na <- sf_obj %>% filter(!is.na(.data[[colname]]))
  
  # rows with missing values
  na_rows <- sf_obj %>% filter(is.na(.data[[colname]]))
  
  # if no NA: return unchanged
  if (nrow(na_rows) == 0) return(sf_obj)
  
  # find nearest non‑NA feature for each NA row
  nearest_id <- st_nearest_feature(na_rows, non_na)
  
  # extract the corresponding values
  replacement_values <- non_na[[colname]][nearest_id]
  
  # fill the NA values
  sf_obj[[colname]][is.na(sf_obj[[colname]])] <- replacement_values
  
  sf_obj
}
```


```{r}
# Downloading the malaria dataset
# library(malariaAtlas)
# d <- getPR(country = "Cameroon", species = "BOTH")
# ggplot2::autoplot(d)
# 
# write_csv(d, "cmr_data/cmr_map_PR.csv")

# africa_pf_pv_PR <- getPR(continent = "Africa", species = "BOTH")
# 
# write_csv(africa_pf_pv_PR, "data/Africa_data/Benchmark data/raw_pr_map.csv")

```


```{r}
GPS_cmr_2011 <- st_read("data/cmr_data/GPS location 2011/CMGE61FL.shp")
GPS_cmr_2022 <- st_read("data/cmr_data/GPS location 2022/CMGE81FL.shp")


shp_cmr <- read_sf("data/cmr_data/CMR_shape_file/cameroun_districts_2024.shp")

cmr_vect <- vect(shp_cmr)

GPS_cmr_coords_2011 <- GPS_cmr_2011 %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_cmr_2011)[, 1],
    latitude  = st_coordinates(GPS_cmr_2011)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  filter(latitude != 0 & longitude != 0)

GPS_cmr_coords_2022 <- GPS_cmr_2022 %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_cmr_2022)[, 1],
    latitude  = st_coordinates(GPS_cmr_2022)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  filter(latitude != 0 & longitude != 0)

# loading the extracted MAP data

raw_prevalence_cmr <- read.csv("data/cmr_data/cmr_map_PR.csv")

raw_prevalence_cmr_with_coord <- raw_prevalence_cmr %>% 
  filter(!is.na(longitude) & !is.na(latitude))

raw_prevalence_cmr_without_coord <- raw_prevalence_cmr %>% 
  filter(is.na(longitude) & is.na(latitude)) %>% 
  dplyr::select(-c(longitude, latitude))


map_without_coords_cmr <- raw_prevalence_cmr_without_coord %>%
  mutate(cluster_num = as.integer(substr(dhs_id, 9, 16)))

map_recovered_cmr_2011 <- map_without_coords_cmr %>%
  left_join(GPS_cmr_coords_2011, by = c("cluster_num" = "DHSCLUST"))


map_recovered_cmr_2022 <- map_without_coords_cmr %>%
  left_join(GPS_cmr_coords_2022, by = c("cluster_num" = "DHSCLUST"))

cat("Recovered:", sum(!is.na(map_recovered_cmr_2011$longitude)), "\n")
cat("Still missing:", sum(is.na(map_recovered_cmr_2011$longitude)), "\n")


map_full_cmr_2011 <- bind_rows(
  raw_prevalence_cmr_with_coord %>% filter(!is.na(longitude) & !is.na(latitude)),
  map_recovered_cmr_2011 %>% filter(!is.na(longitude) & !is.na(latitude))
)

map_full_cmr_2011 <- map_full_cmr_2011 %>%
  filter(year_start == 2011) %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  filter(species == "P. falciparum")


write_csv(map_full_cmr_2011, file = "data/cmr_data/cmr_MAP_full_raw_prevalence_2011.csv")

map_full_cmr_2022 <- bind_rows(
  raw_prevalence_cmr_with_coord %>% filter(!is.na(longitude) & !is.na(latitude)),
  map_recovered_cmr_2022 %>% filter(!is.na(longitude) & !is.na(latitude))
)

map_full_cmr_2022 <- map_full_cmr_2022 %>%
  filter(year_start == 2022) %>%
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  filter(species == "P. falciparum")


write_csv(map_full_cmr_2022, file = "data/cmr_data/cmr_MAP_full_raw_prevalence_2022.csv")

```


# Filter for the year 2011 for our analysis

```{r}
map_2011_cmr <- st_read("data/cmr_data/cmr_MAP_full_raw_prevalence_2011.csv")

map_2011_cmr_sf <- st_as_sf(map_2011_cmr, coords = c("longitude", "latitude"), crs = 4326)


breaks_pr <- c(0, 0.05, 0.10, 0.20, 0.40, 0.60, 1)
labels_pr <- c(
  "0–5%", "5–10%", "10–20%", "20–40%", "40–60%", ">60%"
)

colors_pr <- c(
  "#ffffcc",  # pale yellow (very low)
  "#ffeda0",  # light yellow
  "#feb24c",  # orange
  "#fd8d3c",  # deep orange
  "#f03b20",  # red
  "#bd0026"   # dark red (very high)
)

map_2011_cmr_sf$pr <- as.numeric(map_2011_cmr_sf$pr)

map_2011_cmr_sf$pr_class <- cut(
  map_2011_cmr_sf$pr,
  breaks = breaks_pr,
  labels = labels_pr,
  include.lowest = TRUE
)



map <- mapview(shp_cmr,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
  mapview(map_2011_cmr_sf,
          zcol = "pr_class",             # categorical variable
          cex = "examined",
          col.regions = colors_pr,       # one color per class
          layer.name = "P. malariae prevalence",
          legend = TRUE)


# Save widget manually
htmlwidgets::saveWidget(map@map, "Outputs/cmr_output/raw_prevalence_map.html", selfcontained = FALSE)

# Capture screenshot with webshot2
webshot2::webshot("Outputs/cmr_output/raw_prevalence_map.html",
                  "Outputs/cmr_output/raw_prevalence_map.png", vwidth = 1200, vheight = 900)

plot(st_geometry(map_2011_cmr_sf), pch = 16, cex = 0.5, main = "2011 Survey Points — Cameroon")
```


# Check for multicollinearity

```{r}
cmr_pf_data_2011 <- st_read("data/cmr_data/benchmarking_spatial_data_cmr_sf.gpkg")
grid <- st_read("data/cmr_data/cmr_map_grid.gpkg")
```


```{r}
num_vars <- cmr_pf_data_2011 %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "rainfall", "temperature", "elevation", "vegetation_index", "pop_density", 
  "dist_to_water", "landcover_water", "landcover_trees", "landcover_builts", 
  "landcover_cropland", "landcover_wetland", "ITN_2011", "effective_treat"))

M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  positive / examined ~ rainfall + temperature + elevation +
    vegetation_index + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland + ITN_2011 + effective_treat,
  data = cmr_pf_data_2011
)

vif(fit)

fit1 <- lm(
  positive / examined ~ rainfall + temperature + pop_density + dist_to_water + landcover_water + landcover_trees +
  landcover_builts + landcover_cropland + landcover_wetland + ITN_2011 + effective_treat + vegetation_index,
  data = cmr_pf_data_2011
)

vif(fit1)
```
We can see that elevation and temperature are highly correlated r = 0.8. So in the first place we will continue with temperature is more directly interpretable for malaria biology. We also notice a strong correlation between land cover (trees) and vegetation index. Here we will also go with vegetation index because it is also one of the covariates than is often use for malaria spatial modelling for it biological interpretability.


# Autocorrelation check
```{r}
# create neighbors based on k nearest neighbors (e.g. k = 4)
coords <- st_coordinates(cmr_pf_data_2011)
nb <- knn2nb(knearneigh(coords, k = 4))
lw <- nb2listw(nb, style = "W")

# Moran's I on prevalence
moran.test(cmr_pf_data_2011$pr, lw)
```
- Our malaria prevalence data is strongly spatially clustered a spatial model is absolutely required.

# Variogram
```{r}
map_2011_cmr_sf_sp <- cmr_pf_data_2011 %>% st_transform(3857)

vgm_emp <- variogram(pr ~ 1, data = map_2011_cmr_sf_sp)   # prevalence vs distance
plot(vgm_emp)

vgm_mod <- fit.variogram(
  vgm_emp,
  model = vgm("Mat", nugget = 0.025, psill = 0.05, range = 100000, kappa = 0.55)
)
vgm_mod

plot(vgm_emp, vgm_mod)
```

```{r}

pts      <- st_transform(cmr_pf_data_2011, 3857)
shp_cmr_m  <- st_transform(shp_cmr, 3857)


mesh <- fmesher::fm_mesh_2d_inla(
  loc      = st_coordinates(pts),
  boundary = fm_as_segm(shp_cmr_m),
  max.edge = c(50e3, 200e3),   # 50 km near data, 200 km outer
  cutoff   = 15e3,             # min distance between points (15 km)
  offset   = c(50e3, 150e3)    # extend mesh 50–150 km beyond country
)

# Extract mesh triangles
mesh_tri <- mesh$graph$tv   # triangle vertices (indices)

# Convert each triangle to a polygon
triangles <- lapply(1:nrow(mesh_tri), function(i) {
  coords <- mesh$loc[mesh_tri[i, ], 1:2]
  coords <- rbind(coords, coords[1, ])  # close polygon
  st_polygon(list(coords))
})

mesh_sf <- st_sf(
  geometry = st_sfc(triangles),
  crs = st_crs(pts)   # use same CRS as your points
)


ggplot() +
  geom_sf(data = mesh_sf, fill = NA, color = "grey70", size = 0.2) +
  geom_sf(data = shp_cmr, fill = NA, color = "black", size = 0.4) +
  geom_sf(data = pts, color = "red", size = 2) +
  theme_minimal() +
  labs(title = "INLA Mesh over Country Boundary")

```

```{r}
spde <- inla.spde2.pcmatern(
  mesh = mesh,
  alpha = 2,
  prior.range = c(90e3, 0.5),    # P(range < 120 km) = 0.5
  prior.sigma = c(0.22, 0.5)      # weakly informative
)

A_est <- inla.spde.make.A(mesh, loc = st_coordinates(pts))


#1. Variables to be scale
vars_to_scale <- c("rainfall", "temperature", "pop_density", 
                   "dist_to_water", "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(pts[[var]], na.rm = TRUE)
  sd_val   <- sd(pts[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  pts[[var]] <- (pts[[var]] - mean_val) / sd_val
}



stk_est <- inla.stack(
  data = list(
    y       = pts$positive,
    Ntrials = pts$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
    intercept        = 1,
    rainfall         = pts$rainfall,
    temperature      = pts$temperature,
    pop_density      = pts$pop_density,
    dist_to_water    = pts$dist_to_water,
    lc_water         = pts$landcover_water,
    lc_trees         = pts$landcover_trees,
    lc_built         = pts$landcover_builts,
    lc_wetland       = pts$landcover_wetland,
    lc_cropland      = pts$landcover_cropland
    )
  ),
  tag = "est"
)


stk_est_intervention <- inla.stack(
  data = list(
    y       = pts$positive,
    Ntrials = pts$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
    intercept        = 1,
    rainfall         = pts$rainfall,
    temperature      = pts$temperature,
    pop_density      = pts$pop_density,
    dist_to_water    = pts$dist_to_water,
    lc_water         = pts$landcover_water,
    lc_trees         = pts$landcover_trees,
    lc_built         = pts$landcover_builts,
    lc_wetland       = pts$landcover_wetland,
    lc_cropland      = pts$landcover_cropland,
    ITN              = pts$ITN_2011,
    effective_treat  = pts$effective_treat
    )
  ),
  tag = "est"
)

```

# Build the model

```{r}

formula <- y ~ -1 + intercept +
  rainfall +
  temperature +
  pop_density +
  dist_to_water +
  lc_water +
  lc_cropland+
  lc_wetland +
  lc_built +
  lc_trees  +
  f(spatial, model = spde)


formula_intervention <- y ~ -1 + intercept +
  rainfall +
  temperature +
  pop_density +
  dist_to_water +
  lc_water +
  lc_cropland+
  lc_wetland +
  lc_built +
  lc_trees  +
  ITN +
  effective_treat+
  f(spatial, model = spde)


res <- inla(
  formula,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_est),
  control.predictor = list(
    A = inla.stack.A(stk_est),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(
    dic = TRUE,
    waic = TRUE,
    cpo = TRUE
  )
)


res_intervention <- inla(
  formula_intervention,
  family = "binomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_est_intervention),
  control.predictor = list(
    A = inla.stack.A(stk_est_intervention),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 1
  ),
  control.compute = list(
    dic = TRUE,
    waic = TRUE,
    cpo = TRUE
  )
)

res$summary.fixed

res$summary.hyperpar


res$dic$dic
res$waic$waic

res_intervention$summary.fixed
res_intervention$summary.hyperpar
```

#checking for overdispersion
```{r}
idx_est <- inla.stack.index(stk_est, "est")$data

fitted_est <- res$summary.fitted.values[idx_est, "mean"]

y <- pts$positive
n <- pts$examined

pearson_res <- (y - n * fitted_est) / sqrt(n * fitted_est * (1 - fitted_est))

n_obs <- length(y)
p_eff <- res$dic$p.eff   # effective number of parameters

phi <- sum(pearson_res^2) / (n_obs - p_eff)
phi

```

# Model diagnostic
```{r}
hist(res_intervention$cpo$pit,
     breaks = 20,
     main = "PIT Histogram",
     xlab = "PIT")

idx_est <- inla.stack.index(stk_est_intervention, "est")$data

fitted_est <- res_intervention$summary.fitted.values[idx_est, "mean"]

df_resid <- as.data.frame(cbind(
  observed = pts$positive / pts$examined,
  fitted   = fitted_est
))

pts$resid <- (df_resid$observed - df_resid$fitted) / sqrt(df_resid$fitted * (1 - df_resid$fitted) / pts$examined)

coords <- st_coordinates(pts)

nb <- dnearneigh(coords, d1 = 0, d2 = 28000)   # 100 km neighbours
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

moran.test(pts$resid,
           lw,
           alternative = "two.sided",
           zero.policy = TRUE)


plot(df_resid$observed, df_resid$fitted,
       xlab = "Observed prevalence",
       ylab = "Fitted prevalence")
abline(0, 1, col = "red")

```

The beta‑binomial SPDE model shows strong but short‑range spatial dependence (posterior range ~ 28 km, σ ~ 1.23). The PIT histogram indicates mild right‑skew, reflecting slight underestimation of very high observed prevalences, but overall calibration is acceptable. Residual Moran’s I is significantly negative (I ≈ –0.07, p ≈ 0.02), consistent with localized over‑ and under‑prediction patterns produced by the short spatial range. Compared to the binomial model, the beta‑binomial likelihood reduces residual spatial autocorrelation and improves PIT calibration. Overall, the model fits well, with only minor miscalibration in the upper tail.

# Build grid for prediction and the prediction model

```{r}
# Build the A‑matrix for prediction locations
A_pred <- inla.spde.make.A(
  mesh = mesh,
  loc = st_coordinates(grid)
)


# Scale the prediction grid using the EXACT same parameters as in the fitting model
# (This ensures a grid cell is measured relative to our 36 observed points)
for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid[[var]] <- (grid[[var]] - mean_val) / sd_val
}



#Build the prediction stack

stk_pred <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid$rainfall,
      temperature      = grid$temperature,
      pop_density      = grid$pop_density,
      dist_to_water    = grid$dist_to_water,
      lc_cropland      = grid$landcover_cropland,
      lc_water         = grid$landcover_water,
      lc_wetland       = grid$landcover_wetland,
      lc_built         = grid$landcover_builts,
      lc_trees         = grid$landcover_trees
    )
  ),
  tag = "pred"
)


stk_pred_intervention <- inla.stack(
  data = list(y = NA, Ntrials = NA),
  A    = list(A_pred, 1),
  effects = list(
    spatial = 1:spde$n.spde,
    data.frame(
      intercept        = 1,
      rainfall         = grid$rainfall,
      temperature      = grid$temperature,
      pop_density      = grid$pop_density,
      dist_to_water    = grid$dist_to_water,
      lc_cropland      = grid$landcover_cropland,
      lc_water         = grid$landcover_water,
      lc_wetland       = grid$landcover_wetland,
      lc_built         = grid$landcover_builts,
      lc_trees         = grid$landcover_trees,
      ITN              = grid$ITN_2011,
      effective_treat  = grid$effective_treat
    )
  ),
  tag = "pred"
)


# Combine estimation + prediction stacks
stk_full <- inla.stack(stk_est, stk_pred)

stk_full_intervention <- inla.stack(stk_est_intervention, stk_pred)


# Fit the model again with the full stack
res_pred <- inla(
  formula,
  family = "betabinomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full),
  control.predictor = list(
    A = inla.stack.A(stk_full),
    compute = TRUE,
    link = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_pred_intervention <- inla(
  formula,
  family = "betabinomial",
  Ntrials = Ntrials,
  data = inla.stack.data(stk_full_intervention),
  control.predictor = list(
    A = inla.stack.A(stk_full_intervention),
    compute = TRUE,
    link = 1
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


set.seed(123)
samples <- inla.posterior.sample(100, res_pred_intervention)

idx_pred <- inla.stack.index(stk_full_intervention, tag = "pred")$data

pred_mat <- sapply(samples, function(s) {
  lp <- s$latent[idx_pred]                   # linear predictor for grid cells
  plogis(lp)                                 # convert logit  prevalence
})


grid$prev_mean   <- rowMeans(pred_mat)
grid$prev_sd     <- apply(pred_mat, 1, sd)
grid$prev_q025   <- apply(pred_mat, 1, quantile, 0.025)
grid$prev_q975   <- apply(pred_mat, 1, quantile, 0.975)


v <- vect(grid)

template <- rast(
  ext = ext(v),
  resolution = 5000,
  crs = crs(v)
)


grid_rast <- rasterize(
  v,
  template,
  field = c("prev_mean", "prev_q025", "prev_q975")
)


cmr_prev_1 <- ggplot(shp_cmr) +
  geom_spatraster(data = grid_rast,
                  mapping = aes(fill = prev_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2011 with intervention")

ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_test_pf_all_age_with_intervention.pdf",
       plot = cmr_prev_1, width = 6, height = 4, dpi = 300)

pfpr_2011_cmr <- rast("data/cmr_data/MAP_modelled_prevalence/202608_Global_Pf_Parasite_Rate_CMR_2011.tiff") 

names(pfpr_2011_cmr) <- c("Prev_mean", " ", "lci", "uci")

cmr_prev_2 <- ggplot(shp_cmr) +
  geom_spatraster(data = pfpr_2011_cmr,
                  mapping = aes(fill = Prev_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f (2-10years)")+
  theme_void()+
  labs(title = "annual prevalence of P. f (2-10years) in 2011")

ggsave("Outputs/cmr_output/MAP_pr_modeled/2011_cmr_pr_map_pf_2_10.png",
       plot = cmr_prev_2, width = 6, height = 4, dpi = 300)


cmr_vect <- vect(shp_cmr)

prev_dat_cmr_2011 <- terra::extract(pfpr_2011_cmr, cmr_vect)

prev_dat_cmr_2011_test <- terra::extract(grid_rast, vect(shp_cmr))

prev_summary <- prev_dat_cmr_2011 %>%
  group_by(ID) %>%
  summarise(
    PfPR_2_10   = mean(Prev_mean, na.rm = TRUE),
    PfPR_2_10_l = mean(lci, na.rm = TRUE),
    PfPR_2_10_u = mean(uci, na.rm = TRUE)
  )

prev_summary_test <- prev_dat_cmr_2011_test %>%
  group_by(ID) %>%
  summarise(
    PfPR_mean   = mean(prev_mean, na.rm = TRUE),
    PfPR_l      = mean(prev_q025, na.rm = TRUE),
    PfPR_u      = mean(prev_q975, na.rm = TRUE)
  )

prev_dat_cmr_2011_sf <- shp_cmr %>% 
  mutate(PfPR_2_10   =  prev_summary$PfPR_2_10,
         PfPR_2_10_l = prev_summary$PfPR_2_10_l,
         PfPR_2_10_u = prev_summary$PfPR_2_10_u)

prev_dat_cmr_2011_sf_test <- shp_cmr %>% 
  mutate(PfPR_mean   =  prev_summary_test$PfPR_mean,
         PfPR_l      = prev_summary_test$PfPR_l,
         PfPR_u      = prev_summary_test$PfPR_u)


# Conversion of the prevalence to all age group

n <- nrow(prev_dat_cmr_2011_sf)

prev_dat_cmr_2011_sf$PR_all_age <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(100, n),
  parameters   = "Pf_Smith2007"
)

prev_dat_cmr_2011_sf$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10_u,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")

prev_dat_cmr_2011_sf$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = prev_dat_cmr_2011_sf$PfPR_2_10_l,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")


breaks_prev <- c(0, 0.1, 0.2, 0.4, 0.5, 0.6, 0.8, 1)
labels_prev <- c(
  "0–10%", "10–20%", "20–40%", "40–50%",
  "50–60%", "60–80%", "80–100%"
)

colors_prev <- rev(brewer.pal(7, "RdYlGn"))



cmr_prev_3 <- ggplot(prev_dat_cmr_2011_sf) +
  geom_sf(aes(fill = cut(
    PR_all_age,
    breaks = breaks_prev,
    labels = labels_prev,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev,
    name = "Estimated prevalence"
  ) +
  labs(title = "Estimated P. falciparum in prevalence\n in 2011 from MAP") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15)
  )

ggsave("Outputs/cmr_output/MAP_pr_modeled/2011_cmr_pr_map_pf_all_age_admin2.png",
       plot = cmr_prev_3, width = 6, height = 4, dpi = 300)



cmr_prev_4 <- ggplot(prev_dat_cmr_2011_sf_test) +
  geom_sf(aes(fill = cut(
    PfPR_mean,
    breaks = breaks_prev,
    labels = labels_prev,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev,
    name = "Estimated prevalence"
  ) +
  labs(title = "Estimated P. falciparum in prevalence\n in 2011 usind our model (interventions included)") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15)
  )


ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_pf_all_age_admin2_with_intervention.png",
       plot = cmr_prev_4, width = 6, height = 4, dpi = 300)

prev_dat_cmr_2011_sf_test$PfPR_map <- prev_dat_cmr_2011_sf$PR_all_age
prev_dat_cmr_2011_sf_test$PfPR_map_u <- prev_dat_cmr_2011_sf$PR_all_age_U
prev_dat_cmr_2011_sf_test$PfPR_map_l <- prev_dat_cmr_2011_sf$PR_all_age_L

prev_dat_cmr_2011_sf_test <- prev_dat_cmr_2011_sf_test %>% 
  mutate(diff_PfPR = PfPR_mean - PfPR_map)

breaks_prev_diff <- c(-0.15, -0.05, 0.05, 0.1, 0.15, 0.2, 0.25, 30)
labels_prev_diff <- c(
  "-0.15%<-0.05%", "-0.05%<0.05%", "0.05%<0.10%", "0.10%<0.15%",
  "0.15%<0.20%", "0.20%<0.25%", "0.25<0.30%"
)

colors_prev_diff <- rev(brewer.pal(7, "RdYlGn"))

cmr_prev_5 <- ggplot(prev_dat_cmr_2011_sf_test) +
  geom_sf(aes(fill = cut(
    diff_PfPR,
    breaks = breaks_prev_diff,
    labels = labels_prev_diff,
    include.lowest = TRUE
  )), color = NA) +
  scale_fill_manual(
    values = colors_prev_diff,
    name = "Difference in prevalence"
  ) +
  labs(title = "Difference in prevalence with intervention (2011)") +
  theme_void()+
  theme(
    legend.position = "right",
    plot.title = element_text(size = 15),
    legend.title = element_text(size = 15))


prev_dat_cmr_2011_sf_test$diff_class <- cut(
  prev_dat_cmr_2011_sf_test$diff_PfPR,
  breaks = breaks_prev_diff,
  labels = labels_prev_diff,
  include.lowest = TRUE
)

map1 <- mapview(
  prev_dat_cmr_2011_sf_test,
  zcol = "diff_class",
  col.regions = colors_prev_diff,
  layer.name = "Difference in prevalence (2011)",
  legend = TRUE
  )
  

mapshot(map1, 
        file = "Outputs/cmr_output/Pf_pr_estimated/difference_prev_MAP_vsour_model.html",
        selfcontained = T)

ggsave("Outputs/cmr_output/Pf_pr_estimated/2011_cmr_pr_diff_pf_all_age_admin2_with_intervention.png",
       plot = cmr_prev_5, width = 6, height = 4, dpi = 300)

```


# Extract MAP data

```{r}
#cor(cmr_data_pev_map$prev_mean, cmr_map_pr_pf_2_10$PR_all_age, use = "complete.obs")

cmr_data_pev_map <- cmr_data_pev_map %>% 
  mutate(prev_mean = case_when(is.na(prev_mean) ~ mean(prev_mean, na.rm = TRUE),
                               TRUE ~ prev_mean))

cmr_data_pev_map$PR_all_age <- cmr_map_pr_pf_2_10$PR_all_age
#-----------------------------------------------------------
# 1. BASIC AGREEMENT METRICS
#-----------------------------------------------------------

# Bias
bias <- mean(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age, na.rm = TRUE)

# MAE
mae <- mean(abs(cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age), na.rm = TRUE)

# RMSE
rmse <- sqrt(mean((cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age)^2, na.rm = TRUE))

# Pearson correlation
pearson_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "pearson")

# Spearman rank correlation
spearman_cor <- cor(cmr_data_pev_map$prev_mean, cmr_data_pev_map$PR_all_age, method = "spearman")

# #-----------------------------------------------------------
# # 2. SPATIAL AGREEMENT METRICS
# #-----------------------------------------------------------
# 
# # Create spatial weights
# nb <- poly2nb(cmr_data_pev_map)
# lw <- nb2listw(nb, style = "W")
# 
# # Spatial cross-correlation (correlation of lagged values)
# lag_model <- lag.listw(lw, cmr_data_pev_map$prev_mean)
# lag_map   <- lag.listw(lw, cmr_data_pev_map$PR_all_age)
# spatial_cross_cor <- cor(lag_model, lag_map, use = "complete.obs")
# 
# # Moran's I of residuals
# residuals <- cmr_data_pev_map$prev_mean - cmr_data_pev_map$PR_all_age
# moran_residuals <- moran.test(residuals, lw)
# 
# #-----------------------------------------------------------
# # 3. SEMIVARIOGRAM COMPARISON
# #-----------------------------------------------------------
# 
# library(gstat)
# 
# # Convert sf to sp for gstat
# admin2_sp <- as(cmr_data_pev_map, "Spatial")
# 
# vg_model <- variogram(prev_mean ~ 1, admin2_sp)
# vg_map   <- variogram(PR_all_age ~ 1, admin2_sp)
# 
# #-----------------------------------------------------------
# # 4. DISTRIBUTIONAL COMPARISON
# #-----------------------------------------------------------
# 
# # Q–Q plot data
# qq_data <- data.frame(
#   model = quantile(cmr_data_pev_map$prev_mean, probs = seq(0,1,0.01), na.rm = TRUE),
#   map   = quantile(cmr_data_pev_map$PR_all_age, probs = seq(0,1,0.01), na.rm = TRUE)
# )
# 
# # Distribution overlap (simple measure)
# dens_model <- density(cmr_data_pev_map$prev_mean, na.rm = TRUE)
# dens_map   <- density(cmr_data_pev_map$PR_all_age, na.rm = TRUE)
# 
# # Overlap integral
# x_common <- seq(min(dens_model$x, dens_map$x), max(dens_model$x, dens_map$x), length.out = 1000)
# y_model  <- approx(dens_model$x, dens_model$y, x_common)$y
# y_map    <- approx(dens_map$x, dens_map$y, x_common)$y
# distribution_overlap <- sum(pmin(y_model, y_map)) * (x_common[2] - x_common[1])
# 
# #-----------------------------------------------------------
# # 5. CLASSIFICATION AGREEMENT (TRANSMISSION STRATA)
# #-----------------------------------------------------------
# 
# # Define strata
# cut_points <- c(0, 0.05, 0.30, 1)
# labels <- c("low", "medium", "high")
# 
# cmr_data_pev_map <- cmr_data_pev_map %>%
#   mutate(
#     class_model = cut(prev_mean, breaks = cut_points, labels = labels, include.lowest = TRUE),
#     class_map   = cut(PR_all_age, breaks = cut_points, labels = labels, include.lowest = TRUE)
#   )
# 
# # Confusion matrix
# confusion_matrix <- table(cmr_data_pev_map$class_model, cmr_data_pev_map$class_map)
# 
# # Kappa statistic
# library(irr)
# kappa_stat <- kappam.fleiss(as.data.frame(confusion_matrix))
# 
# #-----------------------------------------------------------
# # 6. ADMIN2 AGGREGATED COMPARISONS
# #-----------------------------------------------------------
# 
# # Admin2 mean prevalence (already aggregated)
# admin2_comparison <- cmr_data_pev_map %>%
#   st_drop_geometry() %>%
#   select(admin2, pfpr_model, pfpr_map_all_age)
# 
# # Rank correlation of districts
# rank_cor <- cor(
#   rank(cmr_data_pev_map$pfpr_model),
#   rank(cmr_data_pev_map$pfpr_map_all_age),
#   method = "spearman"
# )
# 
# #-----------------------------------------------------------
# # 7. POPULATION-WEIGHTED PREVALENCE (if pop column exists)
# #-----------------------------------------------------------
# 
# if("population" %in% colnames(cmr_data_pev_map)) {
#   pwpr_model <- sum(cmr_data_pev_map$pfpr_model * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
#   
#   pwpr_map <- sum(cmr_data_pev_map$pfpr_map_all_age * cmr_data_pev_map$population, na.rm = TRUE) /
#     sum(cmr_data_pev_map$population, na.rm = TRUE)
# }
# 
# #-----------------------------------------------------------
# # 8. UNCERTAINTY COMPARISON (if your model has CI)
# #-----------------------------------------------------------
# 
# if("model_lower" %in% colnames(cmr_data_pev_map) &&
#    "model_upper" %in% colnames(cmr_data_pev_map)) {
#   
#   coverage_probability <- mean(
#     cmr_data_pev_map$pfpr_map_all_age >= cmr_data_pev_map$model_lower &
#       cmr_data_pev_map$pfpr_map_all_age <= cmr_data_pev_map$model_upper,
#     na.rm = TRUE
#   )
# }
# 
# #-----------------------------------------------------------
# # PRINT RESULTS
# #-----------------------------------------------------------
# 
# list(
#   bias = bias,
#   mae = mae,
#   rmse = rmse,
#   pearson_cor = pearson_cor,
#   spearman_cor = spearman_cor,
#   spatial_cross_cor = spatial_cross_cor,
#   moran_residuals = moran_residuals,
#   semivariogram_model = vg_model,
#   semivariogram_map = vg_map,
#   distribution_overlap = distribution_overlap,
#   confusion_matrix = confusion_matrix,
#   kappa_stat = kappa_stat,
#   admin2_comparison = admin2_comparison,
#   rank_cor = rank_cor
# )
# 

```

# Benchmark at the continental level
```{r}
GPS_CI <- st_read("data/Africa_data/Benchmark data/CIGE81FL/CIGE81FL.shp")
GPS_ET <- st_read("data/Africa_data/Benchmark data/ETGE81FLSR/ETGE81FLSR.shp")
GPS_NG <- st_read("data/Africa_data/Benchmark data/NGGE81FL/NGGE81FL.shp")
GPS_NI <- st_read("data/Africa_data/Benchmark data/NIGE81FL/NIGE81FL.shp")
GPS_ML <- st_read("data/Africa_data/Benchmark data/MLGE81FL/MLGE81FL.shp")
GPS_MR <- st_read("data/Africa_data/Benchmark data/MRGE71FL/MRGE71FL.shp")
GPS_MD <- st_read("data/Africa_data/Benchmark data/MDGE81FL/MDGE81FL.shp")
GPS_GN <- st_read("data/Africa_data/Benchmark data/GNGE81FL/GNGE81FL.shp")
GPS_SN <- st_read("data/Africa_data/Benchmark data/SNGE8IFL/SNGE8IFL.shp")
GPS_BF <- st_read("data/Africa_data/Benchmark data/BFGE81FL/BFGE81FL.shp")
GPS_GA <- st_read("data/Africa_data/Benchmark data/GAGE71FL/GAGE71FL.shp")

africa_pf_PR <- read.csv("data/Africa_data/Benchmark data/raw_pr_map.csv")
shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp")

GPS_CI_coords_2021 <- GPS_CI %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_CI)[, 1],
    latitude  = st_coordinates(GPS_CI)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Côte d'Ivoire") %>% 
  filter(latitude != 0 & longitude != 0)

GPS_ET_coords_2021 <- GPS_ET %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_ET)[, 1],
    latitude  = st_coordinates(GPS_ET)[, 2]
  ) %>%
  dplyr::select(SPAFACID, longitude, latitude) %>%
  rename(DHSCLUST = SPAFACID) %>% 
  mutate(country = "Ethiopia") %>% 
  filter(latitude != 0 & longitude != 0)

GPS_NI_coords_2021 <- GPS_NI %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_NI)[, 1],
    latitude  = st_coordinates(GPS_NI)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Niger") %>% 
  filter(latitude != 0 & longitude != 0)

GPS_NG_coords_2021 <- GPS_NG %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_NG)[, 1],
    latitude  = st_coordinates(GPS_NG)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Nigeria") %>%
  filter(latitude != 0 & longitude != 0)


GPS_ML_coords_2021 <- GPS_ML %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_ML)[, 1],
    latitude  = st_coordinates(GPS_ML)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Mali") %>%
  filter(latitude != 0 & longitude != 0)

GPS_MR_coords_2021 <- GPS_MR %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_MR)[, 1],
    latitude  = st_coordinates(GPS_MR)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Mauritania") %>%
  filter(latitude != 0 & longitude != 0)

GPS_MD_coords_2021 <- GPS_MD %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_MD)[, 1],
    latitude  = st_coordinates(GPS_MD)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Madagascar") %>%
  filter(latitude != 0 & longitude != 0)

GPS_GN_coords_2021 <- GPS_GN %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_GN)[, 1],
    latitude  = st_coordinates(GPS_GN)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Guinea") %>%
  filter(latitude != 0 & longitude != 0)

GPS_SN_coords_2021 <- GPS_SN %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_SN)[, 1],
    latitude  = st_coordinates(GPS_SN)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Senegal") %>%
  filter(latitude != 0 & longitude != 0)

GPS_BF_coords_2021 <- GPS_BF %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_BF)[, 1],
    latitude  = st_coordinates(GPS_BF)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Burkina Faso") %>%
  filter(latitude != 0 & longitude != 0)

GPS_GA_coords_2021 <- GPS_GA %>%
  st_drop_geometry() %>%
  mutate(
    longitude = st_coordinates(GPS_GA)[, 1],
    latitude  = st_coordinates(GPS_GA)[, 2]
  ) %>%
  dplyr::select(DHSCLUST, longitude, latitude) %>%
  mutate(country = "Gabon") %>%
  filter(latitude != 0 & longitude != 0)


africa_pf_PR_with_coord_2021 <- africa_pf_PR %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  filter(year_start == 2021 | year_end == 2021)%>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

africa_pf_PR_without_coord_2021 <- africa_pf_PR %>% 
  filter(is.na(longitude) & is.na(latitude)) %>% 
  dplyr::select(-c(longitude, latitude))%>% 
  filter(year_start == 2021 | year_end == 2021)



africa_pf_PR_without_coord_2021 <- africa_pf_PR_without_coord_2021 %>%
  mutate(DHSCLUST = as.integer(substr(dhs_id, 8, 16)))


raw_pr_CI_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_CI_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_NI_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_NI_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_NG_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_NG_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_ML_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_ML_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_MR_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_MR_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_MD_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_MD_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_GN_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_GN_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_SN_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_SN_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_BF_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_BF_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

raw_pr_GA_2021 <- africa_pf_PR_without_coord_2021 %>% 
  left_join(GPS_GA_coords_2021) %>% 
  filter(!is.na(longitude) & !is.na(latitude)) %>% 
  select(country, year_start, year_end, lower_age, upper_age, species, examined,
         positive, pr, method, longitude, latitude)

africa_pf_PR_with_coord_join_2021 <- raw_pr_CI_2021 %>% 
  bind_rows(raw_pr_NI_2021) %>% 
  bind_rows(raw_pr_NG_2021) %>% 
  bind_rows(raw_pr_ML_2021) %>% 
  bind_rows(raw_pr_MR_2021) %>% 
  bind_rows(raw_pr_MD_2021) %>% 
  bind_rows(raw_pr_GN_2021) %>% 
  bind_rows(raw_pr_SN_2021) %>% 
  bind_rows(raw_pr_BF_2021) %>% 
  bind_rows(raw_pr_GA_2021) %>% 
  bind_rows(africa_pf_PR_with_coord_2021) %>% 
  filter(species == "P. falciparum") %>% 
  group_by(longitude, latitude) %>% 
  mutate(examined = sum(examined, na.rm = T),
         positive = sum(positive, na.rm = T),
         pr       = positive/examined) %>% 
  ungroup() %>% 
  distinct()

write_csv(africa_pf_PR_with_coord_join_2021,
          file = "data/Africa_data/africa_MAP_full_raw_prevalence_2021.csv")
```

# Visualization of the observed data

```{r}
africa_pf_PR_with_coord_join_2021 <- read.csv("data/Africa_data/africa_MAP_full_raw_prevalence_2021.csv")

africa_pf_PR_with_coord_join_2021_sf <- st_as_sf(africa_pf_PR_with_coord_join_2021,
                                              coords = c("longitude", "latitude"),
                                              crs = 4326)


breaks_pr <- c(0, 0.05, 0.10, 0.20, 0.40, 0.60, 1)
labels_pr <- c(
  "0–5%", "5–10%", "10–20%", "20–40%", "40–60%", ">60%"
)

colors_pr <- c(
  "#ffffcc",  # pale yellow (very low)
  "#ffeda0",  # light yellow
  "#feb24c",  # orange
  "#fd8d3c",  # deep orange
  "#f03b20",  # red
  "#bd0026"   # dark red (very high)
)

africa_pf_PR_with_coord_join_2021_sf$pr_class <- cut(
  africa_pf_PR_with_coord_join_2021_sf$pr,
  breaks = breaks_pr,
  labels = labels_pr,
  include.lowest = TRUE
)


map1 <- mapview(shp_africa,
        col.regions = "white",
        alpha.regions = 0.2,
        color = "black") +
  mapview(africa_pf_PR_with_coord_join_2021_sf,
          zcol = "pr_class",             # categorical variable
          cex = "examined",
          col.regions = colors_pr,       # one color per class
          layer.name = "P. malariae prevalence",
          legend = TRUE)


# Save widget manually
htmlwidgets::saveWidget(map1@map, "Outputs/africa_output/raw_prevalence_map.html", selfcontained = TRUE)

# Capture screenshot with webshot2
webshot2::webshot("Outputs/africa_output/raw_prevalence_map.html",
                  "Outputs/africa_output/raw_prevalence_map.png", vwidth = 1200, vheight = 900)

```


```{r}
sampled_data_join_sf <- st_read("data/Africa_data/africa_pf_PR_with_coord_join_2021_sf.gpkg")
grid_sf_africa <- st_read("data/Africa_data/grid_data_for_prediction_sf.gpkg")


shp_africa <- st_read("data/Africa_data/shp_africa/Africa_simplified.shp") %>% st_transform(3857)

Africa_union <- st_union(shp_africa )

sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

sampled_data_join_sf_sp     <- as(sampled_data_join_sf, "Spatial")
Africa_union_sp <- as(Africa_union, "Spatial")

coords_Africa <- coordinates(sampled_data_join_sf_sp)


mesh_Africa <- inla.mesh.2d(
  loc      = coords_Africa,
  boundary = inla.sp2segment(Africa_union_sp),
  max.edge = c(500e3, 1000e3),   
  cutoff   = 200e3,
  offset   = c(100e3, 2000e3))

# Extract triangle vertex indices
tri <- mesh_Africa$graph$tv


# Build edges from triangles
edges_list <- list()

for (i in 1:nrow(tri)) {
  idx <- tri[i, ]
  # Each triangle has 3 edges
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(1,2)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(2,3)], 1:2])
  edges_list[[length(edges_list) + 1]] <- st_linestring(mesh_Africa$loc[idx[c(3,1)], 1:2])
}

mesh_edges_sf <- st_as_sf(st_sfc(edges_list, crs = 3857))

coords_df_Africa <- as.data.frame(coords_Africa)
colnames(coords_df_Africa) <- c("X", "Y")

ggplot() +
  geom_sf(data = mesh_edges_sf, fill = NA, color = "grey60", size = 0.3) +
  geom_sf(data = Africa_union, fill = NA, color = "black", size = 0.8) +
  geom_point(data = coords_df_Africa, aes(X, Y), color = "red", size = 1.5) +
  theme_minimal() +
  ggtitle("Mesh for Africa with Boundary")
```

```{r}
num_vars <- sampled_data_join_sf %>% 
  st_drop_geometry() %>% 
  dplyr::select(c(
  "annual_rainfall_2021", "annual_temperature_2021", "elevation", "vegetation_index",
  "pop_density_2020", "dist_to_water", "lc_water", "lc_trees", "annual_ITN_usage_2021",
  "lc_built", "lc_cropland", "lc_wetland"))


M <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(
  M,
  method = "color",
  type = "upper",
  addCoef.col = "black",   
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7
)

fit <- lm(
  pr ~ annual_rainfall_2021 + 
    annual_temperature_2021 + elevation + vegetation_index + pop_density_2020 + 
    dist_to_water + lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021,
  data = sampled_data_join_sf
)

vif(fit)

```

# Variogram
```{r}
sampled_data_join_sf <- sampled_data_join_sf %>% st_transform(3857)

vgm_emp <- variogram(pr ~ 1, data = sampled_data_join_sf)   # prevalence vs distance
plot(vgm_emp)

vgm_mod <- fit.variogram(
  vgm_emp,
  model = vgm("Mat", nugget = 0.04, psill = 0.065, range = 500000, kappa = 0.55)
)
vgm_mod

plot(vgm_emp, vgm_mod)
```


```{r}
#| label: Overdispersion

 # scaling of the varibles

sampled_data_join_sf_scale <- sampled_data_join_sf %>% 
  mutate(annual_rainfall_2021    = as.numeric(scale(annual_rainfall_2021)),
         annual_temperature_2021 = as.numeric(scale(annual_temperature_2021)),
         pop_density_2020        = as.numeric(scale(pop_density_2020)),
         dist_to_water           = as.numeric(scale(dist_to_water)),
         elevation               = as.numeric(scale(elevation)),
         vegetation_index        = as.numeric(scale(vegetation_index)))


glm_binom1 <- glm(
  cbind(positive, examined - positive) ~ 1,
  family = "binomial",
  data = sampled_data_join_sf_scale
)

pearson_res1 <- residuals(glm_binom1, type = "pearson")

phi <- sum(pearson_res1^2) / glm_binom1$df.residual

phi

glm_binom2 <- glm(
  cbind(positive, examined - positive) ~
    annual_rainfall_2021 + pop_density_2020 + dist_to_water + lc_water + lc_trees + lc_cropland +
    lc_built + lc_wetland + annual_temperature_2021 + elevation + vegetation_index,
  family = "binomial",
  data = sampled_data_join_sf_scale
)

pearson_res2 <- residuals(glm_binom2, type = "pearson")

phi <- sum(pearson_res2^2) / glm_binom2$df.residual

phi
```

```{r}
spde_Africa <- inla.spde2.pcmatern(
  mesh = mesh_Africa,
  prior.range = c(500000, 0.5),
  prior.sigma = c(0.24, 0.5))

A_est <- inla.spde.make.A(mesh_Africa, loc = coords_Africa)

#1. Variables to be scale
vars_to_scale <- c("annual_rainfall_2021", "annual_temperature_2021", 
                   "pop_density_2020", "dist_to_water", "elevation")

# 2. Create a named list to store the training parameters (Means and SDs)
scaling_stats <- list()

# 3. Scale the observed data and extract the parameters
for (var in vars_to_scale) {
  # Calculate mean and standard deviation from the 36 observed points
  mean_val <- mean(sampled_data_join_sf[[var]], na.rm = TRUE)
  sd_val   <- sd(sampled_data_join_sf[[var]], na.rm = TRUE)
  
  # Store them for later use on the grid
  scaling_stats[[var]] <- c(mean = mean_val, sd = sd_val)
  
  # Apply the scaling to the observed data
  sampled_data_join_sf[[var]] <- (sampled_data_join_sf[[var]] - mean_val) / sd_val
}


```


```{r}
effects_fixed_tree <- data.frame(
  intercept                    = 1,
  annual_rainfall_2021         = sampled_data_join_sf$annual_rainfall_2021,
  annual_temperature_2021      = sampled_data_join_sf$annual_temperature_2021,
  elevation                    = sampled_data_join_sf$elevation,
  pop_density_2020             = sampled_data_join_sf$pop_density_2020,
  dist_to_water                = sampled_data_join_sf$dist_to_water,
  lc_water                     = sampled_data_join_sf$lc_water,
  lc_trees                     = sampled_data_join_sf$lc_trees,
  lc_built                     = sampled_data_join_sf$lc_built,
  lc_cropland                  = sampled_data_join_sf$lc_cropland,
  lc_wetland                   = sampled_data_join_sf$lc_wetland,
  annual_ITN_usage_2021        = sampled_data_join_sf$annual_ITN_usage_2021
)

effects_fixed_vegetation_index <- data.frame(
  intercept                    = 1,
  annual_rainfall_2021         = sampled_data_join_sf$annual_rainfall_2021,
  annual_temperature_2021      = sampled_data_join_sf$annual_temperature_2021,
  elevation                    = sampled_data_join_sf$elevation,
  vegetation_index            = sampled_data_join_sf$vegetation_index,
  pop_density_2020             = sampled_data_join_sf$pop_density_2020,
  dist_to_water                = sampled_data_join_sf$dist_to_water,
  lc_water                     = sampled_data_join_sf$lc_water,
  lc_built                     = sampled_data_join_sf$lc_built,
  lc_cropland                  = sampled_data_join_sf$lc_cropland,
  lc_wetland                   = sampled_data_join_sf$lc_wetland,
  annual_ITN_usage_2021        = sampled_data_join_sf$annual_ITN_usage_2021
)


stk_est_trees <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$positive,
    Ntrials = sampled_data_join_sf$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_tree
  ),
  tag = "est"
)


stk_est_vegetation_index <- inla.stack(
  data = list(
    y       = sampled_data_join_sf$positive,
    Ntrials = sampled_data_join_sf$examined
  ),
  A = list(A_est, 1),
  effects = list(
    spatial = 1:spde_Africa$n.spde,
    effects_fixed_vegetation_index
  ),
  tag = "est"
)


formula_trees <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 + annual_temperature_2021 + elevation + pop_density_2020 + dist_to_water +
  lc_water + lc_trees + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021 +
  f(spatial, model = spde_Africa)

formula_vegetation_index <- y ~ 
  -1 +
  intercept +
  annual_rainfall_2021 + annual_temperature_2021 + elevation + pop_density_2020 + dist_to_water +
  lc_water + vegetation_index + lc_built + lc_cropland + lc_wetland + annual_ITN_usage_2021 +
  f(spatial, model = spde_Africa)


res_trees <- inla(
  formula_trees,
  family = "betabinomial",
  data = inla.stack.data(stk_est_trees),
  Ntrials = inla.stack.data(stk_est_trees)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_trees),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)


res_vegetation <- inla(
  formula_vegetation_index,
  family = "betabinomial",
  data = inla.stack.data(stk_est_vegetation_index),
  Ntrials = inla.stack.data(stk_est_vegetation_index)$Ntrials,
  control.predictor = list(
    A = inla.stack.A(stk_est_vegetation_index),
    compute = TRUE
  ),
  control.fixed = list(
    mean = 0,
    prec = 0.2
  ),
  control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

```

```{r}
idx_est <- inla.stack.index(stk_est_vegetation_index, "est")$data

fitted_est <- res_vegetation$summary.fitted.values[idx_est, "mean"]

df_resid <- as.data.frame(cbind(
  observed = sampled_data_join_sf$pr,
  fitted   = fitted_est
))

sampled_data_join_sf$resid <- (df_resid$observed - df_resid$fitted) / sqrt(df_resid$fitted * (1 - df_resid$fitted) / sampled_data_join_sf$examined)

coords <- st_coordinates(sampled_data_join_sf)

nb <- dnearneigh(coords, d1 = 0, d2 = 50000)   # 100 km neighbours
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

moran.test(sampled_data_join_sf$resid,
           lw,
           alternative = "two.sided",
           zero.policy = TRUE)

plot(df_resid$observed, df_resid$fitted,
       xlab = "Observed prevalence",
       ylab = "Fitted prevalence")
  abline(0, 1, col = "red")
  
res_vegetation$summary.fixed

```


```{r}
for (var in vars_to_scale) {
  mean_val <- scaling_stats[[var]]["mean"]
  sd_val   <- scaling_stats[[var]]["sd"]
  
  grid_sf_africa[[var]] <- (grid_sf_africa[[var]] - mean_val) / sd_val
}


n_tiles <- 100
tile_id <- cut(seq_len(nrow(grid_sf_africa)),
               breaks = n_tiles,
               labels = FALSE)

pred_mean <- numeric(nrow(grid_sf_africa))
pred_sd   <- numeric(nrow(grid_sf_africa))
pred_q05  <- numeric(nrow(grid_sf_africa))
pred_q95  <- numeric(nrow(grid_sf_africa))

for(tile in seq_len(n_tiles)){
  cat("Tile", tile, "of", n_tiles, "\n")

  idx <- which(tile_id == tile)
  grid_tile <- grid_sf_africa[idx, ]

  A_pred <- inla.spde.make.A(
    mesh = mesh_Africa,
    loc = st_coordinates(grid_tile)
  )

  effects_pred <- data.frame(
    intercept                    = 1,
    annual_rainfall_2021         = grid_tile$annual_rainfall_2021,
    annual_temperature_2021      = grid_tile$annual_temperature_2021,
    elevation                    = grid_tile$elevation,
    vegetation_index             = grid_tile$vegetation_index,
    pop_density_2020             = grid_tile$pop_density_2020,
    dist_to_water                = grid_tile$dist_to_water,
    lc_water                     = grid_tile$lc_water,
    lc_built                     = grid_tile$lc_built,
    lc_cropland                  = grid_tile$lc_cropland,
    lc_wetland                   = grid_tile$lc_wetland,
    annual_ITN_usage_2021        = grid_tile$annual_ITN_usage_2021
    )

  stk_pred <- inla.stack(
    data = list(y = NA, Ntrials = NA),
    A = list(A_pred, 1),
    effects = list(
      spatial = 1:spde_Africa$n.spde,
      effects_pred
    ),
    tag = "pred"
  )

  stk_full <- inla.stack(stk_est_vegetation_index, stk_pred)

  res_tile <- inla(
    formula_vegetation_index,
    family = "betabinomial",
    data = inla.stack.data(stk_full),
    Ntrials = inla.stack.data(stk_full)$Ntrials,

    control.predictor = list(
      A = inla.stack.A(stk_full),
      compute = TRUE,
      link = 1
    ),

    control.mode = list(
      theta = res_vegetation$mode$theta,
      restart = FALSE
    ),

    control.compute = list(config = TRUE)
  )

  pred_idx <- inla.stack.index(stk_full, "pred")$data

  eta_mat <- inla.posterior.sample.eval(
    function() APredictor[pred_idx],
    inla.posterior.sample(200, res_tile)
  )

  prob_mat <- plogis(eta_mat)

  pred_mean[idx] <- rowMeans(prob_mat)
  pred_sd[idx]   <- apply(prob_mat, 1, sd)
  pred_q05[idx]  <- apply(prob_mat, 1, quantile, 0.05)
  pred_q95[idx]  <- apply(prob_mat, 1, quantile, 0.95)

  gc()
}

grid_sf_africa$pred_mean <- pred_mean
grid_sf_africa$pred_sd   <- pred_sd
grid_sf_africa$pred_q05  <- pred_q05
grid_sf_africa$pred_q95  <- pred_q95
```


```{r}
grid_vect <- vect(grid_sf_africa)

global_spatial_limit <- rast("data/Africa_data/spatial_limit_of_transmission/2010_Pf_Limits_Decompressed.geotiff")


################################################################################
#        Rasterize the prediction grid for an easy visualization
################################################################################

limit_3857 <- project(global_spatial_limit, "EPSG:3857", method = "near")
limit_3857 <- crop(limit_3857, vect(shp_africa))
limit_3857 <- mask(limit_3857, vect(shp_africa))

grid_limit <- terra::extract(limit_3857, grid_vect)

grid_sf_africa$limit <- grid_limit[,2]


grid_sf_africa$limit[is.na(grid_sf_africa$limit)] <- 0


grid_sf_africa_mask_unstable <- grid_sf_africa |>
  mutate(
    pred_mean = ifelse(limit == 0, NA, pred_mean),
    pred_sd   = ifelse(limit == 0, NA, pred_sd),
    pred_q05  = ifelse(limit == 0, NA, pred_q05),
    pred_q95  = ifelse(limit == 0, NA, pred_q95)
  )

grid_sf_africa_mask_stable <- grid_sf_africa |>
  mutate(
    pred_mean = ifelse(limit %in% c(0,1), NA, pred_mean),
    pred_sd   = ifelse(limit %in% c(0,1), NA, pred_sd),
    pred_q05  = ifelse(limit %in% c(0,1), NA, pred_q05),
    pred_q95  = ifelse(limit %in% c(0,1), NA, pred_q95)
  )

v1 <- vect(grid_sf_africa_mask_unstable)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)


grid_sf_africa_mask_unstable_raster <- rasterize(
  v1,
  template,
  field = c("pred_mean", "pred_q05", "pred_q95")
)

v1 <- vect(grid_sf_africa_mask_stable)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)

grid_sf_africa_mask_stable_raster <- rasterize(
  v1,
  template,
  field = c("pred_mean", "pred_q05", "pred_q95")
)


################################################################################
#                                   End                                        #
################################################################################


shp_africa_country <- shp_africa %>%
  group_by(country) %>%
  summarise(geometry = st_union(geometry))


Africa_prev_1 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using MAP data")
  

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level_unstable_bench.pdf",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country_1 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using MAP data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_unstable_bench.pdf",
       plot = Africa_prev_country_1, width = 6, height = 4, dpi = 300)

grid_sf_africa_mask_unstable_raster$CI_width <- grid_sf_africa_mask_unstable_raster$pred_q95 - grid_sf_africa_mask_unstable_raster$pred_q05

Africa_prev_country_uncertainty <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster,
                  mapping = aes(fill = CI_width))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "Credible intervale width\n of the estimated prevalence")+
  theme_void()+
  labs(title = "Associated uncertainty to the \n annual prevalence of P. f in 2021 using MAP data")


ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_uncertainty_bench.pdf",
       plot = Africa_prev_country_uncertainty, width = 6, height = 4, dpi = 300)


# setting unstable areas to 0

Africa_prev_2 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using MAP data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level_bench.pdf",
       plot = Africa_prev_2, width = 6, height = 4, dpi = 300)

Africa_prev_country_2 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster,
                  mapping = aes(fill = pred_mean))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 using MAP data")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_bench.pdf",
       plot = Africa_prev_country_2, width = 6, height = 4, dpi = 300)


```

```{r}
pfpr_2021 <- rast("data/prevalence_2000/2025_GBD2024_Global_PfPR_2021.tif") 

shp_africa_country_4326 <- shp_africa_country %>% st_transform(4326)

vect_africa1 <- vect(shp_africa_country_4326)

africa_map_pr_2021 <- crop(pfpr_2021, vect_africa1)
africa_map_pr_2021 <- mask(africa_map_pr_2021, vect_africa1)

africa_map_pr_2021_3857 <- project(africa_map_pr_2021, "EPSG:3857", method = "bilinear")
africa_map_pr_2021_3857_1 <- africa_map_pr_2021_3857[[1]]

names(africa_map_pr_2021_3857_1) <- "PfPR_2_10"


vect_grid_africa <- vect(grid_sf_africa)

grid_map_pr <- terra::extract(africa_map_pr_2021_3857, vect_grid_africa)

grid_map_pr <- grid_map_pr %>% select(-`Mask: Sparsely Populated`)

grid_sf_africa1 <- grid_sf_africa %>%
  mutate(PfPR_2_10   = grid_map_pr[,2],
         PfPR_2_10_L = grid_map_pr[,3],
         PfPR_2_10_U = grid_map_pr[,4]) %>% 
  select(-c(annual_rainfall_2021, annual_temperature_2021, annual_ITN_usage_2021,
            vegetation_index, elevation, dist_to_water, lc_trees, lc_water,
            lc_wetland, lc_cropland, pop_density_2020, lc_built))

numeric_cols <- names(grid_sf_africa1)[sapply(grid_sf_africa1, is.numeric)]

for (col in numeric_cols) {
  grid_sf_africa1 <- fill_by_nearest(grid_sf_africa1, col)
}


# Conversion of the prevalence to all age group

n <- nrow(grid_sf_africa1)

grid_sf_africa1$PR_all_age <- mapply(
  convertPrevalence,
  prev         = grid_sf_africa1$PfPR_2_10,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(100, n),
  parameters   = "Pf_Smith2007"
)

grid_sf_africa1$PR_all_age_U <- mapply(
  convertPrevalence,
  prev         = grid_sf_africa1$PfPR_2_10_U,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")

grid_sf_africa1$PR_all_age_L <- mapply(
  convertPrevalence,
  prev         = grid_sf_africa1$PfPR_2_10_L,
  age_min_in   = rep(2, n),
  age_max_in   = rep(10, n),
  age_min_out  = rep(0, n),
  age_max_out  = rep(99, n),
  parameters   = "Pf_Smith2007")


grid_sf_africa_mask_unstable1 <- grid_sf_africa1 |>
  mutate(
    pred_mean    = ifelse(limit == 0, NA, pred_mean),
    pred_sd      = ifelse(limit == 0, NA, pred_sd),
    pred_q05     = ifelse(limit == 0, NA, pred_q05),
    pred_q95     = ifelse(limit == 0, NA, pred_q95),
    PR_all_age   = ifelse(limit == 0, NA, PR_all_age),
    PR_all_age_L = ifelse(limit == 0, NA, PR_all_age_L),
    PR_all_age_U = ifelse(limit == 0, NA, PR_all_age_U))

grid_sf_africa_mask_stable1 <- grid_sf_africa1 |>
  mutate(
    pred_mean    = ifelse(limit %in% c(0,1), NA, pred_mean),
    pred_sd      = ifelse(limit %in% c(0,1), NA, pred_sd),
    pred_q05     = ifelse(limit %in% c(0,1), NA, pred_q05),
    pred_q95     = ifelse(limit %in% c(0,1), NA, pred_q95),
    PR_all_age   = ifelse(limit %in% c(0,1), NA, PR_all_age),
    PR_all_age_L = ifelse(limit %in% c(0,1), NA, PR_all_age_L),
    PR_all_age_U = ifelse(limit %in% c(0,1), NA, PR_all_age_U))


v1 <- vect(grid_sf_africa_mask_unstable1)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)


grid_sf_africa_mask_unstable_raster1 <- rasterize(
  v1,
  template,
  field = c("pred_mean", "pred_q05", "pred_q95", "PR_all_age", "PR_all_age_L", "PR_all_age_U")
)

v1 <- vect(grid_sf_africa_mask_stable1)

template <- rast(
  ext = ext(v1),
  resolution = 5000,
  crs = crs(v1)
)

grid_sf_africa_mask_stable_raster1 <- rasterize(
  v1,
  template,
 field = c("pred_mean", "pred_q05", "pred_q95", "PR_all_age", "PR_all_age_L", "PR_all_age_U")
)

  
```


```{r}
Africa_prev_1 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster1,
                  mapping = aes(fill = PR_all_age))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 (MAP)")
  

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level_unstable_MAP.png",
       plot = Africa_prev_1, width = 6, height = 4, dpi = 300)

Africa_prev_country_1 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_unstable_raster1,
                  mapping = aes(fill = PR_all_age))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 (MAP)")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_unstable_MAP.png",
       plot = Africa_prev_country_1, width = 6, height = 4, dpi = 300)


# setting unstable areas to 0

Africa_prev_2 <- ggplot(shp_africa) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster1,
                  mapping = aes(fill = PR_all_age))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 (MAP)")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_admin2_level_MAP.png",
       plot = Africa_prev_2, width = 6, height = 4, dpi = 300)

Africa_prev_country_2 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster1,
                  mapping = aes(fill = PR_all_age))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
    na.value = "transparent",
    name = "annual prevalence of P. f")+
  theme_void()+
  labs(title = "annual prevalence of P. f in 2021 (MAP)")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_MAP.png",
       plot = Africa_prev_country_2, width = 6, height = 4, dpi = 300)


grid_sf_africa_mask_stable_raster1$diff_PR <- grid_sf_africa_mask_stable_raster1$pred_mean - grid_sf_africa_mask_stable_raster1$PR_all_age

Africa_prev_country_3 <- ggplot(shp_africa_country) +
  geom_spatraster(data = grid_sf_africa_mask_stable_raster1,
                  mapping = aes(fill = diff_PR))+
  geom_sf(fill = NA)+
  scale_fill_stepsn(
    colours = rev(brewer.pal(5, "RdYlGn")),
    breaks = c(-0.7, 0, 0.05, 0.1, 0.2, 0.56),
    na.value = "transparent",
    name = "Difference in prevalence of P. f")+
  theme_void()+
  labs(title = "Difference in prevalence of P. f in 2021 \n using MAP modeled prevalence vs estimated\n prevalence using our model")

ggsave("Outputs/africa_output/2021_africa_pr_falciparum_map_country_level_diff_MAP.png",
       plot = Africa_prev_country_3, width = 6, height = 4, dpi = 300)

```

