---
title: "Ct Values Analysis: COMAL data"
author: "Ndim Handson NSANI"
format: html
editor: visual
---

```{r}
#| label: loading necessary libraries

library(readxl)
library(ggplot2)
library(tidyverse)
library(tidyterra)
library(terra)
```

```{r}
#| label: loading the data

BEN_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 1)

CON_df <- read_excel("data/COMAL data Benin&Congo.xlsx", sheet = 2)

CMR_df <- read_excel("data/cmr_data/Cameroon2.xlsx")

GAB_df <- read_excel("data/gab_data/Gabon2.xlsx")  
```

```{r}
#| label: cleaning the data 

BEN_Ct <- BEN_df %>% 
  dplyr::select(Village, `Sample ID`, `Household ID`, Sex, `Age (year)`, Season, `Parasitaemia Pf (P/µl)`,
                `Parasitaemia Pm (P/µl)`, `qRT-qPCR Pf positive`, `qRT-qPCR Pm positive`,
                `Pf(Ct) mean`, `Pm(Ct) mean`) %>% 
  rename(Sample_ID        = `Sample ID`,
         Household_ID     = `Household ID`,
         Age              = `Age (year)`,
         Parasitaemia_Pf  = `Parasitaemia Pf (P/µl)`,
         Parasitaemia_Pm  = `Parasitaemia Pm (P/µl)`,
         Positive_pf      = `qRT-qPCR Pf positive`,
         Positive_pm      = `qRT-qPCR Pm positive`,
         Ct_value_pf      = `Pf(Ct) mean`,
         Ct_value_pm      = `Pm(Ct) mean`) %>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 5, 10, Inf),
                                labels = c("<5", "5<10", "10+"),
                                right  = FALSE),
         Parasitaemia_Pf  = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Positive_pf      = case_when(is.na(Positive_pf) ~ "neg",
                                     TRUE ~ "pos"),
         Positive_pm      = case_when(is.na(Positive_pm) ~ "neg",
                                     TRUE ~ "pos"),
         Ct_value_pf      = case_when(is.na(Ct_value_pf) ~ 0,
                                     TRUE ~ Ct_value_pf),
         Ct_value_pm      = case_when(is.na(Ct_value_pm) ~ 0,
                                     TRUE ~ Ct_value_pm),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

CONG_Ct <- CON_df %>% 
  dplyr::select(`Sample ID`, Sex, `Age (year)`,`Parasitaemia Pf (P/µl)`, 
                `Parasitaemia Pm (P/µl)`, `qRT-qPCR Pf positive`,
                `qRT-qPCR Pm positive`, `Pf(Ct) mean`, `Pm(Ct) mean`)%>% 
  rename(Sample_ID        = `Sample ID`,
         Age              = `Age (year)`,
         Parasitaemia_Pf  = `Parasitaemia Pf (P/µl)`,
         Parasitaemia_Pm  = `Parasitaemia Pm (P/µl)`,
         Positive_pf      = `qRT-qPCR Pf positive`,
         Positive_pm      = `qRT-qPCR Pm positive`,
         Ct_value_pf      = `Pf(Ct) mean`,
         Ct_value_pm      = `Pm(Ct) mean`)%>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 5, 10, Inf),
                                labels = c("<5", "5<10", "10+"),
                                right  = FALSE),
         Parasitaemia_Pf  = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Positive_pf      = case_when(is.na(Positive_pf) ~ "neg",
                                     TRUE ~ "pos"),
         Positive_pm      = case_when(is.na(Positive_pm) ~ "neg",
                                     TRUE ~ "pos"),
         Ct_value_pf      = case_when(is.na(Ct_value_pf) ~ 0,
                                     TRUE ~ Ct_value_pf),
         Ct_value_pm      = case_when(is.na(Ct_value_pm) ~ 0,
                                     TRUE ~ Ct_value_pm),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

CMR_Ct <- CMR_df %>% 
  dplyr::select(district, ID, Season, age, parasitemia_pf, parasitemia_pm, pcr_pf,
                pcr_pm, pf_ct, pm_ct) %>% 
  rename(Village         = district,
         Sample_ID       = ID,
         Age             = age,
         Parasitaemia_Pf = parasitemia_pf,
         Parasitaemia_Pm = parasitemia_pm,
         Positive_pf     = pcr_pf,
         Positive_pm     = pcr_pm,
         Ct_value_pf     = pf_ct,
         Ct_value_pm     = pm_ct) %>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 5, 10, Inf),
                                labels = c("<5", "5<10", "10+"),
                                right  = FALSE),
         Parasitaemia_Pf = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))

GAB_Ct <- GAB_df %>% 
  dplyr::select(district, ID, season, age, parasitemia_pf, parasitemia_pm, pcr_pf, 
                pcr_pm, pf_ct, pm_ct) %>% 
  rename(Village         = district,
         Sample_ID       = ID,
         Season          = season,
         Age             = age,
         Parasitaemia_Pf = parasitemia_pf,
         Parasitaemia_Pm = parasitemia_pm,
         Positive_pf     = pcr_pf,
         Positive_pm     = pcr_pm,
         Ct_value_pf     = pf_ct,
         Ct_value_pm     = pm_ct)%>% 
  mutate(Age_group        = cut(Age,
                                breaks = c(0, 5, 10, Inf),
                                labels = c("<5", "5<10", "10+"),
                                right  = FALSE),
         Parasitaemia_Pf = case_when(is.na(Parasitaemia_Pf) ~ 0,
                                     TRUE ~ as.numeric(Parasitaemia_Pf)),
         Parasitaemia_Pm  = case_when(is.na(Parasitaemia_Pm)~0,
                                     TRUE ~ as.numeric(Parasitaemia_Pm)),
         Ct_level_pf      = case_when((Ct_value_pf>0 & Ct_value_pf<=20) ~ "Low",
                                      (Ct_value_pf>20 & Ct_value_pf<=30) ~ "Moderate",
                                      (Ct_value_pf>30) ~ "High",
                                      TRUE ~ "None"),
         Ct_level_pm      = case_when((Ct_value_pm>0 & Ct_value_pm<=20) ~ "Low",
                                      (Ct_value_pm>20 & Ct_value_pm<=30) ~ "Moderate",
                                      (Ct_value_pm>30) ~ "High",
                                      TRUE ~ "None"))
 
```

```{r}
#| label: Demographic and risk factor analysis

boxplot_Ct_Age_Ben <- ggplot(data = BEN_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group in Benin",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/ben_output/Ct_Age_Ben.png", plot = boxplot_Ct_Age_Ben, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_CONG <- ggplot(data = CONG_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group in Congo",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/cong_output/Ct_Age_CONG.png", plot = boxplot_Ct_Age_CONG, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_CMR <- ggplot(data = CMR_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group in Cameroon",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/cmr_output/Ct_Age_CMR.png", plot = boxplot_Ct_Age_CMR, width = 6, height = 4, dpi = 300)


boxplot_Ct_Age_GAB <- ggplot(data = GAB_Ct %>% filter(Ct_value_pm != 0), aes(x= Age_group, y = Ct_value_pm,
                                                       fill = Age_group))+
  geom_boxplot(alpha = 0.6, outlier.fill = NA)+
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(title = "Parasite burden (Ct value) by Age group in Gabon",
       x     = "Age group",
       y     = "PCR Ct value")

ggsave("Outputs/gab_output/Ct_Age_GAB.png", plot = boxplot_Ct_Age_GAB, width = 6, height = 4, dpi = 300)

```

-   In Benin the `5<10` age group is the one with the lowest Ct value meaning they are the one with high parasitemia in the country while `<5` is the one with highest value, then low parasitemia due to acquire natural immunity.

-   In Congo `10+` is the age group with the lowest Ct value and the one with the highest value is `5<10`.

-   In Cameroon `<5` and `5<10` are the ones with the lowest Ct value and the with the highest value is `10+`. For this case with two outliers for `5<10` age group with Ct value greater than `50`.

-   For Gabon we still have `<5` and `5<10` age group which are the ones with the lowest Ct value whereas `10+` is the one with the highest Ct value.


```{r}
#| label: disease burden per age group
  
country_summary <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  group_by(Country) %>%
  summarise(
    n_tested = n(),
    n_positive_pm = sum(Positive_pm == "pos", na.rm = TRUE),
    prevalence_pm = n_positive_pm / n_tested,
    .groups = "drop"
  )


age_prev <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  group_by(Country, Age_group) %>%
  summarise(
    n = n(),
    positive = sum(Positive_pm == "pos", na.rm = TRUE),
    prevalence = positive / n,
    .groups = "drop"
  )



ggplot(age_prev,
       aes(x = reorder(Age_group, prevalence), y = prevalence, fill = Country)) +
  geom_col(position = "dodge")+
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Age group",
    y = "P. malariae prevalence",
    fill = "Country"
  ) +
  theme_minimal()


ggplot(age_prev,
       aes(x = reorder(Country, prevalence),
           y = prevalence,
           fill = Country)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ Age_group, scales = "free_x") +
  labs(
    x = "Country",
    y = "P. malariae prevalence"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
#| label: Ct values at country level 

pm_positive <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  filter(Positive_pm == "pos")

ggplot(pm_positive,
       aes(x = Country, y = Parasitaemia_Pm)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    x = "Country",
    y = "P. malariae parasitaemia"
  ) +
  theme_minimal()


bbplot_countries <- ggplot(pm_positive,
       aes(x = Country, y = Ct_value_pm, fill = Country)) +
  geom_boxplot(alpha = 0.6, outlier.fill = NA) +
  geom_jitter(width = 0.2, alpha = 0.5,  color = "darkslategrey")+
  scale_y_reverse()+
  scale_fill_brewer(palette = "Set2")+
  labs(
    x = "Country",
    y = "P. malariae Ct value"
  )

ggsave("Outputs/Ct_Countries.png", plot = bbplot_countries, width = 6, height = 4, dpi = 300)


box_plot_facet_by_country <- ggplot(pm_positive,
       aes(x = Age_group, y = Ct_value_pm, fill = Age_group)) +
  geom_boxplot(alpha = 0.6, outlier.fill = NA) +
  geom_jitter(width = 0.2, alpha = 0.5,  color = "darkslategrey")+
  scale_y_reverse()+
  facet_wrap(~Country, nrow = 2)+
  scale_fill_brewer(palette = "Set2")+
  labs(
    x = "Country",
    y = "P. malariae Ct value"
  )

ggsave("Outputs/Box_Ct_Countries.png", plot = box_plot_facet_by_country,
       width = 6, height = 4, dpi = 300)

ggplot(pm_positive,
       aes(x = Ct_value_pm, fill = Country)) +
  geom_density(alpha = 0.4) +
  theme_minimal()
```


- So at the country level, Cameroon has the highest Ct level and also the highest
prevalence which means high prevalence might not necessary mean severity of the 
disease but instead a hidden reservoir of the parasite transmission with low
density of parasite load. But if we look closely to countries like Benin, 
Gabon, Congo, we will see that Benin has the prevalence value but all those three
countries have almost the same level of Ct values which mean with the data we
have, we cannot draw an effective conclusion with the data we have.


```{r}
#| label: Age vs Ct values

country_pm <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
)

scarterplot_age_ct <- ggplot(data = country_pm %>% filter(!is.na(Age)), aes(x= Age, y= Ct_value_pm))+
  geom_point(alpha = 0.6) +
  facet_wrap(~Country, nrow = 2)+
  theme_minimal()+
  labs(title = "Relationship between Age and Ct values (P. m)", x = "Age", y = "Ct values")

ggsave("Outputs/scarterplot_age_ct.png", plot = scarterplot_age_ct, width = 6, height = 4, dpi = 300)


cor(country_pm$Age, country_pm$Ct_value_pm, method = "pearson", use = "complete.obs")


country_pm %>%
  group_by(Country) %>%
  summarise(
    prevalence = mean(Positive_pm == "pos", na.rm = TRUE),
    mean_Ct = mean(Ct_value_pm[Positive_pm == "pos"], na.rm = TRUE),
    median_Ct = median(Ct_value_pm[Positive_pm == "pos"], na.rm = TRUE)
  )

ggplot(pm_positive,
       aes(x = Age, y = Ct_value_pm)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ Country) +
  theme_minimal()


co_inf <- bind_rows(
  BEN_Ct  %>% mutate(Country = "Benin"),
  CONG_Ct %>% mutate(Country = "Congo"),
  CMR_Ct  %>% mutate(Country = "Cameroon"),
  GAB_Ct  %>% mutate(Country = "Gabon")
) %>%
  count(Country,Positive_pf, Positive_pm)
```



```{r}

```


```{r}

```


```{r}

```


```{r}

```

